### 2020-07-30 nicolas/PPIPL3/ReportV1.tex
first version of report (incomplete)

%% LyX 2.3.5-1 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[twocolumn,english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{float}
\usepackage{graphicx}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

\makeatother

\usepackage{babel}
\begin{document}
\title{Enhanced Prediction of Protein-Protein Interactions on the Human and
Rice Interactome with Machine Learning}
\author{Nicolás Antonio López Rozo}
\maketitle
\begin{abstract}
Computational tools offer an alternative to identify unmapped, biologically
significant protein-protein interactions (PPI's). A novel approach
by Barabási et al, based on network paths of length three (L3) on
the interactome, outperforms all existing link prediction methods.
L3 can offer therefore a new approach to reveal a smaller subset of
potential interactions for a more affordable experimental validation.

In this study, the previous approach for link prediction is expanded
by using \texttt{XGBoost} module of Machine Learning (ML). For the
feature extraction from the network, the \texttt{Node2Vec} module
was applied, although L3 information is added as a feature for improving
link prediction. Results show that L3 feature is very important for
the ML procedure, although a significant feature importance lies on
the features retrieved from the network structure using \texttt{Node2Vec}.
The mentioned approaches are compared using the human and the rice
interactomes.
\end{abstract}

\section{Introduction}

Proteins, as a fundamental constituents of any living being and their
``building blocks'' in terms of biological functionality, interact
among themselves and with other molecules according to their amino
acid sequences and specific three-dimensional structure and folding.
These interactions also depend on many physical and chemical properties
of each macro-molecule and on the cellular organelle where they operate,
such as pH and temperature. For this reason, experimental ways to
reproduce all of the necessary conditions to validate if an interaction
actually occurs within the cell are generally expensive.

The most common way to validate PPI's is the \emph{Yeast-Two-Hybrid}
technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on the expression of a specific reporter gene that
activates by the binding of a DNA-binding Domain (DB) and an Activation
Domain (AD) of a Transcription Factor. For the Y2H technique, a protein
is fused to the DB domain (known as \emph{bait}) and another one to
the AD (known as \emph{prey}). If the proteins do not interact, then
the reporter gene is not expressed. Otherwise, the reporter gene expression
is activated by the activation domain.

\begin{figure}[h]
\caption{Yeast-2-Hybrid Technique}

\noindent \centering{}\includegraphics[width=1\columnwidth]{Y2H}
\end{figure}

Having several of these Y2H results allows scientists to establish
a PPI network, where all known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
unknown interactions. In this report, three prediction methods are
presented and their results are shown: Common Neighbors (\textbf{CN}),
which uses the length-2 path count; length-3 path count (\textbf{A3});
and degree-normalized length-3 paths score (\textbf{L3}).

\begin{figure}[h]
\caption{Protein-Protein Interaction Prediction Scheme}

\noindent \centering{}\includegraphics[width=0.9\columnwidth]{PPI_prediction}
\end{figure}

Focus of the present study is to evaluate different methods for predicting
protein-protein interactions (PPIs) using the existing knowledge of
the network, which is an undirected graph. The traditional way is
usually based on social networks analysis, more specifically on the
Triadic Closure Principle (TCP), that states that the more common
shared friends that two people have, the more likely that they know
each other. As shown in the next sections, the mentioned approach
fails because it does not consider the structural and chemical properties
of the proteins.

For achieving the described results, several networks are used and
compared using state-of-the-art methods, as well as the proposed ones
(CN, A3, L3). For this report, the human interactome was used (\emph{HI-II-14.txt}).
A curated version of the mentioned network (\emph{HI-TESTED.txt})
was also used in order to compare the influence of research bias on
the network predictions. A massive experimental assay was carried
on and its results are consolidated and used to build a validation
network (\emph{HI-III.txt}).

\section{Materials and Methods}

\subsection{Data Availability}

Human interactome data and base source code were downloaded from the
repository of the length-3 degree normalized paths methodology\cite{Kovacs2019}:
the dataset \emph{HI-II-14.txt} and \emph{HI-TESTED.txt} are used
for prediction and the dataset \emph{HI-III.txt} is used for validation.
Rice interactome information was downloaded from the STRING database
\cite{Szklarczyk2019}, corresponding to the \emph{Oryza sativa} subspecies.
The downloaded file was \emph{4530.protein.links.detailed.v11.0.txt}.
and contains more than 8 million PPIs from several resources. For
the purpose of this study and based on the previous work, only PPIs
with evidence from curated databases were used (rows where \emph{databases}
column has a value greater than zero), resulting in a network with
NNNN nodes and MMMM edges.

\subsection{Code Implementation}

Previous code implementation was adapted from \texttt{C++} to \texttt{Python}
(3.6), in order to unify the algorithms into one single script. For
the purpose of algorithmic validation, the three methods were implemented
from scratch with basic functionalities and data structures of the
\emph{Python} (V3.6) language.

\subsection{Data Preprocessing}

Information for the human interactome was used as-is, which corresponds
to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
$9433$ and $23322$ interactions. For the rice interactome, an additional
preprocessing was performed. The filtered network for rice consists
of $5025$ proteins (nodes) and $164420$ interactions (edges) distributed
among $178$ connected components. The connected component with the
greatest number of edges was selected in this case.

The extracted connected component consists of $n=4390$ nodes and
$m=163319$ edges, which corresponds to $99.33%\%
$ of filtered edges. Further investigation is applied to this network,
which is very similar in number of nodes to the curated information
on the human interactome, although rice network is much more connected.

\subsection{Edge Prediction}

For the interaction prediction for each network, the algorithms described
below were used. It is important to keep in mind how the protein-protein
interaction (PPI) network $G=(V,E)$ is conceptualized: each node
($v_{i}\in V$) represents a protein and each undirected edge ($e_{b}=\{v_{i},v_{j}\},\,e_{b}\in E$)
represents and interaction among proteins $v_{i}$ and $v_{j}$.

\begin{figure}[h]
\caption{Edge Prediction Paradigms}

\noindent \centering{}\includegraphics[width=0.8\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/TCP_extract\string".eps}
\end{figure}

\begin{description}
\item [{Common~Neighbors~(CN)}] This method is based on the Triadic Closure
Principle: ``the more common friends two individuals have, the more
likely that they know each other''. For the implementation of this
method, $A{{}^2}$ matrix is calculated, being $A$ the adjacency
matrix of the network.
\item [{Length-3~Paths~(A3)}] This is the simplest implementation of
the proposed insight of ``if my friends and your friends interact,
then we might interact too''. The calculating is carried on with
$A{{}^3}$, i.e, the third power of the adjacency matrix.
\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
overestimate the importance of some edges due to intermediate hubs
which add many shortcuts in the graph. To address that issue, a degree
normalization for the path $X\rightarrow U\rightarrow V\rightarrow Y$
is applied by considering the degree $k$ of the intermediate nodes
$U$ and $V$, as follows.
\[
p_{XY}=\sum_{U,V}\frac{A_{XU}\cdot A_{UV}\cdot A_{VY}}{\sqrt{k_{U}\cdot k_{V}}}
\]
\\
where $A_{ij}$ represents the value of the adjacency matrix for nodes
$i$ and $j$: 1 if the edge $\{i,j\}$ exists, 0 otherwise.
\end{description}

\subsection{Sampling and Precision Counting}

Sampling proceeds as follows: first, a fraction (10\%) of the edges
is removed at random from the network. Then, for each method (\textbf{CN},
\textbf{A3}, \textbf{L3}) the prediction of the removed fraction is
performed. The new edges are then compared against the original (removed)
edges according to the ranking that each method yields. Edges belonging
to the original connected component are considered positive in the
sense that they are experimentally validated interactions. For each
of the tested networks and a rank $r$, the precision was calculated
as 
\[
P_{r}=\frac{L_{r}}{r}
\]
\\
where $L_{r}$ is the total number of positive edges in the range
$[1,r]$.

The previous procedure was applied for a fraction of 0.90 remaining
edges, and a mean of 10 simulations was obtained.

\subsection{Feature Extraction with \texttt{Node2Vec}}

The \texttt{Node2Vec} module was used for extracting features of the
rice interactome graph. The parameters and considerations for the
model were:
\begin{itemize}
\item All paths in the random walks are equally likely (\texttt{p=1, q=1})
\item Use a modest number of dimensions and threads for calculation (\texttt{dimensions=16,
workers=4})
\item Since length-3 paths are the defining property in this study, there
is no necessity for longer walks. However, it is important to try
out many possible redundant routes and to consider a window of at
least 4 (\texttt{walk\_length=5, num\_walks=300, window=5})
\item Other standard parameters were left with default values (\texttt{min\_count=1,
batch\_words=4})
\item Edge embeddings were calculated using a geometric ratio of the node
embeddings (\texttt{HadamardEmbedder})
\end{itemize}

\subsection{Handcrafted Feature}

Due to the poor results of the \emph{raw} Length-3 counting (\textbf{A3}),
a different approach for this information was carried out in the present
study: As it still gives a lot of information that might be useful
for a predictive routine, this counting was normalized (dividing by
the greatest counting in the \textbf{A3} top predictions) and then
used as a feature for the Machine Learning algorithm. For completeness,
also \textbf{CN} and \textbf{L3} information was used as a possible
feature. Finally, the case were no handcrafted feature was also considered,
that is, only the features extracted from the structure of the network.

\subsection{Feature to Predict: Existence}

The feature to predict corresponds to the possible existence (\emph{True/False})
of a link based on the existing information of the network, using
the network itself in a random sub\_exploration (\texttt{Node2Vec})
as well as in a structured search (A3). This property is evaluated
by taking out a fraction of the edges and then trying to predict for
a given set of possible edges if they have a high probability to belong
to the original network.

\subsection{Machine Learning Algorithm}

The Extreme Gradient Boosting implementation of gradient boosted trees
is applied in this study to evaluate the existence of an edge. Gradient
boosted trees are usually used for supervised learning problems, where
the training data $X_{i}$ has multiple features and pretends to explain
(or predict) a target variable $Y_{i}$. The corresponding implementation
applied for this study is \texttt{XGBoost}, available publicly.

The selected parameters for the model were:
\begin{itemize}
\item \texttt{max\_depth=3}
\item \texttt{colsample\_bytree=0.6}
\item \texttt{eval\_metric='auc'}
\end{itemize}

\subsection{Result Validation}

For applying the test results, a partition of the whole dataset was
made: 80\% of the data was randomly selected and used for training,
while the remaining 20\% was used for validation. The whole training-validation
procedure was applied 5 times.

The chosen metric for validation was the Area under the Curve (\textbf{AUC})
of the Receiver Operating Characteristic (\textbf{ROC}). This curve
corresponds to plot the sensitivity (probability of predicting a real
positive as positive) against 1-specificity (probability of predicting
a real negative as positive). It is worth to remind that AUC values
move in the range $[0,1]$, where 1 is a perfect prediction and 0.5
corresponds to a random guess. Normally, values over 0.8 of AUC are
considered good.

\section{Results}

As explained above, the precision trend for each of the networks are
presented for the predicted top 2000 interactions.

\begin{figure}[h]
\caption{Methods Comparison for \emph{HI-II-14.txt}}

\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
\end{figure}

\begin{figure}[h]
\caption{Methods Comparison for \emph{HI-TESTED.txt}}

\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
\end{figure}

As it can be inferred from the plots, L3-based predictions outperform
their $A{{}^2}$ counterparts. Results also show that L3-score and
$A^{3}$predictions follow a very similar trend.

When analyzing the robustness of each of the networks, the following
values for the Weighted Spectral Distribution were found. For a robustness
reference, the Erdos-Renyi model was used to generate a random network
with the same number of nodes and edges and on those random networks,
the WSD was measured.
\begin{center}
\begin{table}[H]
\caption{Validation of Weighted Spectral Distribution}

\centering{}%
\begin{tabular}{|c|c|c|}
\hline 
Network & Network WSD & Erdos-Renyi WSD\tabularnewline
\hline 
\hline 
HI-II-14 & 393.4939 & 198.7706\tabularnewline
\hline 
HI-TESTED & 423.3902 & 276.1065\tabularnewline
\hline 
HI-III (VALID.) & 373.9369 & 153.5329\tabularnewline
\hline 
\end{tabular}
\end{table}
\par\end{center}

The table shows that the three networks used in this report are robust,
because the are significantly more robust than a network with the
same density generated randomly.

\section{Conclusions}

Taking into account the different results validated in this report,
one can conclude that length-3 path methodologies might work better
on protein-protein interactions than its traditional length-2 (TCP
based) counterparts. On the other hand, it can be seen that degree-normalization
has little effect on the predictions, i.e., non-normalized $A{{}^3}$
matrix predictions are still a good methodology for edge prediction
on PPI networks.

Previous result comes as no surprise when the biological basis of
protein interactions is considered: It is necessary that protein A
and protein B have complementary structures in order to interact,
and when classical length-2 paths are used, the predicted protein
interactions usually have the same structures, and not complementary
ones.

For the analyzed networks, it is also important to highlight their
robustness when compared to a random network with the same density:
if weighted spectral distribution (WSD) is used, then the used networks
are on average twice as robust as their random network versions.

\bibliographystyle{plain}
\bibliography{\string"/home/ocin/Documents/02 Asignaturas/refs\string"}

\end{document}

### 2020-07-30 camila/Report/P5_report.tex
report added

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Plantilla para un artículo en LaTeX en doble columna.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% !TeX spellcheck = es_ANY
%\documentclass[11pt,twocolumn]{article} %doble columna
%\documentclass[11pt]{article}

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.

\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{array}

% Paquetes de la AMS:
\usepackage{amsmath, amsthm, amsfonts, amssymb}

% Atajos.
% Se pueden definir comandos nuevos para acortar cosas que se usan
% frecuentemente. Como ejemplo, aquí se definen la R y la Z dobles que
% suelen representar a los conjuntos de números reales y enteros.
%--------------------------------------------------------------------------

\def\RR{\mathbb{R}}
\def\ZZ{\mathbb{Z}}

% De la misma forma se pueden definir comandos con argumentos. Por
% ejemplo, aquí definimos un comando para escribir el valor absoluto
% de algo más fácilmente.
%--------------------------------------------------------------------------
\newcommand{\abs}[1]{\left\vert#1\right\vert}

%--------------------------------------------------------------------------
\title{Identification of rice genes which respond to saline stress from co-expression networks analysis}
\author{Camila Riccio Rengifo\\
  \small Pontificia Universidad Javeriana Cali\\
}

\begin{document}

\maketitle

\begin{abstract}
A gene co-expression network is an undirected graph , where each node corresponds to a gene, and a pair of nodes is connected if there is a significant co-expression relationship between them, that is, if they show similar expression patterns. These co-expression networks are of biological interest since the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex. Weighted gene co-expression network analysis (WGCNA)  is a method of data mining widely used to build co-expression networks, identify groups (modules) of highly correlated genes and relate these modules to external traits.

The aim of this work is to identify rice genes that respond to saline stress. The methodology is based on WGCNA with a new aproach in the modules detection, using  the hierarchical link clustering (HLC) technique that allows the recognition of overlapping communities, which may have more biological meaning given the overlapping regulatory domains of systems that generate coexpression \cite{gaiteri2014beyond}. Finally the LASSO method is used to select the most significant modules associated with rice phenotypical responses to salt stress. The genes differentially expressed within the selected modules are enriched with gene ontology annotations and their interaction networks are studied.
%\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}

\section*{Introduction}
Introduction text

\section{Methodology}
\subsection{Data pre-processing} 
%-----------------------------
% RNA-seq data
%-----------------------------
RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $p=92$ accessions of the Rice Diversity Panel 1 (with two biological repetitions). \\

The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results. The normalization tecnique used was DESeq2 \cite{love2014moderated}. From the normalized data, the two biological repetitions of each accession were averaged and genes exhibiting low variance (the ratio of upper quantile to lower quantile smaller than 1.5) or low expression (more than 80\% samples with values smaller than 10) were removed. At this point, control and salinity stress treatment data are separated into two matrices $C=[c_{ij}]_{n \times p}$ and $T=[t_{ij}]_{n \times p}$, respectively, where $c_{ij}$ and $t_{ij}$ represent normalized expression level of the gene $i$ in the accession $j$.

Changes in gene expression between control and stress conditions, are measured in terms of log ratios. Matrix $E=[e_{ij}]_{n \times p}$, known as the Log Fold Change matrix, is computed by setting $e_{ij}=\log_2 (t_{ij}/c_{ij})$.

Genes with the ratio of upper quantile to lower quantile larger than $0.25$ were kept. This procedure aims to remove genes with low variance in the differential expression paterns. The final network contains $8928$ genes of the initial $57845$.

\subsection{Network construction}
The co-expression network is constructed based on WGCNA, as explained below.\\

The Log Fold Change matrix is used to construct the weighted co-expression network using the absolute value of the Pearson correlation as the similarity measure between genes. The matrix $S=[s_{ij}]_{n\times n}$ measures the level of concordance between gene expression profiles across experiments and is transformed into an adjacency matrix $A=[a_{ij}]_{n\times n}$ where $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes. In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network. This kind of networks contain many nodes with very few connections and a small number of hubs with high connections. \\

In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node). So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g. $R^2 > 0.85$). Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (rigth) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.

%scale-free beta 1 y 3
\begin{figure}[h]
  \centering
    \includegraphics[clip,width=0.4\textwidth]{Figures/pick_beta.png}
  \caption{Degree distributions}
  \label{fig:beta}
\end{figure}

\subsection{Module detection}	
Once the network has been constructed, the next step is to study its structure and dynamics identifying communities also called modules. Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e. genes) are involved in multiple functions.\\

To identify these overlapping communities we make use of the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}. HLC aproach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities. The algorithm maps links to nodes and connects them if a pair of links shares a node. They compute the similarity between links using the Jaccard index for unweigthed networks and the tanimoto coefficient for weighted networks. With this similarity, they use single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities. Finally, the most relevant communities are established at the maximal partition density, a function  based on link density inside communities.\\

The current network represented by the adjacency matrix $A$, corresponds to a complete and weigthed network of $8,928$ genes (nodes) and $39,850,128$ edges. For computational reasons, this network was transformed into an unweighted one $\hat{A}$, keeping only the connections above the cutoff value of $0.2$. The resulting unweighted network has a total of $5,810$ nodes and $16,875,145$ edges. After applying the HLC algorithm, a total of $4,131$ genes were distributed in $5,143$ overlapping modules of $3$ or more genes. Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.\\

%overlapping percentage
\begin{figure}[h]
  \centering
    \includegraphics[clip,width=0.4\textwidth]{Figures/artificial_modules.png}
  \caption{Overlapping percentage}
  \label{fig:overlap}
\end{figure}

\subsection{Modules association to phenotypic traits}

Our approach to identify the most relevant gene groups (modules) in the response to salt stress, consists in relating the co-expression network with phenotypic data from the samples. We use 3 phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass. These were measured for the 92 genotypes studied, under  control and salt stress conditions. These data can be found in the supplementary information of~\cite{campbell2017allelic}. As shown in Figure \ref{fig:pdata}, there are significant differences in the values of these phenotypic traits between both stress and control conditions. This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.\\

% boxplots
\begin{figure}[h]
  \centering
    \includegraphics[clip,width=0.5\textwidth]{Figures/phenotypic_traits.png}
  \caption{Phenotipic traits distribution under control and salt stress}
  \label{fig:pdata}
\end{figure}

In order to relate the modules to the phenotypic traits, we represent each module with the first principal component of the Log Fold Change submatrix, corresponding to the genes of such module, which can be thought of as an average differential expression profile for each community. These profiles are then associated with each phenotypic trait using the LASSO variable selection method, allowing to identify a set of relevant modules related to the response to salinity conditions in rice plants. See Appendix A for a LASSO method description.\\

%Módulos seleccionados + cantidad de genes por módulo

\subsection{Genes enrichment}
% genes seleccionados con expresión diferencial según análisis DEA. 
% gene ontology de esos genes

\section{Results}
% redes de los genes anotados

\section{Discussion}

\bibliographystyle{acm}
\bibliography{mybibliography}

\newpage
\section*{Appendix}
\subsection{A: Variable selection with LASSO}
LASSO (Least Absolute Shrinkage Selector Operator) is a regularized linear regression technique, a method that combines a regression model with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients. Very usefull in problems where the number of variables (genes) $ n $ is much greater than the number of samples $ p $ ($ n \gg p $). Lasso solves the least squares problem with restriction on the $ L_1$-norm of the coefficient vector:

\begin{equation}
\min \left\lbrace\sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} \right\rbrace , \textrm{sujeto a} \sum_{j=1}^n\abs{\beta_j}\leq s
\end{equation}

Or equivalently minimizing:
\begin{equation}
\sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} + \lambda \sum_{j=1}^n\abs{\beta_j}
\end{equation}
being $ s $, $ \lambda \geq 0 $ the respective penalty parameters for complexity.\\

%LASSO produces parameter estimation and simultaneous variable selection for increasing values of $ \lambda $.

%if $ \lambda = 0 $ the estimator corresponds to the ordinary least squares estimator.

Since the $\lambda$ value determines the degree of penality, the accuracy of the model depends on its choice. Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squered error. With that selected $\lambda$ value, the model is adjusted again, this time using all the observations.\\

In the module gene selection context, the outcome variable correspond to the phenotipic trait, whereas the predictors are the modules, detected by the hierarchical clustering, represented by the first principal component of the module. After running the Lasso regression will select the most significant modules associated with phenotyping data.

\end{document}

### 2020-07-30 mauricio/download_fq_files.sh
Assembly guide and files links to download

@@ -1,0 +1,366 @@
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/009/SRR5495129/SRR5495129.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/000/SRR5495130/SRR5495130.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/003/SRR5495133/SRR5495133.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/004/SRR5495134/SRR5495134.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/005/SRR5495135/SRR5495135.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/006/SRR5495136/SRR5495136.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/007/SRR5495137/SRR5495137.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/008/SRR5495138/SRR5495138.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/009/SRR5495139/SRR5495139.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/000/SRR5495140/SRR5495140.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/001/SRR5495141/SRR5495141.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/002/SRR5495142/SRR5495142.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/003/SRR5495143/SRR5495143.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/004/SRR5495144/SRR5495144.fastq.gz
+wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR549/005/SRR5495145/SRR5495145.fastq.gz
[…]

### 2020-07-30 nicolas/PPIPL3/ReportV1.tex

@@
 %% Do not edit unless you really know what you are doing.
-\documentclass[twocolumn,english]{article}
+\documentclass[onecolumn,english]{article}
 \usepackage[T1]{fontenc}
@@
 \begin{document}
-\title{Enhanced Prediction of Protein-Protein Interactions on the Human and
-Rice Interactome with Machine Learning}
-\author{Nicolás Antonio López Rozo}
+\title{Prediction of Protein-Protein Interactions on the Human and
+Rice Interactome}
+\author{Nicol\'as Antonio L\'opez Rozo}
 \maketitle
 \begin{abstract}
-Computational tools offer an alternative to identify unmapped, biologically
-significant protein-protein interactions (PPI's). A novel approach
-by Barabási et al, based on network paths of length three (L3) on
-the interactome, outperforms all existing link prediction methods.
-L3 can offer therefore a new approach to reveal a smaller subset of
-potential interactions for a more affordable experimental validation.
-
-In this study, the previous approach for link prediction is expanded
-by using \texttt{XGBoost} module of Machine Learning (ML). For the
-feature extraction from the network, the \texttt{Node2Vec} module
-was applied, although L3 information is added as a feature for improving
-link prediction. Results show that L3 feature is very important for
-the ML procedure, although a significant feature importance lies on
-the features retrieved from the network structure using \texttt{Node2Vec}.
-The mentioned approaches are compared using the human and the rice
-interactomes.
+Networks science provides a set of computational tools to identify unmapped protein-protein interactions (PPI's).
+Barab\'asi et al.
+recently introduced a metric (based on network paths of length three) which
+proved to be an important feature in predicting links of the interactome network.
+In other words, paths of length three (L3) enable us to discover hidden link candidates for experimental validation.
+This paper extends the previous approach by Barab\'asi et al.
+using \texttt{XGBoost} module of Machine Learning (ML).
+We train the model using handcrafted features as well as features learned from embeddings.
+For the feature extraction from the network, \texttt{Node2Vec} is applied.
+Results show that while L3 feature remains important for predicting links, so are the features retrieved from the network structure using \texttt{Node2Vec}.
+The mentioned approaches are compared using the human and the rice interactomes.
 \end{abstract}
@@
 compared using state-of-the-art methods, as well as the proposed ones
-(CN, A3, L3). For this report, the human interactome was used (\emph{HI-II-14.txt}).
-A curated version of the mentioned network (\emph{HI-TESTED.txt})
+(CN, A3, L3). For this report, the human interactome was used (\emph{HI-II-14}).
+A curated version of the mentioned network (\emph{HI-TESTED})
 was also used in order to compare the influence of research bias on
@@
 on and its results are consolidated and used to build a validation
-network (\emph{HI-III.txt}).
+network (\emph{HI-III}).
 
@@
 Human interactome data and base source code were downloaded from the
-repository of the length-3 degree normalized paths methodology\cite{Kovacs2019}:
-the dataset \emph{HI-II-14.txt} and \emph{HI-TESTED.txt} are used
-for prediction and the dataset \emph{HI-III.txt} is used for validation.
-Rice interactome information was downloaded from the STRING database
-\cite{Szklarczyk2019}, corresponding to the \emph{Oryza sativa} subspecies.
+repository of the length-3 degree normalized paths methodology~\cite{Kovacs2019}:
+the dataset \emph{HI-II-14} and \emph{HI-TESTED} are used
+for prediction and the dataset \emph{HI-III} is used for validation.
+Rice interactome information was downloaded from the STRING database~\cite{Szklarczyk2019}, corresponding to the \emph{Oryza sativa} subspecies.
 The downloaded file was \emph{4530.protein.links.detailed.v11.0.txt}.
@@
 
-\noindent \centering{}\includegraphics[width=0.8\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/TCP_extract\string".eps}
+\noindent \centering{}\includegraphics[width=0.8\columnwidth]{TCP_extract.png}
 \end{figure}
@@
 
-The selected parameters for the model were:
-\begin{itemize}
-\item \texttt{max\_depth=3}
-\item \texttt{colsample\_bytree=0.6}
-\item \texttt{eval\_metric='auc'}
-\end{itemize}
-
+The selected parameters for the model were \texttt{max\_depth=3}, \texttt{colsample\_bytree=0.6}
+and \texttt{eval\_metric=`auc'}.
 \subsection{Result Validation}
@@
 \begin{figure}[h]
-\caption{Methods Comparison for \emph{HI-II-14.txt}}
+\caption{Methods Comparison for \emph{HI-II-14}}
 
-\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
+\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-ii-14.png}
 \end{figure}
@@
 \begin{figure}[h]
-\caption{Methods Comparison for \emph{HI-TESTED.txt}}
+\caption{Methods Comparison for \emph{HI-TESTED}}
 
-\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
+\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-tested.png}
 \end{figure}
@@
 \bibliographystyle{plain}
-\bibliography{\string"/home/ocin/Documents/02 Asignaturas/refs\string"}
+\bibliography{\string"refs\string"}
 

### 2020-07-30 camila/Report/P5_report.tex

@@
 
-\documentclass[conference]{IEEEtran}
+\documentclass[journal, onecolumn]{IEEEtran}
 \IEEEoverridecommandlockouts
@@
 \begin{abstract}
-A gene co-expression network is an undirected graph , where each node corresponds to a gene, and a pair of nodes is connected if there is a significant co-expression relationship between them, that is, if they show similar expression patterns.
+A gene co-expression network is an undirected graph, where each node corresponds to a gene.
+A pair of nodes is connected if they show similar expression patterns.
 These co-expression networks are of biological interest since the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex.
@@
 The aim of this work is to identify rice genes that respond to saline stress.
-The methodology is based on WGCNA with a new aproach in the modules detection, using  the hierarchical link clustering (HLC) technique that allows the recognition of overlapping communities, which may have more biological meaning given the overlapping regulatory domains of systems that generate coexpression \cite{gaiteri2014beyond}.
+The methodology is based on WGCNA with a new approach in the modules detection, using  the hierarchical link clustering (HLC) technique that allows the recognition of overlapping communities, which may have more biological meaning given the overlapping regulatory domains of systems that generate co-expression \cite{gaiteri2014beyond}.
 Finally the LASSO method is used to select the most significant modules associated with rice phenotypical responses to salt stress.
@@
 %\keywords{First keyword  \and Second keyword \and Another keyword.}
+% need to focus on the main results, need to say what your main findings are - the set of genes, and how they are validated
 \end{abstract}
@@
 The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results.
-The normalization tecnique used was DESeq2 \cite{love2014moderated}.
+The normalization technique used was DESeq2 \cite{love2014moderated}.
 From the normalized data, the two biological repetitions of each accession were averaged and genes exhibiting low variance (the ratio of upper quantile to lower quantile smaller than 1.5) or low expression (more than 80\% samples with values smaller than 10) were removed.
@@
 Genes with the ratio of upper quantile to lower quantile larger than $0.25$ were kept.
-This procedure aims to remove genes with low variance in the differential expression paterns.
+This procedure aims to remove genes with low variance in the differential expression patterns.
 The final network contains $8928$ genes of the initial $57845$.
@@
 $R^2 > 0.85$).
-Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (rigth) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.
+Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.
 
@@
   \centering
-    \includegraphics[clip,width=0.4\textwidth]{Figures/pick_beta.png}
+    \includegraphics[clip,width=0.8\textwidth]{Figures/pick_beta.png}
   \caption{Degree distributions}
@@
 To identify these overlapping communities we make use of the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.
-HLC aproach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities.
+HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities.
 The algorithm maps links to nodes and connects them if a pair of links shares a node.
-They compute the similarity between links using the Jaccard index for unweigthed networks and the tanimoto coefficient for weighted networks.
+They compute the similarity between links using the Jaccard index for unweighted networks and the Tanimoto coefficient for weighted networks.
 With this similarity, they use single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities.
@@
 
-The current network represented by the adjacency matrix $A$, corresponds to a complete and weigthed network of $8,928$ genes (nodes) and $39,850,128$ edges.
+The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8,928$ genes (nodes) and $39,850,128$ edges.
 For computational reasons, this network was transformed into an unweighted one $\hat{A}$, keeping only the connections above the cutoff value of $0.2$.
@@
   \centering
-    \includegraphics[clip,width=0.5\textwidth]{Figures/phenotypic_traits.png}
+    \includegraphics[clip,width=1\textwidth]{Figures/phenotypic_traits.png}
   \caption{Phenotipic traits distribution under control and salt stress}
@@
 LASSO (Least Absolute Shrinkage Selector Operator) is a regularized linear regression technique, a method that combines a regression model with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients.
-Very usefull in problems where the number of variables (genes) $ n $ is much greater than the number of samples $ p $ ($ n \gg p $).
+Very useful in problems where the number of variables (genes) $ n $ is much greater than the number of samples $ p $ ($ n \gg p $).
 Lasso solves the least squares problem with restriction on the $ L_1$-norm of the coefficient vector:
@@
 
-Since the $\lambda$ value determines the degree of penality, the accuracy of the model depends on its choice.
-Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squered error.
+Since the $\lambda$ value determines the degree of penalty, the accuracy of the model depends on its choice.
+Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squared error.
 With that selected $\lambda$ value, the model is adjusted again, this time using all the observations.\\
 
-In the module gene selection context, the outcome variable correspond to the phenotipic trait, whereas the predictors are the modules, detected by the hierarchical clustering, represented by the first principal component of the module.
-After running the Lasso regression will select the most significant modules associated with phenotyping data.
+In the module gene selection context, the outcome variable correspond to the phenotypic trait, whereas the predictors are the modules, detected by the hierarchical clustering, represented by the first principal component of the module.
+After running the Lasso regression will select the most significant modules associated with phenotypic data.
 

### 2020-08-14 camila/Report/article_CR.tex

--- 2020-07-30 camila/Report/P5_report.tex
@@
 
-\begin{abstract}
-A gene co-expression network is an undirected graph, where each node corresponds to a gene.
-A pair of nodes is connected if they show similar expression patterns.
-These co-expression networks are of biological interest since the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex.
-Weighted gene co-expression network analysis (WGCNA)  is a method of data mining widely used to build co-expression networks, identify groups (modules) of highly correlated genes and relate these modules to external traits.
+\section*{Introduction}
+Abiotic stresses are the key factors which negatively influence plant development and productivity.
+They are the main cause of extensive agricultural production losses worldwide.
+One of the most devastating abiotic stresses, causing reduction in the cultivable land, crop quality and productivity is soil salinity.
+It has been estimated that 20\% of total cultivated and 33\% of irrigated agricultural lands worldwide are already affected by high salinity.
+Due to the human activities and natural causes, salinized areas are gradually increasing every year and are expected to reach 50\% by the end of year 2050~\cite{shrivastava2015soil}.
+Salinity effects are the result of elaborated interactions among morphological, physiological, and biochemical processes.
+Those processes are regulated by multiple genes and determine the salt tolerance or susceptibility of the crop~\cite{reddy2017salt}.
+Thus, identifying this group of stress responsive genes may lead to crop improvement in salt tolerance, which is known as a complex quantitative trait.
+Find this target genes is a complex task, because the function of many genes are still not understood and many novel non-coding genes have been discovered.
+Particularly rice (\textit{Oryza sativa}), the major food source around the world, is highly sensitive to salt stress~\cite{chang2019morphological}.
+Therefore, identification of target genes in rice may allow biologist use them as a genetic resource to develop new cultivars with resistance to salinity.\\
 
-The aim of this work is to identify rice genes that respond to saline stress.
-The methodology is based on WGCNA with a new approach in the modules detection, using  the hierarchical link clustering (HLC) technique that allows the recognition of overlapping communities, which may have more biological meaning given the overlapping regulatory domains of systems that generate co-expression \cite{gaiteri2014beyond}.
-Finally the LASSO method is used to select the most significant modules associated with rice phenotypical responses to salt stress.
-The genes differentially expressed within the selected modules are enriched with gene ontology annotations and their interaction networks are studied.
-%\keywords{First keyword  \and Second keyword \and Another keyword.}
-% need to focus on the main results, need to say what your main findings are - the set of genes, and how they are validated
-\end{abstract}
+We propose a methodology to identify stress responsive genes to salt conditions in rice.
+The methodology is based on Weighted Gene Co-expression Network Analysis (WGCNA).
+This is considered an effective and accurate bioinformatic method, using co-expression networks, that has been widely applied in identifying target genes for disease and cancer fields~\cite{tian2018identifying}.
+We follow the WGCNA workflow but with a new approach in the module detection step.
+Our modules are detected using the Hierarchical Link Clustering (HLC) technique~\cite{ahn2010link} that allows the recognition of overlapping communities, which may have more biological meaning given the overlapping regulatory domains of systems that generate co-expression~\cite{gaiteri2014beyond}.
+We conduct a systematic study with a large set of rice data using the proposed methodology.
+RNA-seq data was accessed through GEO database~\cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $92$ accessions of the Rice Diversity Panel 1.
+As the analysis result, 6 modules are detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.
+These genes may act as potential targets for the improvement of salinity tolerance in rice cultivars.
+From those 19 genes, all but 3 genes (associated with $K$ content), were also identified as deferentially expressed ($|LFC| > 2$) for at least one of the 92 accessions, suggesting that those genes are strong candidates as stress responsive genes.
+Only 2 of the 16 diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase.
+In other words, further studies are needed to elucidate the detailed biological function of the remaining 14 genes that have not been named so far,  which may have a potential relevance in stress responsive mechanisms to salt conditions in rice.\\
 
-\section*{Introduction}
-Introduction text
+With the development of high-throughput technologies, including microarrays and RNA sequencing (RNA-seq), genome-wide gene expression can be studied under different environmental stimuli (e.g.
+salt stress).
+Our methodology uses this kind of transcriptomic data measured for two different conditions (control and stress).
+After a process of normalization and filtering of the raw data, a differential expression profile of the genes is built calculating the log fold change (LFC) from control to stress condition.
+The LFC matrix will be the input for the co-expression network construction trough the WGCNA method.
+A similarity matrix is calculated using the absolute value of Pearson's correlation coefficient between pairs of genes.
+Then, the similarity matrix is forced to be a scale-free network, finding a beta exponent such that by raising each entry of the matrix to that value, the probability distribution follows a power law.
+Next, unlike WGCNA, the scale-free network is used to detect overlapping rather than non-overlapping communities, using the HLC technique.
+We also implement a LASSO regression~\cite{tibshirani1996regression} to select the most significant modules associated with rice phenotypical responses to salt stress.
+Finally, for the genes found, we look for previous evidence of important biological implications in tolerance to salt stress.
+That is, the genes deferentially expressed within the selected modules are enriched with gene ontology annotations from QuikGO database~\cite{binns2009quickgo} and their interaction networks reported in STRING database~\cite{szklarczyk2016string} are reviewed.\\
+
+The proposed methodology is modular, since other module detection and selection techniques could be used, instead HLC and LASSO respectively.
+The advantage of using HLC as clustering method is its ability to detect overlapping modules, since biological components are involved in multiple functions and therefore biological communities tend to be highly overlapping ( ).
+On the other hand, LASSO is a regularized regression technique widely used in variable selection, thanks to its ability to obtain zero regression coefficients for the less relevant variables~\cite{desboulets2018review}.
+Additionally, LASSO is especially useful in problems where the number of variables is much larger than the number of samples, which is our case having more than 5000 modules (variables) and 92 accession (samples).
+The combinations of these techniques would allow finding target genes for future biological studies that evaluate their potential as genes that respond to salt stress in rice.
+Furthermore, this study can be extended to other stresses and even to other crops.\\
 
@@
 \subsection{Data pre-processing} 
-%-----------------------------
-% RNA-seq data
-%-----------------------------
-RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $p=92$ accessions of the Rice Diversity Panel 1 (with two biological repetitions).
-\\
-
 The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results.
@@
 From the normalized data, the two biological repetitions of each accession were averaged and genes exhibiting low variance (the ratio of upper quantile to lower quantile smaller than 1.5) or low expression (more than 80\% samples with values smaller than 10) were removed.
-At this point, control and salinity stress treatment data are separated into two matrices $C=[c_{ij}]_{n \times p}$ and $T=[t_{ij}]_{n \times p}$, respectively, where $c_{ij}$ and $t_{ij}$ represent normalized expression level of the gene $i$ in the accession $j$.
+At this point, control and salinity stress treatment data are separated into two matrices $C=[c_{ij}]_{n \times p}$ and $T=[t_{ij}]_{n \times p}$, respectively, where $c_{ij}$ and $t_{ij}$ represent normalized expression level of the gene $i$ in the accession $j$.\\
 
 Changes in gene expression between control and stress conditions, are measured in terms of log ratios.
-Matrix $E=[e_{ij}]_{n \times p}$, known as the Log Fold Change matrix, is computed by setting $e_{ij}=\log_2 (t_{ij}/c_{ij})$.
+Matrix $L=[e_{ij}]_{n \times p}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$.\\
 
@@
 \subsection{Network construction}
-The co-expression network is constructed based on WGCNA, as explained below.\\
-
 The Log Fold Change matrix is used to construct the weighted co-expression network using the absolute value of the Pearson correlation as the similarity measure between genes.
@@
 Once the network has been constructed, the next step is to study its structure and dynamics identifying communities also called modules.
+The idea is to cluster genes with similar expression change patterns.
 Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e.
@@
 
-The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8,928$ genes (nodes) and $39,850,128$ edges.
+The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8928$ genes (nodes) and $39850128$ edges.
 For computational reasons, this network was transformed into an unweighted one $\hat{A}$, keeping only the connections above the cutoff value of $0.2$.
-The resulting unweighted network has a total of $5,810$ nodes and $16,875,145$ edges.
-After applying the HLC algorithm, a total of $4,131$ genes were distributed in $5,143$ overlapping modules of $3$ or more genes.
+The resulting unweighted network has a total of $5810$ nodes and $16875145$ edges.
+After applying the HLC algorithm, a total of $4131$ genes were distributed in $5143$ overlapping modules of $3$ or more genes.
 Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.\\
@@
 \subsection{Modules association to phenotypic traits}
-
 Our approach to identify the most relevant gene groups (modules) in the response to salt stress, consists in relating the co-expression network with phenotypic data from the samples.
@@
 
-%Módulos seleccionados + cantidad de genes por módulo
+In our case study, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.\\
 
@@
 % gene ontology de esos genes
-
-\section{Results}
-% redes de los genes anotados
+% redes de interacción PP de los genes nombrados
 

### 2020-08-14 nicolas/PPIPL3/ReportV1.tex

@@
 \usepackage{graphicx}
+\usepackage{todonotes}
 
@@
 \begin{abstract}
-Networks science provides a set of computational tools to identify unmapped protein-protein interactions (PPI's).
-Barab\'asi et al.
-recently introduced a metric (based on network paths of length three) which
-proved to be an important feature in predicting links of the interactome network.
-In other words, paths of length three (L3) enable us to discover hidden link candidates for experimental validation.
-This paper extends the previous approach by Barab\'asi et al.
-using \texttt{XGBoost} module of Machine Learning (ML).
-We train the model using handcrafted features as well as features learned from embeddings.
-For the feature extraction from the network, \texttt{Node2Vec} is applied.
-Results show that while L3 feature remains important for predicting links, so are the features retrieved from the network structure using \texttt{Node2Vec}.
+Previous Network-based efforts to predict unmapped protein-protein interactions (PPI's) suggest that proteins with multiple paths of length~3 (L3) are more likely to be connected.
+This paper extends this so-called L3 principle by taking into account feature extraction and using \texttt{XGBoost} techniques for prediction.
+In particular, we train the model using handcrafted features as well as features learned from embeddings using \texttt{Node2Vec}.
+Our main result shows that while L3 remains an important principle for predicting links, the approach is outperform by using embedded features.
 The mentioned approaches are compared using the human and the rice interactomes.
+
+% need include a feature that compares the dimension of the embedding is useful for prediction with L3 to show that they independent measures - we do not want to be learning L3 through node2vec
 \end{abstract}
@@
 \section{Introduction}
-
+% need to add other relevant references 
 Proteins, as a fundamental constituents of any living being and their
@@
 and protein B have complementary structures in order to interact,
-and when classical length-2 paths are used, the predicted protein
+and when classical paths of length~2 are used, the predicted protein
 interactions usually have the same structures, and not complementary
@@
 
+% please update the conclusions - WSD does not seem relevant 
 For the analyzed networks, it is also important to highlight their
@@
 \bibliographystyle{plain}
-\bibliography{\string"refs\string"}
+\bibliography{refs.bib}
 

### 2020-08-21 nicolas/PPIPL3/ReportV2.tex

--- 2020-08-14 nicolas/PPIPL3/ReportV1.tex
@@
-%% LyX 2.3.5-1 created this file.  For more info, see http://www.lyx.org/.
+%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
 %% Do not edit unless you really know what you are doing.
-\documentclass[onecolumn,english]{article}
+\documentclass[english]{article}
 \usepackage[T1]{fontenc}
 \usepackage[latin9]{inputenc}
+\usepackage{verbatim}
 \usepackage{float}
 \usepackage{graphicx}
-\usepackage{todonotes}
 
@@
 \begin{document}
-\title{Prediction of Protein-Protein Interactions on the Human and
-Rice Interactome}
-\author{Nicol\'as Antonio L\'opez Rozo}
+\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
+\author{Nicolás Antonio López Rozo}
 \maketitle
 \begin{abstract}
-Previous Network-based efforts to predict unmapped protein-protein interactions (PPI's) suggest that proteins with multiple paths of length~3 (L3) are more likely to be connected.
-This paper extends this so-called L3 principle by taking into account feature extraction and using \texttt{XGBoost} techniques for prediction.
-In particular, we train the model using handcrafted features as well as features learned from embeddings using \texttt{Node2Vec}.
-Our main result shows that while L3 remains an important principle for predicting links, the approach is outperform by using embedded features.
-The mentioned approaches are compared using the human and the rice interactomes.
 
-% need include a feature that compares the dimension of the embedding is useful for prediction with L3 to show that they independent measures - we do not want to be learning L3 through node2vec
+Previous Network-based efforts to predict unmapped protein-protein
+interactions (PPI's) suggest that proteins with multiple paths of
+length 3 (L3) are more likely to be connected. This paper extends
+this so-called L3 principle by taking into account feature extraction
+and using \texttt{XGBoost} techniques for prediction. In particular,
+we train the model using handcrafted features as well as features
+learned from embeddings using \texttt{Node2Vec}. Our main result shows
+that while L3 remains an important principle for predicting links,
+the approach is outperform by using embedded features. The mentioned
+approaches are compared using the human and the rice interactomes.
 \end{abstract}
+\begin{comment}
+need include a feature that compares the dimension of the embedding
+is useful for prediction with L3 to show that they independent measures
+- we do not want to be learning L3 through node2vec
+\end{comment}
 
 \section{Introduction}
-% need to add other relevant references 
-Proteins, as a fundamental constituents of any living being and their
-``building blocks'' in terms of biological functionality, interact
-among themselves and with other molecules according to their amino
-acid sequences and specific three-dimensional structure and folding.
-These interactions also depend on many physical and chemical properties
-of each macro-molecule and on the cellular organelle where they operate,
-such as pH and temperature. For this reason, experimental ways to
-reproduce all of the necessary conditions to validate if an interaction
-actually occurs within the cell are generally expensive.
+
+\begin{comment}
+Need to add other relevant references
+\end{comment}
+
+Proteins are the chief actors of biological functionality inside of
+the cell, and as such, do not usually work as single substances but
+rather as a part of dynamic networks, interacting with other substances
+for a wide variety of purposes \cite{Lin2017}.
+
+Protein-protein interactions (PPIs) have a key role in a variety of
+biological processes such as signal transduction, homeostasis control,
+stress responses, plant defense and organ formation. On the molecular
+level, PPIs have importance in protein phosphorylation, transcriptional
+co-factor recruitment, transporter activation and many others, thus
+playing an essential role in many physiological and developmental
+processes in virtually all organisms \cite{Zhang_2010_PPI}.
+
+Prediction of potentially relevant, yet unexplored PPIs is a current
+research topic on bioinformatics and therefore, several authors have
+proposed different methods for extrapolating information from the
+existing PPI networks. As presented by Kovacs et al (2019), information
+related to counting paths of length 3 (L3) seems to outperform state-of-the-art
+methods when it comes to predict interactions among proteins for a
+variety of model organisms such as yeast (\emph{S. cerevisiae}), Arabidopsis
+(\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
+fission yeast (\emph{S. pombe}) and mouse (\emph{M. musculus}), as
+well as on the human interactome \cite{Kovacs2019}.
 
@@
 activates by the binding of a DNA-binding Domain (DB) and an Activation
-Domain (AD) of a Transcription Factor. For the Y2H technique, a protein
-is fused to the DB domain (known as \emph{bait}) and another one to
-the AD (known as \emph{prey}). If the proteins do not interact, then
-the reporter gene is not expressed. Otherwise, the reporter gene expression
-is activated by the activation domain.
+Domain (AD) of a Transcription Factor that binds to an Upstream Activation
+Sequence (UAS). For the Y2H technique, a protein is fused to the DB
+domain (known as \emph{bait}) and another one to the AD (known as
+\emph{prey}). If the proteins do not interact, then the reporter gene
+is not expressed. Otherwise, the reporter gene expression is activated
+by the activation domain.
+
+\begin{comment}
+NL: I will make this figure myself
+\end{comment}
 
@@
 presented and their results are shown: Common Neighbors (\textbf{CN}),
-which uses the length-2 path count; length-3 path count (\textbf{A3});
-and degree-normalized length-3 paths score (\textbf{L3}).
-
-\begin{figure}[h]
-\caption{Protein-Protein Interaction Prediction Scheme}
-
-\noindent \centering{}\includegraphics[width=0.9\columnwidth]{PPI_prediction}
-\end{figure}
+which uses the length-2 path count; raw count of paths of length 3
+(\textbf{A3}); and degree-normalized length-3 paths score (\textbf{L3}).
 
@@
 shared friends that two people have, the more likely that they know
-each other. As shown in the next sections, the mentioned approach
-fails because it does not consider the structural and chemical properties
-of the proteins.
+each other. As shown by previous studies, the mentioned approach fails
+because it does not consider the structural and chemical properties
+of the proteins \cite{Kovacs2019}.
 
-For achieving the described results, several networks are used and
-compared using state-of-the-art methods, as well as the proposed ones
-(CN, A3, L3). For this report, the human interactome was used (\emph{HI-II-14}).
-A curated version of the mentioned network (\emph{HI-TESTED})
-was also used in order to compare the influence of research bias on
-the network predictions. A massive experimental assay was carried
-on and its results are consolidated and used to build a validation
-network (\emph{HI-III}).
+For achieving the described results, human and rice PPI networks are
+used and compared using state-of-the-art methods, as well as the proposed
+ones (CN, A3, L3). In the case of the human network, human interactome
+(\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED})
+were used. A massive experimental assay was carried on and its results
+were consolidated and used to build a validation network (\emph{HI-III}).
 
@@
 Human interactome data and base source code were downloaded from the
-repository of the length-3 degree normalized paths methodology~\cite{Kovacs2019}:
-the dataset \emph{HI-II-14} and \emph{HI-TESTED} are used
-for prediction and the dataset \emph{HI-III} is used for validation.
-Rice interactome information was downloaded from the STRING database~\cite{Szklarczyk2019}, corresponding to the \emph{Oryza sativa} subspecies.
+repository of the length-3 degree normalized paths methodology \cite{Kovacs2019}:
+the dataset \emph{HI-II-14} and \emph{HI-TESTED} are used for prediction
+and the dataset \emph{HI-III} is used for validation.
+
+Rice interactome information was downloaded from the STRING database
+\cite{Szklarczyk2019}, corresponding to the \emph{Oryza sativa} subspecies.
 The downloaded file was \emph{4530.protein.links.detailed.v11.0.txt}.
 and contains more than 8 million PPIs from several resources. For
-the purpose of this study and based on the previous work, only PPIs
-with evidence from curated databases were used (rows where \emph{databases}
-column has a value greater than zero), resulting in a network with
-NNNN nodes and MMMM edges.
+the purpose of this study and based on previous work, only PPIs with
+evidence from curated databases were used (i.e. rows where the column
+\emph{databases} has a value greater than zero), resulting in a network
+with $5025$ nodes and $164420$ edges.
 
@@
 Previous code implementation was adapted from \texttt{C++} to \texttt{Python}
-(3.6), in order to unify the algorithms into one single script. For
+(V3.6), in order to unify the algorithms into one single script. For
 the purpose of algorithmic validation, the three methods were implemented
 from scratch with basic functionalities and data structures of the
-\emph{Python} (V3.6) language.
+\texttt{Python} language.
 
@@
 to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
-$9433$ and $23322$ interactions. For the rice interactome, an additional
-preprocessing was performed. The filtered network for rice consists
-of $5025$ proteins (nodes) and $164420$ interactions (edges) distributed
-among $178$ connected components. The connected component with the
-greatest number of edges was selected in this case.
+$9433$ and $23322$ interactions.
 
-The extracted connected component consists of $n=4390$ nodes and
-$m=163319$ edges, which corresponds to $99.33%\%
+For the rice interactome, an additional preprocessing was performed.
+The filtered network for rice consists of $5025$ proteins (nodes)
+and $164420$ interactions (edges) distributed among $178$ connected
+components. The connected component with the greatest number of edges
+was selected in this case. The extracted connected component consists
+of $n=4390$ nodes and $m=163319$ edges, which corresponds to $99.33%\%
 $ of filtered edges. Further investigation is applied to this network,
@@
 represents and interaction among proteins $v_{i}$ and $v_{j}$.
-
-\begin{figure}[h]
-\caption{Edge Prediction Paradigms}
-
-\noindent \centering{}\includegraphics[width=0.8\columnwidth]{TCP_extract.png}
-\end{figure}
-
 \begin{description}
@@
 
-\subsection{Sampling and Precision Counting}
+\subsection{Sampling Procedure}
 
-Sampling proceeds as follows: first, a fraction (10\%) of the edges
-is removed at random from the network. Then, for each method (\textbf{CN},
-\textbf{A3}, \textbf{L3}) the prediction of the removed fraction is
-performed. The new edges are then compared against the original (removed)
-edges according to the ranking that each method yields. Edges belonging
-to the original connected component are considered positive in the
-sense that they are experimentally validated interactions. For each
-of the tested networks and a rank $r$, the precision was calculated
-as 
-\[
-P_{r}=\frac{L_{r}}{r}
-\]
-\\
-where $L_{r}$ is the total number of positive edges in the range
-$[1,r]$.
-
-The previous procedure was applied for a fraction of 0.90 remaining
-edges, and a mean of 10 simulations was obtained.
+For each network of protein interactions, the following procedure
+was performed 10 times in order to address the stochastic nature of
+the process and have a consensus:
+\begin{itemize}
+\item A percentage of interactions is removed at random from the network
+(20\%).
+\item The same amount of removed interactions are then predicted using the
+main methods for prediction mentioned by Kovacs et al (2019): Common
+Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
+and the Length-3 degree-normalized score (\textbf{L3}).
+\item A test dataset is created as follows: all removed edges are included
+(as observed positives for the ML algorithm) and from the predicted
+edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
+the previous classification (observed negatives), a random subset
+is chosen such that the dataset is balanced, that is, the amount of
+observed positive labels is equal to the observed negative labels.
+\item Once the dataset is ready, it is randomly partitioned: 80\% is used
+for \texttt{XGBoost} model training and 20\% is used for validation.
+It is important to have in mind that balanced distribution of the
+positive and negative labels in the datasets was satisfied.
+\end{itemize}
 
@@
 
-The selected parameters for the model were \texttt{max\_depth=3}, \texttt{colsample\_bytree=0.6}
-and \texttt{eval\_metric=`auc'}.
+The selected parameters for the model were:\texttt{ max\_depth=3},
+\texttt{colsample\_bytree=0.6} and \texttt{eval\_metric=''auc''}.
+
 \subsection{Result Validation}
 
-For applying the test results, a partition of the whole dataset was
-made: 80\% of the data was randomly selected and used for training,
-while the remaining 20\% was used for validation. The whole training-validation
-procedure was applied 5 times.
+As mentioned before, 80\% of the final dataset was randomly selected
+and used for training, while the remaining 20\% was used for validation.
+The whole training-validation procedure was applied 10 times.
 
@@
 
-\section{Results}
+\section{Results and Discussion}
 
-As explained above, the precision trend for each of the networks are
-presented for the predicted top 2000 interactions.
+\subsection{Rice Interactome}
+
+For the rice interactome, the different model-features combinations
+were trained and validated. After executing the mentioned routines,
+the results are shown in Figure \ref{F1}. First, one should have
+a baseline of comparison, which in this case corresponds to \texttt{Node2Vec}
+without any additional feature included. The plot below shows those
+results, and one can see that its mean performance using the AUC metric
+is 0.52, and that the results among the 10 repetitions are consistent.
+This results mean that the model using only the default features perform
+barely as good as a random choice of the labels.
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F1}Summary ROC curves for \texttt{Node2Vec} model alone}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.95\columnwidth]{Only_ML/ROConlyML_SUMMARY}
+\end{figure}
+
+\section{(PENDING FROM HERE ON)}
 
@@
 
-\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-ii-14.png}
+\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
 \end{figure}
@@
 
-\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-tested.png}
+\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
 \end{figure}
@@
 and protein B have complementary structures in order to interact,
-and when classical paths of length~2 are used, the predicted protein
+and when classical paths of length 2 are used, the predicted protein
 interactions usually have the same structures, and not complementary
@@
 
-% please update the conclusions - WSD does not seem relevant 
+\begin{comment}
+please update the conclusions - WSD does not seem relevant
+\end{comment}
+
 For the analyzed networks, it is also important to highlight their
@@
 \bibliographystyle{plain}
-\bibliography{refs.bib}
+\bibliography{refs}
 

### 2020-08-28 camila/Report/article_CR.tex

@@
 
-\documentclass[journal, onecolumn]{IEEEtran}
+\documentclass[12pt,journal, onecolumn]{IEEEtran}
 \IEEEoverridecommandlockouts
@@
 \usepackage{amsmath, amsthm, amsfonts, amssymb}
+
+\usepackage{mathtools}
 
@@
 The proposed methodology is modular, since other module detection and selection techniques could be used, instead HLC and LASSO respectively.
-The advantage of using HLC as clustering method is its ability to detect overlapping modules, since biological components are involved in multiple functions and therefore biological communities tend to be highly overlapping ( ).
+The advantage of using HLC as clustering method is its ability to detect overlapping modules, since biological components are involved in multiple functions and therefore biological communities tend to be highly overlapping.
 On the other hand, LASSO is a regularized regression technique widely used in variable selection, thanks to its ability to obtain zero regression coefficients for the less relevant variables~\cite{desboulets2018review}.
@@
 \section{Methodology}
+
+This methodology uses as input data RNA-seq read counts, representing gene expression levels.
+More precisely, $n$ gene expression profiles of an organism, measured for $m$ different genotypes under control and treatment conditions, and $r$ biological replicates.
+This data can be represented as a matrix $D=[d_{ij}]$, where $d_{ij} \in \mathbb{N}_0$ for all $i = 1,2,..n$ and $j=1,2,..,2mr$.
+To discover key genes and their interaction for treatment-tolerance related phenotypes, the methodology also requires a set of $p$ phenotypic traits, measured for the $m$ genotypes.
+The phenotypic data can be seen as a matrix $P=[p_{jk}]$, where $p_{jk} \in \mathbb{R}$ for all $k = 1,2,...,p$.
+
 \subsection{Data pre-processing}
+
 The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results.
-The normalization technique used was DESeq2 \cite{love2014moderated}.
-From the normalized data, the two biological repetitions of each accession were averaged and genes exhibiting low variance (the ratio of upper quantile to lower quantile smaller than 1.5) or low expression (more than 80\% samples with values smaller than 10) were removed.
-At this point, control and salinity stress treatment data are separated into two matrices $C=[c_{ij}]_{n \times p}$ and $T=[t_{ij}]_{n \times p}$, respectively, where $c_{ij}$ and $t_{ij}$ represent normalized expression level of the gene $i$ in the accession $j$.\\
+In order to correct library size and RNA composition bias, the suggested normalization technique is DESeq2 \cite{love2014moderated}.
+Then, from the normalized data, average the biological replicates of each accession and remove genes exhibiting low variance or low expression.
+Next, separate control and salinity stress treatment data into two matrices $C=[c_{ij}]_{n \times m}$ and $T=[t_{ij}]_{n \times m}$, respectively, where $c_{ij}$ and $t_{ij}$ represent normalized expression level of the gene $i$ in the accession $j$.\\
 
-Changes in gene expression between control and stress conditions, are measured in terms of log ratios.
-Matrix $L=[e_{ij}]_{n \times p}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$.\\
+Next step is to compile the control and treatment data by measuring the changes in expression levels in terms of logarithmic ratios.
+Matrix $L=[\ell_{ij}]_{n \times p}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$.
+Finally, filter $L$ matrix by removing rows (genes) with low variance in the differential expression patterns.
 
-Genes with the ratio of upper quantile to lower quantile larger than $0.25$ were kept.
-This procedure aims to remove genes with low variance in the differential expression patterns.
-The final network contains $8928$ genes of the initial $57845$.
+\subsection{Co-expression network construction}
 
-\subsection{Network construction}
-The Log Fold Change matrix is used to construct the weighted co-expression network using the absolute value of the Pearson correlation as the similarity measure between genes.
-The matrix $S=[s_{ij}]_{n\times n}$ measures the level of concordance between gene expression profiles across experiments and is transformed into an adjacency matrix $A=[a_{ij}]_{n\times n}$ where $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes.
+The Log fold change matrix $L$ is used to build the co-expression network following the first steps of the WGCNA methodology.
+First, measure the level of concordance between gene differential expression profiles across samples.
+Use the absolute value of the Pearson correlation coefficient as the similarity measure between genes and store the values in the similarity matrix $S=[s_{ij}]_{n\times n}$.\\
+
+Then, transform $S$ into an adjacency matrix $A=[a_{ij}]_{n\times n}$ where $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes.
 In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network.
 This kind of networks contain many nodes with very few connections and a small number of hubs with high connections.
-\\
-
 In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node).
@@
 $R^2 > 0.85$).
+\\
+
+Finally, prepare matrix $A$ to apply the clustering algorithm, transforming it into an unweighted network $\hat{A}$.
+To determine the Pearson Correlation Coefficient (PCC) cutoff for finding biologically relevant co-expressed modules, we use the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values.\\
+
+The matrix $\hat{A}$ can be seen as an undirected graph $G=(V,E)$ where ${V=\{v_1,v_2,\ldots,v_n\}}$ is a set of vertices or nodes and ${E=\{e_1,e_2,\ldots,e_p\}}$ is a set of edges or links that connect the vertices.
+In this genomic context, the graph $G$ represents a co-expression network, where each node corresponds to a gene and a pair of genes is connected if they show similar differential expression patterns.
+
+\subsection{Co-expression module identification}
+
+Next step in the methodology is to study the co-expression network structure and dynamics identifying communities also called modules.
+The idea is to cluster genes with similar differential expression change patterns.
+Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e.
+genes) are involved in multiple functions.
+Thus, unlike WGCNA, the co-expression network $G$ is used to detect overlapping rather than non-overlapping communities, using the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.\\
+
+HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities.
+The algorithm maps links to nodes and connects them if a pair of links shares a node.
+They compute the similarity between links using the Jaccard index for unweighted networks.
+With this similarity, they use single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities.
+Finally, the most relevant communities are established at the maximal partition density, a function  based on link density inside communities.
+
+\subsection{Modules association to phenotypic traits}
+
+To identify the most relevant gene groups (modules), associated with the phenotypic response to a specific treatment in an organism, we use a LASSO based approach.
+Each module can be represented by a eigengene, which is defined as the first principal component of such module.
+A eigengene can be thought of as an average differential expression profile for each community and is computed from the Log Fold Change matrix rows corresponding to the genes belonging to a specific module.
+\\
+
+These profiles are then associated with each phenotypic trait using the least absolute shrinkage and selection operator (LASSO).
+LASSO combines a regression model with a procedure of contraction of some parameters towards zero, imposing a restriction or a penalty on the regression coefficients.
+This technique allows selection of the most relevant variables.
+In our context, the eigenges act as regressor variables and each phenotypic trait its used as an outcome variable.
+\\
+
+The output after applying LASSO is a set of modules for each phenotypic trait. The union of genes belonging to the selected modules are the target genes for downstream analysis.
+
+\subsection{Genes enrichment}
+
+For the genes found, first identify the differential expressed ones, that is genes showing an absolute value of the log fold change greater than 2 ($|\ell_{ij}|\geq 2$) for at least one sample.
+This represents genes whose level of expression is quadrupled (up or down) from control to treatment condition, suggesting that those genes are strong candidates as treatment responsive genes.\\
+
+You also can perform a functional category enrichment looking for the Gene Ontology (GO) annotations from databases like QuikGO~\cite{binns2009quickgo}.
+This annotations can provide evidence of biological implications of the target genes in the treatment-tolerance mechanisms.\\
+
+Those named genes (with GO annotations), can be used to performe another relevant analysis reviewing their reported protein-protein interaction networks using the STRING database~\cite{szklarczyk2016string}.
+The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge transfer between organisms, and from interactions aggregated from other (primary) databases.
+This information elucidates how the selected genes are involved in functional pathways that can be related with the treatment of interest.
+
+\section{Case study}
+RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $m=92$ accessions of the Rice Diversity Panel 1, with two biological repetitions ($r=2$).
+\\
+
+Genes exhibiting low variance (the ratio of upper quantile to lower quantile smaller than 1.5) or low expression (more than 80\% samples with values smaller than 10) were removed.
+The final network contains $8928$ genes of the initial $57845$.\\
+
 Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.
@@
 
-\subsection{Module detection}	
-Once the network has been constructed, the next step is to study its structure and dynamics identifying communities also called modules.
-The idea is to cluster genes with similar expression change patterns.
-Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e.
-genes) are involved in multiple functions.\\
-
-To identify these overlapping communities we make use of the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.
-HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities.
-The algorithm maps links to nodes and connects them if a pair of links shares a node.
-They compute the similarity between links using the Jaccard index for unweighted networks and the Tanimoto coefficient for weighted networks.
-With this similarity, they use single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities.
-Finally, the most relevant communities are established at the maximal partition density, a function  based on link density inside communities.\\
-
 The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8928$ genes (nodes) and $39850128$ edges.
@@
 After applying the HLC algorithm, a total of $4131$ genes were distributed in $5143$ overlapping modules of $3$ or more genes.
-Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.\\
+Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.
 
@@
 
-\subsection{Modules association to phenotypic traits}
-Our approach to identify the most relevant gene groups (modules) in the response to salt stress, consists in relating the co-expression network with phenotypic data from the samples.
-We use 3 phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass.
-These were measured for the 92 genotypes studied, under  control and salt stress conditions.
-These data can be found in the supplementary information of~\cite{campbell2017allelic}.
 As shown in Figure \ref{fig:pdata}, there are significant differences in the values of these phenotypic traits between both stress and control conditions.
-This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.\\
+This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.
 
@@
 
-In order to relate the modules to the phenotypic traits, we represent each module with the first principal component of the Log Fold Change submatrix, corresponding to the genes of such module, which can be thought of as an average differential expression profile for each community.
-These profiles are then associated with each phenotypic trait using the LASSO variable selection method, allowing to identify a set of relevant modules related to the response to salinity conditions in rice plants.
-See Appendix A for a LASSO method description.\\
+We use 3 phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass.
+These were measured for the 92 genotypes studied, under  control and salt stress conditions.
+These data can be found in the supplementary information of~\cite{campbell2017allelic}.\\
 
-In our case study, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.\\
-
-\subsection{Genes enrichment}
-% genes seleccionados con expresión diferencial según análisis DEA. 
-% gene ontology de esos genes
-% redes de interacción PP de los genes nombrados
+Finally, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.
+From those 19 genes, all but 3 genes (associated with $K$ content), were also identified as deferentially expressed ($|LFC| > 2$) for at least one of the 92 accessions.
+Only 2 of the 16 diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase.
 

### 2020-09-11 camila/Report/article_CR.tex
paper draft

@@
 \usepackage{array}
+
+\usepackage{subcaption} % subfigures
 
@@
 These genes may act as potential targets for the improvement of salinity tolerance in rice cultivars.
-From those 19 genes, all but 3 genes (associated with $K$ content), were also identified as deferentially expressed ($|LFC| > 2$) for at least one of the 92 accessions, suggesting that those genes are strong candidates as stress responsive genes.
+From those 19 genes, all but 3 genes (associated with $K$ content), were also identified as deferentially expressed (with a log fold change value greater than $2$) for at least one of the 92 accessions, suggesting that those genes are strong candidates as stress responsive genes.
 Only 2 of the 16 diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase.
@@
 
-\section{Methodology}
+\section{Preliminaries}
 
-This methodology uses as input data RNA-seq read counts, representing gene expression levels.
-More precisely, $n$ gene expression profiles of an organism, measured for $m$ different genotypes under control and treatment conditions, and $r$ biological replicates.
-This data can be represented as a matrix $D=[d_{ij}]$, where $d_{ij} \in \mathbb{N}_0$ for all $i = 1,2,..n$ and $j=1,2,..,2mr$.
-To discover key genes and their interaction for treatment-tolerance related phenotypes, the methodology also requires a set of $p$ phenotypic traits, measured for the $m$ genotypes.
-The phenotypic data can be seen as a matrix $P=[p_{jk}]$, where $p_{jk} \in \mathbb{R}$ for all $k = 1,2,...,p$.
+\subsection{Co-expression network}
+A network is an undirected graph $G=(V,E)$ where ${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of vertices or nodes and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of edges or links that connect the vertices.
+In a gene co-expression network each node corresponds to a gene and a pair of genes is connected if they show similar differential expression patterns.
+These co-expression networks are of biological interest since the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex.\\
 
-\subsection{Data pre-processing}
+When the network is simple and unweighted, it can be represented by an adjacency matrix $A \in \{0,1\}^{n \times n}$.
+This adjacency matrix is symmetric, with a positive one in the positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever there is an edge connecting vertices $v_i$ and $v_j$, and zeros elsewhere.\\
 
-The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results.
-In order to correct library size and RNA composition bias, the suggested normalization technique is DESeq2 \cite{love2014moderated}.
-Then, from the normalized data, average the biological replicates of each accession and remove genes exhibiting low variance or low expression.
-Next, separate control and salinity stress treatment data into two matrices $C=[c_{ij}]_{n \times m}$ and $T=[t_{ij}]_{n \times m}$, respectively, where $c_{ij}$ and $t_{ij}$ represent normalized expression level of the gene $i$ in the accession $j$.\\
+\subsection{Hierarchical Link Clustering}
 
-Next step is to compile the control and treatment data by measuring the changes in expression levels in terms of logarithmic ratios.
-Matrix $L=[\ell_{ij}]_{n \times p}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$.
-Finally, filter $L$ matrix by removing rows (genes) with low variance in the differential expression patterns.
-
-\subsection{Co-expression network construction}
-
-The Log fold change matrix $L$ is used to build the co-expression network following the first steps of the WGCNA methodology.
-First, measure the level of concordance between gene differential expression profiles across samples.
-Use the absolute value of the Pearson correlation coefficient as the similarity measure between genes and store the values in the similarity matrix $S=[s_{ij}]_{n\times n}$.\\
-
-Then, transform $S$ into an adjacency matrix $A=[a_{ij}]_{n\times n}$ where $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes.
-In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network.
-This kind of networks contain many nodes with very few connections and a small number of hubs with high connections.
-In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node).
-So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g.
-$R^2 > 0.85$).
-\\
-
-Finally, prepare matrix $A$ to apply the clustering algorithm, transforming it into an unweighted network $\hat{A}$.
-To determine the Pearson Correlation Coefficient (PCC) cutoff for finding biologically relevant co-expressed modules, we use the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values.\\
-
-The matrix $\hat{A}$ can be seen as an undirected graph $G=(V,E)$ where ${V=\{v_1,v_2,\ldots,v_n\}}$ is a set of vertices or nodes and ${E=\{e_1,e_2,\ldots,e_p\}}$ is a set of edges or links that connect the vertices.
-In this genomic context, the graph $G$ represents a co-expression network, where each node corresponds to a gene and a pair of genes is connected if they show similar differential expression patterns.
-
-\subsection{Co-expression module identification}
-
-Next step in the methodology is to study the co-expression network structure and dynamics identifying communities also called modules.
-The idea is to cluster genes with similar differential expression change patterns.
-Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e.
-genes) are involved in multiple functions.
-Thus, unlike WGCNA, the co-expression network $G$ is used to detect overlapping rather than non-overlapping communities, using the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.\\
-
+Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn et al.~\cite{ahn2010link}.
 HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities.
 The algorithm maps links to nodes and connects them if a pair of links shares a node.
-They compute the similarity between links using the Jaccard index for unweighted networks.
-With this similarity, they use single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities.
-Finally, the most relevant communities are established at the maximal partition density, a function  based on link density inside communities.
+The similarity between two links $e_{ik}$ and $e_{jk}$ is computed using the Jaccard index
+\begin{equation*}
+\label{eq:jaccard}
+S(e_{ik},e_{jk}) = \frac{\vert n_+(i) \cap n_+(j) \vert}{\vert n_+(i) \cup n_+(j) \vert},
+\end{equation*}
+where $n_+(i)$ denotes the set of node $i$ and its neighbors.\\
 
-\subsection{Modules association to phenotypic traits}
+With this similarity, the method uses single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities.
+Hierarchical clustering methods repeatedly merge groups until all elements are members of a single cluster.
+To find meaningful communities it is crucial to know where to partition the dendrogram.
+Thus, the most relevant communities are established at the maximal partition density $D$, a function  based on link density inside communities that measures the quality of a link partition.
+The partition density has a single global maximum along the dendrogram in almost all cases, because the value is just the average density at the top of the dendrogram (a single giant community with every link and node) and it is very small at the bottom of the dendrogram (most communities consists of a single link).
+$D = 1$ when every community is a fully connected clique and $D = 0$ when each community is a tree.
+If a link community is less dense than a tree (when the community subgraph has disconnected components), then that community will give a negative contribution to $D$.
+Computing $D$ at each level of the link dendrogram allows to pick the best level to cut (although meaningful structure exists above and below that threshold).
+This way, the final output corresponds to node clusters, where each node can participate in multiple communities.\\
 
-To identify the most relevant gene groups (modules), associated with the phenotypic response to a specific treatment in an organism, we use a LASSO based approach.
-Each module can be represented by a eigengene, which is defined as the first principal component of such module.
-A eigengene can be thought of as an average differential expression profile for each community and is computed from the Log Fold Change matrix rows corresponding to the genes belonging to a specific module.
-\\
-
-These profiles are then associated with each phenotypic trait using the least absolute shrinkage and selection operator (LASSO).
-LASSO combines a regression model with a procedure of contraction of some parameters towards zero, imposing a restriction or a penalty on the regression coefficients.
-This technique allows selection of the most relevant variables.
-In our context, the eigenges act as regressor variables and each phenotypic trait its used as an outcome variable.
-\\
-
-The output after applying LASSO is a set of modules for each phenotypic trait. The union of genes belonging to the selected modules are the target genes for downstream analysis.
-
-\subsection{Genes enrichment}
-
-For the genes found, first identify the differential expressed ones, that is genes showing an absolute value of the log fold change greater than 2 ($|\ell_{ij}|\geq 2$) for at least one sample.
-This represents genes whose level of expression is quadrupled (up or down) from control to treatment condition, suggesting that those genes are strong candidates as treatment responsive genes.\\
-
-You also can perform a functional category enrichment looking for the Gene Ontology (GO) annotations from databases like QuikGO~\cite{binns2009quickgo}.
-This annotations can provide evidence of biological implications of the target genes in the treatment-tolerance mechanisms.\\
-
-Those named genes (with GO annotations), can be used to performe another relevant analysis reviewing their reported protein-protein interaction networks using the STRING database~\cite{szklarczyk2016string}.
-The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge transfer between organisms, and from interactions aggregated from other (primary) databases.
-This information elucidates how the selected genes are involved in functional pathways that can be related with the treatment of interest.
-
-\section{Case study}
-RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $m=92$ accessions of the Rice Diversity Panel 1, with two biological repetitions ($r=2$).
-\\
-
-Genes exhibiting low variance (the ratio of upper quantile to lower quantile smaller than 1.5) or low expression (more than 80\% samples with values smaller than 10) were removed.
-The final network contains $8928$ genes of the initial $57845$.\\
-
-Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.
-
-%scale-free beta 1 y 3
-\begin{figure}[h]
-  \centering
-    \includegraphics[clip,width=0.8\textwidth]{Figures/pick_beta.png}
-  \caption{Degree distributions}
-  \label{fig:beta}
-\end{figure}
-
-The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8928$ genes (nodes) and $39850128$ edges.
-For computational reasons, this network was transformed into an unweighted one $\hat{A}$, keeping only the connections above the cutoff value of $0.2$.
-The resulting unweighted network has a total of $5810$ nodes and $16875145$ edges.
-After applying the HLC algorithm, a total of $4131$ genes were distributed in $5143$ overlapping modules of $3$ or more genes.
-Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.
-
-%overlapping percentage
-\begin{figure}[h]
-  \centering
-    \includegraphics[clip,width=0.4\textwidth]{Figures/artificial_modules.png}
-  \caption{Overlapping percentage}
-  \label{fig:overlap}
-\end{figure}
-
-As shown in Figure \ref{fig:pdata}, there are significant differences in the values of these phenotypic traits between both stress and control conditions.
-This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.
-
-% boxplots
-\begin{figure}[h]
-  \centering
-    \includegraphics[clip,width=1\textwidth]{Figures/phenotypic_traits.png}
-  \caption{Phenotipic traits distribution under control and salt stress}
-  \label{fig:pdata}
-\end{figure}
-
-We use 3 phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass.
-These were measured for the 92 genotypes studied, under  control and salt stress conditions.
-These data can be found in the supplementary information of~\cite{campbell2017allelic}.\\
-
-Finally, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.
-From those 19 genes, all but 3 genes (associated with $K$ content), were also identified as deferentially expressed ($|LFC| > 2$) for at least one of the 92 accessions.
-Only 2 of the 16 diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase.
-
-\section{Discussion}
-
-\bibliographystyle{acm}
-\bibliography{mybibliography}
-
-\newpage
-\section*{Appendix}
-\subsection{A: Variable selection with LASSO}
-LASSO (Least Absolute Shrinkage Selector Operator) is a regularized linear regression technique, a method that combines a regression model with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients.
-Very useful in problems where the number of variables (genes) $ n $ is much greater than the number of samples $ p $ ($ n \gg p $).
+\subsection{Least Absolute Shrinkage Selector Operator}
+Least Absolute Shrinkage Selector Operator (LASSO) is a regularized linear regression technique, a method that combines a regression model with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients.
+Very useful in problems where the number of variables (genes) $ n $ is much greater than the number of samples $ m $ ($ n \gg m $).
 Lasso solves the least squares problem with restriction on the $ L_1$-norm of the coefficient vector:
@@
 \begin{equation}
-\min \left\lbrace\sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} \right\rbrace , \textrm{sujeto a} \sum_{j=1}^n\abs{\beta_j}\leq s
+\min \left\lbrace\sum_{i=1}^{m}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} \right\rbrace , \textrm{subject to} \sum_{j=1}^n\abs{\beta_j}\leq s
 \end{equation}
@@
 Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squared error.
-With that selected $\lambda$ value, the model is adjusted again, this time using all the observations.\\
+With that selected $\lambda$ value, the model is adjusted again, using all the observations.\\
 
-In the module gene selection context, the outcome variable correspond to the phenotypic trait, whereas the predictors are the modules, detected by the hierarchical clustering, represented by the first principal component of the module.
-After running the Lasso regression will select the most significant modules associated with phenotypic data.
+\section{Methodology}
+
+This methodology uses as input data RNA-seq read counts, representing gene expression levels.
+More precisely, $n_0$ gene expression profiles of an organism, measured for $m$ different genotypes under control and treatment conditions, and $r$ biological replicates.
+This raw data can be represented as a matrix $D_0 \in {\mathbb{N}_0}^{n_0 \times 2mr}$. 
+In order to discover key genes and their interaction with phenotypes related to treatment tolerance, the methodology also requires a set of $p$ phenotypic traits, measured for the $m$ genotypes.
+The phenotypic data can be seen as a matrix $P \in \mathbb{R}^{2m \times p}$ containing two phenotypic values per genotype, one under control condition and the second under treatment condition.\\
+
+\subsection{Data pre-processing}
+
+The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results.
+In order to correct library size and RNA composition bias, the suggested normalization technique is DESeq2 \cite{love2014moderated}.
+From the normalized data $D_1 \in \mathbb{R}^{n_0 \times 2mr}$, average the biological replicates of each genotype, so that the data matrix becomes $D_2 \in \mathbb{R}^{n_0 \times 2m}$.
+Next, remove genes exhibiting low variance or low expression, getting a new number of $n_1 \leq n_0$ genes.
+Separate control and treatment data into two matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1 \times m}$, respectively.
+Each matrix entry, $c_{ij}$ for control and $t_{ij}$ for treatment, represent the normalized expression level of gene $i$ in accession $j$.\\
+
+Now, compile the control and treatment data by measuring the changes in expression levels in terms of logarithmic ratios.
+Matrix $L_0 \in \mathbb{R}^{n_1 \times m}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$.
+Repeat this process on the phenotypic data separating the control and treatment data into matrices $P_c$ and $P_t$ of dimensions $m \times p$.
+Calculate the corresponding logarithmic ratios, obtaining the matrix $P_\ell \in \mathbb{R}^{m \times p}$.
+Finally, filter $L_0$ matrix by removing rows (genes) with low variance in the differential expression patterns, getting a new matrix $L_1$ of dimensions $n_2 \times m$, with $n_2 \leq n_1$.\\
+
+\subsection{Co-expression network construction}
+
+The Log fold change matrix $L_1$ is used to build the co-expression network following the first steps of the WGCNA methodology.
+First, measure the level of concordance between gene differential expression profiles across samples.
+Use the absolute value of the Pearson correlation coefficient as the similarity measure between genes and store the values in the similarity matrix $S\in \mathbb{R_{+}}^{n_2 \times n_2}$.\\
+
+Then, transform $S$ into an adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes.
+In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network.
+This kind of networks contain many nodes with very few connections and a small number of hubs with high connections.
+In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node).
+So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g.
+$R^2 > 0.85$).
+\\
+
+Finally, prepare matrix $A$ to apply the clustering algorithm, transforming it into an unweighted network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$.
+To determine the Pearson Correlation Coefficient (PCC) cutoff for finding biologically relevant co-expressed modules, we use the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values.
+Above the cutoff, the entries of matrix $A$ become $1$ and when the value is below the threshold it becomes zero.\\
+
+\subsection{Co-expression module identification}
+
+Next step in the methodology is to study the co-expression network structure and dynamics identifying communities also called modules.
+The idea is to cluster genes with similar differential expression change patterns.
+Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e.
+genes) are involved in multiple functions.
+Thus, unlike WGCNA, the adjacency matrix $\hat{A}$, from the co-expression network $G$, is used to detect overlapping rather than non-overlapping communities, using the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.\\
+
+%HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple overlapping communities.
+The algorithm maps links to nodes and connects them if a pair of links shares a node.
+The similarity between links is computed using the Jaccard index.
+With this similarity, they use single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities.
+Finally, the most relevant communities are established at the maximal partition density, a function  based on link density inside communities.\\
+
+The HLC algorithm organizes the $n_2$ genes of matrix $\hat{A}$ into $c$ communities, where each gene can belong to one, multiple, or no community.
+This information is represented as an affiliation matrix $F \in \{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ if node $i$ is member of cluster $u$.\\
+
+\subsection{Modules association to phenotypic traits}
+
+To identify the most relevant gene groups (modules), associated with the phenotypic response to a specific treatment in an organism, we use a LASSO based approach.
+Each module can be represented by a eigengene, which is defined as the first principal component of such module.
+A eigengene can be thought of as an average differential expression profile for each community and is computed from the Log Fold Change Matrix $L_1$ and the affiliation matrix $F$.
+Given a module $u$, use the affiliation matrix to identify the genes belonging to $u$ and then select the corresponding rows of the matrix $L_1$ to compute the first principal component of $u$.
+Each principal component would be a column of the matrix $M \in \mathbb{R}^{m \times c}$.\\
+
+% the Log Fold Change matrix rows corresponding to the genes belonging to a specific module. \\
+
+These profiles are then associated with each phenotypic trait using the least absolute shrinkage and selection operator (LASSO).
+LASSO combines a regression model with a procedure of contraction of some parameters towards zero, imposing a restriction or a penalty on the regression coefficients.
+This technique allows selection of the most relevant variables.
+In our context, the eigenges (the columns of $M$) act as regressor variables and each phenotypic trait (each column of $P_\ell$) is used as an outcome variable.
+\\
+
+The output after applying LASSO is a set of modules for each phenotypic trait. The union of genes belonging to the selected modules are the target genes for downstream analysis.
+
+\subsection{Genes enrichment}
+
+For the genes found, first identify the differential expressed ones, that is genes showing an absolute value of the log fold change greater than 2 ($|\ell_{ij}|\geq 2$) for at least one sample.
+This represents genes whose level of expression is quadrupled (up or down) from control to treatment condition, suggesting that those genes are strong candidates as treatment responsive genes.\\
+
+You also can perform a functional category enrichment searching for the Gene Ontology (GO) annotations from databases like QuikGO~\cite{binns2009quickgo}.
+This annotations can provide evidence of biological implications of the target genes in the treatment-tolerance mechanisms.\\
+
+Those named genes (with GO annotations), can be used to perform another relevant analysis reviewing their reported protein-protein interaction networks using the STRING database~\cite{szklarczyk2016string}.
+The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge transfer between organisms, and from interactions aggregated from other (primary) databases.
+This information elucidates how the selected genes are involved in functional pathways that can be related with the treatment of interest.\\
+
+\section{Case study}
+RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n_0=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $m=92$ accessions of the Rice Diversity Panel 1, with $r=2$ biological replicates.
+On the other hand, we use $p=3 $ phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass.
+This traits were measured for the same $92$ genotypes, under  control and salt stress conditions, and can be found in the supplementary information of~\cite{campbell2017allelic}.\\
+
+\subsection{Data pre-processing}
+The DESeq normalization was applied to the raw data and the biological replicates were averaged.
+Genes exhibiting low variance were identified as those with ratio of upper quantile to lower quantile smaller than $1.5$, and were removed.
+Genes with low expression were also removed, corresponding to those having more than $80\%$ samples with values smaller than $10$.
+After this filtering process, $n_1 = 8928$ genes were kept.\\
+
+From the compiled matrix $L_0$ we remove genes whose difference between upper quantile and lower quantile is greater than $0.25$.
+So the resulting matrix $L_1$ contains the log ratios of $n_2 = 8928$ genes.
+We also compute the logarithmic ratios of the phenotipic data, for the $92$ accessions and the $3$ traits.\\
+
+\subsection{Co-expression network construction}
+We used the Log Fold Change matrix $L_1$ to compute the corresponding similarity matrix. 
+We found that for this network, the smallest integer such that the $R^2 \geq 0.8 $ is $\beta=3$.
+Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.
+
+%scale-free beta 1 y 3
+\begin{figure}[h]
+  \centering
+    \includegraphics[clip,width=0.8\textwidth]{Figures/pick_beta.png}
+  \caption{Degree distributions}
+  \label{fig:beta}
+\end{figure}
+
+The resulting adjacency matrix $A$ represents a complete graph $G=(V,E)$, with $|V| = 8928$ genes (nodes) and $|E| = 39850128$ edges.
+We choose $0.2$ as cutoff value, keeping only the connections above this threshold and removing the isolated nodes, so the new adjacency matrix $\hat{A}$ has $n_3 = 5810$ genes and accounts for $16875145$ edges.\\
+
+%The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8928$ genes (nodes) and $39850128$ edges.
+For computational reasons, this network was transformed into an unweighted one $\hat{A}$, keeping only the connections above the cutoff value of $0.2$.
+The resulting unweighted network has a total of $5810$ nodes and $16875145$ edges.
+
+\subsection{Co-expression module identification}
+After applying the HLC algorithm, a total of $4131$ genes were distributed in $c = 5143$ overlapping modules of $3$ or more genes.
+Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.
+
+%overlapping percentage
+\begin{figure}[h]
+  \centering
+    \includegraphics[clip,width=0.5\textwidth]{Figures/artificial_modules.png}
+  \caption{Overlapping percentage}
+  \label{fig:overlap}
+\end{figure}
+
+\subsection{Modules association to phenotypic trits}
+The phenotypic traits under study are shoot $K^+$ content, root biomass and shoot biomass.
+Figure \ref{fig:pdata} shows that there are significant differences in the values of these phenotypic traits between stress and control conditions.
+This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.
+
+% boxplots
+\begin{figure}[h]
+  \centering
+    \includegraphics[clip,width=1\textwidth]{Figures/phenotypic_traits.png}
+  \caption{Phenotipic traits distribution under control and salt stress}
+  \label{fig:pdata}
+\end{figure}
+
+Using the affiliation matrix $F$ derived from HLC output and the Log Fold Change matrix $L_1$, we build the matrix $M$ computing the egengene for each of the $c = 5143$ modules.
+We apply the LASSO technique using each of the phenotypic traits as the outcome variable, one at a time.
+As shown in Figure~\ref{fig:cross-val} we perform cross-validation for each phenotypical trait, in order to select the corresponding regularization parameter $\lambda$ that minimizes the mean-squared error.\\
+
+\begin{figure}[h]
+\centering
+   \begin{subfigure}{0.3\linewidth} \centering
+     \includegraphics[width=1\textwidth]{Figures/lambda_kshoot_2.png}
+     \caption{$K$ shoot, $\lambda = 0.013$}\label{fig:lambda_kshoot}
+   \end{subfigure}
+   \begin{subfigure}{0.3\linewidth} \centering
+     \includegraphics[width=1\textwidth]{Figures/lambda_BMshoot_2.png}
+     \caption{Shoot biomass, $\lambda = 0.025$}\label{fig:lambda_BMshoot}
+   \end{subfigure}
+   \begin{subfigure}{0.3\linewidth} \centering
+     \includegraphics[width=1\textwidth]{Figures/lambda_BMroot.png}
+     \caption{Root biomass, $\lambda = 0.035$}\label{fig:lambda_BMroot}
+   \end{subfigure}
+\caption{Cross-validation of the LASSO regularization parameter $\lambda$, for each phenotypic trait.} \label{fig:cross-val}
+\end{figure}
+
+We finally adjust the three models using all observations and the corresponding $\lambda$ values.
+As result, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot $K$ content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.\\
+
+\subsection{Genes enrichment}
+From the $19$ genes selected by LASSO, all but $3$ genes (associated with $K$ content), were also identified as deferentially expressed ($|\ell_{ij}| \geq 2$) for at least one of the $92$ accessions.
+This suggest that those genes are strong candidates as stress responsive genes to salt conditions in rice.\\
+
+According to the Quickgo database, only $2$ of the $16$ diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase.
+Figure~\ref{fig:3d} shows their corresponding 3D protein structure.
+
+% Figure 3d protein strucures
+\begin{figure}[h]
+\centering
+   \begin{subfigure}{0.49\linewidth} \centering
+     \includegraphics[width=0.5\textwidth]{Figures/structure_LOC_Os12g27254.png}
+     \caption{Spermidine hydroxycinnamoyltransferase 2}\label{fig:structure_SHT2}
+   \end{subfigure}
+   \begin{subfigure}{0.49\linewidth} \centering
+     \includegraphics[width=0.5\textwidth]{Figures/structure_LOC_Os12g37260.png}
+     \caption{Lipoxygenase}\label{fig:structure_Lipoxygenase}
+   \end{subfigure}
+\caption{3D protein structure of named genes selected by LASSO.} \label{fig:3d}
+\end{figure}
+
+On the one hand, SHT2 contributes to the natural variation of spermidine-based phenolamides in rice cultivars.
+On the other hand plant lipoxygenase may be involved in a number of diverse aspects of plant physiology including growth and development, pest resistance, and senescence or responses to wounding.This protein is involved in the pathway oxylipin biosynthesis, which is part of Lipid metabolism.\\
+
+Previous studies (\cite{gupta2013plant, hou2015persimmon, mittova2002salt, peng2019novel, roychoudhury2011amelioration}) provide evidence of biological implications of sperimidine and lipoxygenase in tolerance to salt stress in other plants or even in rice cultivars.
+However, further studies are needed to elucidate the detailed biological function of the remaining 14 genes that have not been named so far,  which may have a potential relevance in stress responsive mechanisms to salt conditions in rice.
+
+%We also use the STRING database to visualize the reported interaction networks of the 2 named genes (see Figure X).
+
+%Figure STRING networks
+
+\section{Discussion}
+
+The methodology described in this manuscript provide an aproach to discover key genes responding to a specific treatment in an organism, linking transcriptomic with phenotypic data, and identifying overlaping gene modules.
+Similar approximations have been proposed~(\cite{langfelder2008wgcna, du2019network}), but they only focus on disjoint (non-overlapping) communities.
+However, it is well known that communities in real networks are overlapping.
+\\
+
+The methodology was applied in a case study with rice under salt stress.
+The results show a group of 14 genes where two of them are related to the response to saline stress according to previous studies, validating the ability of the method to detect this kind of key genes.\\
+
+An advantage of the proposed methodology is its modularity.
+As future work, other module detection and selection techniques could be used, instead HLC and LASSO respectively.
+The combinations of these techniques would allow finding target genes for future biological studies that evaluate their potential as genes that respond to salt stress in rice.
+Furthermore, this study can be extended to other stresses and even to other crops.\\
+
+\bibliographystyle{acm}
+\bibliography{mybibliography}
 

### 2020-09-11 nicolas/PPIPL3/ReportV2.tex

@@
 \begin{document}
-\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
+\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactomes}
 \author{Nicolás Antonio López Rozo}
@@
 
-Previous Network-based efforts to predict unmapped protein-protein
-interactions (PPI's) suggest that proteins with multiple paths of
-length 3 (L3) are more likely to be connected. This paper extends
-this so-called L3 principle by taking into account feature extraction
-and using \texttt{XGBoost} techniques for prediction. In particular,
-we train the model using handcrafted features as well as features
-learned from embeddings using \texttt{Node2Vec}. Our main result shows
-that while L3 remains an important principle for predicting links,
-the approach is outperform by using embedded features. The mentioned
-approaches are compared using the human and the rice interactomes.
+Network analysis for  predict unmapped protein-protein interactions (PPI) suggest that the higher the number of paths of
+length~3 (L3) between two proteins, the more likely they are to interact.
+This paper extends previous work based on the L3 principle by taking into account the representation learning of node features of the PPI network.
+In particular, we train an XGBoost model using L3 and handcrafted features, as well as embeddings from \texttt{Node2Vec}.
+Our main result shows that while L3 is an important feature for predicting links, best performance is achieved when combined with edge embeddings.
+The proposed approached is evaluated for the human and rice interactomes.
 \end{abstract}
@@
 
-Proteins are the chief actors of biological functionality inside of
-the cell, and as such, do not usually work as single substances but
-rather as a part of dynamic networks, interacting with other substances
-for a wide variety of purposes \cite{Lin2017}.
+Proteins are the key actors of biological processes inside cells. As they carry out a variety of tasks, cells do not work as single agents but
+rather as a part of dynamic networks of protein-protein interaction (PPI)~\cite{Lin2017}.
+PPI networks play a key role in a variety of interdependent mechanisms, including signal transduction, homeostasis control, stress responses, plant defense and organ formation.
+At the molecular level, PPIs play an essential role in many physiological and developmental processes, including protein phosphorylation, transcriptional co-factor recruitment, transporter activation~\cite{Zhang_2010_PPI}.
 
-Protein-protein interactions (PPIs) have a key role in a variety of
-biological processes such as signal transduction, homeostasis control,
-stress responses, plant defense and organ formation. On the molecular
-level, PPIs have importance in protein phosphorylation, transcriptional
-co-factor recruitment, transporter activation and many others, thus
-playing an essential role in many physiological and developmental
-processes in virtually all organisms \cite{Zhang_2010_PPI}.
+Prediction of potentially relevant yet unexplored PPIs is a current research topic on bioinformatics. Several authors have
+proposed different methods for extrapolating information from the existing PPI networks.
+Kovacs et al.
+(2019) introduces an approach which relies on counting paths of length 3 (L3) to predict interactions among proteins for a variety of model organisms.
+The proposed approach outperform previous methods in different PPI networks of yeast (\emph{S.
+cerevisiae}), Arabidopsis (\emph{A.
+thaliana}), worm (\emph{C.
+elegans}), fly (\emph{D.
+melanogaster}), fission yeast (\emph{S.
+pombe}) and mouse (\emph{M.
+musculus}), as well as for the human interactome \cite{Kovacs2019}.
 
-Prediction of potentially relevant, yet unexplored PPIs is a current
-research topic on bioinformatics and therefore, several authors have
-proposed different methods for extrapolating information from the
-existing PPI networks. As presented by Kovacs et al (2019), information
-related to counting paths of length 3 (L3) seems to outperform state-of-the-art
-methods when it comes to predict interactions among proteins for a
-variety of model organisms such as yeast (\emph{S. cerevisiae}), Arabidopsis
-(\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
-fission yeast (\emph{S. pombe}) and mouse (\emph{M. musculus}), as
-well as on the human interactome \cite{Kovacs2019}.
-
-The most common way to validate PPI's is the \emph{Yeast-Two-Hybrid}
-technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
-which is based on the expression of a specific reporter gene that
-activates by the binding of a DNA-binding Domain (DB) and an Activation
-Domain (AD) of a Transcription Factor that binds to an Upstream Activation
-Sequence (UAS). For the Y2H technique, a protein is fused to the DB
-domain (known as \emph{bait}) and another one to the AD (known as
-\emph{prey}). If the proteins do not interact, then the reporter gene
-is not expressed. Otherwise, the reporter gene expression is activated
-by the activation domain.
+A most common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid} technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
+which is based on (i) the expression of a specific reporter gene that is activated by the binding of a DNA-binding Domain (DB); and (ii) an activation domain (AD) of a transcription factor that binds to an upstream activation sequence (UAS).
+For the Y2H technique, a protein is fused to the DB domain (known as \emph{bait}) and another one to the AD (known as \emph{prey}).
+If the proteins do not interact, then the reporter gene
+is not expressed. Otherwise, the reporter gene expression is activated by the activation domain.
 
@@
 NL: I will make this figure myself
+JF: Good idea
 \end{comment}
@@
 
-Having several of these Y2H results allows scientists to establish
-a PPI network, where all known interactions for each protein are represented.
-Several algorithms are proposed over these networks in order to predict
-unknown interactions. In this report, three prediction methods are
-presented and their results are shown: Common Neighbors (\textbf{CN}),
-which uses the length-2 path count; raw count of paths of length 3
-(\textbf{A3}); and degree-normalized length-3 paths score (\textbf{L3}).
+Based on the outcome of numerous Y2H experiments PPI networks can be constructed, in which known interactions for each protein are represented.
+Several algorithms are proposed over these networks in order to predict yet unknown interactions.
+This paper evaluates predictions based on three measures: Common Neighbors (\textbf{CN}), based on the count of paths of length 2; the raw count of paths of length 3
+(\textbf{A3}); and the degree-normalized count of paths of length 3(\textbf{L3}).
 
-Focus of the present study is to evaluate different methods for predicting
-protein-protein interactions (PPIs) using the existing knowledge of
-the network, which is an undirected graph. The traditional way is
-usually based on social networks analysis, more specifically on the
-Triadic Closure Principle (TCP), that states that the more common
-shared friends that two people have, the more likely that they know
-each other. As shown by previous studies, the mentioned approach fails
-because it does not consider the structural and chemical properties
+\begin{comment}
+JF: Yet you do more than evaluating this three measures so this has to be updated. 
+JF: Not sure we should use L3 above since it is being redefined as a normalized measure here!
+\end{comment}
+
+The focus of this paper is to evaluate different methods for predicting PPIs using the existing knowledge of the network of interactions, which is represented as an undirected graph.
+The traditional way is of approaching the problem is usually based on social networks analysis, more specifically on the triadic closure principle (TCP), that states that the more common shared friends that two people have, the more likely that they know each other.
+As shown by previous studies, the mentioned approach fails because it does not consider the structural and chemical properties
 of the proteins \cite{Kovacs2019}.
 
-For achieving the described results, human and rice PPI networks are
-used and compared using state-of-the-art methods, as well as the proposed
-ones (CN, A3, L3). In the case of the human network, human interactome
-(\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED})
-were used. A massive experimental assay was carried on and its results
-were consolidated and used to build a validation network (\emph{HI-III}).
+\begin{comment}
+JF: Not sure this is the place to introduce TCP.
+You are already stating what the paper will do.
+Sees like TCP, should be introduced before the L3 measure is introduces, as part of the history of past approaches.
+Also TCP is not referred to later on...
+\end{comment}
 
+For achieving the described results, human and rice PPI networks are used and compared using state-of-the-art methods, as well as the proposed ones (CN, A3, L3).
+In the case of the human network, human interactome (\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED}) were used.
+A massive experimental assay was carried on and its results were consolidated and used to build a validation network (\emph{HI-III}).
+
+%% read throughout the document but edited up here. I prefer to read the results before editing the materials and methods section. 
 \section{Materials and Methods}
@@
 
-\begin{figure}[h]
-\caption{Methods Comparison for \emph{HI-II-14}}
+% cannot compile doc given the way images are being imported
 
-\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
-\end{figure}
-
-\begin{figure}[h]
-\caption{Methods Comparison for \emph{HI-TESTED}}
-
-\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
-\end{figure}
+%\begin{figure}[h]
+%\caption{Methods Comparison for \emph{HI-II-14}}
+%
+%\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
+%\end{figure}
+%
+%\begin{figure}[h]
+%\caption{Methods Comparison for \emph{HI-TESTED}}
+%
+%\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
+%\end{figure}
 

### 2020-09-24 nicolas/PPIPL3/ReportV3.tex
Report - Version 3

--- 2020-09-11 nicolas/PPIPL3/ReportV2.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
 %% A simple dot to overcome graphicx limitations
@@
 \begin{document}
-\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactomes}
+\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
 \author{Nicolás Antonio López Rozo}
@@
 
-Network analysis for  predict unmapped protein-protein interactions (PPI) suggest that the higher the number of paths of
-length~3 (L3) between two proteins, the more likely they are to interact.
-This paper extends previous work based on the L3 principle by taking into account the representation learning of node features of the PPI network.
-In particular, we train an XGBoost model using L3 and handcrafted features, as well as embeddings from \texttt{Node2Vec}.
-Our main result shows that while L3 is an important feature for predicting links, best performance is achieved when combined with edge embeddings.
-The proposed approached is evaluated for the human and rice interactomes.
+Network analyses for predicting unmapped protein-protein interactions
+(PPI) suggest that the higher the number of paths of length\textasciitilde 3
+(L3) between two proteins, the more likely they are to interact. This
+paper extends previous work based on the L3 principle by taking into
+account the representation learning of node features of the PPI network.
+In particular, we train an \texttt{XGBoost} model using L3 and handcrafted
+features, as well as embeddings from \texttt{Node2Vec}. Our main result
+shows that while L3 is an important feature for predicting links,
+best performance is achieved when combined with edge embeddings. The
+proposed approach is evaluated for the human and rice interactomes.
 \end{abstract}
@@
 
-Proteins are the key actors of biological processes inside cells. As they carry out a variety of tasks, cells do not work as single agents but
-rather as a part of dynamic networks of protein-protein interaction (PPI)~\cite{Lin2017}.
-PPI networks play a key role in a variety of interdependent mechanisms, including signal transduction, homeostasis control, stress responses, plant defense and organ formation.
-At the molecular level, PPIs play an essential role in many physiological and developmental processes, including protein phosphorylation, transcriptional co-factor recruitment, transporter activation~\cite{Zhang_2010_PPI}.
+Proteins are the key actors of biological processes inside cells.
+As they carry out a variety of tasks, proteins do not work as single
+agents but rather as a part of dynamic networks of protein-protein
+interactions (PPI) \cite{Lin2017}. PPI networks play a key role in
+a variety of interdependent mechanisms, including signal transduction,
+homeostasis control, stress responses, plant defense and organ formation.
+At the molecular level, PPIs play an essential role in many physiological
+and developmental processes, including protein phosphorylation, transcriptional
+co-factor recruitment and transporter activation \cite{Zhang_2010_PPI}.
 
-Prediction of potentially relevant yet unexplored PPIs is a current research topic on bioinformatics. Several authors have
-proposed different methods for extrapolating information from the existing PPI networks.
-Kovacs et al.
-(2019) introduces an approach which relies on counting paths of length 3 (L3) to predict interactions among proteins for a variety of model organisms.
-The proposed approach outperform previous methods in different PPI networks of yeast (\emph{S.
-cerevisiae}), Arabidopsis (\emph{A.
-thaliana}), worm (\emph{C.
-elegans}), fly (\emph{D.
-melanogaster}), fission yeast (\emph{S.
-pombe}) and mouse (\emph{M.
+Prediction of potentially relevant yet unexplored PPIs is a current
+research topic on bioinformatics. Several authors have proposed different
+methods for extrapolating information from the existing PPI networks.
+Kovacs et al. (2019) introduces an approach which relies on counting
+paths of length 3 (L3) to predict interactions among proteins for
+a variety of model organisms. The proposed approach outperform previous
+methods in different PPI networks of yeast (\emph{S. cerevisiae}),
+Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D.
+melanogaster}), fission yeast (\emph{S. pombe}) and mouse (\emph{M.
 musculus}), as well as for the human interactome \cite{Kovacs2019}.
 
-A most common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid} technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
-which is based on (i) the expression of a specific reporter gene that is activated by the binding of a DNA-binding Domain (DB); and (ii) an activation domain (AD) of a transcription factor that binds to an upstream activation sequence (UAS).
-For the Y2H technique, a protein is fused to the DB domain (known as \emph{bait}) and another one to the AD (known as \emph{prey}).
-If the proteins do not interact, then the reporter gene
-is not expressed. Otherwise, the reporter gene expression is activated by the activation domain.
+A common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid}
+technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
+which is based on the expression of a specific reporter gene that
+activates by the binding of a DNA-binding Domain (DB) and an Activation
+Domain (AD) of a Transcription Factor that binds to an Upstream Activation
+Sequence (UAS). For the Y2H technique, a protein is fused to the DB
+domain (known as \emph{bait}) and another one to the AD (known as
+\emph{prey}). If the proteins do not interact, then the reporter gene
+is not expressed. Otherwise, the reporter gene expression is activated
+by the activation domain.
 
@@
 NL: I will make this figure myself
+
 JF: Good idea
@@
 
-Based on the outcome of numerous Y2H experiments PPI networks can be constructed, in which known interactions for each protein are represented.
-Several algorithms are proposed over these networks in order to predict yet unknown interactions.
-This paper evaluates predictions based on three measures: Common Neighbors (\textbf{CN}), based on the count of paths of length 2; the raw count of paths of length 3
-(\textbf{A3}); and the degree-normalized count of paths of length 3(\textbf{L3}).
+Based on the outcome of numerous Y2H experiments PPI networks can
+be constructed, in which known interactions for each protein are represented.
+Several algorithms are proposed over these networks in order to predict
+yet unknown interactions. This paper evaluates predictions based on
+three measures: Common Neighbors (\textbf{CN}), based on the count
+of paths of length 2; the raw count of paths of length 3 (\textbf{A3});
+and the degree-normalized count of paths of length 3 (\textbf{L3}).
+Furthermore
 
 \begin{comment}
-JF: Yet you do more than evaluating this three measures so this has to be updated. 
-JF: Not sure we should use L3 above since it is being redefined as a normalized measure here!
+JF: Yet you do more than evaluating this three measures so this has
+to be updated.
+
+JF: Not sure we should use L3 above since it is being redefined as
+a normalized measure here!
 \end{comment}
 
-The focus of this paper is to evaluate different methods for predicting PPIs using the existing knowledge of the network of interactions, which is represented as an undirected graph.
-The traditional way is of approaching the problem is usually based on social networks analysis, more specifically on the triadic closure principle (TCP), that states that the more common shared friends that two people have, the more likely that they know each other.
-As shown by previous studies, the mentioned approach fails because it does not consider the structural and chemical properties
-of the proteins \cite{Kovacs2019}.
+The focus of this paper is to evaluate different methods for predicting
+PPIs using the existing knowledge of the network of interactions,
+which is represented as an undirected graph. The traditional way of
+approaching the problem is usually based on social networks analysis,
+more specifically on the triadic closure principle (TCP), that states
+that the more common shared friends that two people have, the more
+likely that they know each other. As shown by previous studies, the
+mentioned approach fails because it does not consider the structural
+and chemical properties of the proteins \cite{Kovacs2019}.
+
+For achieving the described results, human and rice PPI networks are
+used and compared using state-of-the-art methods, as well as the proposed
+ones (CN, A3, L3). In the case of the human network, human interactome
+(\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED})
+were used. A massive experimental assay was carried on and its results
+were consolidated and used to build a validation network (\emph{HI-III}).
 
 \begin{comment}
-JF: Not sure this is the place to introduce TCP.
-You are already stating what the paper will do.
-Sees like TCP, should be introduced before the L3 measure is introduces, as part of the history of past approaches.
-Also TCP is not referred to later on...
+JF: read throughout the document but edited up here. I prefer to read
+the results before editing the materials and methods section. 
 \end{comment}
 
-For achieving the described results, human and rice PPI networks are used and compared using state-of-the-art methods, as well as the proposed ones (CN, A3, L3).
-In the case of the human network, human interactome (\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED}) were used.
-A massive experimental assay was carried on and its results were consolidated and used to build a validation network (\emph{HI-III}).
-
-%% read throughout the document but edited up here. I prefer to read the results before editing the materials and methods section. 
 \section{Materials and Methods}
@@
 to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
-$9433$ and $23322$ interactions.
+$9433$ and $23322$ interactions, respectively.
 
@@
 \end{itemize}
+It is worth mentioning that some exploration experiments were carried
+out with interactions removal percentages of 2, 5, 10 and 20\%, as
+well as train-test partitions of 15-85, 80-20 and 75-25, and the explained
+parameter set was selected because it either represented a marginal
+gain (results not shown) or because it was a common parameter selection
+in related literature.
 
@@
 results, and one can see that its mean performance using the AUC metric
-is 0.52, and that the results among the 10 repetitions are consistent.
+is $0.52$, and that the results among the 10 repetitions are consistent.
 This results mean that the model using only the default features perform
-barely as good as a random choice of the labels.
+barely as good as a random choice of the labels. This result can also
+be assessed when looking at the confusion matrix, where a precision
+of $0.5168$ and a recall of $0.4589$ can be derived.
 
@@
 \par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.95\columnwidth]{Only_ML/ROConlyML_SUMMARY}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_ML/ROConlyML_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_ML/CMonlyML_SUMMARY}
 \end{figure}
 
-\section{(PENDING FROM HERE ON)}
+Next experiment carried out was to add each prediction method score
+as a prediction feature, that is, use the score as the eleventh input
+feature for the ML algorithm. Results for the Common Neighbors method
+(\textbf{CN}), which uses paths of length 2, are shown in Figure \ref{F2}.
+It can be observed again that all 10 experiments have small variability
+among them and perform significantly better than the baseline, with
+an area under the ROC curve of $0.90$, with small standard deviation.
+When looking at the confusion matrix for this model, turns out that
+most of the guesses are true positives and true negatives, resulting
+in a precision of $0.9285$ and a recall of $0.7420$. Same analyses
+are done with the count of paths of length 3 (\textbf{A3}) and with
+the degree-normalized length-3 score (\textbf{L3}) and results are
+presented in figures \ref{F3} and \ref{F4}, respectively, resulting
+also in area under the curve of 0.9 for both cases.
 
-% cannot compile doc given the way images are being imported
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F2}Results for \texttt{Node2Vec} model with CN feature}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA2_SUMMARY}
+\end{figure}
 
-%\begin{figure}[h]
-%\caption{Methods Comparison for \emph{HI-II-14}}
-%
-%\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
-%\end{figure}
-%
-%\begin{figure}[h]
-%\caption{Methods Comparison for \emph{HI-TESTED}}
-%
-%\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
-%\end{figure}
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F3}Results fir \texttt{Node2Vec} model with A3 feature}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA3_SUMMARY}
+\end{figure}
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F4}Results for \texttt{Node2Vec} model with L3 feature}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CML3_SUMMARY}
+\end{figure}
+
+The precision when using the count of paths of length 3 (A3) is $0.8587$
+while the recall is $0.7685$. It is interesting to observe that precision
+decreased $7.5\%$ when compared to the previous case, but recall
+increased $3.6\%$. A similar situation is obtained for the model
+with the normalized score (L3), whose precision decreased to $0.8936$
+($-3.8\%$) and recall increased to $0.7610$ ($+2.6\%$). It can
+be seen also that the ROC curve rapidly achieves higher values of
+true positive rate in the A2-featured model when compared to the other
+models.
+
+Finally, an relevant evaluation on the models that include a handcrafted
+feature is necessary: How important is the appended feature for the
+model result? The answer can be observed by the feature importance
+plots for CN, A3 and L3 in Figure \ref{F5-importance}. Each plot
+resembles the number of times that each feature creates a bifurcation
+in the underlying decision trees that \texttt{XGBoost} uses. The more
+bifurcations, the higher the importance a feature has on the model
+itself. All models have the appended feature as the most important
+by a large margin.
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F5-importance}Importance for \texttt{Node2Vec} models: with
+L3 feature}
+\par\end{centering}
+\noindent \begin{raggedleft}
+\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot 1}
+\par\end{raggedleft}
+\noindent \begin{raggedleft}
+\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot 1}
+\par\end{raggedleft}
+\noindent \raggedleft{}\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot 1}
+\end{figure}
+
+\subsection{Human Interactome}
+
+(PENDING FROM HERE)
+
+\begin{figure}[h]
+\caption{Methods Comparison for \emph{HI-II-14}}
+
+\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-ii-14\lyxdot txt}
+\end{figure}
+
+\begin{figure}[h]
+\caption{Methods Comparison for \emph{HI-TESTED}}
+
+\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-tested\lyxdot txt}
+\end{figure}
 
@@
 $A^{3}$predictions follow a very similar trend.
-
-When analyzing the robustness of each of the networks, the following
-values for the Weighted Spectral Distribution were found. For a robustness
-reference, the Erdos-Renyi model was used to generate a random network
-with the same number of nodes and edges and on those random networks,
-the WSD was measured.
-\begin{center}
-\begin{table}[H]
-\caption{Validation of Weighted Spectral Distribution}
-
-\centering{}%
-\begin{tabular}{|c|c|c|}
-\hline 
-Network & Network WSD & Erdos-Renyi WSD\tabularnewline
-\hline 
-\hline 
-HI-II-14 & 393.4939 & 198.7706\tabularnewline
-\hline 
-HI-TESTED & 423.3902 & 276.1065\tabularnewline
-\hline 
-HI-III (VALID.) & 373.9369 & 153.5329\tabularnewline
-\hline 
-\end{tabular}
-\end{table}
-\par\end{center}
-
-The table shows that the three networks used in this report are robust,
-because the are significantly more robust than a network with the
-same density generated randomly.
 
@@
 \begin{comment}
-please update the conclusions - WSD does not seem relevant
+JF: please update the conclusions - WSD does not seem relevant
+
+NL: Done
 \end{comment}
-
-For the analyzed networks, it is also important to highlight their
-robustness when compared to a random network with the same density:
-if weighted spectral distribution (WSD) is used, then the used networks
-are on average twice as robust as their random network versions.
 

### 2020-09-29 camila/Report/article_CR.tex
new paper draft

@@
 %--------------------------------------------------------------------------
-\title{Identification of rice genes which respond to saline stress from co-expression networks analysis}
+\title{Identification of rice genes which respond to saline stress from co-expression network analysis}
 \author{Camila Riccio Rengifo\\
@@
 \subsection{Co-expression network}
-A network is an undirected graph $G=(V,E)$ where ${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of vertices or nodes and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of edges or links that connect the vertices.
-In a gene co-expression network each node corresponds to a gene and a pair of genes is connected if they show similar differential expression patterns.
-These co-expression networks are of biological interest since the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex.\\
-
-When the network is simple and unweighted, it can be represented by an adjacency matrix $A \in \{0,1\}^{n \times n}$.
-This adjacency matrix is symmetric, with a positive one in the positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever there is an edge connecting vertices $v_i$ and $v_j$, and zeros elsewhere.\\
+A network is an undirected graph $G=(V,E)$ where ${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of \textit{vertices} or \textit{nodes} and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of \textit{edges} or \textit{links} that connect vertices.
+In a gene co-expression network, each node corresponds to a gene.
+A pair of genes is connected if they show similar differential expression patterns.
+A simple and unweighted network can be represented by an adjacency matrix $A \in \{0,1\}^{n \times n}$ that is symmetric  with a positive one in the positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever there is an edge connecting vertices $v_i$ and $v_j$, and zeros elsewhere.
+Co-expression networks are of biological interest because the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex.\\
 
@@
 
-Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn et al.~\cite{ahn2010link}.
-HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities.
+The Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn et al.~\cite{ahn2010link}.
+The HLC approach represents communities as groups of links (rather than nodes), each node inherits all memberships of its links and can thus belong to multiple, overlapping communities.
 The algorithm maps links to nodes and connects them if a pair of links shares a node.
 The similarity between two links $e_{ik}$ and $e_{jk}$ is computed using the Jaccard index
-\begin{equation*}
+\begin{equation}
 \label{eq:jaccard}
-S(e_{ik},e_{jk}) = \frac{\vert n_+(i) \cap n_+(j) \vert}{\vert n_+(i) \cup n_+(j) \vert},
-\end{equation*}
-where $n_+(i)$ denotes the set of node $i$ and its neighbors.\\
+S(e_{ik},e_{jk}) = \frac{\vert n(i) \cap n(j) \vert}{\vert n(i) \cup n(j) \vert},
+\end{equation}
+where $n(i)$ denotes the set corresponding to node $i$ and its neighbors.
+The algorithm uses single-linkage hierarchical clustering to build a dendrogram in which each leaf is a link from the original network and branches represent link communities.
+Hierarchical clustering algorithms repeatedly merge groups until all elements are members of a single cluster.\\
 
-With this similarity, the method uses single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities.
-Hierarchical clustering methods repeatedly merge groups until all elements are members of a single cluster.
-To find meaningful communities it is crucial to know where to partition the dendrogram.
-Thus, the most relevant communities are established at the maximal partition density $D$, a function  based on link density inside communities that measures the quality of a link partition.
-The partition density has a single global maximum along the dendrogram in almost all cases, because the value is just the average density at the top of the dendrogram (a single giant community with every link and node) and it is very small at the bottom of the dendrogram (most communities consists of a single link).
-$D = 1$ when every community is a fully connected clique and $D = 0$ when each community is a tree.
-If a link community is less dense than a tree (when the community subgraph has disconnected components), then that community will give a negative contribution to $D$.
-Computing $D$ at each level of the link dendrogram allows to pick the best level to cut (although meaningful structure exists above and below that threshold).
-This way, the final output corresponds to node clusters, where each node can participate in multiple communities.\\
+For the purpose of finding meaningful communities it is crucial to know where to partition the dendrogram.
+In this case, the most relevant communities are established at the maximal partition density $D$, a function  based on link density inside communities measuring the quality of a link partition.
+The partition density has a single global maximum along the dendrogram in almost all cases, because the value is the average density at the top of the dendrogram (a single giant community with every link and node) and it is very small at the bottom of the dendrogram (most communities consists of a single link).
+It is the case that $D = 1$ when every community is a fully connected clique and $D = 0$ when each community is a tree.
+If a community is less dense than a tree (when the community subgraph has  disconnected components), then that community will give a negative contribution to $D $, which can take values less than zero.The minimum density inside a community is $-2/3$,  given by one community of two disconnected edges.
+Since $D$ is the average of the intra-community density, there is a lower bound of $D=-2/3$.
+Computing $D$ at each level of the link dendrogram can help the purpose of picking the best level to cut (although meaningful structure could exist above or below the threshold).
+The output of the process is a set of node clusters, where each node can participate in multiple communities.\\
 
 \subsection{Least Absolute Shrinkage Selector Operator}
-Least Absolute Shrinkage Selector Operator (LASSO) is a regularized linear regression technique, a method that combines a regression model with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients.
-Very useful in problems where the number of variables (genes) $ n $ is much greater than the number of samples $ m $ ($ n \gg m $).
-Lasso solves the least squares problem with restriction on the $ L_1$-norm of the coefficient vector:
+The Least Absolute Shrinkage Selector Operator (LASSO) is a regularized linear regression technique.
+It combines a regression model, with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients.
+In other words, Lasso solves the least squares problem with restriction on the $ L_1$-norm of the coefficient vector.
+It can be especially useful to solve problems where the number of variables (e.g., genes) $ n $ is much greater than the number of samples $ m $ (i.e., $ n \gg m $).\\
+
+Consider a dataset consisting of $m$ samples, each of which consists of $n$ covariates and a single outcome.
+Let $y_i$ be the outcome and $x_i := (x_1,...,x_n)$ be the covariate vector for the $i-th$ sample.
+Then, the objective of LASSO is to solve
 
@@
 
-Or equivalently minimizing:
+Or, equivalently, in the so-called Lagrangian form, it minimizes
+
 \begin{equation}
@@
 \end{equation}
-being $ s $, $ \lambda \geq 0 $ the respective penalty parameters for complexity.\\
+
+where $s$ is the regularization penalty and $\lambda \geq 0$ is the corresponding Lagrange multiplier.\\
 
@@
 Since the $\lambda$ value determines the degree of penalty, the accuracy of the model depends on its choice.
-Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squared error.
-With that selected $\lambda$ value, the model is adjusted again, using all the observations.\\
+Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squared error.\\
 
@@
 
-This methodology uses as input data RNA-seq read counts, representing gene expression levels.
-More precisely, $n_0$ gene expression profiles of an organism, measured for $m$ different genotypes under control and treatment conditions, and $r$ biological replicates.
-This raw data can be represented as a matrix $D_0 \in {\mathbb{N}_0}^{n_0 \times 2mr}$. 
-In order to discover key genes and their interaction with phenotypes related to treatment tolerance, the methodology also requires a set of $p$ phenotypic traits, measured for the $m$ genotypes.
-The phenotypic data can be seen as a matrix $P \in \mathbb{R}^{2m \times p}$ containing two phenotypic values per genotype, one under control condition and the second under treatment condition.\\
+The proposed methodology uses RNA-seq read counts, representing gene expression levels, as input data.
+More precisely, it uses $n_0$ gene expression profiles of an organism, measured for $m$ different genotypes under control and treatment conditions, and $r$ biological replicates.
+This raw data is represented as a matrix $D_0 \in {\mathbb{N}_0}^{n_0 \times 2mr}$.
+In order to discover key genes and their interaction with phenotypes related to treatment tolerance, the approach also requires a set of $p$ phenotypic traits, measured for the $m$ genotypes.
+The phenotypic data is seen as a matrix $P \in \mathbb{R}^{2m \times p}$ containing two phenotypic values per genotype, one under control condition and the second one under treatment condition.
+The whole methodology can be followed in the flow chart of Figure~\ref{fig:flow_chart}.\\
+
+% flow chart
+\begin{figure}[]
+  \centering
+    \includegraphics[clip,width=1\textwidth]{Figures/flow_chart.pdf}
+  \caption{Proposed methodology flow chart, showing the five macro-processes: \textbf{A}.~Data pre-processing, \textbf{B}.~Co-expression network contruction, \textbf{C}.~Co-expression module identification, \textbf{D}.~Modules association to phenotypic traits, and \textbf{E}.~Genes enrichment.}
+  \label{fig:flow_chart}
+\end{figure}
 
@@
 
-The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results.
+The RNA-seq data cannot be directly interpreted.
+Therefore, a normalization process is applied to deal with the problem of some biases affecting the quantification results.
 In order to correct library size and RNA composition bias, the suggested normalization technique is DESeq2 \cite{love2014moderated}.
-From the normalized data $D_1 \in \mathbb{R}^{n_0 \times 2mr}$, average the biological replicates of each genotype, so that the data matrix becomes $D_2 \in \mathbb{R}^{n_0 \times 2m}$.
+The normalized data is represented as a matrix $D_1 \in \mathbb{R}^{n_0 \times 2mr}$.
+The biological replicates of each genotype are averaged and the resulting data is represented as a matrix $D_2 \in \mathbb{R}^{n_0 \times 2m}$.
+Next, the genes exhibiting low variance or low expression are removed from $D_2$, thus identifying a subset of size $n_1 \leq n_0$ of the original genes.
+The control and treatment data is separated into the matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1 \times m}$, respectively.
+The matrix entries $c_{ij}$ in $C$ and $t_{ij}$ in $T$ represent, respectively, the normalized expression level of gene $i$ in accession $j$.
+Control and treatment data is also separated from phenotypic data $P$, obtaining the $P_c$ and $P_t$ matrices, of dimensions $m \times p$.\\
+
+%From the normalized data $D_1 \in \mathbb{R}^{n_0 \times 2mr}$, average the biological replicates of each genotype, so that the data matrix becomes $D_2 \in \mathbb{R}^{n_0 \times 2m}$.
 Next, remove genes exhibiting low variance or low expression, getting a new number of $n_1 \leq n_0$ genes.
@@
 
-Now, compile the control and treatment data by measuring the changes in expression levels in terms of logarithmic ratios.
+At this point, the changes in expression levels and phenotypic values, between control and treatment conditions, are measured in terms of logarithmic ratios.
+In the case of expression levels, the log ratios are represented in a matrix $L_0 \in \mathbb{R}^{n_1 \times m}$, known as the Log Fold Change matrix, where $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$.
+Similarly, the log ratios of the phenotypic data are computed and represented in a matrix $P_\ell \in \mathbb{R}^{m \times p}$.\\
+
+The final step of the data pre-processing is to filter $L_0$ by removing rows (e.g., genes) with low variance in the differential expression patterns, getting a new matrix $L_1$ of dimensions $n_2 \times m$, with $n_2 \leq n_1$.\\
+
+%Now, compile the control and treatment data by measuring the changes in expression levels in terms of logarithmic ratios.
 Matrix $L_0 \in \mathbb{R}^{n_1 \times m}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$.
@@
 
-The Log fold change matrix $L_1$ is used to build the co-expression network following the first steps of the WGCNA methodology.
-First, measure the level of concordance between gene differential expression profiles across samples.
-Use the absolute value of the Pearson correlation coefficient as the similarity measure between genes and store the values in the similarity matrix $S\in \mathbb{R_{+}}^{n_2 \times n_2}$.\\
+The Log fold change matrix $L_1$ is used to build the co-expression network following the first two steps of the WGCNA methodology~\cite{langfelder2008wgcna}.
+First, the level of concordance between gene differential expression profiles across samples is measured.
+To this end, the absolute value of the Pearson correlation coefficient is used as the similarity measure between genes and the resulting values are stored in the similarity matrix $S\in \mathbb{R_{+}}^{n_2 \times n_2}$.\\
 
-Then, transform $S$ into an adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes.
+And second, the matrix $S$ is transformed into and adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes.
 In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network.
@@
 
-Finally, prepare matrix $A$ to apply the clustering algorithm, transforming it into an unweighted network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$.
-To determine the Pearson Correlation Coefficient (PCC) cutoff for finding biologically relevant co-expressed modules, we use the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values.
-Above the cutoff, the entries of matrix $A$ become $1$ and when the value is below the threshold it becomes zero.\\
+%Then, transform $S$ into an adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes.
+In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network.
+This kind of networks contain many nodes with very few connections and a small number of hubs with high connections.
+In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node).
+So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g.
+$R^2 > 0.85$).
+\\
 
@@
 genes) are involved in multiple functions.
-Thus, unlike WGCNA, the adjacency matrix $\hat{A}$, from the co-expression network $G$, is used to detect overlapping rather than non-overlapping communities, using the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.\\
+Thus, unlike WGCNA, the adjacency matrix $A$, is used to detect overlapping (rather than non-overlapping) communities, using the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.\\
 
@@
 
-The HLC algorithm organizes the $n_2$ genes of matrix $\hat{A}$ into $c$ communities, where each gene can belong to one, multiple, or no community.
+As a previous step, matrix $A$ has to be modified, in order to apply the clustering algorithm, transforming it into an unweighted network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$.
+To this end, the Pearson Correlation Coefficient (PCC) cutoff is determined using the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values.
+Above the cutoff, the entries of matrix $A$ become $1$, and below the cutoff $A$ values becomes zero.\\
+
+%To determine the Pearson Correlation Coefficient (PCC) cutoff for finding biologically relevant co-expressed modules, we use the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values.
+Above the cutoff, the entries of matrix $A$ become $1$ and when the value is below the threshold it becomes zero.\\
+
+Now, the HLC algorithm can be applied.
+It organizes the $n_2$ genes of matrix $\hat{A}$ into $c$ communities, where each gene can belong to one, multiple, or no community.
 This information is represented as an affiliation matrix $F \in \{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ if node $i$ is member of cluster $u$.\\
@@
 
-To identify the most relevant gene groups (modules), associated with the phenotypic response to a specific treatment in an organism, we use a LASSO based approach.
-Each module can be represented by a eigengene, which is defined as the first principal component of such module.
+To identify the most relevant groups (modules) of genes, associated with the phenotypic response to a specific treatment in an organism, the proposed methodology uses a LASSO based approach.
+Each module is represented by a eigengene, which is defined as the first principal component of such module.
 A eigengene can be thought of as an average differential expression profile for each community and is computed from the Log Fold Change Matrix $L_1$ and the affiliation matrix $F$.
-Given a module $u$, use the affiliation matrix to identify the genes belonging to $u$ and then select the corresponding rows of the matrix $L_1$ to compute the first principal component of $u$.
-Each principal component would be a column of the matrix $M \in \mathbb{R}^{m \times c}$.\\
+Given a module $u$, the affiliation matrix is used to identify the genes belonging to $u$ and then the corresponding rows of the matrix $L_1$ are selected to compute the first principal component of $u$.
+Each principal component becomes a column of the matrix $M \in \mathbb{R}^{m \times c}$.\\
 
@@
 These profiles are then associated with each phenotypic trait using the least absolute shrinkage and selection operator (LASSO).
-LASSO combines a regression model with a procedure of contraction of some parameters towards zero, imposing a restriction or a penalty on the regression coefficients.
+%LASSO combines a regression model with a procedure of contraction of some parameters towards zero, imposing a restriction or a penalty on the regression coefficients. 
 This technique allows selection of the most relevant variables.
-In our context, the eigenges (the columns of $M$) act as regressor variables and each phenotypic trait (each column of $P_\ell$) is used as an outcome variable.
-\\
+In our context, the eigenges (the columns of $M$) act as regressor variables and each phenotypic trait (each column of $P_\ell$) is used as an outcome variable at a time.\\
 
-The output after applying LASSO is a set of modules for each phenotypic trait. The union of genes belonging to the selected modules are the target genes for downstream analysis.
+The output after applying LASSO is a set of modules for each phenotypic trait. The union of genes belonging to the selected modules are the target genes for downstream analysis.\\
 
@@
 
-For the genes found, first identify the differential expressed ones, that is genes showing an absolute value of the log fold change greater than 2 ($|\ell_{ij}|\geq 2$) for at least one sample.
+The objective of this part of the analysis is to characterize the genes found with complementary information, helping to elucidate their possible behavior and role in the response to the studied treatment.\\
+
+A simple analysis to made with the selected genes, using the previous calculations, is to identify the differentially expressed ones.
+That is, identify genes showing an absolute value of the log fold change greater than 2 ($|\ell_{ij}|\geq 2$) for at least one sample.
 This represents genes whose level of expression is quadrupled (up or down) from control to treatment condition, suggesting that those genes are strong candidates as treatment responsive genes.\\
 
-You also can perform a functional category enrichment searching for the Gene Ontology (GO) annotations from databases like QuikGO~\cite{binns2009quickgo}.
+A widley used approach in this step is to perform a functional category enrichment, searching for the Gene Ontology (GO) annotations from databases like QuikGO~\cite{binns2009quickgo}.
 This annotations can provide evidence of biological implications of the target genes in the treatment-tolerance mechanisms.\\
 
-Those named genes (with GO annotations), can be used to perform another relevant analysis reviewing their reported protein-protein interaction networks using the STRING database~\cite{szklarczyk2016string}.
+Addicionally, QuickG0 can be used to identify genes with reported protein products, which can be used to perform another relevant analysis reviewing their reported protein-protein interaction networks using the STRING database~\cite{szklarczyk2016string}.
 The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge transfer between organisms, and from interactions aggregated from other (primary) databases.
@@
 
+% Otras caracteristicas pueden buscarse en otras bases de datos o hacer una busqueda más exahustiva en articulos de estudios previos sobre estos genes.
+
 \section{Case study}
-RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n_0=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $m=92$ accessions of the Rice Diversity Panel 1, with $r=2$ biological replicates.
-On the other hand, we use $p=3 $ phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass.
-This traits were measured for the same $92$ genotypes, under  control and salt stress conditions, and can be found in the supplementary information of~\cite{campbell2017allelic}.\\
+The RNA-seq data was obtained from GEO database \cite{GEOAcces90:online} (accession number GSE98455).
+This data corresponds to $n_0=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $m=92$ accessions of the Rice Diversity Panel 1, with $r=2$ biological replicates.
+A total of $p=3 $ phenotypic traits are used: shoot $K^+$ content, root biomass and shoot biomass.
+These traits were measured for the same $92$ genotypes, under  control and salt stress conditions, and can be found in the supplementary information of~\cite{campbell2017allelic}.\\
 
 \subsection{Data pre-processing}
-The DESeq normalization was applied to the raw data and the biological replicates were averaged.
-Genes exhibiting low variance were identified as those with ratio of upper quantile to lower quantile smaller than $1.5$, and were removed.
-Genes with low expression were also removed, corresponding to those having more than $80\%$ samples with values smaller than $10$.
-After this filtering process, $n_1 = 8928$ genes were kept.\\
+DESeq2 normalization was applied to the raw data and the biological replicates were averaged.
+Genes exhibiting low variance were identified as those with ratio of upper quantile to lower quantile smaller than $1.5$, and were removed from the normalized data.
+Genes with low expression, corresponding to those having more than $80\%$ samples with values smaller than $10$, were also removed.
+A total of $n_1 = 8928$ genes were kept after this filtering process.\\
 
-From the compiled matrix $L_0$ we remove genes whose difference between upper quantile and lower quantile is greater than $0.25$.
-So the resulting matrix $L_1$ contains the log ratios of $n_2 = 8928$ genes.
-We also compute the logarithmic ratios of the phenotipic data, for the $92$ accessions and the $3$ traits.\\
+From the Log Fold Change matrix $L_0$, genes whose difference between upper quantile and lower quantile was greater than $0.25$, were removed.
+Therefore, the resulting matrix $L_1$ contains the log ratios of $n_2 = 8928$ genes.
+The logarithmic ratios of the phenotipic data, for the $92$ accessions and the $3$ traits, was also computed.\\
 
 \subsection{Co-expression network construction}
-We used the Log Fold Change matrix $L_1$ to compute the corresponding similarity matrix. 
-We found that for this network, the smallest integer such that the $R^2 \geq 0.8 $ is $\beta=3$.
-Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.
+The Log Fold Change matrix $L_1$ is used to compute the corresponding similarity matrix. 
+For this network, it was observed that $\beta=3$ is the smallest integer such that the $R^2 \geq 0.8 $.
+Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right), which is the degree distribution of a scale-free network with $R^2 = 0.8$ with $\beta = 3$.
 
@@
 
-The resulting adjacency matrix $A$ represents a complete graph $G=(V,E)$, with $|V| = 8928$ genes (nodes) and $|E| = 39850128$ edges.
-We choose $0.2$ as cutoff value, keeping only the connections above this threshold and removing the isolated nodes, so the new adjacency matrix $\hat{A}$ has $n_3 = 5810$ genes and accounts for $16875145$ edges.\\
+The resulting adjacency matrix $A$ represents a complete graph $G=(V,E)$, with $|V| = 8928$ genes and $|E| = 39850128$ edges.
+The cutoff value was set to $0.2$, thus keeping only the connections above this threshold and removing the isolated nodes.
+The resulting adjacency matrix $\hat{A}$ has $n_3 = 5810$ genes and accounts for $16875145$ edges.\\
 
@@
 
-\subsection{Modules association to phenotypic trits}
+\subsection{Modules association to phenotypic traits}
 The phenotypic traits under study are shoot $K^+$ content, root biomass and shoot biomass.
-Figure \ref{fig:pdata} shows that there are significant differences in the values of these phenotypic traits between stress and control conditions.
-This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.
+Figure \ref{fig:pdata} suggests that there are significant differences in the values of these phenotypic traits between stress and control conditions.
+This supports the working hypothesis that these three variables represent tolerance-associated traits in rice under salt stress.
 
@@
 
-Using the affiliation matrix $F$ derived from HLC output and the Log Fold Change matrix $L_1$, we build the matrix $M$ computing the egengene for each of the $c = 5143$ modules.
+Using the affiliation matrix $F$ derived from the HLC output and the Log Fold Change matrix $L_1$, a matrix $M$ is built by computing the egengene for each of the $c = 5143$ modules.
+The LASSO technique was applied by using each of the phenotypic traits as the outcome variable, one at a time.
+As shown in Figure~\ref{fig:cross-val} cross-validation was performed for each phenotypical trait in order to select the corresponding regularization parameter $\lambda$ that minimizes the mean-squared error.\\
+
+%we build the matrix $M$ computing the egengene for each of the $c = 5143$ modules.
 We apply the LASSO technique using each of the phenotypic traits as the outcome variable, one at a time.
@@
 
-We finally adjust the three models using all observations and the corresponding $\lambda$ values.
+FInally, three LASSO models were adjusted by using the corresponding $\lambda$ and phenotypical data with the egengens of matrix $M$.
 As result, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot $K$ content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.\\
@@
 
-On the one hand, SHT2 contributes to the natural variation of spermidine-based phenolamides in rice cultivars.
-On the other hand plant lipoxygenase may be involved in a number of diverse aspects of plant physiology including growth and development, pest resistance, and senescence or responses to wounding.This protein is involved in the pathway oxylipin biosynthesis, which is part of Lipid metabolism.\\
-
-Previous studies (\cite{gupta2013plant, hou2015persimmon, mittova2002salt, peng2019novel, roychoudhury2011amelioration}) provide evidence of biological implications of sperimidine and lipoxygenase in tolerance to salt stress in other plants or even in rice cultivars.
+Uniprot database~\cite{uniprot2018uniprot} reports, on the one hand, that SHT2 contributes to the natural variation of spermidine-based phenolamides in rice cultivars.
+On the other hand, is reported that plant lipoxygenase may be involved in a number of diverse aspects of plant physiology including growth and development, pest resistance, and senescence or responses to wounding.
+This protein is involved in the pathway oxylipin biosynthesis, which is part of Lipid metabolism.
+Aditionally, previous studies (\cite{gupta2013plant, hou2015persimmon, mittova2002salt, peng2019novel, roychoudhury2011amelioration}) provide evidence of biological implications of sperimidine and lipoxygenase in tolerance to salt stress in other plants or even in rice cultivars.
 However, further studies are needed to elucidate the detailed biological function of the remaining 14 genes that have not been named so far,  which may have a potential relevance in stress responsive mechanisms to salt conditions in rice.
@@
 
-The methodology described in this manuscript provide an aproach to discover key genes responding to a specific treatment in an organism, linking transcriptomic with phenotypic data, and identifying overlaping gene modules.
-Similar approximations have been proposed~(\cite{langfelder2008wgcna, du2019network}), but they only focus on disjoint (non-overlapping) communities.
-However, it is well known that communities in real networks are overlapping.
-\\
+This manuscript provides a detailed description of a network analysis-based methodology to 
+%The methodology described in this manuscript provides network analysis-based approach to
+the discovery of key genes responding to a specific treatment in an organism, linking transcriptomic with phenotypic data, and identifying overlaping gene modules.\\
+
+The proposed methodology follows the principles of WGCNA.
+Starts from the level of thousands of genes, identifies biologically interesting gene modules, and identifies key genes in the treatment pathways for further validation.
+In WGCNA the whole workflow is structured in a modular way.
+The proposed methodology takes advantage of this property modifying and exploring new techniques inside some processes.\\
+
+An important modification to the WGCNA methodology is in the pre-processing of the input data.
+Our approach is designed to integrate expression data measured under two different conditions (control and treatment), instead of working with them independently or considering a single condition.
+For this purpose, an approach similar to that proposed in~\cite{du2019network} is used, where the control and treatment data are compiled in a single matrix using the log fold change measure.
+Thus, the input to construct the co-expression network is not the expression data but the changes in the expression levels from one condition to the other, allowing to capture the signal of changes caused by the treatment.\\
+
+The most innovative feature of the proposed methodology is the module detection technique. 
+The co-expression network is calculated as in WGCNA until the scale-free network is obtained.
+In our approach, this network is used to apply the HLC algorithm, which is capable of detecting overlapping communities.
+The usual approximation and its different variations, only focuses on disjoint (non-overlapping) communities, but it is well known that communities in real networks are overlapping.
+Thus the approach presented in this work can be seen as a generalization of the previous approaches.\\
+
+%Similar approximations have been proposed in, e.g., \cite{langfelder2008wgcna} and \cite{du2019network}, but they only focus on disjoint (non-overlapping) communities.
+It is well known that communities in real networks are overlapping, and thus the approach presented in this work can be seen as a generalization of the previous approaches.\\
+
+% La tercera modificación es incluir LASSO como método para relacionar los datos phenotypicos con los transcriptomicos.
 
@@
 
-An advantage of the proposed methodology is its modularity.
-As future work, other module detection and selection techniques could be used, instead HLC and LASSO respectively.
+As future work, other overlapping module detection and selection techniques could be used, instead HLC and LASSO respectively.
 The combinations of these techniques would allow finding target genes for future biological studies that evaluate their potential as genes that respond to salt stress in rice.

### 2020-10-05 mauricio/Recombination/imputed_data.py.py
programs added

@@ -1,0 +1,57 @@
+def read_sequence(fastafile):
+    '''Returns the whole sequence of a fasta file in only one line'''
+    with open(fastafile,'r') as f:
+        f.readline()  
+        seq = ''
+        for line in f.readlines():
+            seq += line.strip('\n').upper()
+    return seq
+
+chr_genome1 = read_sequence('Osat_IR64_AGI_NSD_chrOK.id_chr01.fasta')
+chr_genome2 = read_sequence('Osat_Azucena_AGI_chrOK.id_chr01.fasta')
+
+input_file = open('datos_recombinación.txt','r')
+output_file = open('imputed_recombination_chr01_1M.txt','w')
+
+w_size = 1_000_000
+w_end = w_size
+
+lines = input_file.readlines()
+i=0
+while i < len(lines)-1:
+    
+    if lines[i][0]=='*':
+        pos1, samples1 = lines[i].split()
+        pos2, samples2 = lines[i+1].split()
+        name1,pos1 = pos1.split('_')
+        name2,pos2 = pos2.split('_')
+        pos1, pos2 = int(pos1), int(pos2)
+        
+        while w_end < pos1:
+            output_file.write(name1+'@_'+str(w_end)+' '+samples1+'\n')
+            w_end += w_size
+        
+        output_file.write(lines[i])
+        if pos1 == w_end:
+            w_end += w_size
+        
+        while w_end < pos2:
+            if w_end-pos1 < pos2-w_end:
+                output_file.write(name1+'@_'+str(w_end)+' '+samples1+'\n')
+            else:
+                output_file.write(name2+'@_'+str(w_end)+' '+samples2+'\n')
+            w_end += w_size
+
+    else:
+        output_file.write(lines[i])
+    
+    i+=1
+
+output_file.write(lines[i])
+
+while w_end < max(len(chr_genome1),len(chr_genome2)):
+    output_file.write(name2+'@_'+str(w_end)+' '+samples2+'\n')
+    w_end += w_size
+
+input_file.close()
+output_file.close()

### 2020-10-05 camila/Report/article_CR.tex

@@
 Figure \ref{fig:pdata} suggests that there are significant differences in the values of these phenotypic traits between stress and control conditions.
-This supports the working hypothesis that these three variables represent tolerance-associated traits in rice under salt stress.
+This supports the working hypothesis that these three variables represent tolerance-associated traits in rice under salt stress.\\
 
@@
 
-The proposed methodology follows the principles of WGCNA.
+The proposed methodology is inspired by the workflow suggested by the WGCNA methodology~\cite{langfelder2008wgcna}.
+The main steps are the preprocessing of the gene expression data, the construction of a co-expression network, the detection of modules within the network, relate the modules with external information (e.g.
+phenotypic data) and an enrichment of the identified key genes.
+The methodology is structured in a modular way, which allows modifying and exploring different techniques in each of the steps of the process.
+\\
+
+%The proposed methodology follows the principles of WGCNA.
 Starts from the level of thousands of genes, identifies biologically interesting gene modules, and identifies key genes in the treatment pathways for further validation.
@@
 
-An important modification to the WGCNA methodology is in the pre-processing of the input data.
-Our approach is designed to integrate expression data measured under two different conditions (control and treatment), instead of working with them independently or considering a single condition.
+The proposed approach is designed to integrate expression data measured under two different conditions (control and treatment), unlike the usually co-expression-based approaches which work with both conditions independently or consider a single condition.
 For this purpose, an approach similar to that proposed in~\cite{du2019network} is used, where the control and treatment data are compiled in a single matrix using the log fold change measure.
@@
 
-The most innovative feature of the proposed methodology is the module detection technique. 
+Another important innovation in the proposed methodology is the module detection technique.
 The co-expression network is calculated as in WGCNA until the scale-free network is obtained.
-In our approach, this network is used to apply the HLC algorithm, which is capable of detecting overlapping communities.
-The usual approximation and its different variations, only focuses on disjoint (non-overlapping) communities, but it is well known that communities in real networks are overlapping.
+This network is then used to apply the HLC algorithm, a clustering technique capable of detecting overlapping communities.
+Several approaches of module detection from gene expression have been proposed and were evaluated in~\cite{saelens2018comprehensive}.
+Most of them focuses only on disjoint (non-overlapping) communities and the described techniques dealing with overlap are not clustering but biclustering and decomposition methods.
+It is well known that communities in real networks are overlapping.
 Thus the approach presented in this work can be seen as a generalization of the previous approaches.\\

### 2020-10-08 camila/report/concl.tex
first complete paper draft

\section{Concluding Remarks}
\label{sec.concl}

This manuscript provides a detailed description of a network-based analysis
methodology for the discovery of key genes responding to
a specific treatment in an organism. It links transcriptomic with
phenotypic data, and identifies overlaping gene modules.

The proposed methodology is inspired by the workflow suggested in the
WGCNA methodology~\cite{langfelder2008wgcna}. The main steps are the
preprocessing of the gene expression data, the construction of a
co-expression network, the detection of modules within the network,
the relation of modules with external information (e.g. phenotypic data)
and the enrichment of the identified key genes with additional information.
The WGCNA methodology and therefore the proposal as well, are structured
in a modular way, which allows modifying and  exploring different techniques
in each step of the process.

The proposed approach is designed to integrate expression data
measured under two different conditions (namely, control and treatment),
unlike the usually co-expression-based approaches which work with both
conditions independently or consider only a single condition. For this
purpose, an approach similar to that proposed in~\cite{du2019network}
is used, where the control and treatment data are compiled in a single
matrix using the log fold change measure. Thus, the input to construct
the co-expression network is not the expression data, but instead the changes
in the expression levels from one condition to the other, making room for capturing
the signal of changes caused by the treatment.

Another important feature in the proposed methodology is the module
detection technique. The co-expression network is computed as in
WGCNA until a scale-free network is obtained. This network is then
used to apply the HLC algorithm, a clustering technique capable of
detecting overlapping communities. Several approaches of module
detection from gene expression have been proposed and were evaluated
in~\cite{saelens2018comprehensive}. Most of them focuses only on
disjoint (non-overlapping) communities, and the described techniques
dealing with overlaps are not clustering but biclustering and
decomposition methods. It is well known that communities in real
networks are overlapping~\cite{palla2005uncovering}. Thus, the approach
presented in this work can be seen as a generalization of the previous 
approaches such as WGCNA.

The proposed methodology was applied in a case study with rice under salt
stress. The results show a group of 14 genes in which $2$ of them are
related to the response to saline stress according to previous
studies, validating the ability of the method to detect this kind of
key genes. As future work, other overlapping module detection and selection
techniques should be used instead HLC and LASSO, respectively. The
combination of these techniques would allow finding target genes for
future biological studies that evaluate their potential as genes that
respond to salt stress in rice. Also, laboratory experimentation needs to be
conducted on via to verify the findings of this paper in relation to salinity 
stress. Finally, this study can be extended to other stresses and even to other 
crops.

### 2020-10-08 camila/report/prelim.tex
first complete paper draft

\section{Preliminaries}
\label{sec.prelim}

This section presents preliminaries on networks, hierarchical link
clustering and the LASSO linear regression technique.

\subsection{Co-expression network}

A network is an undirected graph $G=(V,E)$ where
${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of \textit{vertices} or
\textit{nodes} and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of
\textit{edges} or \textit{links} that connect vertices. In a gene
co-expression network, each node corresponds to a gene. A pair of
genes is connected if they show similar expression
patterns. A simple and unweighted network can be represented by an
adjacency matrix $A \in \{0,1\}^{n \times n}$ that is symmetric with a
positive one in the positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever
there is an edge connecting vertices $v_i$ and $v_j$, and zeros
elsewhere. Co-expression networks are of biological interest because
the co-expressed genes are usually controlled by the same
transcriptional regulatory pathway, functionally related or members of
the same pathway or metabolic complex.

\subsection{Hierarchical Link Clustering}

The Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn
et al.~\cite{ahn2010link}. The HLC approach represents communities as
groups of links (rather than nodes), and each node inherits all
memberships of its links and can thus belong to multiple, overlapping
communities. It maps links to nodes and connects them if a pair of
links shares a node. The similarity between two links $e_{ik}$ and
$e_{jk}$ is computed using the Jaccard index
%
\begin{equation}\label{eq:jaccard}
  S(e_{ik},e_{jk}) = \frac{\vert n(i) \cap n(j) \vert}{\vert n(i) \cup n(j) \vert},
\end{equation}
%
where $n(i)$ denotes the set containing exactly node $i$ and its
neighbors. The algorithm uses single-linkage hierarchical clustering
to build a dendrogram in which each leaf is a link from the original
network and branches represent link communities. Hierarchical
clustering algorithms repeatedly merge groups until all elements are
members of a single cluster. For the purpose of finding meaningful
communities, it is crucial to know where to partition the
dendrogram. In this case, the most relevant communities are
established at the maximal partition density $D$, a function based on
link density inside communities measuring the quality of a link
partition. The partition density $D$ has a single global maximum along
the dendrogram in almost all cases, because its value is the average
density at the top of the dendrogram (a single giant community with
every link and node) and it is very small at the bottom of the
dendrogram (most communities consists of a single link). In
particular, it is the case that $D = 1$ when every community is a
fully connected clique and $D = 0$ when each community is a tree. If a
community is less dense than a tree (i.e. when the community subgraph
has disconnected components), then such a community contribute
negatively to $D$, which can take negative values. The minimum density
inside a community is $-2/3$, given by one community of two
disconnected edges. Since $D$ is the average of the intra-community
density, there is a lower bound of $-2/3$ for $D$. Computing $D$ at
each level of the link dendrogram can help the purpose of picking the
best level to cut, although meaningful structure could exist above or
below the threshold. The output of cutting is a set of node clusters,
where each node can participate in multiple communities.

\subsection{Least Absolute Shrinkage Selector Operator}

The Least Absolute Shrinkage Selector Operator (LASSO) is a
regularized linear regression technique. It combines a regression
model with a procedure of contraction of some parameters towards zero
and selection of variables, imposing a restriction or a penalty on the
regression coefficients. In other words, LASSO solves the least
squares problem with restriction on the $ L_1$-norm of the coefficient
vector. It can be especially useful to solve problems where the number
of variables (e.g., genes) $ n $ is much greater than the number of
samples $ m $ (i.e., $ n \gg m $).

Consider a dataset consisting of $m$ samples, each of which consists
of $n$ covariates and a single outcome. Let $y_i$ be the outcome and
$x_i := (x_1,...,x_n)$ be the covariate vector for the $i$-th
sample. The objective of LASSO is to solve
%
\begin{equation}
\min \left\lbrace\sum_{i=1}^{m}{\left( y_i-\sum_{j=1}^n{\beta_j
    x_{ij}}\right)^2} \right\rbrace \quad , \quad \textrm{subject to}
\quad \sum_{j=1}^n\abs{\beta_j}\leq s.
\end{equation}

Equivalently, in the Lagrangian form, it minimizes

\begin{equation}
  \sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} +
  \lambda \sum_{j=1}^n\abs{\beta_j}
\end{equation}
%
where $s$ is the regularization penalty and $\lambda \geq 0$ is the
corresponding Lagrange multiplier. Since the $\lambda$ value
determines the degree of penalty, the accuracy of the model depends on
its choice. Cross-validation is often used to select the
regularization parameter, choosing the one that minimizes the
mean-squared error.

### 2020-10-08 camila/report/framework.tex
first complete paper draft

\section{The Workflow}
\label{sec.framework}

The proposed workflow uses RNA-seq read counts, representing gene
expression levels, as input data. More precisely, it uses $n_0$ gene
expression profiles of an organism, measured for $m$ different
genotypes under control and treatment conditions, and $r$ biological
replicates. This raw data is represented as a matrix $D_0 \in
{\mathbb{N}_0}^{n_0 \times 2mr}$. In order to discover key genes and
their interaction with phenotypes related to treatment tolerance, the
approach also requires a set of $p$ phenotypic traits, measured for
the $m$ genotypes. The phenotypic data is seen as a matrix $P \in
\mathbb{R}^{2m \times p}$ containing two phenotypic values per
genotype, one under control condition and a second one under treatment
condition. The proposed workflow is depicted in
Figure~\ref{fig:flow_chart}. This section explains the five
macro-processes (A)-(E) of the proposed workflow. In comparison with
WGCNA, it adds the macro-step (D) and generalizes macro-steps (A)-(C).

% flow chart
\begin{figure}[htpb]
  \centering
    \includegraphics[clip,width=1\textwidth]{figures/flow_chart_2.pdf}
  \caption{The proposed workflow comprising five macro-steps:
    A.~Data pre-processing, B.~Co-expression network
    contruction, C.~Co-expression module identification,
    D.~Modules association to phenotypic traits, and
    E.~Genes enrichment.}
  \label{fig:flow_chart}
\end{figure}

\subsection{Data pre-processing}
The goal of the data pre-processing stage is to build matrices $P_ell$ and $L_1$ representing, respectively, the changes in phenotypic values and expression levels between control and treatment condition, from RNA-seq and phenotypic data found in matrices $D_0$ and $P$, respectively.

The RNA-seq data cannot be directly interpreted. Therefore, a normalization process is applied to deal with the problem of possible biases affecting the quantification of results. The suggested normalization technique for correcting library size and RNA composition bias is DESeq2 \cite{love2014moderated}. The normalized data is represented as a matrix $D_1 \in \mathbb{R}^{n_0 \times 2mr}$, and the biological replicates of each genotype are averaged and represented as a matrix $D_2 \in \mathbb{R}^{n_0 \times 2m}$. The genes exhibiting low variance or low expression are removed from $D_2$, thus identifying a subset of size $n_1 \leq n_0$ of the original genes. The control and treatment data is separated into the matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1 \times m}$, respectively. The matrix entries $c_{ij}$ in $C$ and $t_{ij}$ in $T$ represent, respectively, the normalized expression level of gene $i$ in accession $j$. Control and treatment data is also separated from phenotypic data $P$, obtaining the $P_c$ and $P_t$ matrices of dimensions $m \times p$.

In the above configuration, the changes in expression levels and phenotypic values, between control and treatment conditions, are measured in terms of logarithmic ratios. In the case of expression levels, the log ratios are represented in the Log Fold Change matrix $L_0 \in \mathbb{R}^{n_1 \times m}$, where $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$. Similarly, the log ratios of the phenotypic data are computed and represented in the $P_\ell \in \mathbb{R}^{m \times p}$ matrix.

The final step of the data pre-processing is to filter $L_0$ by removing rows (e.g., genes) with low variance in the differential expression patterns, obtaining a new matrix $L_1$ of dimensions $n_2 \times m$, with $n_2 \leq n_1$.

\subsection{Co-expression network construction}

A gene co-expression network connect genes with similar expression patterns across biological conditions. The purpose of this stage is to describe how to build the co-expression network $A$ from the Log Fold Change matrix $L_1$, capturing the relationship between genes according to the change in expression levels between the two studied conditions. These co-expression patterns are meaningful for the identification of genes not yet associated with the response to the treatment condition.

The Log Fold Change matrix $L_1$ is used to build the co-expression network following the first two steps of the WGCNA methodology~\cite{langfelder2008wgcna}. First, the level of concordance between gene differential expression profiles across samples is measured. To this end, the absolute value of the Pearson correlation coefficient is used as the similarity measure between genes and the resulting values are stored in the similarity matrix $S\in \mathbb{R_{+}}^{n_2 \times n_2}$. Second, the matrix $S$ is transformed into and adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes. In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network. These networks contain many nodes with very few connections and a small number of hubs with high connections. In a strict scale-free network the logarithm of $P(k)$ (i.e., the probability of a node having degree $k$) is approximately inversely proportional to the logarithm of $k$ (i.e., the degree of a node). So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g., $R^2 > 0.85$).

\subsection{Co-expression module identification}

Next step in the workflow is to identifying communities (also called modules)
from the co-expression network structure and dynamics represented in $A$. 
The idea is to cluster genes with similar patterns of differential
expression change. Membership in these modules may overlap in 
biological contexts, because modules may be related to specific 
molecular, cellular or tissue functions and the biological components
(i.e. genes) may be involved in multiple functions. Thus, unlike WGCNA, the adjacency matrix $A$ is used to detect overlapping (rather than non-overlapping) communities, using the Hierarchical Link Clustering (HLC) algorithm (see section~\ref{sec:HLC}).

As a prelimiary step, matrix $A$ is transformed, into an unweighted network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$ before applying the clustering algorithm. To this end, the Pearson Correlation Coefficient (PCC) cutoff is determined using the approach described in~\cite{aoki2007approaches}. The number of nodes, edges, and the network density is determined for different PCC cutoffs. Around the most biological relevant PCC cutoff, the number of nodes presents a linear decrease, and the density of the network reaches its minimum, while below this value the number of edges rapidly increases. Considering this, a cutoff is selected such that gene pairs which have a correlation score higher than the threshold are considered to have significant co-expression relationship. Above the cutoff, the entries of matrix $A$ become $1$ and below the cutoff $A$ values becomes $0$. The application of the HLC algorithm organizes the $n_2$ genes of matrix $\hat{A}$ into $c$ modules, where each gene can belong to one, multiple, or no module at all. This information is represented as an affiliation matrix $F \in \{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ if node $i$ is member of module $u$.

\subsection{Module association to phenotypic traits}

To identify the most relevant groups (modules) of genes, associated
with the phenotypic response to a specific treatment in an organism,
the proposed workflow uses a LASSO based approach. Each module is
represented by an eigengene, which is defined as the first principal
component of such module. An eigengene can be thought of as an average
differential expression profile for each community: it is computed
from the Log Fold Change Matrix $L_1$ and the affiliation matrix
$F$. Given a module $u$, the affiliation matrix is used to identify
the genes belonging to $u$ and then the corresponding rows of the
matrix $L_1$ are selected to compute the first principal component of
$u$. Each principal component becomes a column of the matrix $M \in
\mathbb{R}^{m \times c}$.
These profiles are associated with each phenotypic trait using 
the least absolute shrinkage and selection operator (LASSO). 
In this context, the eigengenes (i.e., the columns of $M$) act as regressor variables
and each phenotypic trait (i.e., each column of $P_\ell$) is used as an outcome variable.

The output after applying LASSO is a set $W_z$ of modules for each
phenotypic trait $z$, where $W_z \subset \{u \mid 1 \leq u \leq c\}$ for
$z= 1,2,..,p$. The target genes $I$ for downstream analysis, which may
be important in the treatment response, are the union of genes belonging
to the selected modules, that is $I = \cup_{z=1}^{p} W_z$, where
$I \subset \{i \mid 1 \leq i \leq n_2\}$.

\subsection{Gene Enrichment}

The goal of this final stage of the process is to characterize the
genes identified in previous stage with additional information, helping to elucidate their
possible behavior and role in the response to the studied treatment.

A simple analysis to made with the selected genes is to identify the
differentially expressed genes in set $I$. That is, to select those genes
in $I$ having an absolute value of the log fold change of at least $2$
($|\ell_{ij}|\geq 2$) for at least one sample. This
represents genes level of expression is quadrupled (up or down)
from control to treatment condition; these genes are
strong candidates for treatment responsive genes.

Also functional category enrichment can be done by, e.g., searching for gene ontology annotations in databases such as QuickGO~\cite{binns2009quickgo}. Such annotations 
can provide evidence of biological implications of the target genes in
the treatment-tolerance mechanisms. Furthermore, QuickG0 can be used to identify genes with reported protein products, which can be used to perform additional relevant
analysis reviewing their reported protein-protein interactions in other databases, such as STRING~\cite{szklarczyk2016string}. The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge
transfer between organisms, and from interactions aggregated from
other (primary) databases. This information can give new insight on how the
selected genes are involved in functional pathways that can be related
with the treatment of interest.

### 2020-10-08 camila/report/intro.tex
first complete paper draft

\section{Introduction}
\label{sec.intro}

Abiotic stresses are key factors that can negatively influence plant
development and productivity. They are a main cause of extensive
agricultural production losses worldwide~\cite{todo}. Soil salinity is
one of the most devastating abiotic stresses, causing reduction in the
cultivable land, crop quality, and productivity. It has been estimated
that 20\% of total cultivated and 33\% of irrigated agricultural lands
worldwide are already affected by high salinity. Moreover, due to the
human activities and natural causes, salinized areas are gradually
increasing every year and are expected to reach 50\% by the end of
year 2050~\cite{shrivastava2015soil}. Salinity tolerance and
susceptibility in plants is known to be the result of elaborated
interactions between morphological, physiological, and biochemical
processes that are regulated, in the end, by multiple genes in
different parts of a genome~\cite{reddy2017salt}. Therefore,
identifying groups of stress responsive genes may lead to crop
improvement in terms of salinity tolerance and, ultimately, contribute
solutions to the general problem of food sustainability in the years
to come.

This paper proposes a workflow to identify stress responsive genes in
organisms, which is known to be a complex quantitative trait. It takes
as input RNA sequencing read counts measured for genotypes under
control and treatment conditions (and representing gene expression
profiles of the target organism), and biological replicates.  In order
to discover key genes and their interaction with phenotypes related to
treatment tolerance, the approach requires a collection of phenotypic
traits (under control and under treatment), measured for the given
genotypes. The output of the workflow is a collection of characterized
genes, potentially relevant to treatment, yielding insight on the
possible behavior of specific genes and the role they may play in
functional pathways in response to the studied treatment of the
organism of interest. The proposed workflow can thus take advantage of
transcriptomic data for different organisms and conditions, based on
the current availability of high-throughput technologies that include
microarrays and RNA sequencing, to study the reaction of organisms
under different environmental stimuli, such as salt stress.

Technically, the proposed approach is both a generalization and an
extension of Weighted Gene Co-expression Network Analysis
(WGCNA)~\cite{todo2}, a widely applied workflow that has been
successfully used for identifying target genes related to diseases and
cancer in several organisms~\cite{tian2018identifying}. The general
idea behind each approach is to identify specific modules in a network
of genes after a sequence of normalization and filtering steps. The
proposed approach is considered a \textit{generalization} of WGCNA
because module detection can now recognize overlapping communities,
which may have more biological meaning given the overlapping
regulatory domains of systems that generate
co-expression~\cite{gaiteri2014beyond}. This is achieved by using
Hierarchical Link Clustering (HLC)~\cite{ahn2010link}. It is also an
\textit{extension} of WGCNA because additional steps and information
are added to the workflow: namely, some networks in the intermediate
steps are forced to be scale-free~\cite{todo3} and LASSO
regression~\cite{tibshirani1996regression} is employed to select the
most significant modules of phenotypical responses to stress.  The
advantage of using HLC as clustering method is its ability to detect
overlapping modules, since biological components are involved in
multiple functions and therefore biological communities tend to be
highly overlapping. On the other hand, LASSO is a regularized
regression technique widely used in variable selection, thanks to its
ability to obtain zero regression coefficients for the less relevant
variables~\cite{desboulets2018review}. Moreover, LASSO is especially
useful in problems where the number of variables is much larger than
the number of samples, which may be the case more often than desired.
The proposed workflow is also modular, since other module detection
and selection techniques could be used, instead HLC and LASSO,
respectively.

The proposed workflow is showcased with a systematic study on rice
(\textit{Oryza sativa}), a major food source that is known to be
highly sensitive to salt stress~\cite{chang2019morphological}. RNA-seq
data was accessed from the GEO database~\cite{GEOAcces90:online}
(accession number GSE98455). It corresponds to $57845$ gene expression
profiles of shoot tissues measured for both control and salt condition
in $92$ accessions of the Rice Diversity Panel 1. As output, 6 modules
are detected as relevant in the response to salt stress in rice: 3
modules of 3 genes each, all associated with shoot K content, 2
modules of 3 genes associated with shoot biomass, and 1 module of 4
genes associated with root biomass. These genes may act as potential
targets for the improvement of salinity tolerance in rice
cultivars. From the 19 genes, all but 3 genes (associated with $K$
content), were also identified as deferentially expressed for at least
one of the 92 accessions, suggesting that those genes are strong
candidates as stress responsive genes. Only 2 of the 16 diferentially
expressed genes, both from the module related with shoot biomass, are
named and have an associated protein product: Spermidine
hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase. In other words,
further studies are needed to elucidate the detailed biological
function of the remaining 14 genes that have not been named so far,
which may have a potential relevance in stress responsive mechanisms
to salt conditions in rice. The goal is that the results reported in
this paper may allow biologist to develop new rice cultivars with
higher resistance to salinity.

### 2020-10-08 camila/report/main.tex
first complete paper draft

\documentclass[runningheads]{llncs}

\input{packages}
\input{macros}

\begin{document}
%
\title{Using Complex Networks for Identifying Potentially New Saline
  Stress Responsive Genes in Rice}
%
\author{Camila Riccio \and Jorge Finke \and Camilo Rocha}
%% \author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
%% Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
%% Third Author\inst{3}\orcidID{2222--3333-4444-5555}}

\authorrunning{C. Riccio et al.}

\institute{Pontificia Universidad Javeriana, Cali, Colombia
  %% \email{lncs@springer.com}
}  
%
\maketitle

\input{abstract}

\input{intro}

\input{prelim}

\input{framework}

\input{case}

\input{concl}

\bibliographystyle{splncs04}
\bibliography{biblio}

\end{document}

### 2020-10-09 camila/report/framework.tex
references added

@@
 genes) may be involved in multiple functions.
-Thus, unlike WGCNA, the adjacency matrix $A$ is used to detect overlapping (rather than non-overlapping) communities, using the Hierarchical Link Clustering (HLC) algorithm (see section~\ref{sec:HLC}).
+Thus, unlike WGCNA, the adjacency matrix $A$ is used to detect overlapping (rather than non-overlapping) communities, using the Hierarchical Link Clustering (HLC) algorithm (see section~\ref{sec.prelim}).
 

### 2020-10-09 camila/report/intro.tex
references added

@@
 development and productivity. They are a main cause of extensive
-agricultural production losses worldwide~\cite{todo}. Soil salinity is
+agricultural production losses worldwide~\cite{mesterhazy2020losses}. Soil salinity is
 one of the most devastating abiotic stresses, causing reduction in the
@@
 extension of Weighted Gene Co-expression Network Analysis
-(WGCNA)~\cite{todo2}, a widely applied workflow that has been
+(WGCNA)~\cite{langfelder2008wgcna}, a widely applied workflow that has been
 successfully used for identifying target genes related to diseases and

### 2020-10-09 nicolas/PPIPL3/ReportV4.tex
Methodology update of validation

--- 2020-09-24 nicolas/PPIPL3/ReportV3.tex
@@
 and the degree-normalized count of paths of length 3 (\textbf{L3}).
-Furthermore
+Furthermore, the Machine Learning (ML) process is assessed for each
+metric as the only feature, for the features obtained by the Node2Vec
+algorithm and for a combination of the Node2Vec and
 
@@
 to be updated.
+
+NL: Done!
 
@@
 (\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED})
-were used. A massive experimental assay was carried on and its results
-were consolidated and used to build a validation network (\emph{HI-III}).
+were used. A massive experimental assay was carried on by other authors
+and its results were consolidated and used to build a validation network
+(\emph{HI-III}).
 
@@
 It is important to have in mind that balanced distribution of the
-positive and negative labels in the datasets was satisfied.
+positive and negative labels in the datasets was satisfied in training,
+but validation is performed using the remaining unbalanced data.
 \end{itemize}
@@
 results, and one can see that its mean performance using the AUC metric
-is $0.52$, and that the results among the 10 repetitions are consistent.
-This results mean that the model using only the default features perform
-barely as good as a random choice of the labels. This result can also
-be assessed when looking at the confusion matrix, where a precision
-of $0.5168$ and a recall of $0.4589$ can be derived.
+is $0.93$, and that the results among the 10 repetitions are consistent.
+This results mean that the model using only the default features performs
+very good. This result can also be assessed when looking at the confusion
+matrix, where a precision of $0.9814$ and a recall of $0.8649$ can
+be derived.
 
@@
 It can be observed again that all 10 experiments have small variability
-among them and perform significantly better than the baseline, with
-an area under the ROC curve of $0.90$, with small standard deviation.
-When looking at the confusion matrix for this model, turns out that
-most of the guesses are true positives and true negatives, resulting
-in a precision of $0.9285$ and a recall of $0.7420$. Same analyses
-are done with the count of paths of length 3 (\textbf{A3}) and with
-the degree-normalized length-3 score (\textbf{L3}) and results are
-presented in figures \ref{F3} and \ref{F4}, respectively, resulting
-also in area under the curve of 0.9 for both cases.
+among them and perform better than the baseline, with an area under
+the ROC curve of $0.98$, with small standard deviation. When looking
+at the confusion matrix for this model, turns out that most of the
+guesses are true positives and true negatives, resulting in a precision
+of $0.9952$ and a recall of $0.9108$. Same analyses are done with
+the count of paths of length 3 (\textbf{A3}) and with the degree-normalized
+length-3 score (\textbf{L3}) and results are presented in figures
+\ref{F3} and \ref{F4}, respectively, resulting also in area under
+the curve of $0.97$ and $0.98$, respectively.
 
@@
 \noindent \begin{centering}
-\caption{\label{F2}Results for \texttt{Node2Vec} model with CN feature}
+\caption{\label{F2}Results for \texttt{Node2Vec} with CN feature}
 \par\end{centering}
@@
 \noindent \begin{centering}
-\caption{\label{F3}Results fir \texttt{Node2Vec} model with A3 feature}
+\caption{\label{F3}Results for \texttt{Node2Vec} with A3 feature}
 \par\end{centering}
@@
 \noindent \begin{centering}
-\caption{\label{F4}Results for \texttt{Node2Vec} model with L3 feature}
+\caption{\label{F4}Results for \texttt{Node2Vec} with L3 feature}
 \par\end{centering}
@@
 
-The precision when using the count of paths of length 3 (A3) is $0.8587$
-while the recall is $0.7685$. It is interesting to observe that precision
-decreased $7.5\%$ when compared to the previous case, but recall
-increased $3.6\%$. A similar situation is obtained for the model
-with the normalized score (L3), whose precision decreased to $0.8936$
-($-3.8\%$) and recall increased to $0.7610$ ($+2.6\%$). It can
-be seen also that the ROC curve rapidly achieves higher values of
-true positive rate in the A2-featured model when compared to the other
-models.
+The precision when using the count of paths of length 3 (A3) is $0.9935$
+while the recall is $0.9055$. It is interesting to observe that precision
+decreased $0.16\%$ and recall decreased $0.58\%$. A similar situation
+is obtained for the model with the normalized score (L3), whose precision
+decreased to $0.9944$ ($-0.08\%$) and recall decreased to $0.8891$
+($-2.38\%$). It can be seen also that the ROC curve achieves high
+values of true positive rate sooner in the A2-featured model than
+on the other models.
+
+Another relevant assessment was carried out to verify the predictive
+power of the metrics. \texttt{XGBoost} models were trained only using
+each metric, for the same 10 experiments. Results are shown in Figures
+\ref{F5}, \ref{F6} and \ref{F7}. It can be observed that XGBoost
+performs worse when only considering CN, A3 and L3 alone, although
+good values of precision (CN:0.9916, A2:0.9807 and A3:0.9838) and
+recall (CN:0.6704, A2:0.7012 and A3:0.7037) are achieved.
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F5}Results for CN feature only}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA2_SUMMARY}
+\end{figure}
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F6}Results for A3 feature alone}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA3_SUMMARY}
+\end{figure}
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F7}Results for L3 feature alone}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyL3_SUMMARY}
+\end{figure}
 
@@
 model result? The answer can be observed by the feature importance
-plots for CN, A3 and L3 in Figure \ref{F5-importance}. Each plot
-resembles the number of times that each feature creates a bifurcation
-in the underlying decision trees that \texttt{XGBoost} uses. The more
-bifurcations, the higher the importance a feature has on the model
-itself. All models have the appended feature as the most important
-by a large margin.
+plots for CN, A3 and L3 in Figure \ref{F8-importance}. Each plot
+resembles the importance gain from each feature in the underlying
+decision trees that \texttt{XGBoost} uses. Usually, the more bifurcations,
+the higher gain a feature has, the more relevance a feature has on
+the model prediction itself. All models have the appended feature
+as the most important by a good margin, although margin for A3 is
+not wide as for A2 or L3.
 
@@
 \noindent \begin{centering}
-\caption{\label{F5-importance}Importance for \texttt{Node2Vec} models: with
-L3 feature}
+\caption{\label{F8-importance}Importance gain plots for \texttt{Node2Vec}
+with each feature}
 \par\end{centering}
 \noindent \begin{raggedleft}
-\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot 1}
+\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}
 \par\end{raggedleft}
 \noindent \begin{raggedleft}
-\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot 1}
+\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
 \par\end{raggedleft}
-\noindent \raggedleft{}\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot 1}
+\noindent \raggedleft{}\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
 \end{figure}
@@
 
-(PENDING FROM HERE)
+As the first step in the evaluation of CN, A3 and L3 methods for edge
+prediction, the full calculation and ranking of all predictions was
+carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
+and \emph{HI-TESTED}) is shown in Figures \ref{fig:HI1} and \ref{fig:HI2},
+in which the x axis represents a rank position $k$ and the y axis
+displays the precision for the top $k$ predictions of each method,
+when assuming interactome \emph{HI-III} as the validation set.
 
 \begin{figure}[h]
-\caption{Methods Comparison for \emph{HI-II-14}}
+\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14}}
 
@@
 \begin{figure}[h]
-\caption{Methods Comparison for \emph{HI-TESTED}}
+\caption{\label{fig:HI2}Methods Comparison for \emph{HI-TESTED}}
 
@@
 
+However, it is of our interest to assess if Machine Learning can boost
+the overall performance on the prediction, so the proposed procedure
+was adapted to the human interactome data: Instead of having to remove
+a fraction of edges from the network in order to predict and validate,
+now the prediction network and the validation networks are given beforehand.
+The rest of the procedure is carried out as mentioned before. Results
+for the Interactome \emph{HI-II-14} with Node2Vec and each metric
+are presented in Figures \ref{F2-1}, \ref{F3-1} and \ref{F4-1}.
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F2-1}\emph{HI-II-14} results for \texttt{Node2Vec} with CN
+feature}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2}
+\end{figure}
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F3-1}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
+feature}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3}
+\end{figure}
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F4-1}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
+feature}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3}
+\end{figure}
+
+It can be seen that for \emph{HI-II-14}, L3 performs marginally better
+than CN, especially when comparing precision ( vs ) and recall ( vs
+), but all of them have good values for area under the ROC curve of
+0.99.
+
 \section{Conclusions}
 
-Taking into account the different results validated in this report,
-one can conclude that length-3 path methodologies might work better
-on protein-protein interactions than its traditional length-2 (TCP
-based) counterparts. On the other hand, it can be seen that degree-normalization
-has little effect on the predictions, i.e., non-normalized $A{{}^3}$
-matrix predictions are still a good methodology for edge prediction
-on PPI networks.
-
-Previous result comes as no surprise when the biological basis of
-protein interactions is considered: It is necessary that protein A
-and protein B have complementary structures in order to interact,
-and when classical paths of length 2 are used, the predicted protein
-interactions usually have the same structures, and not complementary
-ones.
-
-\begin{comment}
-JF: please update the conclusions - WSD does not seem relevant
-
-NL: Done
-\end{comment}
+Based on previous work from other authors\cite{Kovacs2019}, the intrinsic
+knowledge of existing protein-protein interaction networks let scientists
+elucidate promising new interactions to explore. In this study, a
+machine learning algorithm was fed with previously elucidated information
+about the network, as well as results form a state-of-the-art edge
+embedding technique. Even if the baseline of the handcrafted features
+(CN, A3, L3) yields good results, boosting that information with the
+vector embedding, which calculates vector representations of the network
+structure improves the overall performance of each method on its own.
 

### 2020-10-15 nicolas/PPIPL3/ReportV4.tex
Last update on report

@@
 Network analyses for predicting unmapped protein-protein interactions
-(PPI) suggest that the higher the number of paths of length\textasciitilde 3
-(L3) between two proteins, the more likely they are to interact. This
-paper extends previous work based on the L3 principle by taking into
-account the representation learning of node features of the PPI network.
-In particular, we train an \texttt{XGBoost} model using L3 and handcrafted
+(PPI) suggest that the higher the number of paths of length 3 (L3)
+between two proteins, the more likely they are to interact. This paper
+extends previous work based on the L3 principle by taking into account
+the representation learning of node features of the PPI network. In
+particular, we train an \texttt{XGBoost} model using L3 and handcrafted
 features, as well as embeddings from \texttt{Node2Vec}. Our main result
@@
 Several algorithms are proposed over these networks in order to predict
-yet unknown interactions. This paper evaluates predictions based on
-three measures: Common Neighbors (\textbf{CN}), based on the count
-of paths of length 2; the raw count of paths of length 3 (\textbf{A3});
-and the degree-normalized count of paths of length 3 (\textbf{L3}).
-Furthermore, the Machine Learning (ML) process is assessed for each
-metric as the only feature, for the features obtained by the Node2Vec
-algorithm and for a combination of the Node2Vec and
+yet unknown interactions. The focus of this study is to evaluate different
+methods for predicting PPIs using the existing knowledge of the network
+of interactions, which is represented as an undirected graph. The
+traditional way of approaching the problem is usually based on social
+networks analysis, more specifically on the triadic closure principle
+(TCP), that states that the more common shared friends that two people
+have, the more likely that they know each other. As shown by previous
+studies, the mentioned approach fails because it does not consider
+the structural and chemical properties of the proteins \cite{Kovacs2019}.
 
@@
 \end{comment}
-
-The focus of this paper is to evaluate different methods for predicting
-PPIs using the existing knowledge of the network of interactions,
-which is represented as an undirected graph. The traditional way of
-approaching the problem is usually based on social networks analysis,
-more specifically on the triadic closure principle (TCP), that states
-that the more common shared friends that two people have, the more
-likely that they know each other. As shown by previous studies, the
-mentioned approach fails because it does not consider the structural
-and chemical properties of the proteins \cite{Kovacs2019}.
 
@@
 matrix of the network.
-\item [{Length-3~Paths~(A3)}] This is the simplest implementation of
-the proposed insight of ``if my friends and your friends interact,
+\item [{Length-3~Path~Counts~(A3)}] This is the simplest implementation
+of the proposed insight of ``if my friends and your friends interact,
 then we might interact too''. The calculating is carried on with
@@
 As mentioned before, 80\% of the final dataset was randomly selected
-and used for training, while the remaining 20\% was used for validation.
-The whole training-validation procedure was applied 10 times.
+and used for training, while the remaining information was used for
+validation. The whole training-validation procedure was applied 10
+times.
 
@@
 \par\end{centering}
-\noindent \begin{raggedleft}
-\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}
-\par\end{raggedleft}
-\noindent \begin{raggedleft}
-\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
-\par\end{raggedleft}
-\noindent \raggedleft{}\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
+\begin{centering}
+\includegraphics[width=0.7\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}
+\par\end{centering}
+\begin{centering}
+\includegraphics[width=0.7\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
+\par\end{centering}
+\centering{}\includegraphics[width=0.7\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
 \end{figure}
@@
 carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
-and \emph{HI-TESTED}) is shown in Figures \ref{fig:HI1} and \ref{fig:HI2},
-in which the x axis represents a rank position $k$ and the y axis
-displays the precision for the top $k$ predictions of each method,
-when assuming interactome \emph{HI-III} as the validation set.
+and \emph{HI-TESTED}) is shown in Figure \ref{fig:HI1}, in which
+the x axis represents a rank position $k$ and the y axis displays
+the precision for the top $k$ predictions of each method, when assuming
+interactome \emph{HI-III} as the validation set.
 
 \begin{figure}[h]
-\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14}}
+\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14} and \emph{HI-TESTED}}
 
-\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-ii-14\lyxdot txt}
-\end{figure}
-
-\begin{figure}[h]
-\caption{\label{fig:HI2}Methods Comparison for \emph{HI-TESTED}}
-
-\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-tested\lyxdot txt}
+\noindent \centering{}\includegraphics[width=0.48\columnwidth]{hi-ii-14\lyxdot txt}\includegraphics[width=0.48\columnwidth]{hi-tested\lyxdot txt}
 \end{figure}
@@
 for the Interactome \emph{HI-II-14} with Node2Vec and each metric
-are presented in Figures \ref{F2-1}, \ref{F3-1} and \ref{F4-1}.
+are presented in Figures \ref{HI1}, \ref{HI2} and \ref{HI3}.
 
@@
 \noindent \begin{centering}
-\caption{\label{F2-1}\emph{HI-II-14} results for \texttt{Node2Vec} with CN
+\caption{\label{HI1}\emph{HI-II-14} results for \texttt{Node2Vec} with CN
 feature}
@@
 \noindent \begin{centering}
-\caption{\label{F3-1}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
+\caption{\label{HI2}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
 feature}
@@
 \noindent \begin{centering}
-\caption{\label{F4-1}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
+\caption{\label{HI3}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
 feature}
@@
 
-It can be seen that for \emph{HI-II-14}, L3 performs marginally better
-than CN, especially when comparing precision ( vs ) and recall ( vs
-), but all of them have good values for area under the ROC curve of
-0.99.
+It can be seen that for \emph{HI-II-14}, CN performs marginally better
+than L3 when comparing precision ($0.9978$ vs $0.9977$) but worse
+in terms of recall ($0.9346$ vs $0.9598$), although all metrics
+have a good values for area under the ROC curve of 0.99.
+
+On the other hand, the interactome \emph{HI-TESTED} was assessed with
+the same methodology as \emph{HI-II-14} and results are shown in Figures
+\ref{HI1-1}, \ref{HI2-1} and \ref{HI3-1}. It can be seen that in
+terms of false positives, CN performs better than A3 and L3, but in
+terms of false negatives, A3 actually performs better that CN and
+L3. In terms of AUC, metrics results are indistinguishable.
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{HI1-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
+CN feature}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2\lyxdot 2}
+\end{figure}
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{HI2-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
+A3 feature}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3\lyxdot 2}
+\end{figure}
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{HI3-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
+L3 feature}
+\par\end{centering}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3\lyxdot 2}
+\end{figure}
+
+An interesting point from this human datasets can be observed when
+looking at the importance plots, that show a very biased influence
+of the handcrafted features. Figure X shows this behavior for the
+\emph{HI-II-14} dataset, which is very similar to \emph{HI-TESTED}.
+
+\begin{figure}[H]
+\noindent \begin{centering}
+\caption{\label{F8-importance-1}Importance gain plots for \emph{HI-II-14}}
+\par\end{centering}
+\begin{centering}
+\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A2\lyxdot 1}
+\par\end{centering}
+\begin{centering}
+\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A3\lyxdot 1}
+\par\end{centering}
+\centering{}\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot L3\lyxdot 1}
+\end{figure}
 
@@
 
+It was found that applying \texttt{XGBoost} to the rice PPI prediction
+obtained very good results when only using the node embedding technique
+(AUC=0.93) and when only using the handcrafted features based on protein
+neighborhood (AUC \textasciitilde 0.88). Even better results were
+achieved when the handcrafted features and the vector representation
+information were combined (AUC \textasciitilde 0.98). It is important
+to consider that the importance gain of the handcrafted feature in
+each model was high, although importance values for the A3 method
+stand out less from the node embedding information than in CN and
+L3.
+
+On the human interactome, the results show that on both interactomes
+(\emph{HI-II-14} and \emph{HI-TESTED}) all metrics yield very good
+results for the link prediction problem, and therefore this study
+extends previous work by other authors that only consider the predictive
+power of the metrics alone.
+
+It is important to note the difference in feature importance from
+rice and human interactomes. In the former, features have a more balanced
+importance distribution than in the latter dataset, whose model shows
+a very strong dependence on the handcrafted feature.
+
+With this study, a framework for link prediction on protein-protein
+interactions is presented and validated against the available information
+on rice and human interactomes, but can be easily extended to other
+organisms or even to other graphs not necessarily related to biology.
+
 \bibliographystyle{plain}

### 2020-11-06 camila/report/abstract.tex

\begin{abstract}
  Abiotic stresses are a main cause of extensive agricultural
  production losses worldwide.  This paper proposes a workflow to
  identify stress responsive genes in organisms: on input RNA
  sequencing read counts measured for genotypes under control and
  treatment conditions, and biological replicates, it outputs a
  collection of characterized genes, potentially relevant to
  treatment. Technically, the proposed approach is both a
  generalization and an extension of WGCNA. It is showcased with a
  systematic study on rice (\textit{Oryza sativa}), a major food
  source that is known to be highly sensitive to salt stress.  A total
  of 6 modules are detected as relevant in the response to salt stress
  in rice: 3 modules of 3 genes each, all associated with shoot K
  content, 2 modules of 3 genes associated with shoot biomass, and 1
  module of 4 genes associated with root biomass. These genes may act
  as potential targets for the improvement of salinity tolerance in
  rice cultivars.
  %% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}

### 2020-11-06 camila/report/case.tex

\section{Case Study}
\label{sec.case}
This section presents a case study on the identification of genes that may respond to saline stress according to the methodology presented in section~\ref{sec. framework}. 
The RNA-seq data was obtained from GEO database
\cite{GEOAcces90:online} (accession number GSE98455). This data
corresponds to $n_0=57845$ gene expression profiles of shoot tissues
measured for both control and salt condition in $m=92$ accessions of
the Rice Diversity Panel 1, with $r=2$ biological replicates. A total
of $p=3 $ phenotypic traits are used: shoot $K^+$ content, root
biomass and shoot biomass. These traits were measured for the same
$92$ genotypes, under control and salt stress conditions, and can be
found in the supplementary information
of~\cite{campbell2017allelic}.

\subsection{Data pre-processing}

DESeq2 normalization is applied to the raw data and the biological
replicates are averaged. Genes exhibiting low variance are
identified as those with ratio of upper quantile to lower quantile
smaller than $1.5$ and are removed from the normalized data. Genes
with low expression, corresponding to those having more than $80\%$
samples with values smaller than $10$, are also removed. A total of
$n_1 = 8928$ genes are kept after this filtering process.

From the Log Fold Change matrix $L_0$, genes whose difference between
upper quantile and lower quantile greater than $0.25$ are
removed. Therefore, the resulting matrix $L_1$ contains the log ratios
of $n_2 = 8928$ genes. The logarithmic ratios of the phenotipic data,
for the $92$ accessions and the $3$ traits, is also computed.

\subsection{Co-expression network construction}

The Log Fold Change matrix $L_1$ is used to compute the corresponding
similarity matrix.  For this network, it is observed that $\beta=3$
is the smallest integer such that the $R^2 \geq 0.8
$. Figure~\ref{fig:beta} depicts the degree distribution of the
similarity matrix (left) and the degree distribution of the adjacency
matrix (right), which is the degree distribution of a scale-free
network with $R^2 = 0.8$ with $\beta = 3$.

%scale-free beta 1 y 3
\begin{figure}[htpb]
  \centering
    \includegraphics[clip,width=0.8\textwidth]{figures/pick_beta.png}
  \caption{Degree distribution of $A$ (left) and $\hat{A}$ (right)}
  \label{fig:beta}
\end{figure}

The resulting adjacency matrix $A$ represents a complete graph
$G=(V,E)$, with $|V| = 8928$ genes and $|E| = 39850128$ edges.

\subsection{Co-expression module identification}

The adjacency matrix $A$ is transformed into an unweighted network 
$\hat{A}$ applying the approach described in~\cite{aoki2007approaches}, 
based on the density of the network combined with the decreasing number
of nodes and edges with higher PCC values. The cutoff value is set to $0.2$,
thus keeping only the connections above this threshold and removing the 
isolated nodes. The resulting adjacency matrix $\hat{A}$ has $5810$ connected 
genes and accounts for $16875145$ edges.

After applying the HLC algorithm, a total of $4131$ genes are
distributed in $c = 5143$ overlapping modules of at least $3$ genes. Figure~\ref{fig:overlap} presents a histogram of the overlapping
percentage of these genes, measured as the proportion of modules to
which each gene belongs. The first bar of the histogram represents the
genes with zero overlap, corresponding to $28\%$ of the total
genes; remaining $72\%$ represents the genes belonging to more than one
module.

%overlapping percentage
\begin{figure}[htpb]
  \centering
    \includegraphics[clip,width=0.5\textwidth]{figures/artificial_modules.png}
  \caption{Overlapping percentage}
  \label{fig:overlap}
\end{figure}

\subsection{Module association to phenotypic traits}

The phenotypic traits under study are shoot $K^+$ content, root
biomass and shoot biomass. Figure \ref{fig:pdata} suggests that there
are significant differences in the values of these phenotypic traits
between stress and control conditions. This supports the working
hypothesis that these three variables represent tolerance-associated
traits in rice under salt stress.

% boxplots
\begin{figure}[htpb]
  \centering
    \includegraphics[clip,width=1\textwidth]{figures/phenotypic_traits.png}
  \caption{Phenotipic traits distribution under control and salt stress.}
  \label{fig:pdata}
\end{figure}

Using the affiliation matrix $F$ derived from the HLC output and the
Log Fold Change matrix $L_1$, a matrix $M$ is built by computing the
egengene for each of the $c = 5143$ modules. The LASSO technique is
applied by using each of the phenotypic traits as the outcome
variable, one at a time. As shown in Figure~\ref{fig:cross-val},
cross-validation is performed for each phenotypical trait in order to
select the corresponding regularization parameter $\lambda$ that
minimizes the mean-squared error.

\begin{figure}[htpb]
  \centering
  \subfloat[$K$ shoot, $\lambda = 0.013$]{%
    \includegraphics[width=0.33\textwidth]{figures/lambda_kshoot_2.png}
    \label{fig:lambda_kshoot}
  }
  \subfloat[Shoot biomass, $\lambda = 0.025$]{%
    \includegraphics[width=0.33\textwidth]{figures/lambda_BMshoot_2.png}
    \label{fig:lambda_BMshoot}
  }
  \subfloat[Root biomass, $\lambda = 0.035$]{%
    \includegraphics[width=0.33\textwidth]{figures/lambda_BMroot.png}
    \label{fig:lambda_BMroot}
  }
  \caption{Cross-validation of the LASSO regularization parameter
    $\lambda$, for each phenotypic trait.}
  \label{fig:cross-val}
\end{figure}

Finally, three LASSO models are adjusted by using the corresponding
$\lambda$ and phenotypical data with the egengens of matrix $M$. As
result, 6 modules are detected as relevant in the response to salt
stress in rice: 3 modules of 3 genes, each associated with shoot
$K$ content; 2 modules of 3 genes associated with shoot biomass; and 1
module of 4 genes associated with root biomass.

\subsection{Gene enrichment}

From the $19$ genes selected by LASSO, all but $3$ genes (the ones associated to $K$ content), are also identified as deferentially expressed ($|\ell_{ij}| \geq 2$) for at least one of the $92$ accessions. This suggest that those genes are strong candidates as stress responsive genes to salt conditions in rice. According to the Quickgo database, only $2$ of the $16$ diferentially expressed genes (both from the module related with shoot biomass) are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase. Figure~\ref{fig:3d} shows their corresponding 3D protein structures.

% Figure 3d protein strucures
\begin{figure}[hptb]
  \centering
  \subfloat[Spermidine hydroxycinnamoyltransferase 2]{%
    \includegraphics[width=0.45\textwidth]{figures/structure_LOC_Os12g27254.png}
    \label{fig:structure_SHT2}
  }\hfill
  \subfloat[Lipoxygenase]{%
    \includegraphics[width=0.45\textwidth]{figures/structure_LOC_Os12g37260.png}
    \label{fig:structure_Lipoxygenase}
  }
  \caption{3D protein structure of named genes selected by LASSO, available from~\cite{szklarczyk2016string}.}
  \label{fig:3d}
\end{figure}

Uniprot database~\cite{uniprot2018uniprot} reports, on the one hand,
that SHT2 contributes to the natural variation of spermidine-based
phenolamides in rice cultivars. On the other hand, it is reported that
plant lipoxygenase may be involved in a number of diverse aspects of
plant physiology including growth and development, pest resistance,
and senescence or responses to wounding~\cite{uniprot2018uniprot}. This protein is involved in
the pathway oxylipin biosynthesis, which is part of Lipid
metabolism. Aditionally, previous studies
%
in~\cite{gupta2013plant,hou2015persimmon,mittova2002salt,peng2019novel,roychoudhury2011amelioration}
%
provide evidence of biological implications of sperimidine and
lipoxygenase in tolerance to salt stress in other plants or even in
rice cultivars. However, further studies are needed to elucidate the
detailed biological function of the remaining 14 genes that have not
been named so far, which may have a potential relevance in stress
responsive mechanisms to salt conditions in rice.

### 2020-11-18 miguel/xgbstrc/model/plots.py
new code added

@@ -1,0 +1,207 @@
+#!/usr/bin/env python
+# coding: utf-8
+
+# Imported Libraries
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+
+# Default colors
+COLORS = ['#1f77b4', '#ff7f0e', '#2ca02c', '#d62728', '#9467bd', '#8c564b', '#7f7f7f', '#bcbd22', '#17becf']
+
+
+#!/usr/bin/env python
+# coding: utf-8
+
+# Gene function prediction - Plot function
+# Miguel Romero, nov 3rd
+
+import os
+import sys
+import numpy as np
+import pandas as pd
+
+# Ploting
+import seaborn as sns
+from matplotlib import rc
+from matplotlib import pyplot as plt
+
+rc('font', family='serif', size=18)
+rc('text', usetex=False)
+
+# Default colors
+COLORS = ['#1f77b4', '#ff7f0e', '#2ca02c', '#d62728', '#9467bd', '#8c564b',
+          '#7f7f7f', '#bcbd22', '#17becf']
+
+
+
+# plot auc roc curve
+def plot_roc(fpr, tpr, auc, filename, path=None):
+  fig, ax = plt.subplots(figsize=(6.5,6.5))
+  plt.plot(fpr, tpr, lw=2, label='AUC = {0:.2f}'.format(auc))
+  plt.plot([0, 1], [0, 1], color='navy', lw=2, linestyle='--')
+  plt.xlim([0.0, 1.0])
+  plt.ylim([0.0, 1.0])
+  plt.xlabel('1 - Specificity')
+  plt.ylabel('Sensitivity')
+  plt.legend(loc='lower right')
+  figname = ('{0}/'.format(path) if path != None else '') + filename
+  plt.savefig(figname, format='pdf', dpi=600)
+  plt.close()
+
+
+# plot multiple auc roc curves in one figure
+def plot_mroc(fprl, tprl, aucl, labels, filename, path=None):
+  fig, ax = plt.subplots(figsize=(6.5,6.5))
+  for fpr, tpr, auc, l in zip(fprl, tprl, aucl, labels):
+    plt.plot(fpr, tpr, lw=2, label='{0} AUC = {1:.2f}'.format(l, auc))
+  plt.plot([0, 1], [0, 1], color='navy', lw=2, linestyle='--')
+  plt.xlim([0.0, 1.0])
+  plt.ylim([0.0, 1.0])
+  plt.xlabel('1 - Specificity')
[…]

### 2020-12-07 camila/report/abstract.tex
bmc version

@@
 \begin{abstract}
-  Abiotic stresses are a main cause of extensive agricultural
-  production losses worldwide.  This paper proposes a workflow to
-  identify stress responsive genes in organisms: on input RNA
-  sequencing read counts measured for genotypes under control and
-  treatment conditions, and biological replicates, it outputs a
-  collection of characterized genes, potentially relevant to
+  This paper proposes a workflow to identify genes responding to a
+  specific treatment in an organism, such as abiotic stresses, a main
+  cause of extensive agricultural production losses worldwide.  On
+  input RNA sequencing read counts (measured for genotypes under
+  control and treatment conditions) and biological replicates, it
+  outputs a collection of characterized genes, potentially relevant to
   treatment. Technically, the proposed approach is both a
-  generalization and an extension of WGCNA. It is showcased with a
-  systematic study on rice (\textit{Oryza sativa}), a major food
-  source that is known to be highly sensitive to salt stress.  A total
-  of 6 modules are detected as relevant in the response to salt stress
-  in rice: 3 modules of 3 genes each, all associated with shoot K
-  content, 2 modules of 3 genes associated with shoot biomass, and 1
-  module of 4 genes associated with root biomass. These genes may act
-  as potential targets for the improvement of salinity tolerance in
-  rice cultivars.
-  %% \keywords{First keyword  \and Second keyword \and Another keyword.}
+  generalization and an extension of WGCNA; its main goal is to
+  identify specific modules in a network of genes after a sequence of
+  normalization and filtering steps. In this work, module detection is
+  achieved by using Hierarchical Link Clustering, which can recognize
+  overlapping communities and thus have more biological meaning given
+  the overlapping regulatory domains of systems that generate
+  co-expression. Additional steps and information are also added to
+  the workflow, where some networks in the intermediate steps are
+  forced to be scale-free and LASSO regression is employed to select
+  the most significant modules of phenotypical responses to stress.
+  %
+  %% Biologists may use these genes to develop new crop cultivars with
+  %% higher resistance to a specific treatment.
+  %
+  Finally, the workflow is showcased with a systematic study on rice
+  (\textit{Oryza sativa}), a major food source that is known to be
+  highly sensitive to salt stress: a total of 6 modules are detected
+  as relevant in the response to salt stress in rice;
+  %
+  %% : 3 modules of 3 genes each, all associated with shoot K content,
+  %% 2 modules of 3 genes associated with shoot biomass, and 1 module
+  %% of 4 genes associated with root biomass.
+  %
+  these genes may act as potential targets for the improvement of
+  salinity tolerance in rice cultivars.
+  %
+  The proposed workflow has the potential to ultimately reduce the
+  search-space for candidate genes responding to a specific treatment,
+  which can considerably optimize the effort, time, and money invested
+  by researchers in the experimental validation of stress responsive
+  genes.
+  \keywords{stress responsive genes \and co-expression network 
+            \and overlapping communities \and phenotypic traits
+            \and LASSO regression.}
 \end{abstract}

### 2020-12-07 camila/report/intro.tex
bmc version

--- 2020-10-08 camila/report/intro.tex
@@
-\section{Introduction}
+\section*{Introduction}
 \label{sec.intro}
 
-Abiotic stresses are key factors that can negatively influence plant
-development and productivity. They are a main cause of extensive
-agricultural production losses worldwide~\cite{todo}. Soil salinity is
-one of the most devastating abiotic stresses, causing reduction in the
-cultivable land, crop quality, and productivity. It has been estimated
-that 20\% of total cultivated and 33\% of irrigated agricultural lands
-worldwide are already affected by high salinity. Moreover, due to the
-human activities and natural causes, salinized areas are gradually
-increasing every year and are expected to reach 50\% by the end of
-year 2050~\cite{shrivastava2015soil}. Salinity tolerance and
-susceptibility in plants is known to be the result of elaborated
+Stresses are key factors that influence plant
+development, often associated to extensive
+losses in agricultural production~\cite{mesterhazy2020losses, shrivastava2015soil}. 
+Soil salinity is one of the
+most devastating abiotic stresses. According to~\cite{shrivastava2015soil}, 
+soil salinity contributes to a significant reduction in areas of cultivable
+land and crop quality. The study estimates that 
+20\% of the total cultivated land worldwide and 
+33\% of the total irrigated agricultural land
+is affected by high salinity. 
+By the end of 2050 areas of high salinity are
+expected to reach 50\% of the cultivated land~\cite{shrivastava2015soil}. 
+\vspace{0.5cm}
+
+Salinity tolerance and
+susceptibility are the result of elaborated
 interactions between morphological, physiological, and biochemical
-processes that are regulated, in the end, by multiple genes in
-different parts of a genome~\cite{reddy2017salt}. Therefore,
-identifying groups of stress responsive genes may lead to crop
-improvement in terms of salinity tolerance and, ultimately, contribute
-solutions to the general problem of food sustainability in the years
-to come.
+processes, which are regulated by multiple genes in
+various parts of the plant genome~\cite{reddy2017salt}. Consequently,
+identifying groups of responsive genes is an important step in efforts to
+improve crop varieties in terms of salinity tolerance.
+This paper proposes a workflow to identify stress responsive genes
+associated with a complex quantitative trait.
+\vspace{0.5cm}
 
-This paper proposes a workflow to identify stress responsive genes in
-organisms, which is known to be a complex quantitative trait. It takes
-as input RNA sequencing read counts measured for genotypes under
-control and treatment conditions (and representing gene expression
-profiles of the target organism), and biological replicates.  In order
-to discover key genes and their interaction with phenotypes related to
-treatment tolerance, the approach requires a collection of phenotypic
-traits (under control and under treatment), measured for the given
-genotypes. The output of the workflow is a collection of characterized
-genes, potentially relevant to treatment, yielding insight on the
-possible behavior of specific genes and the role they may play in
-functional pathways in response to the studied treatment of the
-organism of interest. The proposed workflow can thus take advantage of
-transcriptomic data for different organisms and conditions, based on
-the current availability of high-throughput technologies that include
-microarrays and RNA sequencing, to study the reaction of organisms
-under different environmental stimuli, such as salt stress.
+To discover which genes are associated with a phenotypic response to
+treatment the workflow takes as input
+the gene expression profiles of the target organism, specifically, 
+the RNA sequencing read counts (measured under control
+and treatment conditions)
+of at least two biological replicates per genotype. It also receives phenotypic data, specifically, observable traits, measured for each genotype under the two conditions. The output of the
+workflow is a set of genes which are characterized as potentially relevant
+to treatment. 
+\vspace{0.5cm}
 
-Technically, the proposed approach is both a generalization and an
-extension of Weighted Gene Co-expression Network Analysis
-(WGCNA)~\cite{todo2}, a widely applied workflow that has been
-successfully used for identifying target genes related to diseases and
-cancer in several organisms~\cite{tian2018identifying}. The general
-idea behind each approach is to identify specific modules in a network
-of genes after a sequence of normalization and filtering steps. The
-proposed approach is considered a \textit{generalization} of WGCNA
-because module detection can now recognize overlapping communities,
-which may have more biological meaning given the overlapping
-regulatory domains of systems that generate
-co-expression~\cite{gaiteri2014beyond}. This is achieved by using
-Hierarchical Link Clustering (HLC)~\cite{ahn2010link}. It is also an
-\textit{extension} of WGCNA because additional steps and information
-are added to the workflow: namely, some networks in the intermediate
-steps are forced to be scale-free~\cite{todo3} and LASSO
-regression~\cite{tibshirani1996regression} is employed to select the
-most significant modules of phenotypical responses to stress.  The
-advantage of using HLC as clustering method is its ability to detect
-overlapping modules, since biological components are involved in
-multiple functions and therefore biological communities tend to be
-highly overlapping. On the other hand, LASSO is a regularized
-regression technique widely used in variable selection, thanks to its
-ability to obtain zero regression coefficients for the less relevant
-variables~\cite{desboulets2018review}. Moreover, LASSO is especially
-useful in problems where the number of variables is much larger than
-the number of samples, which may be the case more often than desired.
-The proposed workflow is also modular, since other module detection
-and selection techniques could be used, instead HLC and LASSO,
-respectively.
+Broadly speaking, the workflow provides a framework that yields insight into the possible behavior of specific
+genes and the role they play in functional pathways in response to
+treatment. 
+It takes advantage of
+the current availability of high-throughput technologies, which
+enables the access to transcriptomic data of organisms under different conditions, 
+and a better understanding of their reaction under different
+environmental stimuli.
+\vspace{0.5cm}
 
-The proposed workflow is showcased with a systematic study on rice
-(\textit{Oryza sativa}), a major food source that is known to be
+The proposed approach is both a generalization and an
+extension of the widely applied workflow 
+for identifying target genes called 
+Weighted Gene Co-expression Network Analysis
+(WGCNA)~\cite{langfelder2008wgcna, tian2018identifying}.
+Like WGCNA, the general idea behind the proposed
+approach is to identify, after a sequence of normalization
+and filtering steps, specific modules of overlapping communities
+underlying the co-expression network of genes. 
+The proposed approach is considered a \textit{generalization}
+of WGCNA because module detection recognizes overlapping communities
+using Hierarchical Link Clustering (HLC)~\cite{ahn2010link} algorithm.
+Conceptually, the generalization takes into account
+the overlapping nature of the regulatory domains of the
+systems that generate the co-expression network~\cite{gaiteri2014beyond}.
+More specifically, overlapping modules allow for 
+scenarios where biological components  are involved in
+multiple functions.
+\vspace{0.5cm}
+
+The workflow is also an \textit{extension} of WGCNA
+because two additional constraints are considered:
+networks in the intermediate steps are forced to be
+scale-free~\cite{barabasi2003scale}, and LASSO
+regression~\cite{tibshirani1996regression} selects the
+most relevant modules of responsive genes.
+The regularized regression technique of LASSO
+forces the less relevant variables to be associated to regression
+coefficients of value zero~\cite{desboulets2018review}, 
+which is of interest
+in scenarios where the number of variables is much larger than
+the number of samples. This condition is satisfied when the target variables represent the
+overlapping communities (obtained with HLC) and the samples represent genotype data,
+which is usually a small set due to the high cost of the RNA sequencing process. 
+Finally, the proposed workflow is also modular, since other module detection
+and selection techniques could be explored instead HLC and LASSO.
+\vspace{0.5cm}
+
+The approach is showcased with a systematic study on rice
+(\textit{Oryza sativa}), a food source that is known to be
 highly sensitive to salt stress~\cite{chang2019morphological}. RNA-seq
-data was accessed from the GEO database~\cite{GEOAcces90:online}
-(accession number GSE98455). It corresponds to $57845$ gene expression
-profiles of shoot tissues measured for both control and salt condition
-in $92$ accessions of the Rice Diversity Panel 1. As output, 6 modules
+data was accessed from the GEO database~\cite{clough2016gene}
+(accession number GSE98455). It represents $57845$ gene expression
+profiles of shoot tissues measured under control and stress conditions
+in $92$ accessions of the Rice Diversity Panel 1~\cite{eizenga2014registration}. A total of 6 modules
 are detected as relevant in the response to salt stress in rice: 3
-modules of 3 genes each, all associated with shoot K content, 2
-modules of 3 genes associated with shoot biomass, and 1 module of 4
-genes associated with root biomass. These genes may act as potential
-targets for the improvement of salinity tolerance in rice
-cultivars. From the 19 genes, all but 3 genes (associated with $K$
-content), were also identified as deferentially expressed for at least
-one of the 92 accessions, suggesting that those genes are strong
-candidates as stress responsive genes. Only 2 of the 16 diferentially
-expressed genes, both from the module related with shoot biomass, are
-named and have an associated protein product: Spermidine
-hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase. In other words,
-further studies are needed to elucidate the detailed biological
-function of the remaining 14 genes that have not been named so far,
-which may have a potential relevance in stress responsive mechanisms
-to salt conditions in rice. The goal is that the results reported in
-this paper may allow biologist to develop new rice cultivars with
-higher resistance to salinity.
+modules, each grouping together 3 genes, are associated to shoot K content; 2
+modules of 3 genes, are associated to shoot biomass; and 1 module of 4
+genes is associated to root biomass. These genes are potential
+targets for experimental validation of salinity tolerance. 
+From the 19 genes, 16
+are also identified as deferentially expressed for at least
+one of the 92 accessions, which re-enforces the labelling of the genes as
+stress responsive. Only 2 of the 16 differentially
+expressed genes, associated to shoot biomass, are
+named and known to produce protein products: spermidine
+hydroxycinnamoyltransferase~2 (SHT2) and lipoxygenase. 
+Further studies are needed to elucidate the detailed biological
+functions of the remaining 14 genes and their role
+in the mechanisms that respond 
+to salt conditions.
+
+\paragraph{Paper Outline.}
+%The remainder of the paper is organized as
+%follows. Section~\ref{sec.prelim} gathers preliminaries on gene
+%co-expression networks, HLC, and LASSO. The proposed workflow is
+%presented in Section~\ref{sec.framework}, with especial focus on the
+%logical steps of the process and the internal structures supporting
+%the approach. Section~\ref{sec.case} presents the case study on the
+%identification of rice genes that are sensitive to salt
+%stress. Section~\ref{sec.concl} draws some conclusions and future research directions.
+
+The remainder of the paper is organized as
+follows. The \hyperref[sec.prelim]{Preliminaries} section gathers foundations on gene
+co-expression networks, HLC, and LASSO. The proposed workflow is
+presented in \hyperref[sec.framework]{the Workflow} section, which enphasizes on the
+logical steps of the data analysis process and the internal structures supporting
+the approach. The \hyperref[sec.case]{Case Study} section presents an application of the workflow for the
+identification of rice genes that are sensitive to salt
+stress. Finally, the \hyperref[sec.concl]{Concluding Remarks} section draws some conclusions and future research directions.

### 2020-12-07 camila/report/main.tex
bmc version

@@
 %
-\title{Using Complex Networks for Identifying Potentially New Saline
-  Stress Responsive Genes in Rice}
+\title{Using Overlapping Communities and Network Structure for
+  Identifying Reduced Groups of Stress Responsive Genes}
+% \titlerunning{ }
+\titlerunning{Using Network Features for Identifying Stress Responsive Genes}
+%
+%% \title{Using complex networks for identifying a reduced group of stress responsive genes}
 %
@@
 \institute{Pontificia Universidad Javeriana, Cali, Colombia
-  %% \email{lncs@springer.com}
+  \email{\{camila.riccio,jfinke,camilo.rocha\}@javerianacali.edu.co}
 }  

### 2020-12-07 camila/report/prelim.tex
bmc version

@@
-\section{Preliminaries}
+\section*{Preliminaries}
 \label{sec.prelim}
 
-This section presents preliminaries on networks, hierarchical link
-clustering and the LASSO linear regression technique.
+This section presents preliminaries on networks, the clustering algorithm HLC,
+and the linear regression technique LASSO.
 
-\subsection{Co-expression network}
+\subsection*{Co-expression network}
 
-A network is an undirected graph $G=(V,E)$ where
-${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of \textit{vertices} or
-\textit{nodes} and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of
-\textit{edges} or \textit{links} that connect vertices. In a gene
-co-expression network, each node corresponds to a gene. A pair of
-genes is connected if they show similar expression
-patterns. A simple and unweighted network can be represented by an
-adjacency matrix $A \in \{0,1\}^{n \times n}$ that is symmetric with a
-positive one in the positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever
-there is an edge connecting vertices $v_i$ and $v_j$, and zeros
-elsewhere. Co-expression networks are of biological interest because
-the co-expressed genes are usually controlled by the same
-transcriptional regulatory pathway, functionally related or members of
-the same pathway or metabolic complex.
+Consider a network as an undirected graph $G=(V,E)$ where
+${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of $n$ \textit{vertices} (or
+\textit{nodes}) and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of $q$
+\textit{edges} (or \textit{links}) that connect vertices. In a
+co-expression network of genes, each node corresponds to a gene, and
+a link indicates similar expression pattern between two genes.
+The network can be represented by an adjacency matrix $A
+\in \{0,1\}^{n \times n}$ that is symmetric. A matrix entry in 
+positions $(v_i,v_j)$ and $(v_j,v_i)$ is equal to $1$ whenever there is an edge
+connecting vertices $v_i$ and $v_j$, and equal to $0$
+otherwise. Co-expression networks are of biological interest because
+adjacent nodes in the network represent co-expressed genes that are
+usually controlled by the same transcriptional regulatory pathway,
+functionally related, or members of the same pathway or metabolic
+complex~\cite{FIONDA2019915}.
 
-\subsection{Hierarchical Link Clustering}
+\subsection*{Hierarchical Link Clustering}
 
-The Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn
-et al.~\cite{ahn2010link}. The HLC approach represents communities as
-groups of links (rather than nodes), and each node inherits all
-memberships of its links and can thus belong to multiple, overlapping
-communities. It maps links to nodes and connects them if a pair of
-links shares a node. The similarity between two links $e_{ik}$ and
-$e_{jk}$ is computed using the Jaccard index
+The Hierarchical Link Clustering (HLC) algorithm partitions
+groups of links (rather than nodes), each node inherits all
+memberships of its links and can belong to multiple, overlapping
+communities~\cite{ahn2010link}. More specifically, HLC evaluates the similarity 
+between links if they share a particular node.
+Consider a pair of links $e_{ik}$ and $e_{jk}$, which are adjacent to node $k$.
+The similarity between $e_{ik}$ and $e_{jk}$ is defined based on the Jaccard index as
 %
 \begin{equation}\label{eq:jaccard}
-  S(e_{ik},e_{jk}) = \frac{\vert n(i) \cap n(j) \vert}{\vert n(i) \cup n(j) \vert},
+  S(e_{ik},e_{jk}) = \frac{\vert \ \eta(i) \cap \eta(j) \ \vert}{\vert \ \eta(i) \cup \eta(j) \ \vert},
 \end{equation}
 %
-where $n(i)$ denotes the set containing exactly node $i$ and its
+where $\eta(i)$ denotes the set containing node $i$ and its
 neighbors. The algorithm uses single-linkage hierarchical clustering
-to build a dendrogram in which each leaf is a link from the original
-network and branches represent link communities. Hierarchical
-clustering algorithms repeatedly merge groups until all elements are
-members of a single cluster. For the purpose of finding meaningful
-communities, it is crucial to know where to partition the
-dendrogram. In this case, the most relevant communities are
-established at the maximal partition density $D$, a function based on
-link density inside communities measuring the quality of a link
-partition. The partition density $D$ has a single global maximum along
-the dendrogram in almost all cases, because its value is the average
-density at the top of the dendrogram (a single giant community with
-every link and node) and it is very small at the bottom of the
-dendrogram (most communities consists of a single link). In
-particular, it is the case that $D = 1$ when every community is a
-fully connected clique and $D = 0$ when each community is a tree. If a
-community is less dense than a tree (i.e. when the community subgraph
-has disconnected components), then such a community contribute
-negatively to $D$, which can take negative values. The minimum density
-inside a community is $-2/3$, given by one community of two
-disconnected edges. Since $D$ is the average of the intra-community
-density, there is a lower bound of $-2/3$ for $D$. Computing $D$ at
-each level of the link dendrogram can help the purpose of picking the
-best level to cut, although meaningful structure could exist above or
-below the threshold. The output of cutting is a set of node clusters,
-where each node can participate in multiple communities.
+to build a dendrogram in which each leaf is a link from the
+network, and branches represent link communities. 
+%Hierarchical
+%clustering algorithm repeatedly merge groups until all elements are
+%members of a single cluster.
+\vspace{0.5cm}
 
-\subsection{Least Absolute Shrinkage Selector Operator}
+The threshold where to cut the dendogram is defined
+based on the average density of links inside communities (partition density).
+For $G=(V,E)$ and a partition of the links into $c$ subsets, the partition density is computed as
 
-The Least Absolute Shrinkage Selector Operator (LASSO) is a
-regularized linear regression technique. It combines a regression
-model with a procedure of contraction of some parameters towards zero
-and selection of variables, imposing a restriction or a penalty on the
+\begin{equation}
+D = \frac{2}{\vert E \vert} \sum_c \vert E_c \vert \frac{\vert E_c \vert - \vert V_c \vert + 1 }{(\vert V_c \vert -1)(\vert V_c \vert -2)}
+\end{equation}
+
+Note that in most cases, the partition density $D$ has a single
+global maximum along the dendrogram. 
+If the dendogram is cut at the top, then $D$ represents the
+average link density of a single
+giant community.  If the dendogram is cut at the
+bottom, then most communities consists of a single
+link. In other words, note that $D = 1$ when every community
+is a clique and $D = 0$ when each community is a
+tree. If a community is less dense than a tree (i.e., when the
+community subgraph has disconnected components), then such a community
+contributes negatively to $D$, which can take negative values. The
+minimum density inside a community is $-2/3$, given by one community
+of two disconnected edges. Since $D$ is the average of the
+intra-community density, there is a lower bound of $-2/3$ for
+$D$. By computing $D$ at each level of the dendrogram, 
+the level that maximizes partition density can be found 
+(nonetheless meaningful
+structure could exist above or below the threshold).
+
+% figure 0
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[clip,width=0.96\textwidth]{figures/figure0.png}
+  \caption[Example of a full link dendrogram (left) and partition density (right)]%
+  {Example of a full link dendrogram (left) and partition density (right), borrowed from~\cite{ahn2010link}.}
+  \label{fig:hlc_density}
+\end{figure}
+
+The output of
+cutting is a set of node clusters, where each node can participate in
+multiple communities.
+
+\subsection*{Least Absolute Shrinkage Selector Operator (LASSO)}
+
+LASSO is a
+regularized linear regression technique. By combining a regression
+model with a procedure of contraction of some parameters towards $0$,
+LASSO imposes a restriction (or a penalty) on
 regression coefficients. In other words, LASSO solves the least
 squares problem with restriction on the $ L_1$-norm of the coefficient
-vector. It can be especially useful to solve problems where the number
-of variables (e.g., genes) $ n $ is much greater than the number of
-samples $ m $ (i.e., $ n \gg m $).
+vector. In particular, the approach is especially useful in scenarios where the number
+of variables $ c $ is much greater than the number of
+samples $ m $ (i.e., $ c \gg m $).
+\vspace{0.5cm}
 
-Consider a dataset consisting of $m$ samples, each of which consists
-of $n$ covariates and a single outcome. Let $y_i$ be the outcome and
-$x_i := (x_1,...,x_n)$ be the covariate vector for the $i$-th
+Consider a dataset of $m$ samples, consisting each
+of $c$ covariates and a single outcome. Let $y_i$ be the outcome and
+$x_i := (x_{i1},...,x_{ic})$ be the covariate vector for the $i$-th
 sample. The objective of LASSO is to solve
@@
 \begin{equation}
-\min \left\lbrace\sum_{i=1}^{m}{\left( y_i-\sum_{j=1}^n{\beta_j
+\min \left\lbrace\sum_{i=1}^{m}{\left( y_i-\sum_{j=1}^c{\beta_j
     x_{ij}}\right)^2} \right\rbrace \quad , \quad \textrm{subject to}
-\quad \sum_{j=1}^n\abs{\beta_j}\leq s.
+\quad \sum_{j=1}^c\abs{\beta_j}\leq s.
 \end{equation}
 
-Equivalently, in the Lagrangian form, it minimizes
+where $s$ is the regularization penalty. 
+Equivalently, in the Lagrangian form, LASSO minimizes
 
 \begin{equation}
-  \sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} +
-  \lambda \sum_{j=1}^n\abs{\beta_j}
+  \sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^c{\beta_j x_{ij}}\right)^2} +
+  \lambda \sum_{j=1}^c\abs{\beta_j}
 \end{equation}
 %
-where $s$ is the regularization penalty and $\lambda \geq 0$ is the
-corresponding Lagrange multiplier. Since the $\lambda$ value
-determines the degree of penalty, the accuracy of the model depends on
-its choice. Cross-validation is often used to select the
-regularization parameter, choosing the one that minimizes the
+where $\lambda \geq 0$ is the
+corresponding Lagrange multiplier. Since the value of the regularization
+parameter $\lambda$
+determines the degree of penalty and the accuracy of the model,
+cross-validation is used to select the
+regularization parameter that minimizes the
 mean-squared error.

### 2020-12-07 camila/report/abstract-bmc.tex
bmc version

\begin{abstract} % abstract
\parttitle{Background} %if any
This paper proposes a workflow to identify which genes respond to
  specific treatments in plants.
  The workflow takes as
  input the RNA sequencing read counts and phenotypical data of different genotypes,
  measured under  control and treatment conditions.
  It outputs a reduced group of genes marked as relevant for
  treatment response. Technically, the proposed approach is both a
  generalization and an extension of 
%  Weighted Gene Co-expression Network Analysis (WGCNA).
  WGCNA.
  It aims to 
  identify specific modules of overlapping communities underlying the co-expression network of genes.
  Module detection is
  achieved by using Hierarchical Link Clustering.
  Identifying such modules enables us to take into account
  the overlapping nature of the regulatory domains of the systems that generate
  co-expression. LASSO regression is employed to analyze
  phenotypic responses of modules to treatment.

\parttitle{Results} %if any
  The workflow is applied to rice
  (\textit{Oryza sativa}), a major food source known to be
  highly sensitive to salt stress.
  The workflow identifies 19 rice genes that seem
  relevant in the response to salt stress.
  They are
  distributed across 6 modules: 
  3 modules, each grouping together 3 genes, are associated to shoot K content;
  2 modules of 3 genes, are associated to shoot biomass;
  and 1 module of 4 genes is associated to root biomass.
  These genes represent target genes for the improvement of
  salinity tolerance in rice.

\parttitle{Conclusions}
A more effective framework to reduce the
  search-space for target genes that respond to a specific treatment is introduced.
  It facilitates experimental validation by restraining efforts to
  a smaller subset of genes of high potential relevance.

\end{abstract}

### 2020-12-07 camila/report/bmc_article.tex
bmc version

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% https://www.biomedcentral.com/getpublished                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% https://miktex.org/                                             %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

%%% Load packages
\usepackage{amsthm,amsmath}
\usepackage{amsfonts}

\usepackage{graphicx}
%\usepackage[caption=false]{subfig}

%\RequirePackage[numbers]{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{lmodern} %font sizes
\usepackage{hyperref} % ref unlabeled sections

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\def\includegraphic{}
%\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\newcommand{\abs}[1]{\left\vert#1\right\vert}
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{METHODOLOGY}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Using Overlapping Communities and Network Structure for
  Identifying Reduced Groups of Stress Responsive Genes}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
  addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
  corref={aff1},                      % id of corresponding address, if any
% noteref={n1},                        % id's of article notes, if any
  email={camila.riccio@javerianacali.edu.co}   % email address
]{\inits{C.R.}\fnm{Camila} \snm{Riccio}}
\author[
  addressref={aff2},
  email={jfinke@javerianacali.edu.co}
]{\inits{J.F.}\fnm{Jorge} \snm{Finke}}
\author[
  addressref={aff2},
  email={camilo.rocha@javerianacali.edu.co}
%]{\inits{C.R.}\fnm{Camilo} \snm{Rocha}}
]{\inits{H.C.R.}\fnm{Hernan C.} \snm{Rocha}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                                         % unique id
  \orgdiv{Department of Natural Sciences and Mathematics},  % department, if any
  \orgname{Pontificia Universidad Javeriana},               % university, etc
  \city{Cali},                                              % city
  \cny{Colombia}                                            % country
}

\address[id=aff2]{%                                         % unique id
  \orgdiv{Department of Electronics and Computer Science},  % department, if any
  \orgname{Pontificia Universidad Javeriana},               % university, etc
  \city{Cali},                                              % city
  \cny{Colombia}                                            % country
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{artnotes}
%%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
%\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                           %%
%% The Abstract begins here                  %%
%%                                           %%
%% Please refer to the Instructions for      %%
%% authors on https://www.biomedcentral.com/ %%
%% and include the section headings          %%
%% accordingly for your article type.        %%
%%                                           %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract
\parttitle{Background} %if any
This paper proposes a workflow to identify which genes respond to
  specific treatments in plants, such as abiotic stresses.
  The workflow takes as
  input RNA sequencing read counts and phenotypical data (measured for genotypes under
  control and treatment conditions) with biological replicates. It
  outputs a collection of genes marked as potentially relevant to
  treatment. Technically, the proposed approach is both a
  generalization and an extension of WGCNA. Its main goal is to
  identify, after a sequence of normalization and filtering steps, specific modules in a gene co-expression network.
  Module detection is
  achieved by using Hierarchical Link Clustering, to recognize
  overlapping gene communities. Such communities have a biological meaning given
  the overlapping regulatory domains of systems that generate the patterns of
  co-expression. Additional steps and information are also added to
  the workflow, where networks in intermediate steps are
  forced to be scale-free. Finally, LASSO regression is employed to select
  the most significant modules of phenotypical responses to stress.

\parttitle{Results} %if any
The workflow is showcased with a systematic study on rice
  (\textit{Oryza sativa}), a major food source known to be
  highly sensitive to salt stress. A total of 19 genes distributed across 6 modules are detected
  as relevant in the response to salt stress.
  These genes represent potential targets for the improvement of
  salinity tolerance in rice crops.

\parttitle{Conclusions}
The proposed workflow provides a framework to reduce the
  search-space for candidate genes that respond to a specific treatment.
  It aims to optimize the effort, time, and resources researchers invest
  in the experimental validation of stress responses.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{Stress-responsive genes}
\kwd{co-expression network}
\kwd{overlapping communities}
\kwd{phenotypic traits}
\kwd{LASSO}
\kwd{salinity}
\kwd{rice}
\kwd{\textit{Oryza sativa}}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for two column layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                            %%
%% The Main Body begins here                  %%
%%                                            %%
%% Please refer to the instructions for       %%
%% authors on:                                %%
%% https://www.biomedcentral.com/getpublished %%
%% and include the section headings           %%
%% accordingly for your article type.         %%
%%                                            %%
%% See the Results and Discussion section     %%
%% for details on how to create sub-sections  %%
%%                                            %%
%% use \cite{...} to cite references          %%
%%  \cite{koon} and                           %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}      %%
%%                                            %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Introduction}
\label{sec.intro}

Abiotic stresses are key factors that influence plant
development and productivity. They are often associated to extensive
losses in agricultural production~\cite{mesterhazy2020losses}. 
Soil salinity is one of the
most devastating abiotic stresses. According to~\cite{shrivastava2015soil}, 
soil salinity contributes significantly to the reduction in cultivable
land and crop quality. The study estimates that worldwide 20\%
of the total cultivated land and 33\% of the total irrigated agricultural land
is already affected by high salinity. Moreover, due
to human activities and natural causes, salinized areas are
expected to reach 50\% of the total cultivated land by the
end of 2050~\cite{shrivastava2015soil}. Salinity tolerance and
susceptibility in plants is known to be the result of elaborated
interactions between morphological, physiological, and biochemical
processes that are regulated, in the end, by multiple genes in
various parts of the plant genome~\cite{reddy2017salt}. Therefore,
identifying groups of stress responsive genes is an important step in efforts to
improve crop varieties in terms of salinity tolerance and, more generally,
food sustainability.
\vspace{0.5cm}

This paper proposes a workflow to identify stress responsive genes
(known to be a complex quantitative trait).
To discover which genes are related with the phenotypic response to
treatment, the approach requires a collection of phenotypic
traits, measured for the given genotypes.  The proposed workflow takes as input
gene expression profiles of the target organism, specifically, 
RNA sequencing read counts (measured for genotypes under control
and treatment conditions)
with at least two biological replicates. It also receives phenotypic data, i.e., observable characteristics or traits, measured for each genotype under the two conditions. The output of the
workflow is a collection of genes which are characterized as potentially relevant
to treatment. The workflow provides a framework that yields insight into the possible behavior of specific
genes and the role they play in functional pathways in response to
treatment in the organism of interest. Bluntly put, the
workflow takes advantage of transcriptomic data for
organisms under different conditions, based on the current availability of
high-throughput technologies,
to study the reaction of organisms under different
environmental stimuli.
\vspace{0.5cm}

The proposed approach is both a generalization and an
extension of Weighted Gene Co-expression Network Analysis
(WGCNA)~\cite{langfelder2008wgcna}, a widely applied workflow that has
been successfully used for identifying target genes related to
diseases in several
organisms~\cite{tian2018identifying}. The general idea behind each
approach is to identify, after a sequence of normalization and filtering steps, specific modules in a gene co-expression network. 
The proposed approach
is considered a \textit{generalization} of WGCNA because module
detection can now recognize overlapping communities, which have
a biological meaning given the overlapping regulatory domains of
systems that generate the patterns of co-expression~\cite{gaiteri2014beyond}. This is
achieved by using Hierarchical Link Clustering
(HLC)~\cite{ahn2010link}. 
HLC enables the detection of overlapping modules, which naturally takes
into account that biological components  are involved in
multiple functions and therefore biological communities tend to be
highly overlapping.
The workflow is also an \textit{extension} of WGCNA
because additional constraints are considered:
namely, networks in the intermediate steps are forced to be
scale-free~\cite{barabasi2003scale} and LASSO
regression~\cite{tibshirani1996regression} is employed to select the
most significant modules of phenotypical responses to stress.
LASSO is a regularized
regression technique which forces the less relevant variables to be associated to regression coefficients of value zero~\cite{desboulets2018review}. Moreover, LASSO is especially
useful in problems where the number of variables is much larger than
the number of samples. This is surely the case when the variables are
overlapping gene modules obtained with HLC and the samples are genotypes, which is usually a small set due to the high cost of the RNA sequencing process. 
Finally, the proposed workflow is also modular, since other module detection
and selection techniques could be used, instead HLC and LASSO,
respectively.
\vspace{0.5cm}

The approach is showcased with a systematic study on rice
(\textit{Oryza sativa}), a food source that is known to be
highly sensitive to salt stress~\cite{chang2019morphological}. RNA-seq
data was accessed from the GEO database~\cite{clough2016gene}
(accession number GSE98455). It corresponds to $57845$ gene expression
profiles of shoot tissues measured for both control and stress condition
in $92$ accessions of the Rice Diversity Panel 1~\cite{eizenga2014registration}. A total of 6 modules
are detected as relevant in the response to salt stress in rice: 3
modules of 3 genes each, all associated with shoot K content, 2
modules of 3 genes associated with shoot biomass, and 1 module of 4
genes associated with root biomass. These genes are potential
targets for experimental validation of salinity tolerance in rice
cultivars. From the 19 genes, all but 3 genes (associated with $K$
content) are also identified as deferentially expressed for at least
one of the 92 accessions, suggesting that the identified genes are
stress responsive genes. Only 2 of the 16 differentially
expressed genes, both from the module associated to shoot biomass, are
named and have a known associated protein product: spermidine
hydroxycinnamoyltransferase 2 (SHT2) and lipoxygenase. 
Further studies are needed to elucidate the detailed biological
functions of the remaining 14 genes and their role
in stress responsive mechanisms
to salt conditions in rice. The goal is that the results reported in
this case study may allow biologist to develop new rice cultivars with
higher resistance to salinity.

\paragraph{Paper Outline.}
%The remainder of the paper is organized as
%follows. Section~\ref{sec.prelim} gathers preliminaries on gene
%co-expression networks, HLC, and LASSO. The proposed workflow is
%presented in Section~\ref{sec.framework}, with especial focus on the
%logical steps of the process and the internal structures supporting
%the approach. Section~\ref{sec.case} presents the case study on the
%identification of rice genes that are sensitive to salt
%stress. Section~\ref{sec.concl} draws some conclusions and future research directions.

The remainder of the paper is organized as
follows. \hyperref[sec.prelim]{Preliminaries} gathers foundations on gene
co-expression networks, HLC, and LASSO. The proposed workflow is
presented in \hyperref[sec.framework]{The Workflow} section, with especial focus on the
logical steps of the process and the internal structures supporting
the approach. \hyperref[sec.case]{Case Study} presents the case study on the
identification of rice genes that are sensitive to salt
stress. \hyperref[sec.concl]{Concluding Remarks} draws some conclusions and future research directions.

\section*{Preliminaries}
\label{sec.prelim}

This section presents preliminaries on networks, hierarchical link
clustering, and the LASSO linear regression technique.

\subsection*{Co-expression network}

A network is an undirected graph $G=(V,E)$ where
${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of $n$ \textit{vertices} or
\textit{nodes} and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of $q$
\textit{edges} or \textit{links} that connect vertices. In a gene
co-expression network, each node corresponds to a gene. A pair of
genes is connected if they show similar expression patterns. A simple
and unweighted network can be represented by an adjacency matrix $A
\in \{0,1\}^{n \times n}$ that is symmetric with a positive one in the
positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever there is an edge
connecting vertices $v_i$ and $v_j$, and zeros
elsewhere. Co-expression networks are of biological interest because
adjacent nodes in the network represent co-expressed genes that are
usually controlled by the same transcriptional regulatory pathway,
functionally related, or members of the same pathway or metabolic
complex.

\subsection*{Hierarchical Link Clustering}

The Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn
et al.~\cite{ahn2010link}. It represents communities as
groups of links (rather than nodes), and each node inherits all
memberships of its links and can thus belong to multiple, overlapping
communities. The algorithm maps links to nodes and connects them if a pair of
links shares a node. For a connected pair of links $e_{ik}$ and $e_{jk}$,
the shared node $k$ is called a \textit{keystone} node. 
The similarity between two links $e_{ik}$ and
$e_{jk}$ is computed using the Jaccard index
%
\begin{equation}\label{eq:jaccard}
  S(e_{ik},e_{jk}) = \frac{\vert n(i) \cap n(j) \vert}{\vert n(i) \cup n(j) \vert},
\end{equation}
%
where $n(i)$ denotes the set containing exactly node $i$ and its
neighbors. The algorithm uses single-linkage hierarchical clustering
to build a dendrogram in which each leaf is a link from the original
network, and branches represent link communities. Hierarchical
clustering algorithms repeatedly merge groups until all elements are
members of a single cluster.
\vspace{0.5cm}

For the purpose of finding meaningful communities, it is crucial to
know where to partition a dendrogram. In this case, the most
relevant communities are established at the maximal partition density
$D$, a function based on link density inside communities measuring the
quality of a link partition. The partition density $D$ has a single
global maximum along the dendrogram in almost all cases, because its
value is the average density at the top of the dendrogram (a single
giant community with every link and node) and it is very small at the
bottom of the dendrogram (most communities consists of a single
link). In particular, it is the case that $D = 1$ when every community
is a fully connected clique and $D = 0$ when each community is a
tree. If a community is less dense than a tree (i.e. when the
community subgraph has disconnected components), then such a community
contributes negatively to $D$, which can take negative values. The
minimum density inside a community is $-2/3$, given by one community
of two disconnected edges. Since $D$ is the average of the
intra-community density, there is a lower bound of $-2/3$ for
$D$. Computing $D$ at each level of the link dendrogram can help the
purpose of picking the best level to cut, although meaningful
structure could exist above or below the threshold. The output of
cutting is a set of node clusters, where each node can participate in
multiple communities.

\subsection*{Least Absolute Shrinkage Selector Operator}

The Least Absolute Shrinkage Selector Operator (LASSO) is a
regularized linear regression technique. It combines a regression
model with a procedure of contraction of some parameters towards zero
and selection of variables, imposing a restriction or a penalty on the
regression coefficients. In other words, LASSO solves the least
squares problem with restriction on the $ L_1$-norm of the coefficient
vector. It can be especially useful to solve problems where the number
of variables (e.g., genes) $ n $ is much greater than the number of
samples $ m $ (i.e., $ n \gg m $).
\vspace{0.5cm}

Consider a dataset consisting of $m$ samples, each of which consists
of $n$ covariates and a single outcome. Let $y_i$ be the outcome and
$x_i := (x_{i1},...,x_{in})$ be the covariate vector for the $i$-th
sample. The objective of LASSO is to solve
%
\begin{equation}
\min \left\lbrace\sum_{i=1}^{m}{\left( y_i-\sum_{j=1}^n{\beta_j
    x_{ij}}\right)^2} \right\rbrace \quad , \quad \textrm{subject to}
\quad \sum_{j=1}^n\abs{\beta_j}\leq s.
\end{equation}

Equivalently, in the Lagrangian form, it minimizes

\begin{equation}
  \sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} +
  \lambda \sum_{j=1}^n\abs{\beta_j}
\end{equation}
%
where $s$ is the regularization penalty and $\lambda \geq 0$ is the
corresponding Lagrange multiplier. Since the $\lambda$ value
determines the degree of penalty, the accuracy of the model depends on
its choice. Cross-validation is often used to select the
regularization parameter, finally selecting the one that minimizes the
mean-squared error.

\section*{The Workflow}
\label{sec.framework}

The proposed workflow is depicted in
Figure~\ref{fig:flow_chart}. This section explains the five
macro-processes (A)-(E) of the proposed workflow. In comparison with
WGCNA, it adds the macro-step (D) and generalizes macro-steps (A)-(C).
\vspace{0.5cm}

% figure 1
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.96\textwidth]{figures/figure1.pdf}
  \caption[The proposed workflow comprising five macro-steps]%
  {The proposed workflow comprising five macro-steps:
    A.~Data pre-processing, B.~Co-expression network
    contruction, C.~Co-expression module identification,
    D.~Modules association to phenotypic traits, and
    E.~Gene enrichment.}
  \label{fig:flow_chart}
\end{figure}

The workflow uses RNA-seq read counts, representing gene expression
levels, as input data. More precisely, it uses $n_0$ gene expression
profiles of an organism, measured for $m$ different genotypes under
control and treatment conditions, and $r$ biological replicates. This
raw data is represented as a matrix $D_0 \in {\mathbb{N}_0}^{n_0
  \times 2mr}$. In order to discover key genes and their interaction
with phenotypes related to treatment tolerance, the approach also
requires a set of $p$ phenotypic traits, measured for the $m$
genotypes. The phenotypic data is seen as a matrix $P \in
\mathbb{R}^{2m \times p}$ containing two phenotypic values per
genotype, one under control condition and a second one under treatment
condition.

\subsection*{Data Pre-processing}

The goal of the data pre-processing stage is to build matrices
$P_\ell$ and $L_1$ representing, respectively, the changes in
phenotypic values and expression levels between control and treatment
condition, from RNA-seq and phenotypic data found in matrices $D_0$
and $P$, respectively.
\vspace{0.5cm}

The RNA-seq data cannot be directly interpreted. Therefore, a
normalization process is applied to deal with the problem of possible
biases affecting the quantification of results. The suggested
normalization technique for correcting library size and RNA
composition bias is DESeq2~\cite{love2014moderated}. The normalized
data is represented as a matrix $D_1 \in \mathbb{R}^{n_0 \times 2mr}$,
and the biological replicates of each genotype are averaged and
represented as a matrix $D_2 \in \mathbb{R}^{n_0 \times 2m}$. The
genes exhibiting low variance or low expression are removed from
$D_2$, thus identifying a subset of size $n_1 \leq n_0$ of the
original genes. The control and treatment data is separated into the
matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1
  \times m}$, respectively. The matrix entries $c_{ij}$ in $C$ and
$t_{ij}$ in $T$ represent the normalized expression
level of gene $i$ in accession $j$ under control and treatment 
condition, respectively.
Control and treatment data is also separated from
phenotypic data $P$, obtaining the $P_c$ and $P_t$ matrices of
dimensions $m \times p$.
\vspace{0.5cm}

In the above configuration, the changes in expression levels and
phenotypic values between control and treatment conditions, are
measured in terms of logarithmic ratios. In the case of expression
levels, the log ratios are represented in the Log Fold Change matrix
$L_0 \in \mathbb{R}^{n_1 \times m}$, where $\ell_{ij}=\log_2
(t_{ij}/c_{ij})$. Similarly, the log ratios of the phenotypic data are
computed and represented in the $P_\ell \in \mathbb{R}^{m \times p}$
matrix.
\vspace{0.5cm}

The final step of the data pre-processing is to filter $L_0$ by
removing rows (e.g., genes) with low variance in the differential
expression patterns, obtaining a new matrix $L_1$ of dimensions $n_2
\times m$, with $n_2 \leq n_1$.

\subsection*{Co-expression Network Construction}

A gene co-expression network connect genes with similar expression
patterns across biological conditions. The purpose of this stage is to
describe how to build the co-expression network $A$ from the Log Fold
Change matrix $L_1$, capturing the relationship between genes
according to the change in expression levels between the two studied
conditions. These co-expression patterns are meaningful for the
identification of genes not yet associated with the response to the
treatment condition.
\vspace{0.5cm}

The Log Fold Change matrix $L_1$ is used to build the co-expression
network following the first two steps of the WGCNA
methodology~\cite{langfelder2008wgcna}. First, the level of
concordance between gene differential expression profiles across
samples is measured. To this end, the absolute value of the Pearson
correlation coefficient is used as the similarity measure between
genes and the resulting values are stored in the similarity matrix
$S\in \mathbb{R_{+}}^{n_2 \times n_2}$. Second, the matrix $S$ is
transformed into an adjacency matrix $A \in \mathbb{R_+}^{n_2\times
  n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the
connection strength between each pair of genes. In other words, the
elements of the adjacency matrix are the similarity values up to the
power $\beta > 1$ so that the degree distribution will fit a
scale-free network. These networks contain many nodes with very few
connections and a small number of hubs with high connections. In a
strict scale-free network, the logarithm of $P(k)$ (i.e., the
probability of a node having degree $k$) is approximately inversely
proportional to the logarithm of $k$ (i.e., the degree of a node).
The parameter $\beta$ is chosen to be the smallest power such that the
$R^2$ of the linear regression between $log_{10}(p(k))$ and
$log_{10}(k)$ is close to $1$ (e.g., $R^2 > 0.85$).

\subsection*{Co-expression Module Identification}

The next step in the workflow is to identifying communities (also
called modules) from the co-expression network structure and dynamics
represented in $A$.  The idea is to cluster genes with similar
patterns of differential expression change. Membership in these
modules may overlap in biological contexts, because modules may be
related to specific molecular, cellular, or tissue functions, and the
biological components (i.e., genes) may be involved in multiple
functions. Thus, unlike WGCNA, the adjacency matrix $A$ is used to
detect overlapping (rather than non-overlapping) communities, using
the Hierarchical Link Clustering (HLC) algorithm (see
%Section~\ref{sec.prelim}).
\hyperref[sec.prelim]{Preliminaries}).
\vspace{0.5cm}

As a preliminary step, matrix $A$ is transformed into an unweighted
network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$ before using the
clustering algorithm. To this end, the Pearson Correlation Coefficient
(PCC) cutoff is determined using the approach described
in~\cite{aoki2007approaches}. The number of nodes, edges, and the
network density is determined for different PCC cutoffs. Near the most
biological relevant PCC cutoff, the number of nodes presents a linear
decrease and the density of the network reaches its minimum, while
below this value the number of edges rapidly increases. Following this
observation, a cutoff is selected such that gene pairs which have a
correlation score higher than the threshold are considered to have
important co-expression relationship. Above the cutoff, the entries of
matrix $A$ become $1$ and below the cutoff $A$ values become $0$. The
HLC algorithm organizes the $n_2$ genes of matrix $\hat{A}$ into $c$
modules, where each gene can belong to zero, one or multiple modules.
This information is represented as an affiliation matrix $F \in
\{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ iff node $i$ is member of
module $u$ (it is 0, otherwise).

\subsection*{Module Association to Phenotypic Traits}

To identify the most relevant modules of genes, associated
with the phenotypic response to a specific treatment in an organism,
the proposed workflow uses LASSO. Specifically, each module is
represented by an eigengene, which is defined as the first principal
component of such module. An eigengene can be seen as an average
differential expression profile for each community: it is computed
from the Log Fold Change Matrix $L_1$ and the affiliation matrix
$F$. Given a module $u$, the affiliation matrix is used to identify
the genes belonging to $u$ and then the corresponding rows of the
matrix $L_1$ are selected to compute the first principal component of
$u$. Each principal component becomes a column of the matrix $M \in
\mathbb{R}^{m \times c}$.  These profiles are then associated with each
phenotypic trait using
LASSO.  In this context, the eigengenes (i.e., the columns
of $M$) act as regressor variables and each phenotypic trait (i.e.,
each column of $P_\ell$) is used as an outcome variable.
\vspace{0.5cm}

The output after applying LASSO is a set $W_z$ of modules for each
phenotypic trait $z$, where $W_z \subset \{u \mid 1 \leq u \leq c\}$
for $z= 1,2,..,p$. The target genes $I$ for downstream analysis, which
may be important in the treatment response, are the union of genes
belonging to the selected modules; that is $I = \cup_{z=1}^{p} W_z$,
where $I \subset \{i \mid 1 \leq i \leq n_2\}$.

\subsection*{Gene Enrichment}

The goal of this final stage of the process is to annotate with
additional information the genes identified in previous stages,
helping to elucidate their possible behavior and role in the response
to the studied treatment.
\vspace{0.5cm}

A crucial step is to identify the differentially expressed genes in
set $I$. That is, to select those genes in $I$ having an absolute
value of the Log Fold Change of at least $2$ ($|\ell_{ij}|\geq 2$) for
at least one sample. This represents genes whose expression level is
quadrupled (up or down) from control to treatment condition; they are the
strong candidates for treatment responsive genes.
\vspace{0.5cm}

Also, functional category enrichment can be done by, e.g., searching
for gene ontology annotations in databases such as
QuickGO~\cite{binns2009quickgo}. Such annotations can provide evidence
of biological implications of the target genes in the
treatment-tolerance mechanisms. Furthermore, QuickG0 can be used to
identify genes with reported protein products, which can be used to
perform additional relevant analysis reviewing their reported
protein-protein interactions in other databases, such as
STRING~\cite{szklarczyk2016string}. The interactions include direct
(physical) and indirect (functional) associations, and they stem from
computational prediction, knowledge transfer between organisms, and
interactions aggregated from other (primary) databases. This
information can give new insights on how the selected genes are
involved in functional pathways that can be related to the treatment
of interest.

%\section*{Results}
\section*{Identifying Potential Saline Stress Responsive Genes in Rice}
\label{sec.case}

This section presents a case study on the identification of genes in
\textit{Oryza sativa} that may respond to saline stress following the
approach presented in %Section~\ref{sec.framework}.
\hyperref[sec.framework]{The Workflow}.
\vspace{0.5cm}

The RNA-seq data used for the experiments summarized in this section
was obtained from GEO database \cite{clough2016gene} (accession
number GSE98455). This data corresponds to $n_0=57\,845$ gene
expression profiles of shoot tissues measured for both control and
salt condition in $m=92$ accessions of the Rice Diversity Panel 1~\cite{eizenga2014registration},
with $r=2$ biological replicates. A total of $p=3 $ phenotypic traits
are used: shoot $K^+$ content, root biomass, and shoot biomass. These
traits were measured for the same $92$ genotypes, under control and
salt stress conditions, and can be found in the supplementary
information of~\cite{campbell2017allelic}.

\subsection*{Data Pre-processing}

DESeq2 normalization is applied to the raw data and the biological
replicates are averaged. Genes exhibiting low variance are
identified as those with ratio of upper quantile to lower quantile
smaller than $1.5$ and are removed from the normalized data. Genes
with low expression, corresponding to those having more than $80\%$
samples with values smaller than $10$, are also removed. A total of
$n_1 = 9\,414$ genes are kept after this filtering process.
\vspace{0.5cm}

From the Log Fold Change matrix $L_0$, genes whose difference between
upper quantile and lower quantile greater than $0.25$ are
removed. Therefore, the resulting matrix $L_1$ contains the log ratios
of $n_2 = 8\,928$ genes. The logarithmic ratios of the phenotypic data,
for the $92$ accessions and the $3$ traits, are also computed.

\subsection*{Co-expression Network Construction}

The Log Fold Change matrix $L_1$ is used to compute the corresponding
similarity matrix.  For this network, it is observed that $\beta=3$ is
the smallest integer such that the $R^2 \geq 0.8
$. Figure~\ref{fig:beta} depicts the degree distribution of the
similarity matrix (left) and the degree distribution of the adjacency
matrix (right), which is the degree distribution of a scale-free
network with $R^2 = 0.8$ with $\beta = 3$.
\vspace{0.5cm}

%figure 2
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.8\textwidth]{figures/figure2.png}
  \caption{Degree distribution of $A$ (left) and $\hat{A}$ (right).}
  \label{fig:beta}
\end{figure}

The resulting adjacency matrix $A$ represents a complete graph
$G=(V,E)$, with $|V| = 8\,928$ genes ($|E| = 39\,850\,128$ edges).

\subsection*{Co-expression Module Identification}

The adjacency matrix $A$ is transformed into an unweighted network
$\hat{A}$ applying the approach described
in~\cite{aoki2007approaches}. The cutoff value is set to $0.2$, 
based on the density of the network
combined with the decreasing number of nodes and edges with higher PCC
values. Thus keeping only the
connections above this threshold and removing the isolated nodes. The
resulting adjacency matrix $\hat{A}$ has $5\,810$ connected genes and
accounts for $16\,875\,145$ edges.
\vspace{0.5cm}

After applying the HLC algorithm, a total of $4\,131$ genes are
distributed in $c = 5\,143$ overlapping modules of at least $3$
genes. Figure~\ref{fig:overlap} presents a histogram of the
overlapping percentage of these genes, measured as the proportion of
modules to which each gene belongs. The first bar of the histogram
represents the genes with zero overlap, corresponding to $28\%$ of the
total genes; the remaining $72\%$ represents the genes belonging to
more than one module.

%figure 3
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.5\textwidth]{figures/figure3.png}
  \caption{Overlapping percentage of genes after applying HLC.}
  \label{fig:overlap}
\end{figure}

\subsection*{Module Association to Phenotypic Traits}

The phenotypic traits under study are shoot $K^+$ content, root
biomass, and shoot biomass. Figure~\ref{fig:pdata} suggests that there
are significant differences in the values of these phenotypic traits
between stress and control conditions. This supports the working
hypothesis that these three variables represent tolerance-associated
traits in rice under salt stress.
\vspace{0.5cm}

%figure 4
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.9\textwidth]{figures/figure4.png}
  \caption{Phenotipic traits distribution under control and salt stress.}
  \label{fig:pdata}
\end{figure}

Using the affiliation matrix $F$ derived from the HLC output and the
Log Fold Change matrix $L_1$, a matrix $M$ is built by computing the
eigengene for each of the $c = 5\,143$ modules. The LASSO technique is
applied by using each of the phenotypic traits as the outcome
variable, one at a time. As shown in Figure~\ref{fig:cross-val},
cross-validation is performed for each phenotypical trait in order to
select the corresponding regularization parameter $\lambda$ that
minimizes the mean-squared error.
\vspace{0.5cm}

%figure 5
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.96\textwidth]{figures/figure5.pdf}
  \caption{Cross-validation of the LASSO regularization parameter
    $\lambda$, for each phenotypic trait.}
  \label{fig:cross-val}
\end{figure}

Finally, three LASSO models are adjusted by using the corresponding
$\lambda$ and phenotypical data with the eigengenes of matrix $M$. As
result, 6 modules are detected as relevant in the response to salt
stress in rice: 3 modules of 3 genes, each associated with shoot $K$
content; 2 modules of 3 genes associated with shoot biomass; and 1
module of 4 genes associated with root biomass (see
Figure~\ref{fig:final_genes}).

\subsection*{Gene Enrichment}

From the $19$ genes selected by LASSO, all but $3$ genes (the ones
associated to $K$ content), are also identified as deferentially
expressed ($|\ell_{ij}| \geq 2$) for at least one of the $92$
accessions. This suggests that those genes are strong candidates as
stress responsive genes to salt conditions in rice.
\vspace{0.5cm}

Figure~\ref{fig:final_genes} summarizes how from the initial
$n_0=57\,845$ genes, the proposed workflow identifies a reduced set of
$19$ genes. First, $48\,431$ genes are discarded after filtering the
normalized expression data $D_2$ and then $486$ additional genes are
discarded when filtering the Log Fold Change matrix $L_0$, to finally
arrive at $19$ genes, of which $16$ are differentially expressed.
\vspace{0.5cm}

% figure 6
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.8\textwidth]{figures/figure6.pdf}
   \caption[Venn diagram for the case study in rice]%
   {Venn diagram representing the number of genes selected at
    different stages of the proposed workflow for the case study in
    rice.}
  \label{fig:final_genes}
\end{figure}

According to the Quickgo database, only $2$ of the $16$ differentially
expressed genes (both from the module related to shoot biomass) are
named and have an associated protein product: spermidine
hydroxycinnamoyltransferase 2 (SHT2) and lipoxygenase.
Figure~\ref{fig:3d} shows their corresponding 3D protein structures.
\vspace{0.5cm}

%figure 7
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.8\textwidth]{figures/figure7.pdf}
  \caption{3D protein structure of named genes selected by LASSO, borrowed from~\cite{szklarczyk2016string}.}
  \label{fig:3d}
\end{figure}

The Uniprot database~\cite{uniprot2018uniprot} reports, on the one
hand, that SHT2 contributes to the natural variation of
spermidine-based phenolamides in rice cultivars. On the other hand, it
is reported in~\cite{uniprot2018uniprot} that plant lipoxygenase may
be involved in a number of diverse aspects of plant physiology
including growth and development, pest resistance, and senescence or
responses to wounding. This protein is involved in the pathway
oxylipin biosynthesis, which is part of Lipid metabolism. Additionally,
previous studies
%
in~\cite{gupta2013plant,hou2015persimmon,mittova2002salt,peng2019novel,roychoudhury2011amelioration}
%
provide evidence of biological implications of sperimidine and
lipoxygenase in tolerance to salt stress in other plants or even in
rice cultivars.

As a conclusion, the results presented in this section suggest that
further studies are needed to elucidate the detailed biological
function of the remaining 14 genes that have not been named so far in
the literature.  They may have the potential to intervene in stress
responsive mechanisms to salt conditions in rice.

\section*{Concluding Remarks}
\label{sec.concl}

This manuscript provides a detailed description of a network-based
analysis workflow for the discovery of key genes responding to a
specific treatment in an organism. It links transcriptomic with
phenotypic data and identifies overlapping gene modules.
\vspace{0.5cm}

The proposed approach is inspired by the workflow suggested in the
WGCNA~\cite{langfelder2008wgcna}. Its main steps are the preprocessing
of the gene expression data, the construction of a co-expression
network, the detection of modules within the network, the relation of
modules with external information (e.g., phenotypic data), and the
enrichment of the identified key genes with additional information.
Both approaches are structured in a modular way, which allows
modifying and exploring different techniques in each step of the
workflow.
\vspace{0.5cm}

The proposed workflow is designed to integrate expression data
measured under two different conditions (namely, control and
treatment), unlike the usually co-expression-based approaches which
work with both conditions independently or consider only a single
condition. For this purpose, an approach similar to that proposed
in~\cite{du2019network} is used, where the control and treatment data
are compiled in a single matrix using the Log Fold Change
measure. Thus, the input to construct the co-expression network is not
the expression data, but instead the changes in the expression levels
from one condition to the other, making room for capturing the signal
of changes caused by the treatment.
\vspace{0.5cm}

An important feature in the proposed workflow is the module
detection technique. The co-expression network is computed, as in
WGCNA, until a scale-free network is obtained. In the proposed
approach, this network is then used to apply the HLC algorithm, a
clustering technique capable of detecting overlapping
communities. Several approaches of module detection from gene
expression have been proposed and were evaluated
in~\cite{saelens2018comprehensive}. Most of them focus mainly on
disjoint (non-overlapping) communities; the techniques described
dealing with overlaps are not clustering, but bi-clustering and
decomposition methods. It is well known that communities in real
networks, including biological ones, are likely
overlap~\cite{palla2005uncovering}. Thus, the approach presented
in this work can be seen as a generalization of the previous
approaches, such as WGCNA, with the potential to deal with genes
associated to multiple biological processes.
\vspace{0.5cm}

The approach was applied in a case study with rice under salt
stress. The results show a group of 14 genes, of which only $2$ of
them have been previously related to saline stress response in other
studies. As future work, other overlapping module detection and
selection techniques should be used instead HLC and LASSO,
respectively. The combination of these techniques would allow finding
target genes for future biological studies that evaluate their
potential as genes that respond to salt stress in rice, and other
crops and stresses. In-vivo laboratory experimentation needs to be
conducted to validate the findings of this paper in relation to
salinity stress.
\vspace{0.5cm}

Finally, the workflow is presented as a protocol capable of
considerably reducing the number of genes detected as relevant in the
response to stress of choice. Other traditionally used methods for
this purpose tend to generate a large list of candidate genes, thus
limiting subsequent efforts in experimental validation. In this sense,
the proposed workflow can help in reducing such efforts in time and
money invested by researchers in the experimental validation of
stress-responsive genes.

%\section*{Appendix}
%Text for this section\ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Acknowledgements}%% if any
Not applicable

\section*{Funding}%% if any
This work was funded by the OMICAS program: Optimización Multiescala In-silico de
Cultivos Agrícolas Sostenibles (Infraestructura y Validación en Arroz y Caña de Azúcar),
anchored at the Pontificia Universidad Javeriana in Cali and funded within the Colombian
Scientific Ecosystem by The World Bank, the Colombian Ministry of Science, Technology and
Innovation, the Colombian Ministry of Education and the Colombian Ministry of Industry 
and Turism, and ICETEX, under GRANT ID: FP44842-217-2018.

\section*{Abbreviations}%% if any
\textbf{HLC:} Hierarchical Link Clustering\\
\textbf{LASSO:} Least Absolute Shrinkage Selector Operator\\
\textbf{PCC:} Pearson Correlation Coefficient\\
\textbf{RNA-seq:} RNA sequencing\\
\textbf{SHT2:} Spermidine hydroxycinnamoyltransferase 2\\
\textbf{WGCNA:} Weighted Gene Co-expression Network Analysis\\

\section*{Availability of data and materials}%% if any
The datasets analysed during the current study are publicy available. They can be found in the following locations:
\begin{itemize}
\item RNA-seq data of salt stress in rice is available on the GEO (GSE98455).
\item Phenotypic data of salt stress in rice is a subset of the supplementary file 1 included in ~\cite{campbell2017allelic}.
\end{itemize}

\section*{Ethics approval and consent to participate}%% if any
Not applicable

\section*{Competing interests}
The authors declare that they have no competing interests.

\section*{Consent for publication}%% if any
Not applicable

\section*{Authors' contributions}
J.F. and C.R. proposed the original idea. 
J.F. provide advice on algorithms concepts and implementation.
C.R. structured the methodology and performed the analysis. 
C.R., J.F., and H.C.R. wrote the manuscript.
All authors read and approved the final manuscript.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{biblio}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

%\section*{Figures}
%
%%flow chart
%\begin{figure}[h!]
%  \centering
%    \includegraphics[clip,width=1\textwidth]{figures/figure1.pdf}
%  \caption[The proposed workflow comprising five macro-steps]%
%  {The proposed workflow comprising five macro-steps:
%    A.~Data pre-processing, B.~Co-expression network
%    contruction, C.~Co-expression module identification,
%    D.~Modules association to phenotypic traits, and
%    E.~Genes enrichment.}
%  \label{fig:flow_chart}
%\end{figure}
%
%%scale-free beta 1 y 3
%\begin{figure}[h!]
%  \centering
%    \includegraphics[clip,width=0.8\textwidth]{figures/figure2.png}
%  \caption{Degree distribution of $A$ (left) and $\hat{A}$ (right)}
%  \label{fig:beta}
%\end{figure}
%
%%overlapping percentage
%\begin{figure}[h!]
%  \centering
%    \includegraphics[clip,width=0.5\textwidth]{figures/figure3.png}
%  \caption{Overlapping percentage}
%  \label{fig:overlap}
%\end{figure}
%
%% boxplots
%\begin{figure}[h!]
%  \centering
%    \includegraphics[clip,width=0.9\textwidth]{figures/figure4.png}
%  \caption{Phenotipic traits distribution under control and salt stress.}
%  \label{fig:pdata}
%\end{figure}
%
%\begin{figure}[h!]
%  \centering
%    \includegraphics[clip,width=0.9\textwidth]{figures/figure5.pdf}
%  \caption{Cross-validation of the LASSO regularization parameter
%    $\lambda$, for each phenotypic trait.}
%  \label{fig:cross-val}
%\end{figure}
%
%\begin{figure}[h!]
%  \centering
%    \includegraphics[clip,width=0.8\textwidth]{figures/figure6.pdf}
%   \caption[Venn diagram for the case study in rice]%
%   {Venn diagram representing the number of genes selected at
%    different stages of the proposed workflow for the case study in
%    rice.}
%  \label{fig:final_genes}
%\end{figure}
%
%% Figure 3d protein strucures
%\begin{figure}[hptb]
%  \centering
%    \includegraphics[clip,width=0.8\textwidth]{figures/figure7.pdf}
%  \caption{3D protein structure of named genes selected by LASSO, available from~\cite{szklarczyk2016string}.}
%  \label{fig:3d}
%\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
%\section*{Tables}
%\begin{table}[h!]
%\caption{Sample table title. This is where the description of the table should go}
%  \begin{tabular}{cccc}
%    \hline
%    & B1  &B2   & B3\\ \hline
%    A1 & 0.1 & 0.2 & 0.3\\
%    A2 & ... & ..  & .\\
%    A3 & ..  & .   & .\\ \hline
%  \end{tabular}
%\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\section*{Additional Files}
%  \subsection*{Additional file 1 --- Sample additional file title}
%    Additional file descriptions text (including details of how to
%    view the file, if it is in a non-standard format or the file extension).  This might
%    refer to a multi-page table or a figure.
%
%  \subsection*{Additional file 2 --- Sample additional file title}
%    Additional file descriptions text.

\end{backmatter}
\end{document}

### 2020-12-07 camila/report/case.tex
bmc version

@@
-\section{Case Study}
+%\section*{Results}
+\section*{Identifying Potential Saline Stress Responsive Genes in Rice}
 \label{sec.case}
-This section presents a case study on the identification of genes that may respond to saline stress according to the methodology presented in section~\ref{sec. framework}. 
-The RNA-seq data was obtained from GEO database
-\cite{GEOAcces90:online} (accession number GSE98455). This data
-corresponds to $n_0=57845$ gene expression profiles of shoot tissues
-measured for both control and salt condition in $m=92$ accessions of
-the Rice Diversity Panel 1, with $r=2$ biological replicates. A total
-of $p=3 $ phenotypic traits are used: shoot $K^+$ content, root
-biomass and shoot biomass. These traits were measured for the same
-$92$ genotypes, under control and salt stress conditions, and can be
-found in the supplementary information
-of~\cite{campbell2017allelic}.
 
-\subsection{Data pre-processing}
+This section presents a case study, applying the approach introduced in 
+\hyperref[sec.framework]{the Workflow} section, to identify genes in
+\textit{Oryza sativa} that respond to saline stress.
+\vspace{0.5cm}
+
+The RNA-seq data are available from the
+GEO database \cite{clough2016gene} (accession
+number GSE98455). It corresponds to $n_0=57\,845$ gene
+expression profiles of shoot tissues measured for control and
+salt conditions in $m=92$ accessions of the Rice Diversity Panel 1~\cite{eizenga2014registration},
+with $r=2$ biological replicates. A total of $p=3 $ phenotypic traits
+are used: shoot $K$ content, root biomass, and shoot biomass. These
+traits were measured for the same $92$ genotypes, under control and
+treatment conditions, and can be found in the supplementary
+information for~\cite{campbell2017allelic}.
+
+\subsection*{A. Data Pre-processing}
 
@@
 with low expression, corresponding to those having more than $80\%$
-samples with values smaller than $10$, are also removed. A total of
-$n_1 = 8928$ genes are kept after this filtering process.
+samples with values smaller than $10$, are also removed. After this
+filtering process a total of $n_1 = 9\,414$ genes are kept for further analysis.
+\vspace{0.5cm}
 
 From the Log Fold Change matrix $L_0$, genes whose difference between
-upper quantile and lower quantile greater than $0.25$ are
+upper quantile and lower quantile is greater than $0.25$ are
 removed. Therefore, the resulting matrix $L_1$ contains the log ratios
-of $n_2 = 8928$ genes. The logarithmic ratios of the phenotipic data,
-for the $92$ accessions and the $3$ traits, is also computed.
+of $n_2 = 8\,928$ genes. The logarithmic ratios of the phenotypic data,
+for the $92$ accessions and the $3$ traits, are also computed.
 
-\subsection{Co-expression network construction}
+\subsection*{B. Construction of the Co-expression Network}
 
 The Log Fold Change matrix $L_1$ is used to compute the corresponding
-similarity matrix.  For this network, it is observed that $\beta=3$
-is the smallest integer such that the $R^2 \geq 0.8
+similarity matrix.  For this network, it is observed that $\beta=3$ is
+the smallest integer such that the $R^2 \geq 0.8
 $. Figure~\ref{fig:beta} depicts the degree distribution of the
@@
 network with $R^2 = 0.8$ with $\beta = 3$.
+\vspace{0.5cm}
 
-%scale-free beta 1 y 3
-\begin{figure}[htpb]
+%figure 2
+\begin{figure}[htbp]
   \centering
-    \includegraphics[clip,width=0.8\textwidth]{figures/pick_beta.png}
-  \caption{Degree distribution of $A$ (left) and $\hat{A}$ (right)}
+    \includegraphics[clip,width=0.8\textwidth]{figures/figure2.png}
+  \caption{Degree distribution of the network represented by $S$ (left) and $A$ (right).}
   \label{fig:beta}
@@
 The resulting adjacency matrix $A$ represents a complete graph
-$G=(V,E)$, with $|V| = 8928$ genes and $|E| = 39850128$ edges.
+$G=(V,E)$, with $|V| = 8\,928$ genes ($|E| = 39\,850\,128$ edges).
 
-\subsection{Co-expression module identification}
+\subsection*{C. Identification of Co-expression Modules}
 
 The adjacency matrix $A$ is transformed into an unweighted network
-$\hat{A}$ applying the approach described in~\cite{aoki2007approaches}, 
-based on the density of the network combined with the decreasing number
-of nodes and edges with higher PCC values. The cutoff value is set to $0.2$,
-thus keeping only the connections above this threshold and removing the 
-isolated nodes. The resulting adjacency matrix $\hat{A}$ has $5810$ connected 
-genes and accounts for $16875145$ edges.
+$\hat{A}$ applying the approach described
+in~\cite{aoki2007approaches}. The cutoff value is set to $0.2$, 
+based on the density of the network
+combined with the decreasing number of nodes and edges with higher PCC
+values. Thus keeping only the
+connections above this threshold and removing any isolated nodes. The
+resulting adjacency matrix $\hat{A}$ has $5\,810$ connected genes and
+accounts for $614\,501$ edges.
+\vspace{0.5cm}
 
-After applying the HLC algorithm, a total of $4131$ genes are
-distributed in $c = 5143$ overlapping modules of at least $3$ genes. Figure~\ref{fig:overlap} presents a histogram of the overlapping
-percentage of these genes, measured as the proportion of modules to
-which each gene belongs. The first bar of the histogram represents the
-genes with zero overlap, corresponding to $28\%$ of the total
-genes; remaining $72\%$ represents the genes belonging to more than one
-module.
+After applying the HLC algorithm, a total of $4\,131$ genes are
+distributed in $c = 5\,143$ overlapping modules of at least $3$
+genes. Figure~\ref{fig:overlap} presents a histogram of the
+overlapping percentage of these genes, measured as the proportion of
+modules to which each gene belongs. The first bar of the histogram
+represents the genes with zero overlap, corresponding to $28\%$ of the
+total genes; the remaining $72\%$ represents the genes belonging to
+more than one module.
 
-%overlapping percentage
-\begin{figure}[htpb]
+%figure 3
+\begin{figure}[htbp]
   \centering
-    \includegraphics[clip,width=0.5\textwidth]{figures/artificial_modules.png}
-  \caption{Overlapping percentage}
+    \includegraphics[clip,width=0.95\textwidth]{figures/figure3.pdf}
+  \caption{Overlapping percentage of genes after applying HLC.}
   \label{fig:overlap}
@@
 
-\subsection{Module association to phenotypic traits}
+\subsection*{D. Detection of Modules Association to Phenotypic Traits}
 
-The phenotypic traits under study are shoot $K^+$ content, root
-biomass and shoot biomass. Figure \ref{fig:pdata} suggests that there
+The phenotypic traits under study are shoot $K$ content, root
+biomass, and shoot biomass. Figure~\ref{fig:pdata} suggests that there
 are significant differences in the values of these phenotypic traits
@@
 traits in rice under salt stress.
+\vspace{0.5cm}
 
-% boxplots
-\begin{figure}[htpb]
+%figure 4
+\begin{figure}[htbp]
   \centering
-    \includegraphics[clip,width=1\textwidth]{figures/phenotypic_traits.png}
+    \includegraphics[clip,width=0.9\textwidth]{figures/figure4.pdf}
   \caption{Phenotipic traits distribution under control and salt stress.}
@@
 Log Fold Change matrix $L_1$, a matrix $M$ is built by computing the
-egengene for each of the $c = 5143$ modules. The LASSO technique is
+eigengene for each of the $c = 5\,143$ modules. The LASSO technique is
 applied by using each of the phenotypic traits as the outcome
@@
 minimizes the mean-squared error.
+\vspace{0.5cm}
 
-\begin{figure}[htpb]
+%figure 5
+\begin{figure}[htbp]
   \centering
-  \subfloat[$K$ shoot, $\lambda = 0.013$]{%
-    \includegraphics[width=0.33\textwidth]{figures/lambda_kshoot_2.png}
-    \label{fig:lambda_kshoot}
-  }
-  \subfloat[Shoot biomass, $\lambda = 0.025$]{%
-    \includegraphics[width=0.33\textwidth]{figures/lambda_BMshoot_2.png}
-    \label{fig:lambda_BMshoot}
-  }
-  \subfloat[Root biomass, $\lambda = 0.035$]{%
-    \includegraphics[width=0.33\textwidth]{figures/lambda_BMroot.png}
-    \label{fig:lambda_BMroot}
-  }
+    \includegraphics[clip,width=0.96\textwidth]{figures/figure5.pdf}
   \caption{Cross-validation of the LASSO regularization parameter
@@
 Finally, three LASSO models are adjusted by using the corresponding
-$\lambda$ and phenotypical data with the egengens of matrix $M$. As
+$\lambda$ and phenotypical data with the eigengenes of matrix $M$. As
 result, 6 modules are detected as relevant in the response to salt
-stress in rice: 3 modules of 3 genes, each associated with shoot
-$K$ content; 2 modules of 3 genes associated with shoot biomass; and 1
-module of 4 genes associated with root biomass.
+stress in rice: 3 modules of 3 genes, each associated with shoot $K$
+content; 2 modules of 3 genes associated with shoot biomass; and 1
+module of 4 genes associated with root biomass (see
+Figure~\ref{fig:final_genes}).
 
-\subsection{Gene enrichment}
+\subsection*{E. Gene Enrichment}
 
-From the $19$ genes selected by LASSO, all but $3$ genes (the ones associated to $K$ content), are also identified as deferentially expressed ($|\ell_{ij}| \geq 2$) for at least one of the $92$ accessions.
-This suggest that those genes are strong candidates as stress responsive genes to salt conditions in rice.
-According to the Quickgo database, only $2$ of the $16$ diferentially expressed genes (both from the module related with shoot biomass) are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase.
+From the $19$ genes selected by LASSO, all but $3$ genes (the ones
+associated to $K$ content), are also identified as deferentially
+expressed ($|\ell_{ij}| \geq 2$) for at least one of the $92$
+accessions. These genes are strong 
+candidates target genes in rice.
+%candidates as stress responsive genes to salt conditions in rice.
+\vspace{0.5cm}
+
+Figure~\ref{fig:final_genes} summarizes how from the initial
+$n_0=57\,845$ genes, the proposed workflow identifies a reduced set of
+$19$ genes. First, $48\,431$ genes are discarded after filtering the
+normalized expression data $D_2$ and then $486$ additional genes are
+discarded when filtering the Log Fold Change matrix $L_0$, to finally
+arrive at $19$ genes, of which $16$ are differentially expressed.
+\vspace{0.5cm}
+
+% figure 6
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[clip,width=0.8\textwidth]{figures/figure6.pdf}
+   \caption[Venn diagram for the case study in rice]%
+   {Venn diagram representing the number of genes selected at
+    different stages of the proposed workflow for the case study in
+    rice.}
+  \label{fig:final_genes}
+\end{figure}
+
+According to the Quickgo database, only $2$ of the $16$ differentially
+expressed genes (both from the module related to shoot biomass) are
+named and have an associated protein product: spermidine
+hydroxycinnamoyltransferase 2 (SHT2) and lipoxygenase.
 Figure~\ref{fig:3d} shows their corresponding 3D protein structures.
+\vspace{0.5cm}
 
-% Figure 3d protein strucures
-\begin{figure}[hptb]
+%figure 7
+\begin{figure}[htbp]
   \centering
-  \subfloat[Spermidine hydroxycinnamoyltransferase 2]{%
-    \includegraphics[width=0.45\textwidth]{figures/structure_LOC_Os12g27254.png}
-    \label{fig:structure_SHT2}
-  }\hfill
-  \subfloat[Lipoxygenase]{%
-    \includegraphics[width=0.45\textwidth]{figures/structure_LOC_Os12g37260.png}
-    \label{fig:structure_Lipoxygenase}
-  }
-  \caption{3D protein structure of named genes selected by LASSO, available from~\cite{szklarczyk2016string}.}
+    \includegraphics[clip,width=0.8\textwidth]{figures/figure7.pdf}
+  \caption{3D protein structure of named genes selected by LASSO, borrowed from~\cite{szklarczyk2016string}.}
   \label{fig:3d}
@@
 
-Uniprot database~\cite{uniprot2018uniprot} reports, on the one hand,
-that SHT2 contributes to the natural variation of spermidine-based
-phenolamides in rice cultivars. On the other hand, it is reported that
-plant lipoxygenase may be involved in a number of diverse aspects of
-plant physiology including growth and development, pest resistance,
-and senescence or responses to wounding~\cite{uniprot2018uniprot}. This protein is involved in
-the pathway oxylipin biosynthesis, which is part of Lipid
-metabolism. Aditionally, previous studies
+The Uniprot database~\cite{uniprot2018uniprot} reports, on the one
+hand, that SHT2 contributes to the natural variation of
+spermidine-based phenolamides in rice cultivars. On the other hand, it
+is reported in~\cite{uniprot2018uniprot} that plant lipoxygenase may
+be involved in a number of diverse aspects of plant physiology
+including growth and development, pest resistance, and senescence or
+responses to wounding. This protein is involved in the pathway
+oxylipin biosynthesis, which is part of Lipid metabolism. Additionally,
+previous studies
 %
@@
 lipoxygenase in tolerance to salt stress in other plants or even in
-rice cultivars. However, further studies are needed to elucidate the
-detailed biological function of the remaining 14 genes that have not
-been named so far, which may have a potential relevance in stress
+rice cultivars.
+
+As a conclusion, the results presented in this section suggest that
+further studies are needed to elucidate the detailed biological
+function of the remaining 14 genes that have not been named so far in
+the literature.  They may have the potential to intervene in stress
 responsive mechanisms to salt conditions in rice.

### 2020-12-07 camila/report/concl.tex
references added
bmc version

@@
-\section{Concluding Remarks}
+\section*{Concluding Remarks}
 \label{sec.concl}
 
-This manuscript provides a detailed description of a network-based analysis
-methodology for the discovery of key genes responding to
-a specific treatment in an organism. It links transcriptomic with
-phenotypic data, and identifies overlaping gene modules.
+This manuscript provides a detailed description of a network-based
+analysis workflow for the discovery of key genes responding to a
+specific treatment in an organism. It links transcriptomic with
+phenotypic data and identifies overlapping gene modules.
+\vspace{0.5cm}
 
-The proposed methodology is inspired by the workflow suggested in the
-WGCNA methodology~\cite{langfelder2008wgcna}. The main steps are the
-preprocessing of the gene expression data, the construction of a
-co-expression network, the detection of modules within the network,
-the relation of modules with external information (e.g. phenotypic data)
-and the enrichment of the identified key genes with additional information.
-The WGCNA methodology and therefore the proposal as well, are structured
-in a modular way, which allows modifying and  exploring different techniques
-in each step of the process.
+The proposed approach is inspired by the workflow suggested in the
+WGCNA~\cite{langfelder2008wgcna}. Its main steps are the preprocessing
+of the gene expression data, the construction of a co-expression
+network, the detection of modules within the network, the relation of
+modules with external information (e.g., phenotypic data), and the
+enrichment of the identified key genes with additional information.
+Both approaches are structured in a modular way, which allows
+modifying and exploring different techniques in each step of the
+workflow.
+\vspace{0.5cm}
 
-The proposed approach is designed to integrate expression data
-measured under two different conditions (namely, control and treatment),
-unlike the usually co-expression-based approaches which work with both
-conditions independently or consider only a single condition. For this
-purpose, an approach similar to that proposed in~\cite{du2019network}
-is used, where the control and treatment data are compiled in a single
-matrix using the log fold change measure. Thus, the input to construct
-the co-expression network is not the expression data, but instead the changes
-in the expression levels from one condition to the other, making room for capturing
-the signal of changes caused by the treatment.
+The proposed workflow is designed to integrate expression data
+measured under two different conditions (namely, control and
+treatment), unlike the usually co-expression-based approaches which
+work with both conditions independently or consider only a single
+condition. For this purpose, an approach similar to that proposed
+in~\cite{du2019network} is used, where the control and treatment data
+are compiled in a single matrix using the Log Fold Change
+measure. Thus, the input to construct the co-expression network is not
+the expression data, but instead the changes in the expression levels
+from one condition to the other, making room for capturing the signal
+of changes caused by the treatment.
+\vspace{0.5cm}
 
-Another important feature in the proposed methodology is the module
-detection technique. The co-expression network is computed as in
-WGCNA until a scale-free network is obtained. This network is then
-used to apply the HLC algorithm, a clustering technique capable of
-detecting overlapping communities. Several approaches of module
-detection from gene expression have been proposed and were evaluated
-in~\cite{saelens2018comprehensive}. Most of them focuses only on
-disjoint (non-overlapping) communities, and the described techniques
-dealing with overlaps are not clustering but biclustering and
+An important feature in the proposed workflow is the module
+detection technique. The co-expression network is computed, as in
+WGCNA, until a scale-free network is obtained. In the proposed
+approach, this network is then used to apply the HLC algorithm, a
+clustering technique capable of detecting overlapping
+communities. Several approaches of module detection from gene
+expression have been proposed and were evaluated
+in~\cite{saelens2018comprehensive}. Most of them focus mainly on
+disjoint (non-overlapping) communities; the techniques described
+dealing with overlaps are not clustering, but bi-clustering and
 decomposition methods. It is well known that communities in real
-networks are overlapping~\cite{palla2005uncovering}. Thus, the approach
-presented in this work can be seen as a generalization of the previous 
-approaches such as WGCNA.
+networks, including biological ones, are likely
+overlap~\cite{palla2005uncovering}. Thus, the approach presented
+in this work can be seen as a generalization of the previous
+approaches, such as WGCNA, with the potential to deal with genes
+associated to multiple biological processes.
+\vspace{0.5cm}
 
-The proposed methodology was applied in a case study with rice under salt
-stress. The results show a group of 14 genes in which $2$ of them are
-related to the response to saline stress according to previous
-studies, validating the ability of the method to detect this kind of
-key genes. As future work, other overlapping module detection and selection
-techniques should be used instead HLC and LASSO, respectively. The
-combination of these techniques would allow finding target genes for
-future biological studies that evaluate their potential as genes that
-respond to salt stress in rice. Also, laboratory experimentation needs to be
-conducted on via to verify the findings of this paper in relation to salinity 
-stress. Finally, this study can be extended to other stresses and even to other 
-crops.
+The approach was applied in a case study with rice under salt
+stress. The results show a group of 14 genes, of which only $2$ of
+them have been previously related to saline stress response in other
+studies. As future work, other overlapping module detection and
+selection techniques should be used instead HLC and LASSO,
+respectively. The combination of these techniques would allow finding
+target genes for future biological studies that evaluate their
+potential as genes that respond to salt stress in rice, and other
+crops and stresses. In-vivo laboratory experimentation needs to be
+conducted to validate the findings of this paper in relation to
+salinity stress.
+\vspace{0.5cm}
+
+Finally, the workflow is presented as a protocol capable of
+considerably reducing the number of genes detected as relevant in the
+response to stress of choice. Other traditionally used methods for
+this purpose tend to generate a large list of candidate genes, thus
+limiting subsequent efforts in experimental validation. In this sense,
+the proposed workflow can help in reducing such efforts in time and
+money invested by researchers in the experimental validation of
+stress-responsive genes.

### 2020-12-07 camila/report/declarations.tex
bmc version

--- 2020-12-07 camila/report/bmc_article.tex
@@
-%% BioMed_Central_Tex_Template_v1.06
-%%                                      %
-%  bmc_article.tex            ver: 1.06 %
-%                                       %
-
-%%IMPORTANT: do not delete the first line of this template
-%%It must be present to enable the BMC Submission system to
-%%recognise this template!!
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                     %%
-%%  LaTeX template for BioMed Central  %%
-%%     journal article submissions     %%
-%%                                     %%
-%%          <8 June 2012>              %%
-%%                                     %%
-%%                                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% For instructions on how to fill out this Tex template           %%
-%% document please refer to Readme.html and the instructions for   %%
-%% authors page on the biomed central website                      %%
-%% https://www.biomedcentral.com/getpublished                      %%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%% BioMed Central currently use the MikTex distribution of         %%
-%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
-%% https://miktex.org/                                             %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%% additional documentclass options:
-%  [doublespacing]
-%  [linenumbers]   - put the line numbers on margins
-
-%%% loading packages, author definitions
-
-%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass{bmcart}
-
-%%% Load packages
-\usepackage{amsthm,amsmath}
-\usepackage{amsfonts}
-
-\usepackage{graphicx}
-%\usepackage[caption=false]{subfig}
-
-%\RequirePackage[numbers]{natbib}
-%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
-%\RequirePackage{hyperref}
-\usepackage[utf8]{inputenc} %unicode support
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage{lmodern} %font sizes
-\usepackage{hyperref} % ref unlabeled sections
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                             %%
-%%  If you wish to display your graphics for   %%
-%%  your own use using includegraphic or       %%
-%%  includegraphics, then comment out the      %%
-%%  following two lines of code.               %%
-%%  NB: These line *must* be included when     %%
-%%  submitting to BMC.                         %%
-%%  All figure files must be submitted as      %%
-%%  separate graphics through the BMC          %%
-%%  submission process, not included in the    %%
-%%  submitted article.                         %%
-%%                                             %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\def\includegraphic{}
-%\def\includegraphics{}
-
-%%% Put your definitions there:
-\startlocaldefs
-\newcommand{\abs}[1]{\left\vert#1\right\vert}
-\endlocaldefs
-
-%%% Begin ...
-\begin{document}
-
-%%% Start of article front matter
-\begin{frontmatter}
-
-\begin{fmbox}
-\dochead{METHODOLOGY}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{Using Overlapping Communities and Network Structure for
-  Identifying Reduced Groups of Stress Responsive Genes}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Specify information, if available,       %%
-%% in the form:                             %%
-%%   <key>={<id1>,<id2>}                    %%
-%%   <key>=                                 %%
-%% Comment or delete the keys which are     %%
-%% not used. Repeat \author command as much %%
-%% as required.                             %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\author[
-  addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-  corref={aff1},                      % id of corresponding address, if any
-% noteref={n1},                        % id's of article notes, if any
-  email={camila.riccio@javerianacali.edu.co}   % email address
-]{\inits{C.R.}\fnm{Camila} \snm{Riccio}}
-\author[
-  addressref={aff2},
-  email={jfinke@javerianacali.edu.co}
-]{\inits{J.F.}\fnm{Jorge} \snm{Finke}}
-\author[
-  addressref={aff2},
-  email={camilo.rocha@javerianacali.edu.co}
-%]{\inits{C.R.}\fnm{Camilo} \snm{Rocha}}
-]{\inits{H.C.R.}\fnm{Hernan C.} \snm{Rocha}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%% Repeat \address commands as much as      %%
-%% required.                                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\address[id=aff1]{%                                         % unique id
-  \orgdiv{Department of Natural Sciences and Mathematics},  % department, if any
-  \orgname{Pontificia Universidad Javeriana},               % university, etc
-  \city{Cali},                                              % city
-  \cny{Colombia}                                            % country
-}
-
-\address[id=aff2]{%                                         % unique id
-  \orgdiv{Department of Electronics and Computer Science},  % department, if any
-  \orgname{Pontificia Universidad Javeriana},               % university, etc
-  \city{Cali},                                              % city
-  \cny{Colombia}                                            % country
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter short notes here                   %%
-%%                                          %%
-%% Short notes will be after addresses      %%
-%% on first page.                           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\begin{artnotes}
-%%\note{Sample of title note}     % note to the article
-%\note[id=n1]{Equal contributor} % note, connected to author
-%\end{artnotes}
-
-\end{fmbox}% comment this for two column layout
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                           %%
-%% The Abstract begins here                  %%
-%%                                           %%
-%% Please refer to the Instructions for      %%
-%% authors on https://www.biomedcentral.com/ %%
-%% and include the section headings          %%
-%% accordingly for your article type.        %%
-%%                                           %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstractbox}
-
-\begin{abstract} % abstract
-\parttitle{Background} %if any
-This paper proposes a workflow to identify which genes respond to
-  specific treatments in plants, such as abiotic stresses.
-  The workflow takes as
-  input RNA sequencing read counts and phenotypical data (measured for genotypes under
-  control and treatment conditions) with biological replicates. It
-  outputs a collection of genes marked as potentially relevant to
-  treatment. Technically, the proposed approach is both a
-  generalization and an extension of WGCNA. Its main goal is to
-  identify, after a sequence of normalization and filtering steps, specific modules in a gene co-expression network.
-  Module detection is
-  achieved by using Hierarchical Link Clustering, to recognize
-  overlapping gene communities. Such communities have a biological meaning given
-  the overlapping regulatory domains of systems that generate the patterns of
-  co-expression. Additional steps and information are also added to
-  the workflow, where networks in intermediate steps are
-  forced to be scale-free. Finally, LASSO regression is employed to select
-  the most significant modules of phenotypical responses to stress.
-
-\parttitle{Results} %if any
-The workflow is showcased with a systematic study on rice
-  (\textit{Oryza sativa}), a major food source known to be
-  highly sensitive to salt stress. A total of 19 genes distributed across 6 modules are detected
-  as relevant in the response to salt stress.
-  These genes represent potential targets for the improvement of
-  salinity tolerance in rice crops.
-
-\parttitle{Conclusions}
-The proposed workflow provides a framework to reduce the
-  search-space for candidate genes that respond to a specific treatment.
-  It aims to optimize the effort, time, and resources researchers invest
-  in the experimental validation of stress responses.
-
-\end{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The keywords begin here                  %%
-%%                                          %%
-%% Put each keyword in separate \kwd{}.     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{keyword}
-\kwd{Stress-responsive genes}
-\kwd{co-expression network}
-\kwd{overlapping communities}
-\kwd{phenotypic traits}
-\kwd{LASSO}
-\kwd{salinity}
-\kwd{rice}
-\kwd{\textit{Oryza sativa}}
-\end{keyword}
-
-% MSC classifications codes, if any
-%\begin{keyword}[class=AMS]
-%\kwd[Primary ]{}
-%\kwd{}
-%\kwd[; secondary ]{}
-%\end{keyword}
-
-\end{abstractbox}
-%
-%\end{fmbox}% uncomment this for two column layout
-
-\end{frontmatter}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                            %%
-%% The Main Body begins here                  %%
-%%                                            %%
-%% Please refer to the instructions for       %%
-%% authors on:                                %%
-%% https://www.biomedcentral.com/getpublished %%
-%% and include the section headings           %%
-%% accordingly for your article type.         %%
-%%                                            %%
-%% See the Results and Discussion section     %%
-%% for details on how to create sub-sections  %%
-%%                                            %%
-%% use \cite{...} to cite references          %%
-%%  \cite{koon} and                           %%
-%%  \cite{oreg,khar,zvai,xjon,schn,pond}      %%
-%%                                            %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
-% <put your article body there>
-
-%%%%%%%%%%%%%%%%
-%% Background %%
-%%
-\section*{Introduction}
-\label{sec.intro}
-
-Abiotic stresses are key factors that influence plant
-development and productivity. They are often associated to extensive
-losses in agricultural production~\cite{mesterhazy2020losses}. 
-Soil salinity is one of the
-most devastating abiotic stresses. According to~\cite{shrivastava2015soil}, 
-soil salinity contributes significantly to the reduction in cultivable
-land and crop quality. The study estimates that worldwide 20\%
-of the total cultivated land and 33\% of the total irrigated agricultural land
-is already affected by high salinity. Moreover, due
-to human activities and natural causes, salinized areas are
-expected to reach 50\% of the total cultivated land by the
-end of 2050~\cite{shrivastava2015soil}. Salinity tolerance and
-susceptibility in plants is known to be the result of elaborated
-interactions between morphological, physiological, and biochemical
-processes that are regulated, in the end, by multiple genes in
-various parts of the plant genome~\cite{reddy2017salt}. Therefore,
-identifying groups of stress responsive genes is an important step in efforts to
-improve crop varieties in terms of salinity tolerance and, more generally,
-food sustainability.
-\vspace{0.5cm}
-
-This paper proposes a workflow to identify stress responsive genes
-(known to be a complex quantitative trait).
-To discover which genes are related with the phenotypic response to
-treatment, the approach requires a collection of phenotypic
-traits, measured for the given genotypes.  The proposed workflow takes as input
-gene expression profiles of the target organism, specifically, 
-RNA sequencing read counts (measured for genotypes under control
-and treatment conditions)
-with at least two biological replicates. It also receives phenotypic data, i.e., observable characteristics or traits, measured for each genotype under the two conditions. The output of the
-workflow is a collection of genes which are characterized as potentially relevant
-to treatment. The workflow provides a framework that yields insight into the possible behavior of specific
-genes and the role they play in functional pathways in response to
-treatment in the organism of interest. Bluntly put, the
-workflow takes advantage of transcriptomic data for
-organisms under different conditions, based on the current availability of
-high-throughput technologies,
-to study the reaction of organisms under different
-environmental stimuli.
-\vspace{0.5cm}
-
-The proposed approach is both a generalization and an
-extension of Weighted Gene Co-expression Network Analysis
-(WGCNA)~\cite{langfelder2008wgcna}, a widely applied workflow that has
-been successfully used for identifying target genes related to
-diseases in several
-organisms~\cite{tian2018identifying}. The general idea behind each
-approach is to identify, after a sequence of normalization and filtering steps, specific modules in a gene co-expression network. 
-The proposed approach
-is considered a \textit{generalization} of WGCNA because module
-detection can now recognize overlapping communities, which have
-a biological meaning given the overlapping regulatory domains of
-systems that generate the patterns of co-expression~\cite{gaiteri2014beyond}. This is
-achieved by using Hierarchical Link Clustering
-(HLC)~\cite{ahn2010link}. 
-HLC enables the detection of overlapping modules, which naturally takes
-into account that biological components  are involved in
-multiple functions and therefore biological communities tend to be
-highly overlapping.
-The workflow is also an \textit{extension} of WGCNA
-because additional constraints are considered:
-namely, networks in the intermediate steps are forced to be
-scale-free~\cite{barabasi2003scale} and LASSO
-regression~\cite{tibshirani1996regression} is employed to select the
-most significant modules of phenotypical responses to stress.
-LASSO is a regularized
-regression technique which forces the less relevant variables to be associated to regression coefficients of value zero~\cite{desboulets2018review}. Moreover, LASSO is especially
-useful in problems where the number of variables is much larger than
-the number of samples. This is surely the case when the variables are
-overlapping gene modules obtained with HLC and the samples are genotypes, which is usually a small set due to the high cost of the RNA sequencing process. 
-Finally, the proposed workflow is also modular, since other module detection
-and selection techniques could be used, instead HLC and LASSO,
-respectively.
-\vspace{0.5cm}
-
-The approach is showcased with a systematic study on rice
-(\textit{Oryza sativa}), a food source that is known to be
-highly sensitive to salt stress~\cite{chang2019morphological}. RNA-seq
-data was accessed from the GEO database~\cite{clough2016gene}
-(accession number GSE98455). It corresponds to $57845$ gene expression
-profiles of shoot tissues measured for both control and stress condition
-in $92$ accessions of the Rice Diversity Panel 1~\cite{eizenga2014registration}. A total of 6 modules
-are detected as relevant in the response to salt stress in rice: 3
-modules of 3 genes each, all associated with shoot K content, 2
-modules of 3 genes associated with shoot biomass, and 1 module of 4
-genes associated with root biomass. These genes are potential
-targets for experimental validation of salinity tolerance in rice
-cultivars. From the 19 genes, all but 3 genes (associated with $K$
-content) are also identified as deferentially expressed for at least
-one of the 92 accessions, suggesting that the identified genes are
-stress responsive genes. Only 2 of the 16 differentially
-expressed genes, both from the module associated to shoot biomass, are
-named and have a known associated protein product: spermidine
-hydroxycinnamoyltransferase 2 (SHT2) and lipoxygenase. 
-Further studies are needed to elucidate the detailed biological
-functions of the remaining 14 genes and their role
-in stress responsive mechanisms
-to salt conditions in rice. The goal is that the results reported in
-this case study may allow biologist to develop new rice cultivars with
-higher resistance to salinity.
-
-\paragraph{Paper Outline.}
-%The remainder of the paper is organized as
-%follows. Section~\ref{sec.prelim} gathers preliminaries on gene
-%co-expression networks, HLC, and LASSO. The proposed workflow is
-%presented in Section~\ref{sec.framework}, with especial focus on the
-%logical steps of the process and the internal structures supporting
-%the approach. Section~\ref{sec.case} presents the case study on the
-%identification of rice genes that are sensitive to salt
-%stress. Section~\ref{sec.concl} draws some conclusions and future research directions.
-
-The remainder of the paper is organized as
-follows. \hyperref[sec.prelim]{Preliminaries} gathers foundations on gene
-co-expression networks, HLC, and LASSO. The proposed workflow is
-presented in \hyperref[sec.framework]{The Workflow} section, with especial focus on the
-logical steps of the process and the internal structures supporting
-the approach. \hyperref[sec.case]{Case Study} presents the case study on the
-identification of rice genes that are sensitive to salt
-stress. \hyperref[sec.concl]{Concluding Remarks} draws some conclusions and future research directions.
-
-\section*{Preliminaries}
-\label{sec.prelim}
-
-This section presents preliminaries on networks, hierarchical link
-clustering, and the LASSO linear regression technique.
-
-\subsection*{Co-expression network}
-
-A network is an undirected graph $G=(V,E)$ where
-${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of $n$ \textit{vertices} or
-\textit{nodes} and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of $q$
-\textit{edges} or \textit{links} that connect vertices. In a gene
-co-expression network, each node corresponds to a gene. A pair of
-genes is connected if they show similar expression patterns. A simple
-and unweighted network can be represented by an adjacency matrix $A
-\in \{0,1\}^{n \times n}$ that is symmetric with a positive one in the
-positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever there is an edge
-connecting vertices $v_i$ and $v_j$, and zeros
-elsewhere. Co-expression networks are of biological interest because
-adjacent nodes in the network represent co-expressed genes that are
-usually controlled by the same transcriptional regulatory pathway,
-functionally related, or members of the same pathway or metabolic
-complex.
-
-\subsection*{Hierarchical Link Clustering}
-
-The Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn
-et al.~\cite{ahn2010link}. It represents communities as
-groups of links (rather than nodes), and each node inherits all
-memberships of its links and can thus belong to multiple, overlapping
-communities. The algorithm maps links to nodes and connects them if a pair of
-links shares a node. For a connected pair of links $e_{ik}$ and $e_{jk}$,
-the shared node $k$ is called a \textit{keystone} node. 
-The similarity between two links $e_{ik}$ and
-$e_{jk}$ is computed using the Jaccard index
-%
-\begin{equation}\label{eq:jaccard}
-  S(e_{ik},e_{jk}) = \frac{\vert n(i) \cap n(j) \vert}{\vert n(i) \cup n(j) \vert},
-\end{equation}
-%
-where $n(i)$ denotes the set containing exactly node $i$ and its
-neighbors. The algorithm uses single-linkage hierarchical clustering
-to build a dendrogram in which each leaf is a link from the original
-network, and branches represent link communities. Hierarchical
-clustering algorithms repeatedly merge groups until all elements are
-members of a single cluster.
-\vspace{0.5cm}
-
-For the purpose of finding meaningful communities, it is crucial to
-know where to partition a dendrogram. In this case, the most
-relevant communities are established at the maximal partition density
-$D$, a function based on link density inside communities measuring the
-quality of a link partition. The partition density $D$ has a single
-global maximum along the dendrogram in almost all cases, because its
-value is the average density at the top of the dendrogram (a single
-giant community with every link and node) and it is very small at the
-bottom of the dendrogram (most communities consists of a single
-link). In particular, it is the case that $D = 1$ when every community
-is a fully connected clique and $D = 0$ when each community is a
-tree. If a community is less dense than a tree (i.e. when the
-community subgraph has disconnected components), then such a community
-contributes negatively to $D$, which can take negative values. The
-minimum density inside a community is $-2/3$, given by one community
-of two disconnected edges. Since $D$ is the average of the
-intra-community density, there is a lower bound of $-2/3$ for
-$D$. Computing $D$ at each level of the link dendrogram can help the
-purpose of picking the best level to cut, although meaningful
-structure could exist above or below the threshold. The output of
-cutting is a set of node clusters, where each node can participate in
-multiple communities.
-
-\subsection*{Least Absolute Shrinkage Selector Operator}
-
-The Least Absolute Shrinkage Selector Operator (LASSO) is a
-regularized linear regression technique. It combines a regression
-model with a procedure of contraction of some parameters towards zero
-and selection of variables, imposing a restriction or a penalty on the
-regression coefficients. In other words, LASSO solves the least
-squares problem with restriction on the $ L_1$-norm of the coefficient
-vector. It can be especially useful to solve problems where the number
-of variables (e.g., genes) $ n $ is much greater than the number of
-samples $ m $ (i.e., $ n \gg m $).
-\vspace{0.5cm}
-
-Consider a dataset consisting of $m$ samples, each of which consists
-of $n$ covariates and a single outcome. Let $y_i$ be the outcome and
-$x_i := (x_{i1},...,x_{in})$ be the covariate vector for the $i$-th
-sample. The objective of LASSO is to solve
-%
-\begin{equation}
-\min \left\lbrace\sum_{i=1}^{m}{\left( y_i-\sum_{j=1}^n{\beta_j
-    x_{ij}}\right)^2} \right\rbrace \quad , \quad \textrm{subject to}
-\quad \sum_{j=1}^n\abs{\beta_j}\leq s.
-\end{equation}
-
-Equivalently, in the Lagrangian form, it minimizes
-
-\begin{equation}
-  \sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} +
-  \lambda \sum_{j=1}^n\abs{\beta_j}
-\end{equation}
-%
-where $s$ is the regularization penalty and $\lambda \geq 0$ is the
-corresponding Lagrange multiplier. Since the $\lambda$ value
-determines the degree of penalty, the accuracy of the model depends on
-its choice. Cross-validation is often used to select the
-regularization parameter, finally selecting the one that minimizes the
-mean-squared error.
-
-\section*{The Workflow}
-\label{sec.framework}
-
-The proposed workflow is depicted in
-Figure~\ref{fig:flow_chart}. This section explains the five
-macro-processes (A)-(E) of the proposed workflow. In comparison with
-WGCNA, it adds the macro-step (D) and generalizes macro-steps (A)-(C).
-\vspace{0.5cm}
-
-% figure 1
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[clip,width=0.96\textwidth]{figures/figure1.pdf}
-  \caption[The proposed workflow comprising five macro-steps]%
-  {The proposed workflow comprising five macro-steps:
-    A.~Data pre-processing, B.~Co-expression network
-    contruction, C.~Co-expression module identification,
-    D.~Modules association to phenotypic traits, and
-    E.~Gene enrichment.}
-  \label{fig:flow_chart}
-\end{figure}
-
-The workflow uses RNA-seq read counts, representing gene expression
-levels, as input data. More precisely, it uses $n_0$ gene expression
-profiles of an organism, measured for $m$ different genotypes under
-control and treatment conditions, and $r$ biological replicates. This
-raw data is represented as a matrix $D_0 \in {\mathbb{N}_0}^{n_0
-  \times 2mr}$. In order to discover key genes and their interaction
-with phenotypes related to treatment tolerance, the approach also
-requires a set of $p$ phenotypic traits, measured for the $m$
-genotypes. The phenotypic data is seen as a matrix $P \in
-\mathbb{R}^{2m \times p}$ containing two phenotypic values per
-genotype, one under control condition and a second one under treatment
-condition.
-
-\subsection*{Data Pre-processing}
-
-The goal of the data pre-processing stage is to build matrices
-$P_\ell$ and $L_1$ representing, respectively, the changes in
-phenotypic values and expression levels between control and treatment
-condition, from RNA-seq and phenotypic data found in matrices $D_0$
-and $P$, respectively.
-\vspace{0.5cm}
-
-The RNA-seq data cannot be directly interpreted. Therefore, a
-normalization process is applied to deal with the problem of possible
-biases affecting the quantification of results. The suggested
-normalization technique for correcting library size and RNA
-composition bias is DESeq2~\cite{love2014moderated}. The normalized
-data is represented as a matrix $D_1 \in \mathbb{R}^{n_0 \times 2mr}$,
-and the biological replicates of each genotype are averaged and
-represented as a matrix $D_2 \in \mathbb{R}^{n_0 \times 2m}$. The
-genes exhibiting low variance or low expression are removed from
-$D_2$, thus identifying a subset of size $n_1 \leq n_0$ of the
-original genes. The control and treatment data is separated into the
-matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1
-  \times m}$, respectively. The matrix entries $c_{ij}$ in $C$ and
-$t_{ij}$ in $T$ represent the normalized expression
-level of gene $i$ in accession $j$ under control and treatment 
-condition, respectively.
-Control and treatment data is also separated from
-phenotypic data $P$, obtaining the $P_c$ and $P_t$ matrices of
-dimensions $m \times p$.
-\vspace{0.5cm}
-
-In the above configuration, the changes in expression levels and
-phenotypic values between control and treatment conditions, are
-measured in terms of logarithmic ratios. In the case of expression
-levels, the log ratios are represented in the Log Fold Change matrix
-$L_0 \in \mathbb{R}^{n_1 \times m}$, where $\ell_{ij}=\log_2
-(t_{ij}/c_{ij})$. Similarly, the log ratios of the phenotypic data are
-computed and represented in the $P_\ell \in \mathbb{R}^{m \times p}$
-matrix.
-\vspace{0.5cm}
-
-The final step of the data pre-processing is to filter $L_0$ by
-removing rows (e.g., genes) with low variance in the differential
-expression patterns, obtaining a new matrix $L_1$ of dimensions $n_2
-\times m$, with $n_2 \leq n_1$.
-
-\subsection*{Co-expression Network Construction}
-
-A gene co-expression network connect genes with similar expression
-patterns across biological conditions. The purpose of this stage is to
-describe how to build the co-expression network $A$ from the Log Fold
-Change matrix $L_1$, capturing the relationship between genes
-according to the change in expression levels between the two studied
-conditions. These co-expression patterns are meaningful for the
-identification of genes not yet associated with the response to the
-treatment condition.
-\vspace{0.5cm}
-
-The Log Fold Change matrix $L_1$ is used to build the co-expression
-network following the first two steps of the WGCNA
-methodology~\cite{langfelder2008wgcna}. First, the level of
-concordance between gene differential expression profiles across
-samples is measured. To this end, the absolute value of the Pearson
-correlation coefficient is used as the similarity measure between
-genes and the resulting values are stored in the similarity matrix
-$S\in \mathbb{R_{+}}^{n_2 \times n_2}$. Second, the matrix $S$ is
-transformed into an adjacency matrix $A \in \mathbb{R_+}^{n_2\times
-  n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the
-connection strength between each pair of genes. In other words, the
-elements of the adjacency matrix are the similarity values up to the
-power $\beta > 1$ so that the degree distribution will fit a
-scale-free network. These networks contain many nodes with very few
-connections and a small number of hubs with high connections. In a
-strict scale-free network, the logarithm of $P(k)$ (i.e., the
-probability of a node having degree $k$) is approximately inversely
-proportional to the logarithm of $k$ (i.e., the degree of a node).
-The parameter $\beta$ is chosen to be the smallest power such that the
-$R^2$ of the linear regression between $log_{10}(p(k))$ and
-$log_{10}(k)$ is close to $1$ (e.g., $R^2 > 0.85$).
-
-\subsection*{Co-expression Module Identification}
-
-The next step in the workflow is to identifying communities (also
-called modules) from the co-expression network structure and dynamics
-represented in $A$.  The idea is to cluster genes with similar
-patterns of differential expression change. Membership in these
-modules may overlap in biological contexts, because modules may be
-related to specific molecular, cellular, or tissue functions, and the
-biological components (i.e., genes) may be involved in multiple
-functions. Thus, unlike WGCNA, the adjacency matrix $A$ is used to
-detect overlapping (rather than non-overlapping) communities, using
-the Hierarchical Link Clustering (HLC) algorithm (see
-%Section~\ref{sec.prelim}).
-\hyperref[sec.prelim]{Preliminaries}).
-\vspace{0.5cm}
-
-As a preliminary step, matrix $A$ is transformed into an unweighted
-network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$ before using the
-clustering algorithm. To this end, the Pearson Correlation Coefficient
-(PCC) cutoff is determined using the approach described
-in~\cite{aoki2007approaches}. The number of nodes, edges, and the
-network density is determined for different PCC cutoffs. Near the most
-biological relevant PCC cutoff, the number of nodes presents a linear
-decrease and the density of the network reaches its minimum, while
-below this value the number of edges rapidly increases. Following this
-observation, a cutoff is selected such that gene pairs which have a
-correlation score higher than the threshold are considered to have
-important co-expression relationship. Above the cutoff, the entries of
-matrix $A$ become $1$ and below the cutoff $A$ values become $0$. The
-HLC algorithm organizes the $n_2$ genes of matrix $\hat{A}$ into $c$
-modules, where each gene can belong to zero, one or multiple modules.
-This information is represented as an affiliation matrix $F \in
-\{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ iff node $i$ is member of
-module $u$ (it is 0, otherwise).
-
-\subsection*{Module Association to Phenotypic Traits}
-
-To identify the most relevant modules of genes, associated
-with the phenotypic response to a specific treatment in an organism,
-the proposed workflow uses LASSO. Specifically, each module is
-represented by an eigengene, which is defined as the first principal
-component of such module. An eigengene can be seen as an average
-differential expression profile for each community: it is computed
-from the Log Fold Change Matrix $L_1$ and the affiliation matrix
-$F$. Given a module $u$, the affiliation matrix is used to identify
-the genes belonging to $u$ and then the corresponding rows of the
-matrix $L_1$ are selected to compute the first principal component of
-$u$. Each principal component becomes a column of the matrix $M \in
-\mathbb{R}^{m \times c}$.  These profiles are then associated with each
-phenotypic trait using
-LASSO.  In this context, the eigengenes (i.e., the columns
-of $M$) act as regressor variables and each phenotypic trait (i.e.,
-each column of $P_\ell$) is used as an outcome variable.
-\vspace{0.5cm}
-
-The output after applying LASSO is a set $W_z$ of modules for each
-phenotypic trait $z$, where $W_z \subset \{u \mid 1 \leq u \leq c\}$
-for $z= 1,2,..,p$. The target genes $I$ for downstream analysis, which
-may be important in the treatment response, are the union of genes
-belonging to the selected modules; that is $I = \cup_{z=1}^{p} W_z$,
-where $I \subset \{i \mid 1 \leq i \leq n_2\}$.
-
-\subsection*{Gene Enrichment}
-
-The goal of this final stage of the process is to annotate with
-additional information the genes identified in previous stages,
-helping to elucidate their possible behavior and role in the response
-to the studied treatment.
-\vspace{0.5cm}
-
-A crucial step is to identify the differentially expressed genes in
-set $I$. That is, to select those genes in $I$ having an absolute
-value of the Log Fold Change of at least $2$ ($|\ell_{ij}|\geq 2$) for
-at least one sample. This represents genes whose expression level is
-quadrupled (up or down) from control to treatment condition; they are the
-strong candidates for treatment responsive genes.
-\vspace{0.5cm}
-
-Also, functional category enrichment can be done by, e.g., searching
-for gene ontology annotations in databases such as
-QuickGO~\cite{binns2009quickgo}. Such annotations can provide evidence
-of biological implications of the target genes in the
-treatment-tolerance mechanisms. Furthermore, QuickG0 can be used to
-identify genes with reported protein products, which can be used to
-perform additional relevant analysis reviewing their reported
-protein-protein interactions in other databases, such as
-STRING~\cite{szklarczyk2016string}. The interactions include direct
-(physical) and indirect (functional) associations, and they stem from
-computational prediction, knowledge transfer between organisms, and
-interactions aggregated from other (primary) databases. This
-information can give new insights on how the selected genes are
-involved in functional pathways that can be related to the treatment
-of interest.
-
-%\section*{Results}
-\section*{Identifying Potential Saline Stress Responsive Genes in Rice}
-\label{sec.case}
-
-This section presents a case study on the identification of genes in
-\textit{Oryza sativa} that may respond to saline stress following the
-approach presented in %Section~\ref{sec.framework}.
-\hyperref[sec.framework]{The Workflow}.
-\vspace{0.5cm}
-
-The RNA-seq data used for the experiments summarized in this section
-was obtained from GEO database \cite{clough2016gene} (accession
-number GSE98455). This data corresponds to $n_0=57\,845$ gene
-expression profiles of shoot tissues measured for both control and
-salt condition in $m=92$ accessions of the Rice Diversity Panel 1~\cite{eizenga2014registration},
-with $r=2$ biological replicates. A total of $p=3 $ phenotypic traits
-are used: shoot $K^+$ content, root biomass, and shoot biomass. These
-traits were measured for the same $92$ genotypes, under control and
-salt stress conditions, and can be found in the supplementary
-information of~\cite{campbell2017allelic}.
-
-\subsection*{Data Pre-processing}
-
-DESeq2 normalization is applied to the raw data and the biological
-replicates are averaged. Genes exhibiting low variance are
-identified as those with ratio of upper quantile to lower quantile
-smaller than $1.5$ and are removed from the normalized data. Genes
-with low expression, corresponding to those having more than $80\%$
-samples with values smaller than $10$, are also removed. A total of
-$n_1 = 9\,414$ genes are kept after this filtering process.
-\vspace{0.5cm}
-
-From the Log Fold Change matrix $L_0$, genes whose difference between
-upper quantile and lower quantile greater than $0.25$ are
-removed. Therefore, the resulting matrix $L_1$ contains the log ratios
-of $n_2 = 8\,928$ genes. The logarithmic ratios of the phenotypic data,
-for the $92$ accessions and the $3$ traits, are also computed.
-
-\subsection*{Co-expression Network Construction}
-
-The Log Fold Change matrix $L_1$ is used to compute the corresponding
-similarity matrix.  For this network, it is observed that $\beta=3$ is
-the smallest integer such that the $R^2 \geq 0.8
-$. Figure~\ref{fig:beta} depicts the degree distribution of the
-similarity matrix (left) and the degree distribution of the adjacency
-matrix (right), which is the degree distribution of a scale-free
-network with $R^2 = 0.8$ with $\beta = 3$.
-\vspace{0.5cm}
-
-%figure 2
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[clip,width=0.8\textwidth]{figures/figure2.png}
-  \caption{Degree distribution of $A$ (left) and $\hat{A}$ (right).}
-  \label{fig:beta}
-\end{figure}
-
-The resulting adjacency matrix $A$ represents a complete graph
-$G=(V,E)$, with $|V| = 8\,928$ genes ($|E| = 39\,850\,128$ edges).
-
-\subsection*{Co-expression Module Identification}
-
-The adjacency matrix $A$ is transformed into an unweighted network
-$\hat{A}$ applying the approach described
-in~\cite{aoki2007approaches}. The cutoff value is set to $0.2$, 
-based on the density of the network
-combined with the decreasing number of nodes and edges with higher PCC
-values. Thus keeping only the
-connections above this threshold and removing the isolated nodes. The
-resulting adjacency matrix $\hat{A}$ has $5\,810$ connected genes and
-accounts for $16\,875\,145$ edges.
-\vspace{0.5cm}
-
-After applying the HLC algorithm, a total of $4\,131$ genes are
-distributed in $c = 5\,143$ overlapping modules of at least $3$
-genes. Figure~\ref{fig:overlap} presents a histogram of the
-overlapping percentage of these genes, measured as the proportion of
-modules to which each gene belongs. The first bar of the histogram
-represents the genes with zero overlap, corresponding to $28\%$ of the
-total genes; the remaining $72\%$ represents the genes belonging to
-more than one module.
-
-%figure 3
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[clip,width=0.5\textwidth]{figures/figure3.png}
-  \caption{Overlapping percentage of genes after applying HLC.}
-  \label{fig:overlap}
-\end{figure}
-
-\subsection*{Module Association to Phenotypic Traits}
-
-The phenotypic traits under study are shoot $K^+$ content, root
-biomass, and shoot biomass. Figure~\ref{fig:pdata} suggests that there
-are significant differences in the values of these phenotypic traits
-between stress and control conditions. This supports the working
-hypothesis that these three variables represent tolerance-associated
-traits in rice under salt stress.
-\vspace{0.5cm}
-
-%figure 4
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[clip,width=0.9\textwidth]{figures/figure4.png}
-  \caption{Phenotipic traits distribution under control and salt stress.}
-  \label{fig:pdata}
-\end{figure}
-
-Using the affiliation matrix $F$ derived from the HLC output and the
-Log Fold Change matrix $L_1$, a matrix $M$ is built by computing the
-eigengene for each of the $c = 5\,143$ modules. The LASSO technique is
-applied by using each of the phenotypic traits as the outcome
-variable, one at a time. As shown in Figure~\ref{fig:cross-val},
-cross-validation is performed for each phenotypical trait in order to
-select the corresponding regularization parameter $\lambda$ that
-minimizes the mean-squared error.
-\vspace{0.5cm}
-
-%figure 5
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[clip,width=0.96\textwidth]{figures/figure5.pdf}
-  \caption{Cross-validation of the LASSO regularization parameter
-    $\lambda$, for each phenotypic trait.}
-  \label{fig:cross-val}
-\end{figure}
-
-Finally, three LASSO models are adjusted by using the corresponding
-$\lambda$ and phenotypical data with the eigengenes of matrix $M$. As
-result, 6 modules are detected as relevant in the response to salt
-stress in rice: 3 modules of 3 genes, each associated with shoot $K$
-content; 2 modules of 3 genes associated with shoot biomass; and 1
-module of 4 genes associated with root biomass (see
-Figure~\ref{fig:final_genes}).
-
-\subsection*{Gene Enrichment}
-
-From the $19$ genes selected by LASSO, all but $3$ genes (the ones
-associated to $K$ content), are also identified as deferentially
-expressed ($|\ell_{ij}| \geq 2$) for at least one of the $92$
-accessions. This suggests that those genes are strong candidates as
-stress responsive genes to salt conditions in rice.
-\vspace{0.5cm}
-
-Figure~\ref{fig:final_genes} summarizes how from the initial
-$n_0=57\,845$ genes, the proposed workflow identifies a reduced set of
-$19$ genes. First, $48\,431$ genes are discarded after filtering the
-normalized expression data $D_2$ and then $486$ additional genes are
-discarded when filtering the Log Fold Change matrix $L_0$, to finally
-arrive at $19$ genes, of which $16$ are differentially expressed.
-\vspace{0.5cm}
-
-% figure 6
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[clip,width=0.8\textwidth]{figures/figure6.pdf}
-   \caption[Venn diagram for the case study in rice]%
-   {Venn diagram representing the number of genes selected at
-    different stages of the proposed workflow for the case study in
-    rice.}
-  \label{fig:final_genes}
-\end{figure}
-
-According to the Quickgo database, only $2$ of the $16$ differentially
-expressed genes (both from the module related to shoot biomass) are
-named and have an associated protein product: spermidine
-hydroxycinnamoyltransferase 2 (SHT2) and lipoxygenase.
-Figure~\ref{fig:3d} shows their corresponding 3D protein structures.
-\vspace{0.5cm}
-
-%figure 7
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[clip,width=0.8\textwidth]{figures/figure7.pdf}
-  \caption{3D protein structure of named genes selected by LASSO, borrowed from~\cite{szklarczyk2016string}.}
-  \label{fig:3d}
-\end{figure}
-
-The Uniprot database~\cite{uniprot2018uniprot} reports, on the one
-hand, that SHT2 contributes to the natural variation of
-spermidine-based phenolamides in rice cultivars. On the other hand, it
-is reported in~\cite{uniprot2018uniprot} that plant lipoxygenase may
-be involved in a number of diverse aspects of plant physiology
-including growth and development, pest resistance, and senescence or
-responses to wounding. This protein is involved in the pathway
-oxylipin biosynthesis, which is part of Lipid metabolism. Additionally,
-previous studies
-%
-in~\cite{gupta2013plant,hou2015persimmon,mittova2002salt,peng2019novel,roychoudhury2011amelioration}
-%
-provide evidence of biological implications of sperimidine and
-lipoxygenase in tolerance to salt stress in other plants or even in
-rice cultivars.
-
-As a conclusion, the results presented in this section suggest that
-further studies are needed to elucidate the detailed biological
-function of the remaining 14 genes that have not been named so far in
-the literature.  They may have the potential to intervene in stress
-responsive mechanisms to salt conditions in rice.
-
-\section*{Concluding Remarks}
-\label{sec.concl}
-
-This manuscript provides a detailed description of a network-based
-analysis workflow for the discovery of key genes responding to a
-specific treatment in an organism. It links transcriptomic with
-phenotypic data and identifies overlapping gene modules.
-\vspace{0.5cm}
-
-The proposed approach is inspired by the workflow suggested in the
-WGCNA~\cite{langfelder2008wgcna}. Its main steps are the preprocessing
-of the gene expression data, the construction of a co-expression
-network, the detection of modules within the network, the relation of
-modules with external information (e.g., phenotypic data), and the
-enrichment of the identified key genes with additional information.
-Both approaches are structured in a modular way, which allows
-modifying and exploring different techniques in each step of the
-workflow.
-\vspace{0.5cm}
-
-The proposed workflow is designed to integrate expression data
-measured under two different conditions (namely, control and
-treatment), unlike the usually co-expression-based approaches which
-work with both conditions independently or consider only a single
-condition. For this purpose, an approach similar to that proposed
-in~\cite{du2019network} is used, where the control and treatment data
-are compiled in a single matrix using the Log Fold Change
-measure. Thus, the input to construct the co-expression network is not
-the expression data, but instead the changes in the expression levels
-from one condition to the other, making room for capturing the signal
-of changes caused by the treatment.
-\vspace{0.5cm}
-
-An important feature in the proposed workflow is the module
-detection technique. The co-expression network is computed, as in
-WGCNA, until a scale-free network is obtained. In the proposed
-approach, this network is then used to apply the HLC algorithm, a
-clustering technique capable of detecting overlapping
-communities. Several approaches of module detection from gene
-expression have been proposed and were evaluated
-in~\cite{saelens2018comprehensive}. Most of them focus mainly on
-disjoint (non-overlapping) communities; the techniques described
-dealing with overlaps are not clustering, but bi-clustering and
-decomposition methods. It is well known that communities in real
-networks, including biological ones, are likely
-overlap~\cite{palla2005uncovering}. Thus, the approach presented
-in this work can be seen as a generalization of the previous
-approaches, such as WGCNA, with the potential to deal with genes
-associated to multiple biological processes.
-\vspace{0.5cm}
-
-The approach was applied in a case study with rice under salt
-stress. The results show a group of 14 genes, of which only $2$ of
-them have been previously related to saline stress response in other
-studies. As future work, other overlapping module detection and
-selection techniques should be used instead HLC and LASSO,
-respectively. The combination of these techniques would allow finding
-target genes for future biological studies that evaluate their
-potential as genes that respond to salt stress in rice, and other
-crops and stresses. In-vivo laboratory experimentation needs to be
-conducted to validate the findings of this paper in relation to
-salinity stress.
-\vspace{0.5cm}
-
-Finally, the workflow is presented as a protocol capable of
-considerably reducing the number of genes detected as relevant in the
-response to stress of choice. Other traditionally used methods for
-this purpose tend to generate a large list of candidate genes, thus
-limiting subsequent efforts in experimental validation. In this sense,
-the proposed workflow can help in reducing such efforts in time and
-money invested by researchers in the experimental validation of
-stress-responsive genes.
-
-%\section*{Appendix}
-%Text for this section\ldots
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Backmatter begins here                   %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{backmatter}
-
 \section*{Acknowledgements}%% if any
@@
 \textbf{SHT2:} Spermidine hydroxycinnamoyltransferase 2\\
-\textbf{WGCNA:} Weighted Gene Co-expression Network Analysis\\
+\textbf{WGCNA:} Weighted Gene Co-expression Network Analysis
 
@@
 All authors read and approved the final manuscript.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{biblio}      % Bibliography file (usually '*.bib' )
-% for author-year bibliography (bmc-mathphys or spbasic)
-% a) write to bib file (bmc-mathphys only)
-% @settings{label, options="nameyear"}
-% b) uncomment next line
-%\nocite{label}
-
-% or include bibliography directly:
-% \begin{thebibliography}
-% \bibitem{b1}
-% \end{thebibliography}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%
-%% Do not use \listoffigures as most will included as separate files
-
-%\section*{Figures}
-%
-%%flow chart
-%\begin{figure}[h!]
-%  \centering
-%    \includegraphics[clip,width=1\textwidth]{figures/figure1.pdf}
-%  \caption[The proposed workflow comprising five macro-steps]%
-%  {The proposed workflow comprising five macro-steps:
-%    A.~Data pre-processing, B.~Co-expression network
-%    contruction, C.~Co-expression module identification,
-%    D.~Modules association to phenotypic traits, and
-%    E.~Genes enrichment.}
-%  \label{fig:flow_chart}
-%\end{figure}
-%
-%%scale-free beta 1 y 3
-%\begin{figure}[h!]
-%  \centering
-%    \includegraphics[clip,width=0.8\textwidth]{figures/figure2.png}
-%  \caption{Degree distribution of $A$ (left) and $\hat{A}$ (right)}
-%  \label{fig:beta}
-%\end{figure}
-%
-%%overlapping percentage
-%\begin{figure}[h!]
-%  \centering
-%    \includegraphics[clip,width=0.5\textwidth]{figures/figure3.png}
-%  \caption{Overlapping percentage}
-%  \label{fig:overlap}
-%\end{figure}
-%
-%% boxplots
-%\begin{figure}[h!]
-%  \centering
-%    \includegraphics[clip,width=0.9\textwidth]{figures/figure4.png}
-%  \caption{Phenotipic traits distribution under control and salt stress.}
-%  \label{fig:pdata}
-%\end{figure}
-%
-%\begin{figure}[h!]
-%  \centering
-%    \includegraphics[clip,width=0.9\textwidth]{figures/figure5.pdf}
-%  \caption{Cross-validation of the LASSO regularization parameter
-%    $\lambda$, for each phenotypic trait.}
-%  \label{fig:cross-val}
-%\end{figure}
-%
-%\begin{figure}[h!]
-%  \centering
-%    \includegraphics[clip,width=0.8\textwidth]{figures/figure6.pdf}
-%   \caption[Venn diagram for the case study in rice]%
-%   {Venn diagram representing the number of genes selected at
-%    different stages of the proposed workflow for the case study in
-%    rice.}
-%  \label{fig:final_genes}
-%\end{figure}
-%
-%% Figure 3d protein strucures
-%\begin{figure}[hptb]
-%  \centering
-%    \includegraphics[clip,width=0.8\textwidth]{figures/figure7.pdf}
-%  \caption{3D protein structure of named genes selected by LASSO, available from~\cite{szklarczyk2016string}.}
-%  \label{fig:3d}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Tables                        %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Use of \listoftables is discouraged.
-%%
-%\section*{Tables}
-%\begin{table}[h!]
-%\caption{Sample table title. This is where the description of the table should go}
-%  \begin{tabular}{cccc}
-%    \hline
-%    & B1  &B2   & B3\\ \hline
-%    A1 & 0.1 & 0.2 & 0.3\\
-%    A2 & ... & ..  & .\\
-%    A3 & ..  & .   & .\\ \hline
-%  \end{tabular}
-%\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\section*{Additional Files}
-%  \subsection*{Additional file 1 --- Sample additional file title}
-%    Additional file descriptions text (including details of how to
-%    view the file, if it is in a non-standard format or the file extension).  This might
-%    refer to a multi-page table or a figure.
-%
-%  \subsection*{Additional file 2 --- Sample additional file title}
-%    Additional file descriptions text.
-
-\end{backmatter}
-\end{document}

### 2020-12-07 camila/report/framework.tex
bmc version

--- 2020-10-08 camila/report/framework.tex
@@
-\section{The Workflow}
+\section*{The Workflow}
 \label{sec.framework}
 
-The proposed workflow uses RNA-seq read counts, representing gene
-expression levels, as input data. More precisely, it uses $n_0$ gene
-expression profiles of an organism, measured for $m$ different
-genotypes under control and treatment conditions, and $r$ biological
-replicates. This raw data is represented as a matrix $D_0 \in
-{\mathbb{N}_0}^{n_0 \times 2mr}$. In order to discover key genes and
-their interaction with phenotypes related to treatment tolerance, the
-approach also requires a set of $p$ phenotypic traits, measured for
-the $m$ genotypes. The phenotypic data is seen as a matrix $P \in
-\mathbb{R}^{2m \times p}$ containing two phenotypic values per
-genotype, one under control condition and a second one under treatment
-condition. The proposed workflow is depicted in
-Figure~\ref{fig:flow_chart}. This section explains the five
-macro-processes (A)-(E) of the proposed workflow. In comparison with
-WGCNA, it adds the macro-step (D) and generalizes macro-steps (A)-(C).
+Figure~\ref{fig:flow_chart} introduces the proposed workflow.
+It can be broken down into five
+macro-processes (A)-(E). Compared to
+WGCNA, the workflow adds the macro-step (D) and generalizes macro-steps (A)-(C).
+\vspace{0.5cm}
 
-% flow chart
-\begin{figure}[htpb]
+% figure 1
+\begin{figure}[htbp]
   \centering
-    \includegraphics[clip,width=1\textwidth]{figures/flow_chart_2.pdf}
-  \caption{The proposed workflow comprising five macro-steps:
+    \includegraphics[clip,width=0.96\textwidth]{figures/figure1.pdf}
+  \caption[The proposed workflow is broken down into five macro-steps]%
+  {The proposed workflow is broken down into five macro-steps:
     A.~Data pre-processing, B.~Co-expression network
-    contruction, C.~Co-expression module identification,
-    D.~Modules association to phenotypic traits, and
-    E.~Genes enrichment.}
+    contruction, C.~Identification of co-expression modules,
+    D.~Detection of modules association to phenotypic traits, and
+    E.~Gene enrichment.}
   \label{fig:flow_chart}
@@
 
-\subsection{Data pre-processing}
-The goal of the data pre-processing stage is to build matrices $P_ell$ and $L_1$ representing, respectively, the changes in phenotypic values and expression levels between control and treatment condition, from RNA-seq and phenotypic data found in matrices $D_0$ and $P$, respectively.
+The input of the workflow includes RNA-seq read counts, representing gene expression
+levels. More precisely, the workflow uses $n_0$ gene expression
+profiles, measured for $m$ different genotypes 
+of $r$ biological replicates (under
+control and treatment conditions). This
+data is represented as matrix $D_0 \in {\mathbb{N}_0}^{n_0
+  \times 2mr}$. To discover key genes and their interaction
+with phenotypes related to treatment, the approach also
+requires a set of $p$ phenotypic traits, measured for $m$
+genotypes. The phenotypic data is captured by matrix $P \in
+\mathbb{R}^{2m \times p}$ which contains two phenotypic values per
+genotype (under control and treatment
+conditions).
 
-The RNA-seq data cannot be directly interpreted.
-Therefore, a normalization process is applied to deal with the problem of possible biases affecting the quantification of results.
-The suggested normalization technique for correcting library size and RNA composition bias is DESeq2 \cite{love2014moderated}.
-The normalized data is represented as a matrix $D_1 \in \mathbb{R}^{n_0 \times 2mr}$, and the biological replicates of each genotype are averaged and represented as a matrix $D_2 \in \mathbb{R}^{n_0 \times 2m}$.
-The genes exhibiting low variance or low expression are removed from $D_2$, thus identifying a subset of size $n_1 \leq n_0$ of the original genes.
-The control and treatment data is separated into the matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1 \times m}$, respectively.
-The matrix entries $c_{ij}$ in $C$ and $t_{ij}$ in $T$ represent, respectively, the normalized expression level of gene $i$ in accession $j$.
-Control and treatment data is also separated from phenotypic data $P$, obtaining the $P_c$ and $P_t$ matrices of dimensions $m \times p$.
+\subsection*{A. Data Pre-processing}
 
-In the above configuration, the changes in expression levels and phenotypic values, between control and treatment conditions, are measured in terms of logarithmic ratios.
-In the case of expression levels, the log ratios are represented in the Log Fold Change matrix $L_0 \in \mathbb{R}^{n_1 \times m}$, where $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$.
-Similarly, the log ratios of the phenotypic data are computed and represented in the $P_\ell \in \mathbb{R}^{m \times p}$ matrix.
+The goal of the data pre-processing stage is to build matrices
+$P_\ell$ and $L_1$ which represent, respectively, the changes in
+phenotypic values and expression levels between control and treatment
+condition. In other words, $P_\ell$ and $L_1$ are constructed from RNA-seq and phenotypic data found in matrices $D_0$
+and $P$.
+\vspace{0.5cm}
 
-The final step of the data pre-processing is to filter $L_0$ by removing rows (e.g., genes) with low variance in the differential expression patterns, obtaining a new matrix $L_1$ of dimensions $n_2 \times m$, with $n_2 \leq n_1$.
+A normalization process is applied to interpret RNA-seq data and handle possible
+biases affecting the quantification of results. Here,  DESeq2~\cite{love2014moderated}
+is used to correct for library size and RNA
+composition bias. The normalized
+data is represented as a matrix $D_1 \in \mathbb{R}^{n_0 \times 2mr}$,
+and the biological replicates of each genotype are averaged and
+represented as a matrix $D_2 \in \mathbb{R}^{n_0 \times 2m}$. The
+genes exhibiting low variance or low expression are removed from
+$D_2$. Consequently, this stage of the approach reduces
+the set of genes from a pool of size $n_0$ to $n_1$.
+The control and treatment data is separated into the
+matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1
+  \times m}$, respectively. The matrix entries $c_{ij}$ in $C$ and
+$t_{ij}$ in $T$ represent the normalized expression
+level of gene $i$ in accession $j$ under control and treatment 
+condition, respectively.
+Control and treatment data is also separated from
+phenotypic data $P$, obtaining the $P_c$ and $P_t$ matrices of
+dimensions $m \times p$.
+\vspace{0.5cm}
 
-\subsection{Co-expression network construction}
+In the above configuration, the changes in expression levels and
+phenotypic values between control and treatment conditions are
+measured in terms of logarithmic ratios. In the case of expression
+levels, the log ratios are represented in the Log Fold Change matrix
+$L_0 \in \mathbb{R}^{n_1 \times m}$, where $\ell_{ij}=\log_2
+(t_{ij}/c_{ij})$. Similarly, the log ratios of the phenotypic data are
+computed and represented in the $P_\ell \in \mathbb{R}^{m \times p}$
+matrix.
+\vspace{0.5cm}
 
-A gene co-expression network connect genes with similar expression patterns across biological conditions.
-The purpose of this stage is to describe how to build the co-expression network $A$ from the Log Fold Change matrix $L_1$, capturing the relationship between genes according to the change in expression levels between the two studied conditions.
-These co-expression patterns are meaningful for the identification of genes not yet associated with the response to the treatment condition.
+The final stage of pre-processing is to filter $L_0$ by
+removing rows (e.g., genes) with low variance in the differential
+expression patterns, obtaining a new matrix $L_1$ of dimensions $n_2
+\times m$, with $n_2 \leq n_1$.
 
-The Log Fold Change matrix $L_1$ is used to build the co-expression network following the first two steps of the WGCNA methodology~\cite{langfelder2008wgcna}.
-First, the level of concordance between gene differential expression profiles across samples is measured.
-To this end, the absolute value of the Pearson correlation coefficient is used as the similarity measure between genes and the resulting values are stored in the similarity matrix $S\in \mathbb{R_{+}}^{n_2 \times n_2}$.
-Second, the matrix $S$ is transformed into and adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes.
-In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network.
-These networks contain many nodes with very few connections and a small number of hubs with high connections.
-In a strict scale-free network the logarithm of $P(k)$ (i.e., the probability of a node having degree $k$) is approximately inversely proportional to the logarithm of $k$ (i.e., the degree of a node).
-So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g., $R^2 > 0.85$).
+\subsection*{B. Construction of the Co-expression Network}
 
-\subsection{Co-expression module identification}
+A gene co-expression network connects genes with similar expression
+patterns across biological conditions. The purpose of this step is to
+describe how to build the co-expression network $A$ from the Log Fold
+Change matrix $L_1$, capturing the relationship between genes
+according to the change in expression levels between the two studied
+conditions. These co-expression patterns are meaningful for the
+identification of genes that are not yet associated to treatment response.
+\vspace{0.5cm}
 
-Next step in the workflow is to identifying communities (also called modules)
-from the co-expression network structure and dynamics represented in $A$. 
-The idea is to cluster genes with similar patterns of differential
-expression change. Membership in these modules may overlap in 
-biological contexts, because modules may be related to specific 
-molecular, cellular or tissue functions and the biological components
-(i.e.
-genes) may be involved in multiple functions.
-Thus, unlike WGCNA, the adjacency matrix $A$ is used to detect overlapping (rather than non-overlapping) communities, using the Hierarchical Link Clustering (HLC) algorithm (see section~\ref{sec:HLC}).
+The Log Fold Change matrix $L_1$ is used to build the co-expression
+network following the first two steps of WGCNA~\cite{langfelder2008wgcna}.
+First, the level of
+concordance between gene differential expression profiles across
+samples is measured. To this end, the absolute value of the Pearson
+correlation coefficient is used as the similarity measure between
+genes and the resulting values are stored in the similarity matrix
+$S\in \mathbb{R_{+}}^{n_2 \times n_2}$. Second, the matrix $S$ is
+transformed into an adjacency matrix $A \in \mathbb{R_+}^{n_2\times
+  n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the
+connection strength between each pair of genes. In other words, the
+elements of the adjacency matrix are the similarity values up to the
+power $\beta > 1$ so that the degree distribution will fit a
+scale-free network. These networks contain many nodes with very few
+connections and a small number of hubs with high connections. In a
+strict scale-free network, the logarithm of $P(k)$ (i.e., the
+probability of a node having degree $k$) is approximately inversely
+proportional to the logarithm of $k$ (i.e., the degree of a node).
+The parameter $\beta$ is chosen to be the smallest value for which the
+$R^2$ of the linear regression between $log_{10}(p(k))$ and
+$log_{10}(k)$ is closest to $1$ (Here, $R^2 > 0.8$).
 
-As a prelimiary step, matrix $A$ is transformed, into an unweighted network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$ before applying the clustering algorithm.
-To this end, the Pearson Correlation Coefficient (PCC) cutoff is determined using the approach described in~\cite{aoki2007approaches}.
-The number of nodes, edges, and the network density is determined for different PCC cutoffs.
-Around the most biological relevant PCC cutoff, the number of nodes presents a linear decrease, and the density of the network reaches its minimum, while below this value the number of edges rapidly increases.
-Considering this, a cutoff is selected such that gene pairs which have a correlation score higher than the threshold are considered to have significant co-expression relationship.
-Above the cutoff, the entries of matrix $A$ become $1$ and below the cutoff $A$ values becomes $0$.
-The application of the HLC algorithm organizes the $n_2$ genes of matrix $\hat{A}$ into $c$ modules, where each gene can belong to one, multiple, or no module at all.
-This information is represented as an affiliation matrix $F \in \{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ if node $i$ is member of module $u$.
+\subsection*{C. Identification of Co-expression Modules}
 
-\subsection{Module association to phenotypic traits}
+The next step in the workflow is to identifying modules of overlapping communities
+from the co-expression network
+represented by $A$.  The idea is to cluster genes with similar
+patterns of differential expression change. Membership in these
+modules may overlap in biological contexts, because modules may be
+related to specific molecular, cellular, or tissue functions, and the
+biological components (i.e., genes) may be involved in multiple
+functions. Unlike WGCNA, the workflow applies 
+the Hierarchical Link Clustering (HLC) algorithm (overviewed in the 
+\hyperref[sec.prelim]{Preliminaries} section) to
+detect overlapping rather than non-overlapping communities.
+%Section~\ref{sec.prelim}).
+\vspace{0.5cm}
 
-To identify the most relevant groups (modules) of genes, associated
-with the phenotypic response to a specific treatment in an organism,
-the proposed workflow uses a LASSO based approach. Each module is
+First, the adjacency matrix $A$ is transformed into an unweighted
+network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$.
+To this end, the Pearson Correlation Coefficient
+(PCC) cutoff is determined using the approach described
+in~\cite{aoki2007approaches}. The number of nodes, edges, and the
+network density is determined for different PCC cutoffs. 
+In a neighborhood of the optimal PCC cutoff
+the number of nodes presents a linear
+decrease and the density of the network reaches its minimum, while
+below this value the number of edges rapidly increases. Following this
+observation, a cutoff is selected such that gene pairs which have a
+correlation score higher than the threshold are considered to have
+a significant level of co-expression. Above the cutoff, the entries of
+matrix $A$ become $1$ and below the cutoff $A$ values become $0$. The
+HLC algorithm organizes the $n_2$ genes of matrix $\hat{A}$ into $c$~modules, where each gene can belong to zero or multiple modules.
+This information is represented as an affiliation matrix $F \in
+\{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ if node $i$ is a member of
+module $u$ (and $f_{iu}=0$ otherwise).
+
+\subsection*{D. Detection of Modules Association to Phenotypic Traits}
+
+To identify the most relevant modules, associated
+with the phenotypic response to a specific treatment,
+the proposed workflow uses LASSO. Specifically, each module is
 represented by an eigengene, which is defined as the first principal
-component of such module. An eigengene can be thought of as an average
+component of such module. An eigengene can be seen as an average
 differential expression profile for each community: it is computed
 from the Log Fold Change Matrix $L_1$ and the affiliation matrix
-$F$. Given a module $u$, the affiliation matrix is used to identify
+$F$. Given a module $u$, the affiliation matrix $F$ is used to identify
 the genes belonging to $u$ and then the corresponding rows of the
@@
 $u$. Each principal component becomes a column of the matrix $M \in
-\mathbb{R}^{m \times c}$.
-These profiles are associated with each phenotypic trait using 
-the least absolute shrinkage and selection operator (LASSO). 
-In this context, the eigengenes (i.e., the columns of $M$) act as regressor variables
-and each phenotypic trait (i.e., each column of $P_\ell$) is used as an outcome variable.
+\mathbb{R}^{m \times c}$.  These profiles are associated with each
+phenotypic trait using
+LASSO.  In this context, the eigengenes (i.e., the columns
+of $M$) act as regressor variables and each phenotypic trait (i.e.,
+each column of $P_\ell$) is used as an outcome variable.
+\vspace{0.5cm}
 
 The output after applying LASSO is a set $W_z$ of modules for each
-phenotypic trait $z$, where $W_z \subset \{u \mid 1 \leq u \leq c\}$ for
-$z= 1,2,..,p$. The target genes $I$ for downstream analysis, which may
-be important in the treatment response, are the union of genes belonging
-to the selected modules, that is $I = \cup_{z=1}^{p} W_z$, where
-$I \subset \{i \mid 1 \leq i \leq n_2\}$.
+phenotypic trait $z$, where $W_z \subset \{u \mid 1 \leq u \leq c\}$
+for $z= 1,2,..,p$. The target genes $I$ for downstream analysis
+are the union of genes
+belonging to the selected modules; that is $I = \cup_{z=1}^{p} W_z$,
+where $I \subset \{i \mid 1 \leq i \leq n_2\}$.
 
-\subsection{Gene Enrichment}
+\subsection*{E. Gene Enrichment}
 
-The goal of this final stage of the process is to characterize the
-genes identified in previous stage with additional information, helping to elucidate their
-possible behavior and role in the response to the studied treatment.
+The goal of this final step of the process is to annotate with
+additional information the genes identified in previous stages,
+helping to elucidate their possible behavior and role in the response
+to the studied treatment.
+\vspace{0.5cm}
 
-A simple analysis to made with the selected genes is to identify the
-differentially expressed genes in set $I$. That is, to select those genes
-in $I$ having an absolute value of the log fold change of at least $2$
-($|\ell_{ij}|\geq 2$) for at least one sample. This
-represents genes level of expression is quadrupled (up or down)
-from control to treatment condition; these genes are
-strong candidates for treatment responsive genes.
+A crucial step is to identify the differentially expressed genes in
+set $I$. That is, to select the genes in $I$ that have an absolute
+value of the Log Fold Change of at least $2$ ($|\ell_{ij}|\geq 2$) for
+at least one sample. This represents genes whose expression level is
+quadrupled (up or down) from the control to treatment condition; they are
+target genes.
+\vspace{0.5cm}
 
-Also functional category enrichment can be done by, e.g., searching for gene ontology annotations in databases such as QuickGO~\cite{binns2009quickgo}. Such annotations 
-can provide evidence of biological implications of the target genes in
-the treatment-tolerance mechanisms. Furthermore, QuickG0 can be used to identify genes with reported protein products, which can be used to perform additional relevant
-analysis reviewing their reported protein-protein interactions in other databases, such as STRING~\cite{szklarczyk2016string}.
-The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge
-transfer between organisms, and from interactions aggregated from
-other (primary) databases. This information can give new insight on how the
-selected genes are involved in functional pathways that can be related
-with the treatment of interest.
+Furthermore, functional category enrichment can be carried out by, e.g., searching
+for gene ontology annotations in databases such as
+QuickGO~\cite{binns2009quickgo}. Such annotations can provide evidence
+of biological implications of the target genes in the
+treatment-tolerance mechanisms. Furthermore, QuickGO can be used to
+identify the protein products of genes, which can be used to
+perform additional analysis that provides new insights into how target genes are
+involved in functional pathways that can be related to treatment.
+Such analysis includes a review of reported
+protein-protein interactions in databases such as
+STRING~\cite{szklarczyk2016string}. The protein interactions include direct
+(physical) and indirect (functional) associations. They stem from
+computational prediction, knowledge transfer between organisms, and
+interactions aggregated from other (primary) databases. The search for unknown interactions would extend the workflow with additional steps.

### 2020-12-07 nicolas/PPIPL3/ReportV5.tex

%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{verbatim}
\usepackage{float}
\usepackage{graphicx}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{multirow}
%\usepackage{table}
%\usepackage{tabular}

\makeatother

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
\author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
\maketitle
\begin{abstract}

Efforts to predict unmapped protein-protein interactions (PPI) through
network analysis suggest that the higher the number of paths of length
3 (L3) between two proteins, the higher the probability of interaction.
This paper extends previous work on predicting protein interactions
based on the L3 principle by taking into account the learning of node
features of the PPI network. In particular, we train an \texttt{XGBoost}
model using handcrafted features (including L3), as well as edge embeddings
from \texttt{Node2Vec}. Our main result shows that while L3 is an
important feature for predicting links, best performance is achieved
when combined with edge embeddings. The proposed approach is evaluated
for the human and rice interactomes and pretends to exploit the intrinsic
network structure. Results for both organisms show higher AUC values
for the combination of automated and handcrafted features than for
each of its components alone. The handcrafted features have more importance
for the prediction model in the human interactome than in rice. The
developed framework can be easily applied to interactomes for other
organisms or even to networks outside of the biological context.
\end{abstract}
\begin{comment}
need include a feature that compares the dimension of the embedding
is useful for prediction with L3 to show that they independent measures
- we do not want to be learning L3 through node2vec
\end{comment}

\section{Introduction}

\begin{comment}
Need to add other relevant references
\end{comment}

Proteins are key actors of biological processes inside cells. As they
carry out a variety of tasks, proteins do not work as single agents
but rather as a part of dynamic networks of protein-protein interactions
(PPI) \cite{Lin2017}. PPI networks play an important role in a variety
of interdependent mechanisms, including signal transduction, homeostasis
control, stress responses, plant defense and organ formation. They
are also involved a variety of physiological and developmental processes,
including protein phosphorylation, transcriptional co-factor recruitment
and transporter activation \cite{Zhang_2010_PPI}.

Different methods for predicting interactions in PPI networks have
been proposed in recent years \cite{Chen2019PPI,Chang2016PPI,Kotlyar2015PPI}.
Kovacs et al. (2019) introduce an approach which relies on counting
the number of paths of length 3 (L3) to predict interactions among
proteins for different organisms. The proposed approach outperforms
previous methods in PPI networks of yeast (\emph{S. cerevisiae}),
Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D.
melanogaster}), fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus})
and the human interactome \cite{Kovacs2019}.

A common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid}
technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on (i) the expression of a specific reporter gene that
activates by the binding of a DNA-binding Domain (DB), and (ii) an
Activation Domain (AD) of a Transcription Factor that binds to an
Upstream Activation Sequence (UAS). For the Y2H technique, a protein
is fused to the DB domain (known as \emph{bait}) and another protein
to the AD (known as \emph{prey}). If the proteins do not interact,
then the reporter gene is not expressed. Otherwise, the reporter gene
expression is activated by the activation domain.

\begin{figure}[h]
\caption{Yeast-2-Hybrid Technique}

\noindent \centering{}\includegraphics[width=0.7\columnwidth]{Y2H}
\end{figure}

Based on the outcome of numerous Y2H experiments, PPI networks which
capture known interactions for each protein can be constructed. Several
algorithms are proposed over these networks to predict unknown interactions.
The focus of this study is to evaluate different methods for predicting
PPIs using the network of known interactions. These interactions are
represented as an undirected graph. A common way of predicting interactions
is usually based on social networks analysis, more specifically on
the triadic closure principle (TCP). TCP states that the higher the
number of common neighboring nodes between two nodes, the higher the
probability that they interact \cite{Goldberg2003SmallWorld}. Previous
studies show that the mentioned approach fails because it does not
consider the structural and chemical properties of the proteins \cite{Kovacs2019,Cannistraci2013Networks}.

\begin{comment}
JF: Yet you do more than evaluating this three measures so this has
to be updated.

NL: Done!

JF: Not sure we should use L3 above since it is being redefined as
a normalized measure here!

JF{*}: state what the results are: ``The contributions of this paper
are...''

Which article predicts PPI based on TCP? add refs
\end{comment}

For achieving the described results, human and rice PPI networks are
compared using state-of-the-art methods, as well as the proposed ones
(A2, A3, L3). In the case of the human network, two different versions
of the human interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
are used for training and another interactome version is used for
validation (\emph{HI-III})\cite{Kovacs2019}. For the rice interactome,
a portion of the known interactions removed and then predicted with
different techniques based on the known interactions. The main contributions
of this paper are i) a general framework for link prediction in non-directed
networks and ii) two applications of this framework to biological
networks with a structural insight

\begin{comment}
JF: read throughout the document but edited up here. I prefer to read
the results before editing the materials and methods section. 
\end{comment}

The remainder of this paper is organized as follows. Section 2 describes
the methodological steps and key milestones in the preparation of
the networks, model parameters and experimental configurations. Section
3 presents the main results of the models for the human and rice interactomes.
Section 4 presents the main conclusions of this study. Finally, Appendix
presents supplementary information and figures which complement the
results described in this paper. Figures in the Appendix can be identified
with a prefix (e.g. S\ref{F2}).

\section{Materials and Methods}

\subsection{Data Availability}

The human interactome data was downloaded from the repository of the
length-3 degree normalized paths methodology \cite{Kovacs2019}: the
dataset \emph{HI-II-14} and \emph{HI-TESTED} are used for prediction.
The dataset \emph{HI-III} is available in the same repository and
is used for validation.

The rice interactome information was downloaded from the STRING database
\cite{Szklarczyk2019}, corresponding to the subspecies \emph{Oryza
sativa}. The file contains more than 8 million PPIs from several sources.
For the purpose of this study, only PPIs with evidence from curated
databases are used. The resulting network contains $5025$ nodes and
$164420$ edges.

\subsection{Code Implementation}

The original code implementation in \texttt{C++} from Kovacs et al
(2019) was adapted to \texttt{Python} (V3.6). For the purpose of algorithmic
validation, the three methods were implemented from scratch with basic
functionalities and data structures of the \texttt{Python} language.
The code is publicly available under the repository github.com/ocinlr/PPIPL3.

\subsection{Data Preprocessing}

Information for the human interactome was used as-is, which corresponds
to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
$9433$ and $23322$ interactions, respectively.

For the rice interactome, an additional preprocessing was performed.
The filtered network for rice consists of $5025$ proteins (nodes)
and $164420$ interactions (edges) distributed among $178$ connected
components. The connected component with the greatest number of edges
was selected. The extracted connected component consists of $n=4390$
nodes and $m=163319$ edges, which corresponds to $99.33$\% of edges.

\subsection{Edge Prediction}

For predicting interactions for each network, the algorithms described
below were used. It is important to keep in mind how the PPI network
$G=(V,E)$ is conceptualized: each node ($v_{i}\in V$) represents
a protein and each undirected edge ($e_{b}=\{v_{i},v_{j}\},\,e_{b}\in E$)
represents an interaction among proteins $v_{i}$ and $v_{j}$.
\begin{description}
\item [{Common~Neighbors~(A2)}] This method is based on the Triadic Closure
Principle which states that the more common friends two individuals
have, the more likely that they know each other. For implementing
this measure, $A{{}^2}$ matrix is calculated, where $A$ is the adjacency
matrix of the network.
\item [{Length-3~Path~Counts~(A3)}] This is a simple implementation
of the notion of ``if my friends and your friends interact, then
we might interact too''. For implementing this measure, $A{{}^3}$
is calculated.
\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
overestimate the importance of some edges due to intermediate hubs
which add many shortcuts in the graph. To address that issue, a degree
normalization for the path $v_{i}\rightarrow v_{m}\rightarrow v_{n}\rightarrow v_{j}$
is applied by considering the degree of the intermediate nodes $k_{m}$
and $k_{n}$, as follows.
\[
p_{i,j}=\sum_{m,n\in|V|}\frac{A_{i,m}\cdot A_{m,n}\cdot A_{n,j}}{\sqrt{k_{m}\cdot k_{n}}}
\]
\\
where $A_{i,j}$ represents the value of the adjacency matrix for
nodes $i$ and $j$.
\end{description}

\subsection{Sampling Procedure}

For each PPI network, the following procedure was performed 10 times
to address the stochastic nature of the process:
\begin{enumerate}
\item A percentage of interactions is removed at random from the network
(20\%).
\item The same amount of removed interactions are then predicted using the
main methods for prediction mentioned by Kovacs et al (2019): Common
Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
and the Length-3 degree-normalized score (\textbf{L3}).
\item A test dataset is created as follows: all removed edges are included
(as observed positives for the ML algorithm) and from the predicted
edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
the previous classification (observed negatives), a random subset
is chosen such that the dataset is balanced, that is, the amount of
observed positive labels is equal to the observed negative labels.
\item Once the dataset is ready, it is randomly partitioned: 80\% is used
for \texttt{XGBoost} model training and 20\% is used for validation.
It is important to have in mind that balanced distribution of the
positive and negative labels in the datasets was satisfied in training,
but validation is performed using the remaining unbalanced data.
\end{enumerate}
It is worth mentioning that some exploration experiments were carried
out with interactions removal percentages of 2, 5, 10 and 20\%, as
well as train-test partitions of 15-85, 80-20 and 75-25, and the explained
parameter set was selected because it either represented a marginal
gain (results not shown) or because it was a common parameter selection
in related literature.

\subsection{Feature Extraction with \texttt{Node2Vec}}

The \texttt{Node2Vec} module was used for extracting features of the
interactomes. The considerations for the model were:
\begin{itemize}
\item All paths in the random walks are equally likely.
\item Use a modest number of dimensions and threads for calculation.
\item Since length-3 paths are the defining property in this study, there
is no necessity for longer walks. However, it is important to try
out many possible redundant routes and to consider a window of at
least 4.
\item Other standard parameters were left with default values.
\item Edge embeddings were calculated using a geometric ratio of the node
embeddings.
\end{itemize}

\subsection{Handcrafted Features}

Due to the poor results of the \emph{raw} Length-3 counting (\textbf{A3}),
a different approach for this information was carried out: As it still
gives a lot of information that might be useful for a predictive routine,
this counting was normalized (dividing by the greatest counting in
the \textbf{A3} top predictions) and then used as a feature for the
Machine Learning algorithm. For completeness, \textbf{A2} and \textbf{L3}
measures were used as a input features. Finally, the case were no
handcrafted feature was also considered, that is, only the features
extracted from the structure of the network were considered.

\subsection{Feature to Predict: Existence}

The feature to predict corresponds to the possible existence (\emph{True/False})
of a link based on the existing information of the network. This is
evaluated by taking out a fraction of the edges and then trying to
predict for a given set of possible edges if they have a high probability
to belong to the original network.

\subsection{Machine Learning Algorithm}

The Extreme Gradient Boosting implementation of gradient boosted trees
is applied to evaluate the existence of an edge. Gradient boosted
trees are usually used for supervised learning problems, where the
training data $X_{i}$ has multiple features and pretends to explain
(or predict) a target variable $Y_{i}$. The corresponding implementation
applied for this study is \texttt{XGBoost}.

\subsection{Result Validation}

As mentioned before, 80\% of the final dataset was randomly selected
and used for training, while the remaining information was used for
validation. The whole training-validation procedure was applied 10
times.

The chosen metric for validation was the Area under the Curve (\textbf{AUC})
of the Receiver Operating Characteristic (\textbf{ROC}). This curve
corresponds to plot the sensitivity (probability of predicting a real
positive as positive) against 1-specificity (probability of predicting
a real negative as positive). AUC values of 1 represent a perfect
prediction and 0.5 corresponds to a random guess. Normally, values
over 0.8 of AUC are considered good.

\section{Results and Discussion}

\subsection{Human Interactome}

As the first step in the evaluation of A2, A3 and L3 methods for link
prediction, the full calculation and ranking of all predictions was
carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
and \emph{HI-TESTED}) is shown in Figure \ref{fig:HI1}, in which
the x axis represents a rank position $k$ and the y axis displays
the precision for the top $k$ predictions of each method, when assuming
interactome \emph{HI-III} as the validation set.

\begin{figure}[h]
\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14} and \emph{HI-TESTED}}

\noindent \centering{}\includegraphics[width=0.48\columnwidth]{hi-ii-14\lyxdot txt}\includegraphics[width=0.48\columnwidth]{hi-tested\lyxdot txt}
\end{figure}

As it can be inferred from the plots, L3-based predictions outperform
their $A{{}^2}$ counterparts. Results also show that L3-score and
$A^{3}$predictions follow a very similar trend.

However, it is of our interest to assess if Machine Learning can boost
the overall performance on the prediction, so the proposed procedure
was adapted to the human interactome data: Instead of having to remove
a fraction of edges from the network in order to predict and validate,
the prediction network and the validation networks are given beforehand.
The rest of the procedure is carried out as mentioned in Section 2.
Table \ref{T1} presents the results for the combinations of Node2Vec
with each feature.

\begin{table}
\caption{\label{T1}Summary statistics for human interactomes}
\includegraphics[width=1\columnwidth]{T1}
\end{table}

In general, performance of all models is above 0.96 in terms of F1-Score.
Although marginally, the A2 feature-enriched model for HI-TESTED performs
better than their counterparts based on paths of length 3. The opposite
is also true for HI-II-14: models enriched with L3 and A3 features
perform better than their A2 counterpart. Results for the Interactome
\emph{HI-II-14} with Node2Vec and each metric are presented in Figures
S\ref{HI1}, S\ref{HI2} and S\ref{HI3}. It can be seen that for
\emph{HI-II-14}, A2 performs marginally better than L3 when comparing
precision but worse in terms of recall, although all metrics have
AUC values of 0.99.

On the other hand, the interactome \emph{HI-TESTED} was assessed with
the same methodology as \emph{HI-II-14} and results are shown in Figures
S\ref{HI1-1}, S\ref{HI2-1} and S\ref{HI3-1}. For this interactome,
A2 performs better in terms of false positives than A3 and L3, but
in terms of false negatives, A3 actually performs better that A2 and
L3. In terms of AUC, metrics results are indistinguishable.

An interesting assessment from this human datasets can be observed
when looking at the importance plots, which show a very biased influence
of the handcrafted features. Figure \ref{F8-importance-H1} shows
this behavior for the \emph{HI-II-14} dataset, which is very similar
to \emph{HI-TESTED} (Figure S\ref{F8-importance-H2}).

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F8-importance-H1}Importance gain plots for \emph{HI-II-14}}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A2\lyxdot 1}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A3\lyxdot 1}
\par\end{centering}
\centering{}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot L3\lyxdot 1}
\end{figure}

\subsection{Rice Interactome}

For the rice interactome, different combinations of features were
trained on the model and validated. The results are shown in Table
\ref{T2}. First, prediction based on \texttt{Node2Vec} without any
additional feature included is used as baseline of comparison. Figure
\ref{F1} shows those results, and one can see that its mean performance
using the AUC is $0.93$. The results for all 10 repetitions of each
model are consistent, meaning the variance among each line is not
significant. The results for each model using only the default features
are satisfactory. This result can also be assessed when looking at
the confusion matrix, where a precision of $0.98$ and a recall of
$0.86$ are obtained.

\begin{table}
\caption{\label{T2}Summary statistics for rice interactome}
\includegraphics[width=1\columnwidth]{T2}
\end{table}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F1}Summary ROC curves for \texttt{Node2Vec} model alone}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_ML/ROConlyML_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_ML/CMonlyML_SUMMARY}
\end{figure}

The next experiment carried out was to add for each prediction method
score as a prediction feature, that is, use the score as an additional
feature for the ML algorithm. Results for the Common Neighbors method
(\textbf{A2}), which uses paths of length 2, are shown in Figure S\ref{F2}.
All 10 experiments have small variability among them and perform better
than the baseline, with AUC of $0.98$. When looking at the confusion
matrix for this model, most guesses are true positives and true negatives,
resulting in a precision of $0.99$ and a recall of $0.91$. The same
analyses are done with the count of paths of length 3 (\textbf{A3})
and with the degree-normalized length-3 score (\textbf{L3}). These
results are presented in figures S\ref{F3} and \ref{F4}, respectively.
The corresponding AUC are $0.97$ and $0.98$.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F4}Results for \texttt{Node2Vec} with L3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CML3_SUMMARY}
\end{figure}

The precision when using the count of paths of length 3 (A3) is $0.99$
while the recall is $0.90$. It is interesting to observe that precision
decreased $0.16\%$ and recall decreased $0.58\%$. A similar situation
is obtained for the model with the normalized score (L3), whose precision
decreased to $0.99$ ($-0.08\%$) and recall decreased to $0.88$
($-2.38\%$). It can be seen also that the ROC curve achieves high
values of true positive rate sooner in the A2-featured model than
on the other models.

Another relevant assessment was carried out to verify the predictive
power of the metrics. \texttt{XGBoost} models were trained only using
each metric, for the same 10 experiments. Results are shown in Figures
S\ref{F5}, S\ref{F6} and \ref{F7}. XGBoost yields lower AUC values
when only considering A2, A3 and L3. Precision and recall for this
assessments, although lower than the previous experiments, are still
considered satisfactory.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F7}Results for L3 feature alone}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyL3_SUMMARY}
\end{figure}

Finally, an relevant evaluation on the models that include a handcrafted
feature is necessary: How important is the appended feature for the
model result? The answer can be observed by the feature importance
plots for A2, A3 and L3 in Figure \ref{F8-importance}. Each plot
resembles the importance gain from each feature in the underlying
decision trees that \texttt{XGBoost} uses. Usually, the more bifurcations,
the higher gain a feature has, the more relevance a feature has on
the model prediction itself. All models have the appended feature
as the most important by a good margin, although margin for A3 is
not wide as for A2 or L3.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F8-importance}Importance gain plots for \texttt{Node2Vec}
with each feature}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
\par\end{centering}
\centering{}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
\end{figure}

\section{Conclusions}

Based on previous work from other authors\cite{Kovacs2019}, the intrinsic
knowledge of existing protein-protein interaction networks let scientists
elucidate promising new interactions to explore. In this study, a
machine learning algorithm was fed with previously elucidated information
about the network, as well as results form a state-of-the-art edge
embedding technique. Even if the baseline of the handcrafted features
(A2, A3, L3) yields good results, boosting that information with the
vector embedding, which calculates vector representations of the network
structure improves the overall performance of each method on its own.

On both human interactomes (\emph{HI-II-14} and \emph{HI-TESTED}),
applying \texttt{XGBoost} with the metrics yield results with high
AUC for the link prediction problem, and therefore this study extends
previous work by other authors that only consider the predictive power
of the metrics alone. In this case, the importance of the handcrafted
features was very significant and stands out of the Node2Vec features.

On the rice interactome, PPI prediction obtained high AUC values when
only using the node embedding technique (AUC=0.93) and when only using
the handcrafted features based on protein neighborhood (AUC \textasciitilde 0.88).
Even higher AUC results were achieved when the handcrafted features
and the vector representation information were combined (AUC \textasciitilde 0.98).

Importance values for the A3 method stand out less from the node embedding
information than in A2 and L3. In this case, the importance gain of
the handcrafted features in each model was not as high in rice as
in the human interactome.

With this study, a framework for link prediction on protein-protein
interactions is presented and validated against the available information
on rice and human interactomes. However, it might be easily extended
to other organisms or even to other graphs not necessarily related
to biology, because the only required information is the network connectivity.
Satisfactory results might be expected when the framework is applied
to networks with a dense adjacency matrix, since calculations of L3
and Node2Vec rely on network exploration for an arbitrary node and
its degree.

\bibliographystyle{plain}
\bibliography{refs}

\pagebreak{}

\appendix

\part{Appendix}

\section{Supplementary Information}

\subsection{Datasets}

In the case of the human interactomes, the files were downloaded as-is
from the repository from Kovacs et al (2019). For the rice interactome,
the downloaded file was \emph{4530.protein.links.detailed.v11.0.txt.}
From this file, only interactions which have a curated origin (meaning
a value different than zero on the column \emph{'databases'}).

\subsection{Node2Vec Parameters}

For all experiments, the following parameters hold:
\begin{itemize}
\item For the (\texttt{p=1}, \texttt{q=1})
\item (\texttt{dimensions=16, workers=4})
\item (\texttt{walk\_length=5}, \texttt{num\_walks=300}, \texttt{window=5})
\item (\texttt{min\_count=1}, \texttt{batch\_words=4})
\item (\texttt{EdgeEmbedding=EdgeHadamardEmbedder})
\end{itemize}

\subsection{XGBoost parameters}
\begin{itemize}
\item \texttt{max\_depth=3}
\item \texttt{colsample\_bytree=0.6}
\item \texttt{eval\_metric=''auc''}
\end{itemize}

\section{Supplementary Figures}

\subsection{Human Interactome}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI1}\emph{HI-II-14} results for \texttt{Node2Vec} with A2
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI2}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI3}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI1-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
A2 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2\lyxdot 2}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI2-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
A3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3\lyxdot 2}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI3-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
L3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3\lyxdot 2}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F8-importance-H2}Importance gain plots for \emph{HI-TESTED}}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A2\lyxdot 2}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A3\lyxdot 2}
\par\end{centering}
\centering{}\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot L3\lyxdot 2}
\end{figure}

\subsection{Rice Interactome}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F2}Results for \texttt{Node2Vec} with A2 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA2_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F3}Results for \texttt{Node2Vec} with A3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA3_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F5}Results for A2 feature only}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA2_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F6}Results for A3 feature alone}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA3_SUMMARY}
\end{figure}

\end{document}

### 2020-12-09 nicolas/PPIPL3/ReportV6.tex
files were missing in git

--- 2020-12-07 nicolas/PPIPL3/ReportV5.tex
@@
 \usepackage{float}
+\usepackage{textcomp}
 \usepackage{graphicx}
@@
 \author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
-email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
+\protect\hphantom{ema.}email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
 Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
@@
 
-Efforts to predict unmapped protein-protein interactions (PPI) through
-network analysis suggest that the higher the number of paths of length
-3 (L3) between two proteins, the higher the probability of interaction.
-This paper extends previous work on predicting protein interactions
-based on the L3 principle by taking into account the learning of node
-features of the PPI network. In particular, we train an \texttt{XGBoost}
-model using handcrafted features (including L3), as well as edge embeddings
-from \texttt{Node2Vec}. Our main result shows that while L3 is an
-important feature for predicting links, best performance is achieved
-when combined with edge embeddings. The proposed approach is evaluated
-for the human and rice interactomes and pretends to exploit the intrinsic
-network structure. Results for both organisms show higher AUC values
-for the combination of automated and handcrafted features than for
-each of its components alone. The handcrafted features have more importance
-for the prediction model in the human interactome than in rice. The
-developed framework can be easily applied to interactomes for other
-organisms or even to networks outside of the biological context.
+Recent developments in network-based prediction of unmapped protein-protein
+interactions (PPIs) reveal that two proteins are more likely to interact,
+the higher the number of paths of length 3 between them, normalized
+by the geometric average of their interactions. This paper extends
+previous work on predicting binary interactions by taking into account
+the learning of features (embeddings) of the PPI network. In particular,
+we implement a gradient boosted decision tree model (\texttt{XGBoost})
+using handcrafted features (including the normalized measure of paths
+of length 3, L3) and embeddings from an algorithm that generates a
+low-dimensional representation of nodes (\texttt{Node2Vec}). Our main
+result shows that while L3 remains an important feature for predicting
+PPIs, a better performance is achieved when L3 is combined with embedding
+features. The proposed approach is validated for the human and rice
+interactomes. For both cases, higher AUC values are obtained when
+both types of features are used to train the model. The developed
+framework can also be applied to interactomes of other organisms for
+which a PPI network become recently available.
 \end{abstract}
-\begin{comment}
-need include a feature that compares the dimension of the embedding
-is useful for prediction with L3 to show that they independent measures
-- we do not want to be learning L3 through node2vec
-\end{comment}
 
@@
 
-\begin{comment}
-Need to add other relevant references
-\end{comment}
+Proteins are key actors of biological processes inside cells. Rather
+than carrying out tasks as single agents, they are part of dynamic
+networks of protein-protein interactions (PPI) \cite{Lin2017}. Such
+networks underlie a variety of interdependent mechanisms, including
+signal transduction, homeostasis control and stress responses. Furthermore,
+PPI networks play an important role in physiological and developmental
+processes such as protein phosphorylation, transcriptional co-factor
+recruitment and transporter activation \cite{Zhang_2010_PPI}.
 
-Proteins are key actors of biological processes inside cells. As they
-carry out a variety of tasks, proteins do not work as single agents
-but rather as a part of dynamic networks of protein-protein interactions
-(PPI) \cite{Lin2017}. PPI networks play an important role in a variety
-of interdependent mechanisms, including signal transduction, homeostasis
-control, stress responses, plant defense and organ formation. They
-are also involved a variety of physiological and developmental processes,
-including protein phosphorylation, transcriptional co-factor recruitment
-and transporter activation \cite{Zhang_2010_PPI}.
-
-Different methods for predicting interactions in PPI networks have
-been proposed in recent years \cite{Chen2019PPI,Chang2016PPI,Kotlyar2015PPI}.
-Kovacs et al. (2019) introduce an approach which relies on counting
-the number of paths of length 3 (L3) to predict interactions among
-proteins for different organisms. The proposed approach outperforms
-previous methods in PPI networks of yeast (\emph{S. cerevisiae}),
-Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D.
-melanogaster}), fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus})
-and the human interactome \cite{Kovacs2019}.
-
-A common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid}
-technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
-which is based on (i) the expression of a specific reporter gene that
-activates by the binding of a DNA-binding Domain (DB), and (ii) an
-Activation Domain (AD) of a Transcription Factor that binds to an
-Upstream Activation Sequence (UAS). For the Y2H technique, a protein
-is fused to the DB domain (known as \emph{bait}) and another protein
-to the AD (known as \emph{prey}). If the proteins do not interact,
-then the reporter gene is not expressed. Otherwise, the reporter gene
-expression is activated by the activation domain.
+A common way to create PPI networks or validate particular protein-protein
+interactions is the \emph{Yeast-Two-Hybrid} technique (also known
+as \emph{two-hybrid screening} or \emph{Y2H}). Figure \ref{Y2H}A
+illustrates the biological basis of Y2H: the expression of a specific
+reporter gene that activates by the binding of a DNA-binding Domain
+(DB) and an Activation Domain (AD) of a Transcription Factor that
+binds to an Upstream Activation Sequence (UAS). To evaluate an interaction,
+the Y2H approach fuses a protein to the DB domain (known as \emph{bait})
+and another protein to the AD (known as \emph{prey}). If the proteins
+interact, then the reporter gene expression is activated by the AD
+(Fig. \ref{Y2H}B). Otherwise, if proteins fail to interact, then
+the reporter gene is not expressed (Fig. \ref{Y2H}C).
 
 \begin{figure}[h]
-\caption{Yeast-2-Hybrid Technique}
+\caption{\label{Y2H}The Yeast-2-Hybrid technique offers an experimental approach
+for constructing PPI networks}
 
@@
 
-Based on the outcome of numerous Y2H experiments, PPI networks which
-capture known interactions for each protein can be constructed. Several
-algorithms are proposed over these networks to predict unknown interactions.
+Based on the outcome of numerous Y2H experiments, undirected networks
+of interactions between proteins can be constructed. However, relying
+only on experimental validation of PPIs is detrimental to research
+due to its hindrance in costs, accuracy, required manpower and time
+\cite{Laraia2015PPI,Macalino2018PPI}. Besides, the number of missing
+interactions for pairs of proteins on many organisms leverage the
+usage of computational methods for prediction of such interactions.
+\begin{comment}
 The focus of this study is to evaluate different methods for predicting
-PPIs using the network of known interactions. These interactions are
-represented as an undirected graph. A common way of predicting interactions
-is usually based on social networks analysis, more specifically on
-the triadic closure principle (TCP). TCP states that the higher the
-number of common neighboring nodes between two nodes, the higher the
-probability that they interact \cite{Goldberg2003SmallWorld}. Previous
-studies show that the mentioned approach fails because it does not
-consider the structural and chemical properties of the proteins \cite{Kovacs2019,Cannistraci2013Networks}.
+PPIs using the network of known interactions. 
+\end{comment}
+
+A common way of predicting interactions is usually based on social
+networks analysis, more specifically on the triadic closure principle
+(TCP). TCP states that the higher the number of common neighboring
+nodes between two nodes, the higher the probability that they interact
+\cite{Goldberg2003SmallWorld}. TCP can be addressed mathematically
+by counting the number of shared neighbors of a pair of nodes, also
+known as the Common Neighbors algorithm. By raising the adjacency
+matrix (A) of the network to the second power (A\texttwosuperior ),
+TCP can be considered for further analyses. However, previous studies
+show that the mentioned approach usually fails because it does not
+consider the structural and chemical properties of the proteins \cite{Cannistraci2013Networks,Kovacs2019}.
+
+A variety of methods for predicting interactions in PPI networks have
+been proposed in recent years \cite{Chang2016PPI,Chen2019PPI,Kotlyar2015PPI}.
+Kovacs et al (2019) introduce a network-based approach which predicts
+the interaction between two proteins based on the number of paths
+of length 3, normalized by the geometric average of their interactions.
+The simplest mathematical representation of this principle consists
+on the third power of the adjacency matrix (A\textthreesuperior ).
+Furthermore, Kovacs et al present a degree-normalized scaling for
+this metric, which reduces bias caused by intermediate hubs within
+the paths of length 3. This handcrafted measure, denoted L3, enables
+the proposed approach to outperform previous methods for predicting
+binary protein interactions in yeast (\emph{S. cerevisiae}), Arabidopsis
+(\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
+fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus}) and humans
+\cite{Kovacs2019}.
 
@@
 
-For achieving the described results, human and rice PPI networks are
-compared using state-of-the-art methods, as well as the proposed ones
-(A2, A3, L3). In the case of the human network, two different versions
-of the human interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
+The focus of this study is to evaluate different methods for predicting
+PPIs using the network of known interactions. To achieve this, human
+and rice PPI networks are compared using the proposed methods (A2,
+A3, L3), as well as a low-dimensional representation of nodes (\texttt{Node2Vec})\cite{Grover_2016}.
+After that, combinations of the two types of methods are tested. In
+the case of the human network, two different versions of the human
+interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
 are used for training and another interactome version is used for
@@
 networks and ii) two applications of this framework to biological
-networks with a structural insight
-
-\begin{comment}
-JF: read throughout the document but edited up here. I prefer to read
-the results before editing the materials and methods section. 
-\end{comment}
+networks with a structural insight.
 
@@
 Importance values for the A3 method stand out less from the node embedding
-information than in A2 and L3. In this case, the importance gain of
-the handcrafted features in each model was not as high in rice as
-in the human interactome.
+information than in A2 and L3. However, handcrafted features are more
+important for the prediction of the human than in the rice interactome.
 

### 2020-12-09 nicolas/PPIPL3/ReportV7.tex
files were missing in git

--- 2020-12-09 nicolas/PPIPL3/ReportV6.tex
@@
 \begin{document}
-\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
+\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactomes}
 \author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
@@
 
-Recent developments in network-based prediction of unmapped protein-protein
-interactions (PPIs) reveal that two proteins are more likely to interact,
-the higher the number of paths of length 3 between them, normalized
-by the geometric average of their interactions. This paper extends
-previous work on predicting binary interactions by taking into account
-the learning of features (embeddings) of the PPI network. In particular,
-we implement a gradient boosted decision tree model (\texttt{XGBoost})
-using handcrafted features (including the normalized measure of paths
-of length 3, L3) and embeddings from an algorithm that generates a
-low-dimensional representation of nodes (\texttt{Node2Vec}). Our main
-result shows that while L3 remains an important feature for predicting
-PPIs, a better performance is achieved when L3 is combined with embedding
-features. The proposed approach is validated for the human and rice
-interactomes. For both cases, higher AUC values are obtained when
-both types of features are used to train the model. The developed
-framework can also be applied to interactomes of other organisms for
-which a PPI network become recently available.
+A recent study in network-based prediction of protein-protein interactions
+(PPIs) reveals that two proteins are more likely to interact, the
+higher the number of paths of length 3 between them (normalized by
+the geometric average of their interactions). This paper extends previous
+work on mapping binary interactions by taking into account the learning
+of features (embeddings) of the PPI network. In particular, we implement
+a gradient boosted decision tree model (\texttt{XGBoost}) using handcrafted
+features (including the normalized measure) and embeddings from an
+algorithm that generates a low-dimensional representation of nodes
+(\texttt{Node2Vec}). Our main result shows that while the measure
+remains an important feature for predicting interactions, better performance
+is achieved when in addition embedding features are considered. The
+proposed approach is validated for the human and rice interactomes.
+For both cases, the combination of both types of features yield higher
+AUC values. The developed framework can also be applied to interactomes
+of other organisms for which PPI networks have recently become available.
 \end{abstract}
@@
 
-A common way to create PPI networks or validate particular protein-protein
-interactions is the \emph{Yeast-Two-Hybrid} technique (also known
+A common way to create PPI networks (or validate particular protein-protein
+interactions) is the \emph{Yeast-Two-Hybrid} technique (also known
 as \emph{two-hybrid screening} or \emph{Y2H}). Figure \ref{Y2H}A
 illustrates the biological basis of Y2H: the expression of a specific
-reporter gene that activates by the binding of a DNA-binding Domain
-(DB) and an Activation Domain (AD) of a Transcription Factor that
-binds to an Upstream Activation Sequence (UAS). To evaluate an interaction,
-the Y2H approach fuses a protein to the DB domain (known as \emph{bait})
-and another protein to the AD (known as \emph{prey}). If the proteins
-interact, then the reporter gene expression is activated by the AD
-(Fig. \ref{Y2H}B). Otherwise, if proteins fail to interact, then
-the reporter gene is not expressed (Fig. \ref{Y2H}C).
+reporter gene is activated by the binding of a DNA-binding Domain
+(DB) and an Activation Domain (AD) of a Transcription Factor, which
+in turn binds to an Upstream Activation Sequence (UAS). To evaluate
+an interaction between two proteins, the Y2H approach fuses one protein
+to the DB domain (known as \emph{bait}) and another protein to the
+AD (known as \emph{prey}). If the proteins interact, the reporter
+gene expression is activated by the AD (Fig. \ref{Y2H}B). Otherwise,
+if proteins fail to interact, the reporter gene is not expressed (Fig.
+\ref{Y2H}C).
 
@@
 \caption{\label{Y2H}The Yeast-2-Hybrid technique offers an experimental approach
-for constructing PPI networks}
+for constructing PPI networks.}
 
@@
 the case of the human network, two different versions of the human
-interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
-are used for training and another interactome version is used for
-validation (\emph{HI-III})\cite{Kovacs2019}. For the rice interactome,
-a portion of the known interactions removed and then predicted with
-different techniques based on the known interactions. The main contributions
-of this paper are i) a general framework for link prediction in non-directed
-networks and ii) two applications of this framework to biological
-networks with a structural insight.
+interactome \cite{Rolland2014Human} are used for training and another
+interactome version is used for validation (\emph{HI-III})\cite{Kovacs2019}.
+For the rice interactome, a portion of the known interactions removed
+and then predicted with different techniques based on the known interactions.
+The main contributions of this paper are i) a general framework for
+link prediction in non-directed networks and ii) two applications
+of this framework to biological networks with a structural insight.
 
@@
 presents supplementary information and figures which complement the
-results described in this paper. Figures in the Appendix can be identified
-with a prefix (e.g. S\ref{F2}).
+results described in this paper.
 

### 2020-12-09 nicolas/PPIPL3/bmc_template/abstract.tex
BMC-formated paper

\parttitle{Background} %if any
Abstract A recent study in network-based prediction of protein-protein interactions (PPIs) reveals that two proteins are more likely to interact, the higher the number of paths of length 3 between them (normalized by the geometric average of their interactions). This paper extends previous work on mapping binary interactions by taking into account the learning of features (embeddings) of the PPI network. In particular, we implement a gradient boosted decision tree model (XGBoost) using handcrafted features (including the normalized measure) and embeddings from an algorithm that generates a low-dimensional representation of nodes (node2vec).

\parttitle{Results} %if any
Our main result shows that while the measure remains an important feature for predicting interactions, better performance is achieved when in addition embedding features are considered. The proposed approach is validated for the human and rice interactomes. For both cases, the combination of both types of features yield higher AUC values.

\parttitle{Conclusions} %if any
As found on this study on both human and rice, when information from handcrafted features based on neighborhood is enhanced with vector representations from random walks, the prediction power of the model improves. Besides, a supervised learning model can be trained for predicting unknown interactions based on such information. Finally, the developed framework can also be applied to interactomes of other organisms for which PPI networks have recently become available.

### 2020-12-09 nicolas/PPIPL3/bmc_template/introduction.tex
BMC-formated paper

--- 2020-12-09 nicolas/PPIPL3/ReportV7.tex
@@
-%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english]{article}
-\usepackage[T1]{fontenc}
-\usepackage[latin9]{inputenc}
-\usepackage{verbatim}
-\usepackage{float}
-\usepackage{textcomp}
-\usepackage{graphicx}
-
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% A simple dot to overcome graphicx limitations
-\newcommand{\lyxdot}{.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
-\usepackage{multirow}
-%\usepackage{table}
-%\usepackage{tabular}
-
-\makeatother
-
-\usepackage{babel}
-\begin{document}
-\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactomes}
-\author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
-\protect\hphantom{ema.}email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
-Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
-\maketitle
-\begin{abstract}
-
-A recent study in network-based prediction of protein-protein interactions
-(PPIs) reveals that two proteins are more likely to interact, the
-higher the number of paths of length 3 between them (normalized by
-the geometric average of their interactions). This paper extends previous
-work on mapping binary interactions by taking into account the learning
-of features (embeddings) of the PPI network. In particular, we implement
-a gradient boosted decision tree model (\texttt{XGBoost}) using handcrafted
-features (including the normalized measure) and embeddings from an
-algorithm that generates a low-dimensional representation of nodes
-(\texttt{Node2Vec}). Our main result shows that while the measure
-remains an important feature for predicting interactions, better performance
-is achieved when in addition embedding features are considered. The
-proposed approach is validated for the human and rice interactomes.
-For both cases, the combination of both types of features yield higher
-AUC values. The developed framework can also be applied to interactomes
-of other organisms for which PPI networks have recently become available.
-\end{abstract}
-
-\section{Introduction}
-
 Proteins are key actors of biological processes inside cells. Rather
@@
 processes such as protein phosphorylation, transcriptional co-factor
-recruitment and transporter activation \cite{Zhang_2010_PPI}.
+recruitment and transporter activation \cite{Zhang2010PPI}.
 
@@
 
-\noindent \centering{}\includegraphics[width=0.7\columnwidth]{Y2H}
+%\noindent \centering{}\includegraphics[width=0.7\columnwidth]{Y2H}
 \end{figure}
@@
 usage of computational methods for prediction of such interactions.
-\begin{comment}
-The focus of this study is to evaluate different methods for predicting
-PPIs using the network of known interactions. 
-\end{comment}
 
@@
 The simplest mathematical representation of this principle consists
-on the third power of the adjacency matrix (A\textthreesuperior ).
+on the third power of the adjacency matrix (A\textthreesuperior).
 Furthermore, Kovacs et al present a degree-normalized scaling for
@@
 \cite{Kovacs2019}.
-
-\begin{comment}
-JF: Yet you do more than evaluating this three measures so this has
-to be updated.
-
-NL: Done!
-
-JF: Not sure we should use L3 above since it is being redefined as
-a normalized measure here!
-
-JF{*}: state what the results are: ``The contributions of this paper
-are...''
-
-Which article predicts PPI based on TCP? add refs
-\end{comment}
 
@@
 
-The remainder of this paper is organized as follows. Section 2 describes
+\paragraph*{Structure:} This paper is organized as follows. \emph{Materials and Methods} describes
 the methodological steps and key milestones in the preparation of
-the networks, model parameters and experimental configurations. Section
-3 presents the main results of the models for the human and rice interactomes.
-Section 4 presents the main conclusions of this study. Finally, Appendix
+the networks, model parameters and experimental configurations.
+\emph{Results} presents the main results of the models for the human and rice interactomes.
+\emph{Conclussions} presents the main conclusions of this study. Finally, \emph{Appendix}
 presents supplementary information and figures which complement the
 results described in this paper.
-
-\section{Materials and Methods}
-
-\subsection{Data Availability}
-
-The human interactome data was downloaded from the repository of the
-length-3 degree normalized paths methodology \cite{Kovacs2019}: the
-dataset \emph{HI-II-14} and \emph{HI-TESTED} are used for prediction.
-The dataset \emph{HI-III} is available in the same repository and
-is used for validation.
-
-The rice interactome information was downloaded from the STRING database
-\cite{Szklarczyk2019}, corresponding to the subspecies \emph{Oryza
-sativa}. The file contains more than 8 million PPIs from several sources.
-For the purpose of this study, only PPIs with evidence from curated
-databases are used. The resulting network contains $5025$ nodes and
-$164420$ edges.
-
-\subsection{Code Implementation}
-
-The original code implementation in \texttt{C++} from Kovacs et al
-(2019) was adapted to \texttt{Python} (V3.6). For the purpose of algorithmic
-validation, the three methods were implemented from scratch with basic
-functionalities and data structures of the \texttt{Python} language.
-The code is publicly available under the repository github.com/ocinlr/PPIPL3.
-
-\subsection{Data Preprocessing}
-
-Information for the human interactome was used as-is, which corresponds
-to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
-$9433$ and $23322$ interactions, respectively.
-
-For the rice interactome, an additional preprocessing was performed.
-The filtered network for rice consists of $5025$ proteins (nodes)
-and $164420$ interactions (edges) distributed among $178$ connected
-components. The connected component with the greatest number of edges
-was selected. The extracted connected component consists of $n=4390$
-nodes and $m=163319$ edges, which corresponds to $99.33$\% of edges.
-
-\subsection{Edge Prediction}
-
-For predicting interactions for each network, the algorithms described
-below were used. It is important to keep in mind how the PPI network
-$G=(V,E)$ is conceptualized: each node ($v_{i}\in V$) represents
-a protein and each undirected edge ($e_{b}=\{v_{i},v_{j}\},\,e_{b}\in E$)
-represents an interaction among proteins $v_{i}$ and $v_{j}$.
-\begin{description}
-\item [{Common~Neighbors~(A2)}] This method is based on the Triadic Closure
-Principle which states that the more common friends two individuals
-have, the more likely that they know each other. For implementing
-this measure, $A{{}^2}$ matrix is calculated, where $A$ is the adjacency
-matrix of the network.
-\item [{Length-3~Path~Counts~(A3)}] This is a simple implementation
-of the notion of ``if my friends and your friends interact, then
-we might interact too''. For implementing this measure, $A{{}^3}$
-is calculated.
-\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
-overestimate the importance of some edges due to intermediate hubs
-which add many shortcuts in the graph. To address that issue, a degree
-normalization for the path $v_{i}\rightarrow v_{m}\rightarrow v_{n}\rightarrow v_{j}$
-is applied by considering the degree of the intermediate nodes $k_{m}$
-and $k_{n}$, as follows.
-\[
-p_{i,j}=\sum_{m,n\in|V|}\frac{A_{i,m}\cdot A_{m,n}\cdot A_{n,j}}{\sqrt{k_{m}\cdot k_{n}}}
-\]
-\\
-where $A_{i,j}$ represents the value of the adjacency matrix for
-nodes $i$ and $j$.
-\end{description}
-
-\subsection{Sampling Procedure}
-
-For each PPI network, the following procedure was performed 10 times
-to address the stochastic nature of the process:
-\begin{enumerate}
-\item A percentage of interactions is removed at random from the network
-(20\%).
-\item The same amount of removed interactions are then predicted using the
-main methods for prediction mentioned by Kovacs et al (2019): Common
-Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
-and the Length-3 degree-normalized score (\textbf{L3}).
-\item A test dataset is created as follows: all removed edges are included
-(as observed positives for the ML algorithm) and from the predicted
-edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
-the previous classification (observed negatives), a random subset
-is chosen such that the dataset is balanced, that is, the amount of
-observed positive labels is equal to the observed negative labels.
-\item Once the dataset is ready, it is randomly partitioned: 80\% is used
-for \texttt{XGBoost} model training and 20\% is used for validation.
-It is important to have in mind that balanced distribution of the
-positive and negative labels in the datasets was satisfied in training,
-but validation is performed using the remaining unbalanced data.
-\end{enumerate}
-It is worth mentioning that some exploration experiments were carried
-out with interactions removal percentages of 2, 5, 10 and 20\%, as
-well as train-test partitions of 15-85, 80-20 and 75-25, and the explained
-parameter set was selected because it either represented a marginal
-gain (results not shown) or because it was a common parameter selection
-in related literature.
-
-\subsection{Feature Extraction with \texttt{Node2Vec}}
-
-The \texttt{Node2Vec} module was used for extracting features of the
-interactomes. The considerations for the model were:
-\begin{itemize}
-\item All paths in the random walks are equally likely.
-\item Use a modest number of dimensions and threads for calculation.
-\item Since length-3 paths are the defining property in this study, there
-is no necessity for longer walks. However, it is important to try
-out many possible redundant routes and to consider a window of at
-least 4.
-\item Other standard parameters were left with default values.
-\item Edge embeddings were calculated using a geometric ratio of the node
-embeddings.
-\end{itemize}
-
-\subsection{Handcrafted Features}
-
-Due to the poor results of the \emph{raw} Length-3 counting (\textbf{A3}),
-a different approach for this information was carried out: As it still
-gives a lot of information that might be useful for a predictive routine,
-this counting was normalized (dividing by the greatest counting in
-the \textbf{A3} top predictions) and then used as a feature for the
-Machine Learning algorithm. For completeness, \textbf{A2} and \textbf{L3}
-measures were used as a input features. Finally, the case were no
-handcrafted feature was also considered, that is, only the features
-extracted from the structure of the network were considered.
-
-\subsection{Feature to Predict: Existence}
-
-The feature to predict corresponds to the possible existence (\emph{True/False})
-of a link based on the existing information of the network. This is
-evaluated by taking out a fraction of the edges and then trying to
-predict for a given set of possible edges if they have a high probability
-to belong to the original network.
-
-\subsection{Machine Learning Algorithm}
-
-The Extreme Gradient Boosting implementation of gradient boosted trees
-is applied to evaluate the existence of an edge. Gradient boosted
-trees are usually used for supervised learning problems, where the
-training data $X_{i}$ has multiple features and pretends to explain
-(or predict) a target variable $Y_{i}$. The corresponding implementation
-applied for this study is \texttt{XGBoost}.
-
-\subsection{Result Validation}
-
-As mentioned before, 80\% of the final dataset was randomly selected
-and used for training, while the remaining information was used for
-validation. The whole training-validation procedure was applied 10
-times.
-
-The chosen metric for validation was the Area under the Curve (\textbf{AUC})
-of the Receiver Operating Characteristic (\textbf{ROC}). This curve
-corresponds to plot the sensitivity (probability of predicting a real
-positive as positive) against 1-specificity (probability of predicting
-a real negative as positive). AUC values of 1 represent a perfect
-prediction and 0.5 corresponds to a random guess. Normally, values
-over 0.8 of AUC are considered good.
-
-\section{Results and Discussion}
-
-\subsection{Human Interactome}
-
-As the first step in the evaluation of A2, A3 and L3 methods for link
-prediction, the full calculation and ranking of all predictions was
-carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
-and \emph{HI-TESTED}) is shown in Figure \ref{fig:HI1}, in which
-the x axis represents a rank position $k$ and the y axis displays
-the precision for the top $k$ predictions of each method, when assuming
-interactome \emph{HI-III} as the validation set.
-
-\begin{figure}[h]
-\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14} and \emph{HI-TESTED}}
-
-\noindent \centering{}\includegraphics[width=0.48\columnwidth]{hi-ii-14\lyxdot txt}\includegraphics[width=0.48\columnwidth]{hi-tested\lyxdot txt}
-\end{figure}
-
-As it can be inferred from the plots, L3-based predictions outperform
-their $A{{}^2}$ counterparts. Results also show that L3-score and
-$A^{3}$predictions follow a very similar trend.
-
-However, it is of our interest to assess if Machine Learning can boost
-the overall performance on the prediction, so the proposed procedure
-was adapted to the human interactome data: Instead of having to remove
-a fraction of edges from the network in order to predict and validate,
-the prediction network and the validation networks are given beforehand.
-The rest of the procedure is carried out as mentioned in Section 2.
-Table \ref{T1} presents the results for the combinations of Node2Vec
-with each feature.
-
-\begin{table}
-\caption{\label{T1}Summary statistics for human interactomes}
-\includegraphics[width=1\columnwidth]{T1}
-\end{table}
-
-In general, performance of all models is above 0.96 in terms of F1-Score.
-Although marginally, the A2 feature-enriched model for HI-TESTED performs
-better than their counterparts based on paths of length 3. The opposite
-is also true for HI-II-14: models enriched with L3 and A3 features
-perform better than their A2 counterpart. Results for the Interactome
-\emph{HI-II-14} with Node2Vec and each metric are presented in Figures
-S\ref{HI1}, S\ref{HI2} and S\ref{HI3}. It can be seen that for
-\emph{HI-II-14}, A2 performs marginally better than L3 when comparing
-precision but worse in terms of recall, although all metrics have
-AUC values of 0.99.
-
-On the other hand, the interactome \emph{HI-TESTED} was assessed with
-the same methodology as \emph{HI-II-14} and results are shown in Figures
-S\ref{HI1-1}, S\ref{HI2-1} and S\ref{HI3-1}. For this interactome,
-A2 performs better in terms of false positives than A3 and L3, but
-in terms of false negatives, A3 actually performs better that A2 and
-L3. In terms of AUC, metrics results are indistinguishable.
-
-An interesting assessment from this human datasets can be observed
-when looking at the importance plots, which show a very biased influence
-of the handcrafted features. Figure \ref{F8-importance-H1} shows
-this behavior for the \emph{HI-II-14} dataset, which is very similar
-to \emph{HI-TESTED} (Figure S\ref{F8-importance-H2}).
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance-H1}Importance gain plots for \emph{HI-II-14}}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A2\lyxdot 1}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A3\lyxdot 1}
-\par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot L3\lyxdot 1}
-\end{figure}
-
-\subsection{Rice Interactome}
-
-For the rice interactome, different combinations of features were
-trained on the model and validated. The results are shown in Table
-\ref{T2}. First, prediction based on \texttt{Node2Vec} without any
-additional feature included is used as baseline of comparison. Figure
-\ref{F1} shows those results, and one can see that its mean performance
-using the AUC is $0.93$. The results for all 10 repetitions of each
-model are consistent, meaning the variance among each line is not
-significant. The results for each model using only the default features
-are satisfactory. This result can also be assessed when looking at
-the confusion matrix, where a precision of $0.98$ and a recall of
-$0.86$ are obtained.
-
-\begin{table}
-\caption{\label{T2}Summary statistics for rice interactome}
-\includegraphics[width=1\columnwidth]{T2}
-\end{table}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F1}Summary ROC curves for \texttt{Node2Vec} model alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_ML/ROConlyML_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_ML/CMonlyML_SUMMARY}
-\end{figure}
-
-The next experiment carried out was to add for each prediction method
-score as a prediction feature, that is, use the score as an additional
-feature for the ML algorithm. Results for the Common Neighbors method
-(\textbf{A2}), which uses paths of length 2, are shown in Figure S\ref{F2}.
-All 10 experiments have small variability among them and perform better
-than the baseline, with AUC of $0.98$. When looking at the confusion
-matrix for this model, most guesses are true positives and true negatives,
-resulting in a precision of $0.99$ and a recall of $0.91$. The same
-analyses are done with the count of paths of length 3 (\textbf{A3})
-and with the degree-normalized length-3 score (\textbf{L3}). These
-results are presented in figures S\ref{F3} and \ref{F4}, respectively.
-The corresponding AUC are $0.97$ and $0.98$.
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F4}Results for \texttt{Node2Vec} with L3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CML3_SUMMARY}
-\end{figure}
-
-The precision when using the count of paths of length 3 (A3) is $0.99$
-while the recall is $0.90$. It is interesting to observe that precision
-decreased $0.16\%$ and recall decreased $0.58\%$. A similar situation
-is obtained for the model with the normalized score (L3), whose precision
-decreased to $0.99$ ($-0.08\%$) and recall decreased to $0.88$
-($-2.38\%$). It can be seen also that the ROC curve achieves high
-values of true positive rate sooner in the A2-featured model than
-on the other models.
-
-Another relevant assessment was carried out to verify the predictive
-power of the metrics. \texttt{XGBoost} models were trained only using
-each metric, for the same 10 experiments. Results are shown in Figures
-S\ref{F5}, S\ref{F6} and \ref{F7}. XGBoost yields lower AUC values
-when only considering A2, A3 and L3. Precision and recall for this
-assessments, although lower than the previous experiments, are still
-considered satisfactory.
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F7}Results for L3 feature alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyL3_SUMMARY}
-\end{figure}
-
-Finally, an relevant evaluation on the models that include a handcrafted
-feature is necessary: How important is the appended feature for the
-model result? The answer can be observed by the feature importance
-plots for A2, A3 and L3 in Figure \ref{F8-importance}. Each plot
-resembles the importance gain from each feature in the underlying
-decision trees that \texttt{XGBoost} uses. Usually, the more bifurcations,
-the higher gain a feature has, the more relevance a feature has on
-the model prediction itself. All models have the appended feature
-as the most important by a good margin, although margin for A3 is
-not wide as for A2 or L3.
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance}Importance gain plots for \texttt{Node2Vec}
-with each feature}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
-\par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
-\end{figure}
-
-\section{Conclusions}
-
-Based on previous work from other authors\cite{Kovacs2019}, the intrinsic
-knowledge of existing protein-protein interaction networks let scientists
-elucidate promising new interactions to explore. In this study, a
-machine learning algorithm was fed with previously elucidated information
-about the network, as well as results form a state-of-the-art edge
-embedding technique. Even if the baseline of the handcrafted features
-(A2, A3, L3) yields good results, boosting that information with the
-vector embedding, which calculates vector representations of the network
-structure improves the overall performance of each method on its own.
-
-On both human interactomes (\emph{HI-II-14} and \emph{HI-TESTED}),
-applying \texttt{XGBoost} with the metrics yield results with high
-AUC for the link prediction problem, and therefore this study extends
-previous work by other authors that only consider the predictive power
-of the metrics alone. In this case, the importance of the handcrafted
-features was very significant and stands out of the Node2Vec features.
-
-On the rice interactome, PPI prediction obtained high AUC values when
-only using the node embedding technique (AUC=0.93) and when only using
-the handcrafted features based on protein neighborhood (AUC \textasciitilde 0.88).
-Even higher AUC results were achieved when the handcrafted features
-and the vector representation information were combined (AUC \textasciitilde 0.98).
-
-Importance values for the A3 method stand out less from the node embedding
-information than in A2 and L3. However, handcrafted features are more
-important for the prediction of the human than in the rice interactome.
-
-With this study, a framework for link prediction on protein-protein
-interactions is presented and validated against the available information
-on rice and human interactomes. However, it might be easily extended
-to other organisms or even to other graphs not necessarily related
-to biology, because the only required information is the network connectivity.
-Satisfactory results might be expected when the framework is applied
-to networks with a dense adjacency matrix, since calculations of L3
-and Node2Vec rely on network exploration for an arbitrary node and
-its degree.
-
-\bibliographystyle{plain}
-\bibliography{refs}
-
-\pagebreak{}
-
-\appendix
-
-\part{Appendix}
-
-\section{Supplementary Information}
-
-\subsection{Datasets}
-
-In the case of the human interactomes, the files were downloaded as-is
-from the repository from Kovacs et al (2019). For the rice interactome,
-the downloaded file was \emph{4530.protein.links.detailed.v11.0.txt.}
-From this file, only interactions which have a curated origin (meaning
-a value different than zero on the column \emph{'databases'}).
-
-\subsection{Node2Vec Parameters}
-
-For all experiments, the following parameters hold:
-\begin{itemize}
-\item For the (\texttt{p=1}, \texttt{q=1})
-\item (\texttt{dimensions=16, workers=4})
-\item (\texttt{walk\_length=5}, \texttt{num\_walks=300}, \texttt{window=5})
-\item (\texttt{min\_count=1}, \texttt{batch\_words=4})
-\item (\texttt{EdgeEmbedding=EdgeHadamardEmbedder})
-\end{itemize}
-
-\subsection{XGBoost parameters}
-\begin{itemize}
-\item \texttt{max\_depth=3}
-\item \texttt{colsample\_bytree=0.6}
-\item \texttt{eval\_metric=''auc''}
-\end{itemize}
-
-\section{Supplementary Figures}
-
-\subsection{Human Interactome}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI1}\emph{HI-II-14} results for \texttt{Node2Vec} with A2
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI2}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI3}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI1-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-A2 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI2-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-A3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI3-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-L3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance-H2}Importance gain plots for \emph{HI-TESTED}}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A2\lyxdot 2}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A3\lyxdot 2}
-\par\end{centering}
-\centering{}\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot L3\lyxdot 2}
-\end{figure}
-
-\subsection{Rice Interactome}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F2}Results for \texttt{Node2Vec} with A2 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA2_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F3}Results for \texttt{Node2Vec} with A3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA3_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F5}Results for A2 feature only}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA2_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F6}Results for A3 feature alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA3_SUMMARY}
-\end{figure}
-
-\end{document}

### 2020-12-11 nicolas/PPIPL3/bmc_template/appendix.tex
BMC-formated document

\subsection*{Node2Vec Parameters}

For all experiments, the following parameters hold:
\begin{itemize}
\item For the (\texttt{p=1}, \texttt{q=1})
\item (\texttt{dimensions=16, workers=4})
\item (\texttt{walk\_length=5}, \texttt{num\_walks=300}, \texttt{window=5})
\item (\texttt{min\_count=1}, \texttt{batch\_words=4})
\item (\texttt{EdgeEmbedding=EdgeHadamardEmbedder})
\end{itemize}

\subsection*{XGBoost parameters}
\begin{itemize}
\item \texttt{max\_depth=3}
\item \texttt{colsample\_bytree=0.6}
\item \texttt{eval\_metric='auc'}
\end{itemize}

\subsection*{Supplementary Figures for Human Interactome}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{HI1}\emph{HI-II-14} results for \texttt{Node2Vec} with A2
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{figures/Human/CM\lyxdot A2}
\end{figure}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{HI2}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{figures/Human/CM\lyxdot A3}
\end{figure}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{HI3}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{figures/Human/CM\lyxdot L3}
\end{figure}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{HI1-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
A2 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Human/ROC\lyxdot A2\lyxdot 2}\includegraphics[width=0.48\columnwidth]{figures/Human/CM\lyxdot A2\lyxdot 2}
\end{figure}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{HI2-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
A3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Human/ROC\lyxdot A3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{figures/Human/CM\lyxdot A3\lyxdot 2}
\end{figure}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{HI3-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
L3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Human/ROC\lyxdot L3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{figures/Human/CM\lyxdot L3\lyxdot 2}
\end{figure}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{F8-importance-H2}Importance gain plots for \emph{HI-TESTED}}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.7\columnwidth]{figures/Human/Imp\lyxdot A2\lyxdot 2}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.7\columnwidth]{figures/Human/Imp\lyxdot A3\lyxdot 2}
\par\end{centering}
\centering{}\includegraphics[width=0.7\columnwidth]{figures/Human/Imp\lyxdot L3\lyxdot 2}
\end{figure}

\subsection*{Supplementary Figures for Rice Interactome}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{F2}Results for \texttt{Node2Vec} with A2 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/CMA2_SUMMARY}
\end{figure}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{F3}Results for \texttt{Node2Vec} with A3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/CMA3_SUMMARY}
\end{figure}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{F5}Results for A2 feature only}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Only_Metric/ROConlyA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{figures/Only_Metric/CMonlyA2_SUMMARY}
\end{figure}

\begin{figure}[h]
\noindent \begin{centering}
\caption{\label{F6}Results for A3 feature alone}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Only_Metric/ROConlyA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{figures/Only_Metric/CMonlyA3_SUMMARY}
\end{figure}

### 2020-12-11 nicolas/PPIPL3/bmc_template/conclusions.tex
BMC-formated document

--- 2020-12-07 nicolas/PPIPL3/ReportV5.tex
@@
-%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english]{article}
-\usepackage[T1]{fontenc}
-\usepackage[latin9]{inputenc}
-\usepackage{verbatim}
-\usepackage{float}
-\usepackage{graphicx}
-
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% A simple dot to overcome graphicx limitations
-\newcommand{\lyxdot}{.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
-\usepackage{multirow}
-%\usepackage{table}
-%\usepackage{tabular}
-
-\makeatother
-
-\usepackage{babel}
-\begin{document}
-\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
-\author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
-email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
-Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
-\maketitle
-\begin{abstract}
-
-Efforts to predict unmapped protein-protein interactions (PPI) through
-network analysis suggest that the higher the number of paths of length
-3 (L3) between two proteins, the higher the probability of interaction.
-This paper extends previous work on predicting protein interactions
-based on the L3 principle by taking into account the learning of node
-features of the PPI network. In particular, we train an \texttt{XGBoost}
-model using handcrafted features (including L3), as well as edge embeddings
-from \texttt{Node2Vec}. Our main result shows that while L3 is an
-important feature for predicting links, best performance is achieved
-when combined with edge embeddings. The proposed approach is evaluated
-for the human and rice interactomes and pretends to exploit the intrinsic
-network structure. Results for both organisms show higher AUC values
-for the combination of automated and handcrafted features than for
-each of its components alone. The handcrafted features have more importance
-for the prediction model in the human interactome than in rice. The
-developed framework can be easily applied to interactomes for other
-organisms or even to networks outside of the biological context.
-\end{abstract}
-\begin{comment}
-need include a feature that compares the dimension of the embedding
-is useful for prediction with L3 to show that they independent measures
-- we do not want to be learning L3 through node2vec
-\end{comment}
-
-\section{Introduction}
-
-\begin{comment}
-Need to add other relevant references
-\end{comment}
-
-Proteins are key actors of biological processes inside cells. As they
-carry out a variety of tasks, proteins do not work as single agents
-but rather as a part of dynamic networks of protein-protein interactions
-(PPI) \cite{Lin2017}. PPI networks play an important role in a variety
-of interdependent mechanisms, including signal transduction, homeostasis
-control, stress responses, plant defense and organ formation. They
-are also involved a variety of physiological and developmental processes,
-including protein phosphorylation, transcriptional co-factor recruitment
-and transporter activation \cite{Zhang_2010_PPI}.
-
-Different methods for predicting interactions in PPI networks have
-been proposed in recent years \cite{Chen2019PPI,Chang2016PPI,Kotlyar2015PPI}.
-Kovacs et al. (2019) introduce an approach which relies on counting
-the number of paths of length 3 (L3) to predict interactions among
-proteins for different organisms. The proposed approach outperforms
-previous methods in PPI networks of yeast (\emph{S. cerevisiae}),
-Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D.
-melanogaster}), fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus})
-and the human interactome \cite{Kovacs2019}.
-
-A common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid}
-technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
-which is based on (i) the expression of a specific reporter gene that
-activates by the binding of a DNA-binding Domain (DB), and (ii) an
-Activation Domain (AD) of a Transcription Factor that binds to an
-Upstream Activation Sequence (UAS). For the Y2H technique, a protein
-is fused to the DB domain (known as \emph{bait}) and another protein
-to the AD (known as \emph{prey}). If the proteins do not interact,
-then the reporter gene is not expressed. Otherwise, the reporter gene
-expression is activated by the activation domain.
-
-\begin{figure}[h]
-\caption{Yeast-2-Hybrid Technique}
-
-\noindent \centering{}\includegraphics[width=0.7\columnwidth]{Y2H}
-\end{figure}
-
-Based on the outcome of numerous Y2H experiments, PPI networks which
-capture known interactions for each protein can be constructed. Several
-algorithms are proposed over these networks to predict unknown interactions.
-The focus of this study is to evaluate different methods for predicting
-PPIs using the network of known interactions. These interactions are
-represented as an undirected graph. A common way of predicting interactions
-is usually based on social networks analysis, more specifically on
-the triadic closure principle (TCP). TCP states that the higher the
-number of common neighboring nodes between two nodes, the higher the
-probability that they interact \cite{Goldberg2003SmallWorld}. Previous
-studies show that the mentioned approach fails because it does not
-consider the structural and chemical properties of the proteins \cite{Kovacs2019,Cannistraci2013Networks}.
-
-\begin{comment}
-JF: Yet you do more than evaluating this three measures so this has
-to be updated.
-
-NL: Done!
-
-JF: Not sure we should use L3 above since it is being redefined as
-a normalized measure here!
-
-JF{*}: state what the results are: ``The contributions of this paper
-are...''
-
-Which article predicts PPI based on TCP? add refs
-\end{comment}
-
-For achieving the described results, human and rice PPI networks are
-compared using state-of-the-art methods, as well as the proposed ones
-(A2, A3, L3). In the case of the human network, two different versions
-of the human interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
-are used for training and another interactome version is used for
-validation (\emph{HI-III})\cite{Kovacs2019}. For the rice interactome,
-a portion of the known interactions removed and then predicted with
-different techniques based on the known interactions. The main contributions
-of this paper are i) a general framework for link prediction in non-directed
-networks and ii) two applications of this framework to biological
-networks with a structural insight
-
-\begin{comment}
-JF: read throughout the document but edited up here. I prefer to read
-the results before editing the materials and methods section. 
-\end{comment}
-
-The remainder of this paper is organized as follows. Section 2 describes
-the methodological steps and key milestones in the preparation of
-the networks, model parameters and experimental configurations. Section
-3 presents the main results of the models for the human and rice interactomes.
-Section 4 presents the main conclusions of this study. Finally, Appendix
-presents supplementary information and figures which complement the
-results described in this paper. Figures in the Appendix can be identified
-with a prefix (e.g. S\ref{F2}).
-
-\section{Materials and Methods}
-
-\subsection{Data Availability}
-
-The human interactome data was downloaded from the repository of the
-length-3 degree normalized paths methodology \cite{Kovacs2019}: the
-dataset \emph{HI-II-14} and \emph{HI-TESTED} are used for prediction.
-The dataset \emph{HI-III} is available in the same repository and
-is used for validation.
-
-The rice interactome information was downloaded from the STRING database
-\cite{Szklarczyk2019}, corresponding to the subspecies \emph{Oryza
-sativa}. The file contains more than 8 million PPIs from several sources.
-For the purpose of this study, only PPIs with evidence from curated
-databases are used. The resulting network contains $5025$ nodes and
-$164420$ edges.
-
-\subsection{Code Implementation}
-
-The original code implementation in \texttt{C++} from Kovacs et al
-(2019) was adapted to \texttt{Python} (V3.6). For the purpose of algorithmic
-validation, the three methods were implemented from scratch with basic
-functionalities and data structures of the \texttt{Python} language.
-The code is publicly available under the repository github.com/ocinlr/PPIPL3.
-
-\subsection{Data Preprocessing}
-
-Information for the human interactome was used as-is, which corresponds
-to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
-$9433$ and $23322$ interactions, respectively.
-
-For the rice interactome, an additional preprocessing was performed.
-The filtered network for rice consists of $5025$ proteins (nodes)
-and $164420$ interactions (edges) distributed among $178$ connected
-components. The connected component with the greatest number of edges
-was selected. The extracted connected component consists of $n=4390$
-nodes and $m=163319$ edges, which corresponds to $99.33$\% of edges.
-
-\subsection{Edge Prediction}
-
-For predicting interactions for each network, the algorithms described
-below were used. It is important to keep in mind how the PPI network
-$G=(V,E)$ is conceptualized: each node ($v_{i}\in V$) represents
-a protein and each undirected edge ($e_{b}=\{v_{i},v_{j}\},\,e_{b}\in E$)
-represents an interaction among proteins $v_{i}$ and $v_{j}$.
-\begin{description}
-\item [{Common~Neighbors~(A2)}] This method is based on the Triadic Closure
-Principle which states that the more common friends two individuals
-have, the more likely that they know each other. For implementing
-this measure, $A{{}^2}$ matrix is calculated, where $A$ is the adjacency
-matrix of the network.
-\item [{Length-3~Path~Counts~(A3)}] This is a simple implementation
-of the notion of ``if my friends and your friends interact, then
-we might interact too''. For implementing this measure, $A{{}^3}$
-is calculated.
-\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
-overestimate the importance of some edges due to intermediate hubs
-which add many shortcuts in the graph. To address that issue, a degree
-normalization for the path $v_{i}\rightarrow v_{m}\rightarrow v_{n}\rightarrow v_{j}$
-is applied by considering the degree of the intermediate nodes $k_{m}$
-and $k_{n}$, as follows.
-\[
-p_{i,j}=\sum_{m,n\in|V|}\frac{A_{i,m}\cdot A_{m,n}\cdot A_{n,j}}{\sqrt{k_{m}\cdot k_{n}}}
-\]
-\\
-where $A_{i,j}$ represents the value of the adjacency matrix for
-nodes $i$ and $j$.
-\end{description}
-
-\subsection{Sampling Procedure}
-
-For each PPI network, the following procedure was performed 10 times
-to address the stochastic nature of the process:
-\begin{enumerate}
-\item A percentage of interactions is removed at random from the network
-(20\%).
-\item The same amount of removed interactions are then predicted using the
-main methods for prediction mentioned by Kovacs et al (2019): Common
-Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
-and the Length-3 degree-normalized score (\textbf{L3}).
-\item A test dataset is created as follows: all removed edges are included
-(as observed positives for the ML algorithm) and from the predicted
-edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
-the previous classification (observed negatives), a random subset
-is chosen such that the dataset is balanced, that is, the amount of
-observed positive labels is equal to the observed negative labels.
-\item Once the dataset is ready, it is randomly partitioned: 80\% is used
-for \texttt{XGBoost} model training and 20\% is used for validation.
-It is important to have in mind that balanced distribution of the
-positive and negative labels in the datasets was satisfied in training,
-but validation is performed using the remaining unbalanced data.
-\end{enumerate}
-It is worth mentioning that some exploration experiments were carried
-out with interactions removal percentages of 2, 5, 10 and 20\%, as
-well as train-test partitions of 15-85, 80-20 and 75-25, and the explained
-parameter set was selected because it either represented a marginal
-gain (results not shown) or because it was a common parameter selection
-in related literature.
-
-\subsection{Feature Extraction with \texttt{Node2Vec}}
-
-The \texttt{Node2Vec} module was used for extracting features of the
-interactomes. The considerations for the model were:
-\begin{itemize}
-\item All paths in the random walks are equally likely.
-\item Use a modest number of dimensions and threads for calculation.
-\item Since length-3 paths are the defining property in this study, there
-is no necessity for longer walks. However, it is important to try
-out many possible redundant routes and to consider a window of at
-least 4.
-\item Other standard parameters were left with default values.
-\item Edge embeddings were calculated using a geometric ratio of the node
-embeddings.
-\end{itemize}
-
-\subsection{Handcrafted Features}
-
-Due to the poor results of the \emph{raw} Length-3 counting (\textbf{A3}),
-a different approach for this information was carried out: As it still
-gives a lot of information that might be useful for a predictive routine,
-this counting was normalized (dividing by the greatest counting in
-the \textbf{A3} top predictions) and then used as a feature for the
-Machine Learning algorithm. For completeness, \textbf{A2} and \textbf{L3}
-measures were used as a input features. Finally, the case were no
-handcrafted feature was also considered, that is, only the features
-extracted from the structure of the network were considered.
-
-\subsection{Feature to Predict: Existence}
-
-The feature to predict corresponds to the possible existence (\emph{True/False})
-of a link based on the existing information of the network. This is
-evaluated by taking out a fraction of the edges and then trying to
-predict for a given set of possible edges if they have a high probability
-to belong to the original network.
-
-\subsection{Machine Learning Algorithm}
-
-The Extreme Gradient Boosting implementation of gradient boosted trees
-is applied to evaluate the existence of an edge. Gradient boosted
-trees are usually used for supervised learning problems, where the
-training data $X_{i}$ has multiple features and pretends to explain
-(or predict) a target variable $Y_{i}$. The corresponding implementation
-applied for this study is \texttt{XGBoost}.
-
-\subsection{Result Validation}
-
-As mentioned before, 80\% of the final dataset was randomly selected
-and used for training, while the remaining information was used for
-validation. The whole training-validation procedure was applied 10
-times.
-
-The chosen metric for validation was the Area under the Curve (\textbf{AUC})
-of the Receiver Operating Characteristic (\textbf{ROC}). This curve
-corresponds to plot the sensitivity (probability of predicting a real
-positive as positive) against 1-specificity (probability of predicting
-a real negative as positive). AUC values of 1 represent a perfect
-prediction and 0.5 corresponds to a random guess. Normally, values
-over 0.8 of AUC are considered good.
-
-\section{Results and Discussion}
-
-\subsection{Human Interactome}
-
-As the first step in the evaluation of A2, A3 and L3 methods for link
-prediction, the full calculation and ranking of all predictions was
-carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
-and \emph{HI-TESTED}) is shown in Figure \ref{fig:HI1}, in which
-the x axis represents a rank position $k$ and the y axis displays
-the precision for the top $k$ predictions of each method, when assuming
-interactome \emph{HI-III} as the validation set.
-
-\begin{figure}[h]
-\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14} and \emph{HI-TESTED}}
-
-\noindent \centering{}\includegraphics[width=0.48\columnwidth]{hi-ii-14\lyxdot txt}\includegraphics[width=0.48\columnwidth]{hi-tested\lyxdot txt}
-\end{figure}
-
-As it can be inferred from the plots, L3-based predictions outperform
-their $A{{}^2}$ counterparts. Results also show that L3-score and
-$A^{3}$predictions follow a very similar trend.
-
-However, it is of our interest to assess if Machine Learning can boost
-the overall performance on the prediction, so the proposed procedure
-was adapted to the human interactome data: Instead of having to remove
-a fraction of edges from the network in order to predict and validate,
-the prediction network and the validation networks are given beforehand.
-The rest of the procedure is carried out as mentioned in Section 2.
-Table \ref{T1} presents the results for the combinations of Node2Vec
-with each feature.
-
-\begin{table}
-\caption{\label{T1}Summary statistics for human interactomes}
-\includegraphics[width=1\columnwidth]{T1}
-\end{table}
-
-In general, performance of all models is above 0.96 in terms of F1-Score.
-Although marginally, the A2 feature-enriched model for HI-TESTED performs
-better than their counterparts based on paths of length 3. The opposite
-is also true for HI-II-14: models enriched with L3 and A3 features
-perform better than their A2 counterpart. Results for the Interactome
-\emph{HI-II-14} with Node2Vec and each metric are presented in Figures
-S\ref{HI1}, S\ref{HI2} and S\ref{HI3}. It can be seen that for
-\emph{HI-II-14}, A2 performs marginally better than L3 when comparing
-precision but worse in terms of recall, although all metrics have
-AUC values of 0.99.
-
-On the other hand, the interactome \emph{HI-TESTED} was assessed with
-the same methodology as \emph{HI-II-14} and results are shown in Figures
-S\ref{HI1-1}, S\ref{HI2-1} and S\ref{HI3-1}. For this interactome,
-A2 performs better in terms of false positives than A3 and L3, but
-in terms of false negatives, A3 actually performs better that A2 and
-L3. In terms of AUC, metrics results are indistinguishable.
-
-An interesting assessment from this human datasets can be observed
-when looking at the importance plots, which show a very biased influence
-of the handcrafted features. Figure \ref{F8-importance-H1} shows
-this behavior for the \emph{HI-II-14} dataset, which is very similar
-to \emph{HI-TESTED} (Figure S\ref{F8-importance-H2}).
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance-H1}Importance gain plots for \emph{HI-II-14}}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A2\lyxdot 1}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A3\lyxdot 1}
-\par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot L3\lyxdot 1}
-\end{figure}
-
-\subsection{Rice Interactome}
-
-For the rice interactome, different combinations of features were
-trained on the model and validated. The results are shown in Table
-\ref{T2}. First, prediction based on \texttt{Node2Vec} without any
-additional feature included is used as baseline of comparison. Figure
-\ref{F1} shows those results, and one can see that its mean performance
-using the AUC is $0.93$. The results for all 10 repetitions of each
-model are consistent, meaning the variance among each line is not
-significant. The results for each model using only the default features
-are satisfactory. This result can also be assessed when looking at
-the confusion matrix, where a precision of $0.98$ and a recall of
-$0.86$ are obtained.
-
-\begin{table}
-\caption{\label{T2}Summary statistics for rice interactome}
-\includegraphics[width=1\columnwidth]{T2}
-\end{table}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F1}Summary ROC curves for \texttt{Node2Vec} model alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_ML/ROConlyML_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_ML/CMonlyML_SUMMARY}
-\end{figure}
-
-The next experiment carried out was to add for each prediction method
-score as a prediction feature, that is, use the score as an additional
-feature for the ML algorithm. Results for the Common Neighbors method
-(\textbf{A2}), which uses paths of length 2, are shown in Figure S\ref{F2}.
-All 10 experiments have small variability among them and perform better
-than the baseline, with AUC of $0.98$. When looking at the confusion
-matrix for this model, most guesses are true positives and true negatives,
-resulting in a precision of $0.99$ and a recall of $0.91$. The same
-analyses are done with the count of paths of length 3 (\textbf{A3})
-and with the degree-normalized length-3 score (\textbf{L3}). These
-results are presented in figures S\ref{F3} and \ref{F4}, respectively.
-The corresponding AUC are $0.97$ and $0.98$.
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F4}Results for \texttt{Node2Vec} with L3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CML3_SUMMARY}
-\end{figure}
-
-The precision when using the count of paths of length 3 (A3) is $0.99$
-while the recall is $0.90$. It is interesting to observe that precision
-decreased $0.16\%$ and recall decreased $0.58\%$. A similar situation
-is obtained for the model with the normalized score (L3), whose precision
-decreased to $0.99$ ($-0.08\%$) and recall decreased to $0.88$
-($-2.38\%$). It can be seen also that the ROC curve achieves high
-values of true positive rate sooner in the A2-featured model than
-on the other models.
-
-Another relevant assessment was carried out to verify the predictive
-power of the metrics. \texttt{XGBoost} models were trained only using
-each metric, for the same 10 experiments. Results are shown in Figures
-S\ref{F5}, S\ref{F6} and \ref{F7}. XGBoost yields lower AUC values
-when only considering A2, A3 and L3. Precision and recall for this
-assessments, although lower than the previous experiments, are still
-considered satisfactory.
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F7}Results for L3 feature alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyL3_SUMMARY}
-\end{figure}
-
-Finally, an relevant evaluation on the models that include a handcrafted
-feature is necessary: How important is the appended feature for the
-model result? The answer can be observed by the feature importance
-plots for A2, A3 and L3 in Figure \ref{F8-importance}. Each plot
-resembles the importance gain from each feature in the underlying
-decision trees that \texttt{XGBoost} uses. Usually, the more bifurcations,
-the higher gain a feature has, the more relevance a feature has on
-the model prediction itself. All models have the appended feature
-as the most important by a good margin, although margin for A3 is
-not wide as for A2 or L3.
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance}Importance gain plots for \texttt{Node2Vec}
-with each feature}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
-\par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
-\end{figure}
-
-\section{Conclusions}
-
 Based on previous work from other authors\cite{Kovacs2019}, the intrinsic
@@
 its degree.
-
-\bibliographystyle{plain}
-\bibliography{refs}
-
-\pagebreak{}
-
-\appendix
-
-\part{Appendix}
-
-\section{Supplementary Information}
-
-\subsection{Datasets}
-
-In the case of the human interactomes, the files were downloaded as-is
-from the repository from Kovacs et al (2019). For the rice interactome,
-the downloaded file was \emph{4530.protein.links.detailed.v11.0.txt.}
-From this file, only interactions which have a curated origin (meaning
-a value different than zero on the column \emph{'databases'}).
-
-\subsection{Node2Vec Parameters}
-
-For all experiments, the following parameters hold:
-\begin{itemize}
-\item For the (\texttt{p=1}, \texttt{q=1})
-\item (\texttt{dimensions=16, workers=4})
-\item (\texttt{walk\_length=5}, \texttt{num\_walks=300}, \texttt{window=5})
-\item (\texttt{min\_count=1}, \texttt{batch\_words=4})
-\item (\texttt{EdgeEmbedding=EdgeHadamardEmbedder})
-\end{itemize}
-
-\subsection{XGBoost parameters}
-\begin{itemize}
-\item \texttt{max\_depth=3}
-\item \texttt{colsample\_bytree=0.6}
-\item \texttt{eval\_metric=''auc''}
-\end{itemize}
-
-\section{Supplementary Figures}
-
-\subsection{Human Interactome}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI1}\emph{HI-II-14} results for \texttt{Node2Vec} with A2
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI2}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI3}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI1-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-A2 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI2-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-A3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI3-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-L3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance-H2}Importance gain plots for \emph{HI-TESTED}}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A2\lyxdot 2}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A3\lyxdot 2}
-\par\end{centering}
-\centering{}\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot L3\lyxdot 2}
-\end{figure}
-
-\subsection{Rice Interactome}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F2}Results for \texttt{Node2Vec} with A2 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA2_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F3}Results for \texttt{Node2Vec} with A3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA3_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F5}Results for A2 feature only}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA2_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F6}Results for A3 feature alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA3_SUMMARY}
-\end{figure}
-
-\end{document}

### 2020-12-11 nicolas/PPIPL3/bmc_template/declarations.tex
BMC-formated document

--- 2020-12-07 camila/report/declarations.tex
@@
 \section*{Acknowledgements}%% if any
-Not applicable
+N.L. would like to thank Camila Riccio and Miguel Romero-Diaz for their help 
+during the research.
 
@@
 \section*{Abbreviations}%% if any
-\textbf{HLC:} Hierarchical Link Clustering\\
-\textbf{LASSO:} Least Absolute Shrinkage Selector Operator\\
-\textbf{PCC:} Pearson Correlation Coefficient\\
-\textbf{RNA-seq:} RNA sequencing\\
-\textbf{SHT2:} Spermidine hydroxycinnamoyltransferase 2\\
-\textbf{WGCNA:} Weighted Gene Co-expression Network Analysis
+\textbf{PPI:} Protein-Protein Interaction\\
+\textbf{Y2H:} Yeast-2-Hybrid screening\\
+\textbf{AUC:} Area under the curve\\
+\textbf{ROC:} Receiver Operating Characteristic\\
+\textbf{FP:} False Positive\\
+\textbf{FN:} False Negative\\
+\textbf{TP:} True Positive\\
+\textbf{TN:} True Negative
 
@@
 \begin{itemize}
-\item RNA-seq data of salt stress in rice is available on the GEO (GSE98455).
-\item Phenotypic data of salt stress in rice is a subset of the supplementary file 1 included in ~\cite{campbell2017allelic}.
+\item Human interactome information was retrieved of the supplementary files included in ~\cite{Kovacs2019}.
+\item Rice Interactomes were downloaded from STRING database of protein interactions (4530.protein.links.detailed.v11.0.txt).
+Only interactions which have a curated origin (meaning a value different than zero on the column \emph{'databases'})
 \end{itemize}
@@
 \section*{Authors' contributions}
-J.F. and C.R. proposed the original idea. 
-J.F. provide advice on algorithms concepts and implementation.
-C.R. structured the methodology and performed the analysis. 
-C.R., J.F., and H.C.R. wrote the manuscript.
+N.L. proposed the original idea. 
+J.F. provide advice on algorithms, libraries and implementation.
+N.L. performed the implementation and analysis. 
+N.L., J.F., and C.R. wrote the manuscript.
 All authors read and approved the final manuscript.

### 2020-12-11 nicolas/PPIPL3/bmc_template/introduction.tex
BMC-formated document

@@
 an interaction between two proteins, the Y2H approach fuses one protein
-to the DB domain (known as \emph{bait}) and another protein to the
+to the DB domain (known as \emph{bait}) and the other to the
 AD (known as \emph{prey}). If the proteins interact, the reporter
@@
 for constructing PPI networks.}
-
-%\noindent \centering{}\includegraphics[width=0.7\columnwidth]{Y2H}
+\centering
+	\includegraphics[width=0.7\columnwidth]{../Y2H}
 \end{figure}
 
-Based on the outcome of numerous Y2H experiments, undirected networks
-of interactions between proteins can be constructed. However, relying
-only on experimental validation of PPIs is detrimental to research
-due to its hindrance in costs, accuracy, required manpower and time
-\cite{Laraia2015PPI,Macalino2018PPI}. Besides, the number of missing
-interactions for pairs of proteins on many organisms leverage the
-usage of computational methods for prediction of such interactions.
+Based on the outcome of Y2H experiments, numerous networks
+of interactions between proteins have been constructed \cite{Weimann2013Y2H,Rajagopala2015Y2H,Shokri2019TFY2H}.
+However, identifying protein interactions through Y2H experiments is costly in terms of time
+and resources\cite{Laraia2015PPI,Macalino2018PPI}. For an organism with 2000
+unique proteins, for example, a detailed PPI network requires the experimental
+validation of around 2 million interactions. The number of missing
+interactions for pairs of proteins on many organisms require the
+use of computational tools for predicting such interactions.
 
-A common way of predicting interactions is usually based on social
-networks analysis, more specifically on the triadic closure principle
+Different methods for predicting interactions in PPI networks have
+been proposed \cite{Chang2016PPI,Chen2019PPI,Kotlyar2015PPI}. A 
+common approach for predicting PPIs is based the triadic closure principle
 (TCP). TCP states that the higher the number of common neighboring
@@
 known as the Common Neighbors algorithm. By raising the adjacency
-matrix (A) of the network to the second power (A\texttwosuperior ),
-TCP can be considered for further analyses. However, previous studies
-show that the mentioned approach usually fails because it does not
+matrix (A) of the PPI network to the second power (A\texttwosuperior ),
+the provides an insight into which proteins may interact with each other. However, previous studies
+show that this mentioned approach alone usually fails because it does not
 consider the structural and chemical properties of the proteins \cite{Cannistraci2013Networks,Kovacs2019}.
 
-A variety of methods for predicting interactions in PPI networks have
-been proposed in recent years \cite{Chang2016PPI,Chen2019PPI,Kotlyar2015PPI}.
 Kovacs et al (2019) introduce a network-based approach which predicts
-the interaction between two proteins based on the number of paths
-of length 3, normalized by the geometric average of their interactions.
-The simplest mathematical representation of this principle consists
-on the third power of the adjacency matrix (A\textthreesuperior).
-Furthermore, Kovacs et al present a degree-normalized scaling for
-this metric, which reduces bias caused by intermediate hubs within
-the paths of length 3. This handcrafted measure, denoted L3, enables
-the proposed approach to outperform previous methods for predicting
-binary protein interactions in yeast (\emph{S. cerevisiae}), Arabidopsis
-(\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
+an interaction between two proteins based on the number of paths
+of length 3. The number of paths of length 3 is computed by raising
+the adjacency matrix of the network to the third power (A\textthreesuperior).
+The degree-normalized measure, denoted as L3,  mitigates bias caused by
+intermediate hubs within the paths of length 3. The measure L3 enables
+the approach in Kovacs et al (2019) to outperform previous methods for predicting
+binary protein interactions for various organisms, including yeast (\emph{S. cerevisiae}),
+Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
 fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus}) and humans
@@
 
-The focus of this study is to evaluate different methods for predicting
-PPIs using the network of known interactions. To achieve this, human
-and rice PPI networks are compared using the proposed methods (A2,
-A3, L3), as well as a low-dimensional representation of nodes (\texttt{Node2Vec})\cite{Grover_2016}.
-After that, combinations of the two types of methods are tested. In
-the case of the human network, two different versions of the human
-interactome \cite{Rolland2014Human} are used for training and another
-interactome version is used for validation (\emph{HI-III})\cite{Kovacs2019}.
-For the rice interactome, a portion of the known interactions removed
-and then predicted with different techniques based on the known interactions.
-The main contributions of this paper are i) a general framework for
-link prediction in non-directed networks and ii) two applications
-of this framework to biological networks with a structural insight.
+Our work extends previous analyses by using handcrafted features (A2,
+A3, L3) as well as a low-dimensional representation of nodes 
+(\texttt{Node2Vec})\cite{Grover_2016} for PPI networks in the human and rice 
+interactomes. To evaluate the proposed approach, we try to predict interactions
+using a state-of-the-art Machine Learning algorithm called \texttt{XGBoost}\cite{2016ChenXGB},
+which uses a gradient boosted decision tree model to weigh the features.
+Our main results show that adding the handcrafted features derived from the
+network connectivity to the Machine Learning model improve the prediction
+power of the models. This property can be assessed when looking at the their
+importance in the model.  The main contributions of this paper are i) a
+general framework for link prediction in non-directed networks and ii) two
+applications of this framework to biological networks with a structural
+insight.
 
-\paragraph*{Structure:} This paper is organized as follows. \emph{Materials and Methods} describes
+\paragraph*{Document structure:} This paper is organized as follows. \emph{Materials and Methods} describes
 the methodological steps and key milestones in the preparation of
@@
 \emph{Results} presents the main results of the models for the human and rice interactomes.
-\emph{Conclussions} presents the main conclusions of this study. Finally, \emph{Appendix}
+\emph{Conclusions} presents the main conclusions of this study. Finally, the \emph{appendix}
 presents supplementary information and figures which complement the

### 2020-12-11 nicolas/PPIPL3/bmc_template/materials.tex
BMC-formated document

--- 2020-12-07 nicolas/PPIPL3/ReportV5.tex
@@
-%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english]{article}
-\usepackage[T1]{fontenc}
-\usepackage[latin9]{inputenc}
-\usepackage{verbatim}
-\usepackage{float}
-\usepackage{graphicx}
-
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% A simple dot to overcome graphicx limitations
-\newcommand{\lyxdot}{.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
-\usepackage{multirow}
-%\usepackage{table}
-%\usepackage{tabular}
-
-\makeatother
-
-\usepackage{babel}
-\begin{document}
-\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
-\author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
-email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
-Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
-\maketitle
-\begin{abstract}
-
-Efforts to predict unmapped protein-protein interactions (PPI) through
-network analysis suggest that the higher the number of paths of length
-3 (L3) between two proteins, the higher the probability of interaction.
-This paper extends previous work on predicting protein interactions
-based on the L3 principle by taking into account the learning of node
-features of the PPI network. In particular, we train an \texttt{XGBoost}
-model using handcrafted features (including L3), as well as edge embeddings
-from \texttt{Node2Vec}. Our main result shows that while L3 is an
-important feature for predicting links, best performance is achieved
-when combined with edge embeddings. The proposed approach is evaluated
-for the human and rice interactomes and pretends to exploit the intrinsic
-network structure. Results for both organisms show higher AUC values
-for the combination of automated and handcrafted features than for
-each of its components alone. The handcrafted features have more importance
-for the prediction model in the human interactome than in rice. The
-developed framework can be easily applied to interactomes for other
-organisms or even to networks outside of the biological context.
-\end{abstract}
-\begin{comment}
-need include a feature that compares the dimension of the embedding
-is useful for prediction with L3 to show that they independent measures
-- we do not want to be learning L3 through node2vec
-\end{comment}
-
-\section{Introduction}
-
-\begin{comment}
-Need to add other relevant references
-\end{comment}
-
-Proteins are key actors of biological processes inside cells. As they
-carry out a variety of tasks, proteins do not work as single agents
-but rather as a part of dynamic networks of protein-protein interactions
-(PPI) \cite{Lin2017}. PPI networks play an important role in a variety
-of interdependent mechanisms, including signal transduction, homeostasis
-control, stress responses, plant defense and organ formation. They
-are also involved a variety of physiological and developmental processes,
-including protein phosphorylation, transcriptional co-factor recruitment
-and transporter activation \cite{Zhang_2010_PPI}.
-
-Different methods for predicting interactions in PPI networks have
-been proposed in recent years \cite{Chen2019PPI,Chang2016PPI,Kotlyar2015PPI}.
-Kovacs et al. (2019) introduce an approach which relies on counting
-the number of paths of length 3 (L3) to predict interactions among
-proteins for different organisms. The proposed approach outperforms
-previous methods in PPI networks of yeast (\emph{S. cerevisiae}),
-Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D.
-melanogaster}), fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus})
-and the human interactome \cite{Kovacs2019}.
-
-A common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid}
-technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
-which is based on (i) the expression of a specific reporter gene that
-activates by the binding of a DNA-binding Domain (DB), and (ii) an
-Activation Domain (AD) of a Transcription Factor that binds to an
-Upstream Activation Sequence (UAS). For the Y2H technique, a protein
-is fused to the DB domain (known as \emph{bait}) and another protein
-to the AD (known as \emph{prey}). If the proteins do not interact,
-then the reporter gene is not expressed. Otherwise, the reporter gene
-expression is activated by the activation domain.
-
-\begin{figure}[h]
-\caption{Yeast-2-Hybrid Technique}
-
-\noindent \centering{}\includegraphics[width=0.7\columnwidth]{Y2H}
-\end{figure}
-
-Based on the outcome of numerous Y2H experiments, PPI networks which
-capture known interactions for each protein can be constructed. Several
-algorithms are proposed over these networks to predict unknown interactions.
-The focus of this study is to evaluate different methods for predicting
-PPIs using the network of known interactions. These interactions are
-represented as an undirected graph. A common way of predicting interactions
-is usually based on social networks analysis, more specifically on
-the triadic closure principle (TCP). TCP states that the higher the
-number of common neighboring nodes between two nodes, the higher the
-probability that they interact \cite{Goldberg2003SmallWorld}. Previous
-studies show that the mentioned approach fails because it does not
-consider the structural and chemical properties of the proteins \cite{Kovacs2019,Cannistraci2013Networks}.
-
-\begin{comment}
-JF: Yet you do more than evaluating this three measures so this has
-to be updated.
-
-NL: Done!
-
-JF: Not sure we should use L3 above since it is being redefined as
-a normalized measure here!
-
-JF{*}: state what the results are: ``The contributions of this paper
-are...''
-
-Which article predicts PPI based on TCP? add refs
-\end{comment}
-
-For achieving the described results, human and rice PPI networks are
-compared using state-of-the-art methods, as well as the proposed ones
-(A2, A3, L3). In the case of the human network, two different versions
-of the human interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
-are used for training and another interactome version is used for
-validation (\emph{HI-III})\cite{Kovacs2019}. For the rice interactome,
-a portion of the known interactions removed and then predicted with
-different techniques based on the known interactions. The main contributions
-of this paper are i) a general framework for link prediction in non-directed
-networks and ii) two applications of this framework to biological
-networks with a structural insight
-
-\begin{comment}
-JF: read throughout the document but edited up here. I prefer to read
-the results before editing the materials and methods section. 
-\end{comment}
-
-The remainder of this paper is organized as follows. Section 2 describes
-the methodological steps and key milestones in the preparation of
-the networks, model parameters and experimental configurations. Section
-3 presents the main results of the models for the human and rice interactomes.
-Section 4 presents the main conclusions of this study. Finally, Appendix
-presents supplementary information and figures which complement the
-results described in this paper. Figures in the Appendix can be identified
-with a prefix (e.g. S\ref{F2}).
-
-\section{Materials and Methods}
-
-\subsection{Data Availability}
+\subsection*{Data Availability}
 
@@
 
-\subsection{Code Implementation}
+\subsection*{Code Implementation}
 
@@
 
-\subsection{Data Preprocessing}
+\subsection*{Data Preprocessing}
 
@@
 
-\subsection{Edge Prediction}
+\subsection*{Edge Prediction}
 
@@
 represents an interaction among proteins $v_{i}$ and $v_{j}$.
-\begin{description}
-\item [{Common~Neighbors~(A2)}] This method is based on the Triadic Closure
+%\begin{description}\paragraph*{Common~Neighbors~(A2)} 
+%\item [{Common~Neighbors~(A2)}] :This method is based on the Triadic Closure
+\paragraph*{Common~Neighbors~(A2):} This method is based on the Triadic Closure
 Principle which states that the more common friends two individuals
@@
 matrix of the network.
-\item [{Length-3~Path~Counts~(A3)}] This is a simple implementation
+%\item [{Length-3~Path~Counts~(A3)}] This is a simple implementation
+\paragraph*{Length-3~Path~Counts~(A3):} This is a simple implementation
 of the notion of ``if my friends and your friends interact, then
@@
 is calculated.
-\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
+%\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
+\paragraph*{Degree-normalized~L-3~Score~(L3):} The previous approach might
 overestimate the importance of some edges due to intermediate hubs
@@
 nodes $i$ and $j$.
-\end{description}
+%\end{description}
 
-\subsection{Sampling Procedure}
+\subsection*{Sampling Procedure}
 
@@
 
-\subsection{Feature Extraction with \texttt{Node2Vec}}
+\subsection*{Feature Extraction with \texttt{Node2Vec}}
 
@@
 
-\subsection{Handcrafted Features}
+\subsection*{Handcrafted Features}
 
@@
 
-\subsection{Feature to Predict: Existence}
+\subsection*{Feature to Predict: Existence}
 
@@
 
-\subsection{Machine Learning Algorithm}
+\subsection*{Machine Learning Algorithm}
 
@@
 
-\subsection{Result Validation}
+\subsection*{Result Validation}
 
@@
 over 0.8 of AUC are considered good.
-
-\section{Results and Discussion}
-
-\subsection{Human Interactome}
-
-As the first step in the evaluation of A2, A3 and L3 methods for link
-prediction, the full calculation and ranking of all predictions was
-carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
-and \emph{HI-TESTED}) is shown in Figure \ref{fig:HI1}, in which
-the x axis represents a rank position $k$ and the y axis displays
-the precision for the top $k$ predictions of each method, when assuming
-interactome \emph{HI-III} as the validation set.
-
-\begin{figure}[h]
-\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14} and \emph{HI-TESTED}}
-
-\noindent \centering{}\includegraphics[width=0.48\columnwidth]{hi-ii-14\lyxdot txt}\includegraphics[width=0.48\columnwidth]{hi-tested\lyxdot txt}
-\end{figure}
-
-As it can be inferred from the plots, L3-based predictions outperform
-their $A{{}^2}$ counterparts. Results also show that L3-score and
-$A^{3}$predictions follow a very similar trend.
-
-However, it is of our interest to assess if Machine Learning can boost
-the overall performance on the prediction, so the proposed procedure
-was adapted to the human interactome data: Instead of having to remove
-a fraction of edges from the network in order to predict and validate,
-the prediction network and the validation networks are given beforehand.
-The rest of the procedure is carried out as mentioned in Section 2.
-Table \ref{T1} presents the results for the combinations of Node2Vec
-with each feature.
-
-\begin{table}
-\caption{\label{T1}Summary statistics for human interactomes}
-\includegraphics[width=1\columnwidth]{T1}
-\end{table}
-
-In general, performance of all models is above 0.96 in terms of F1-Score.
-Although marginally, the A2 feature-enriched model for HI-TESTED performs
-better than their counterparts based on paths of length 3. The opposite
-is also true for HI-II-14: models enriched with L3 and A3 features
-perform better than their A2 counterpart. Results for the Interactome
-\emph{HI-II-14} with Node2Vec and each metric are presented in Figures
-S\ref{HI1}, S\ref{HI2} and S\ref{HI3}. It can be seen that for
-\emph{HI-II-14}, A2 performs marginally better than L3 when comparing
-precision but worse in terms of recall, although all metrics have
-AUC values of 0.99.
-
-On the other hand, the interactome \emph{HI-TESTED} was assessed with
-the same methodology as \emph{HI-II-14} and results are shown in Figures
-S\ref{HI1-1}, S\ref{HI2-1} and S\ref{HI3-1}. For this interactome,
-A2 performs better in terms of false positives than A3 and L3, but
-in terms of false negatives, A3 actually performs better that A2 and
-L3. In terms of AUC, metrics results are indistinguishable.
-
-An interesting assessment from this human datasets can be observed
-when looking at the importance plots, which show a very biased influence
-of the handcrafted features. Figure \ref{F8-importance-H1} shows
-this behavior for the \emph{HI-II-14} dataset, which is very similar
-to \emph{HI-TESTED} (Figure S\ref{F8-importance-H2}).
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance-H1}Importance gain plots for \emph{HI-II-14}}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A2\lyxdot 1}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A3\lyxdot 1}
-\par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot L3\lyxdot 1}
-\end{figure}
-
-\subsection{Rice Interactome}
-
-For the rice interactome, different combinations of features were
-trained on the model and validated. The results are shown in Table
-\ref{T2}. First, prediction based on \texttt{Node2Vec} without any
-additional feature included is used as baseline of comparison. Figure
-\ref{F1} shows those results, and one can see that its mean performance
-using the AUC is $0.93$. The results for all 10 repetitions of each
-model are consistent, meaning the variance among each line is not
-significant. The results for each model using only the default features
-are satisfactory. This result can also be assessed when looking at
-the confusion matrix, where a precision of $0.98$ and a recall of
-$0.86$ are obtained.
-
-\begin{table}
-\caption{\label{T2}Summary statistics for rice interactome}
-\includegraphics[width=1\columnwidth]{T2}
-\end{table}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F1}Summary ROC curves for \texttt{Node2Vec} model alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_ML/ROConlyML_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_ML/CMonlyML_SUMMARY}
-\end{figure}
-
-The next experiment carried out was to add for each prediction method
-score as a prediction feature, that is, use the score as an additional
-feature for the ML algorithm. Results for the Common Neighbors method
-(\textbf{A2}), which uses paths of length 2, are shown in Figure S\ref{F2}.
-All 10 experiments have small variability among them and perform better
-than the baseline, with AUC of $0.98$. When looking at the confusion
-matrix for this model, most guesses are true positives and true negatives,
-resulting in a precision of $0.99$ and a recall of $0.91$. The same
-analyses are done with the count of paths of length 3 (\textbf{A3})
-and with the degree-normalized length-3 score (\textbf{L3}). These
-results are presented in figures S\ref{F3} and \ref{F4}, respectively.
-The corresponding AUC are $0.97$ and $0.98$.
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F4}Results for \texttt{Node2Vec} with L3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CML3_SUMMARY}
-\end{figure}
-
-The precision when using the count of paths of length 3 (A3) is $0.99$
-while the recall is $0.90$. It is interesting to observe that precision
-decreased $0.16\%$ and recall decreased $0.58\%$. A similar situation
-is obtained for the model with the normalized score (L3), whose precision
-decreased to $0.99$ ($-0.08\%$) and recall decreased to $0.88$
-($-2.38\%$). It can be seen also that the ROC curve achieves high
-values of true positive rate sooner in the A2-featured model than
-on the other models.
-
-Another relevant assessment was carried out to verify the predictive
-power of the metrics. \texttt{XGBoost} models were trained only using
-each metric, for the same 10 experiments. Results are shown in Figures
-S\ref{F5}, S\ref{F6} and \ref{F7}. XGBoost yields lower AUC values
-when only considering A2, A3 and L3. Precision and recall for this
-assessments, although lower than the previous experiments, are still
-considered satisfactory.
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F7}Results for L3 feature alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyL3_SUMMARY}
-\end{figure}
-
-Finally, an relevant evaluation on the models that include a handcrafted
-feature is necessary: How important is the appended feature for the
-model result? The answer can be observed by the feature importance
-plots for A2, A3 and L3 in Figure \ref{F8-importance}. Each plot
-resembles the importance gain from each feature in the underlying
-decision trees that \texttt{XGBoost} uses. Usually, the more bifurcations,
-the higher gain a feature has, the more relevance a feature has on
-the model prediction itself. All models have the appended feature
-as the most important by a good margin, although margin for A3 is
-not wide as for A2 or L3.
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance}Importance gain plots for \texttt{Node2Vec}
-with each feature}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
-\par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
-\end{figure}
-
-\section{Conclusions}
-
-Based on previous work from other authors\cite{Kovacs2019}, the intrinsic
-knowledge of existing protein-protein interaction networks let scientists
-elucidate promising new interactions to explore. In this study, a
-machine learning algorithm was fed with previously elucidated information
-about the network, as well as results form a state-of-the-art edge
-embedding technique. Even if the baseline of the handcrafted features
-(A2, A3, L3) yields good results, boosting that information with the
-vector embedding, which calculates vector representations of the network
-structure improves the overall performance of each method on its own.
-
-On both human interactomes (\emph{HI-II-14} and \emph{HI-TESTED}),
-applying \texttt{XGBoost} with the metrics yield results with high
-AUC for the link prediction problem, and therefore this study extends
-previous work by other authors that only consider the predictive power
-of the metrics alone. In this case, the importance of the handcrafted
-features was very significant and stands out of the Node2Vec features.
-
-On the rice interactome, PPI prediction obtained high AUC values when
-only using the node embedding technique (AUC=0.93) and when only using
-the handcrafted features based on protein neighborhood (AUC \textasciitilde 0.88).
-Even higher AUC results were achieved when the handcrafted features
-and the vector representation information were combined (AUC \textasciitilde 0.98).
-
-Importance values for the A3 method stand out less from the node embedding
-information than in A2 and L3. In this case, the importance gain of
-the handcrafted features in each model was not as high in rice as
-in the human interactome.
-
-With this study, a framework for link prediction on protein-protein
-interactions is presented and validated against the available information
-on rice and human interactomes. However, it might be easily extended
-to other organisms or even to other graphs not necessarily related
-to biology, because the only required information is the network connectivity.
-Satisfactory results might be expected when the framework is applied
-to networks with a dense adjacency matrix, since calculations of L3
-and Node2Vec rely on network exploration for an arbitrary node and
-its degree.
-
-\bibliographystyle{plain}
-\bibliography{refs}
-
-\pagebreak{}
-
-\appendix
-
-\part{Appendix}
-
-\section{Supplementary Information}
-
-\subsection{Datasets}
-
-In the case of the human interactomes, the files were downloaded as-is
-from the repository from Kovacs et al (2019). For the rice interactome,
-the downloaded file was \emph{4530.protein.links.detailed.v11.0.txt.}
-From this file, only interactions which have a curated origin (meaning
-a value different than zero on the column \emph{'databases'}).
-
-\subsection{Node2Vec Parameters}
-
-For all experiments, the following parameters hold:
-\begin{itemize}
-\item For the (\texttt{p=1}, \texttt{q=1})
-\item (\texttt{dimensions=16, workers=4})
-\item (\texttt{walk\_length=5}, \texttt{num\_walks=300}, \texttt{window=5})
-\item (\texttt{min\_count=1}, \texttt{batch\_words=4})
-\item (\texttt{EdgeEmbedding=EdgeHadamardEmbedder})
-\end{itemize}
-
-\subsection{XGBoost parameters}
-\begin{itemize}
-\item \texttt{max\_depth=3}
-\item \texttt{colsample\_bytree=0.6}
-\item \texttt{eval\_metric=''auc''}
-\end{itemize}
-
-\section{Supplementary Figures}
-
-\subsection{Human Interactome}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI1}\emph{HI-II-14} results for \texttt{Node2Vec} with A2
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI2}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI3}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI1-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-A2 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI2-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-A3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI3-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-L3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance-H2}Importance gain plots for \emph{HI-TESTED}}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A2\lyxdot 2}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A3\lyxdot 2}
-\par\end{centering}
-\centering{}\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot L3\lyxdot 2}
-\end{figure}
-
-\subsection{Rice Interactome}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F2}Results for \texttt{Node2Vec} with A2 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA2_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F3}Results for \texttt{Node2Vec} with A3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA3_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F5}Results for A2 feature only}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA2_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F6}Results for A3 feature alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA3_SUMMARY}
-\end{figure}
-
-\end{document}

### 2020-12-11 nicolas/PPIPL3/bmc_template/results.tex
BMC-formated document

--- 2020-12-07 nicolas/PPIPL3/ReportV5.tex
@@
-%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english]{article}
-\usepackage[T1]{fontenc}
-\usepackage[latin9]{inputenc}
-\usepackage{verbatim}
-\usepackage{float}
-\usepackage{graphicx}
-
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% A simple dot to overcome graphicx limitations
-\newcommand{\lyxdot}{.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
-\usepackage{multirow}
-%\usepackage{table}
-%\usepackage{tabular}
-
-\makeatother
-
-\usepackage{babel}
-\begin{document}
-\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
-\author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
-email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
-Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
-\maketitle
-\begin{abstract}
-
-Efforts to predict unmapped protein-protein interactions (PPI) through
-network analysis suggest that the higher the number of paths of length
-3 (L3) between two proteins, the higher the probability of interaction.
-This paper extends previous work on predicting protein interactions
-based on the L3 principle by taking into account the learning of node
-features of the PPI network. In particular, we train an \texttt{XGBoost}
-model using handcrafted features (including L3), as well as edge embeddings
-from \texttt{Node2Vec}. Our main result shows that while L3 is an
-important feature for predicting links, best performance is achieved
-when combined with edge embeddings. The proposed approach is evaluated
-for the human and rice interactomes and pretends to exploit the intrinsic
-network structure. Results for both organisms show higher AUC values
-for the combination of automated and handcrafted features than for
-each of its components alone. The handcrafted features have more importance
-for the prediction model in the human interactome than in rice. The
-developed framework can be easily applied to interactomes for other
-organisms or even to networks outside of the biological context.
-\end{abstract}
-\begin{comment}
-need include a feature that compares the dimension of the embedding
-is useful for prediction with L3 to show that they independent measures
-- we do not want to be learning L3 through node2vec
-\end{comment}
-
-\section{Introduction}
-
-\begin{comment}
-Need to add other relevant references
-\end{comment}
-
-Proteins are key actors of biological processes inside cells. As they
-carry out a variety of tasks, proteins do not work as single agents
-but rather as a part of dynamic networks of protein-protein interactions
-(PPI) \cite{Lin2017}. PPI networks play an important role in a variety
-of interdependent mechanisms, including signal transduction, homeostasis
-control, stress responses, plant defense and organ formation. They
-are also involved a variety of physiological and developmental processes,
-including protein phosphorylation, transcriptional co-factor recruitment
-and transporter activation \cite{Zhang_2010_PPI}.
-
-Different methods for predicting interactions in PPI networks have
-been proposed in recent years \cite{Chen2019PPI,Chang2016PPI,Kotlyar2015PPI}.
-Kovacs et al. (2019) introduce an approach which relies on counting
-the number of paths of length 3 (L3) to predict interactions among
-proteins for different organisms. The proposed approach outperforms
-previous methods in PPI networks of yeast (\emph{S. cerevisiae}),
-Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D.
-melanogaster}), fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus})
-and the human interactome \cite{Kovacs2019}.
-
-A common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid}
-technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
-which is based on (i) the expression of a specific reporter gene that
-activates by the binding of a DNA-binding Domain (DB), and (ii) an
-Activation Domain (AD) of a Transcription Factor that binds to an
-Upstream Activation Sequence (UAS). For the Y2H technique, a protein
-is fused to the DB domain (known as \emph{bait}) and another protein
-to the AD (known as \emph{prey}). If the proteins do not interact,
-then the reporter gene is not expressed. Otherwise, the reporter gene
-expression is activated by the activation domain.
-
-\begin{figure}[h]
-\caption{Yeast-2-Hybrid Technique}
-
-\noindent \centering{}\includegraphics[width=0.7\columnwidth]{Y2H}
-\end{figure}
-
-Based on the outcome of numerous Y2H experiments, PPI networks which
-capture known interactions for each protein can be constructed. Several
-algorithms are proposed over these networks to predict unknown interactions.
-The focus of this study is to evaluate different methods for predicting
-PPIs using the network of known interactions. These interactions are
-represented as an undirected graph. A common way of predicting interactions
-is usually based on social networks analysis, more specifically on
-the triadic closure principle (TCP). TCP states that the higher the
-number of common neighboring nodes between two nodes, the higher the
-probability that they interact \cite{Goldberg2003SmallWorld}. Previous
-studies show that the mentioned approach fails because it does not
-consider the structural and chemical properties of the proteins \cite{Kovacs2019,Cannistraci2013Networks}.
-
-\begin{comment}
-JF: Yet you do more than evaluating this three measures so this has
-to be updated.
-
-NL: Done!
-
-JF: Not sure we should use L3 above since it is being redefined as
-a normalized measure here!
-
-JF{*}: state what the results are: ``The contributions of this paper
-are...''
-
-Which article predicts PPI based on TCP? add refs
-\end{comment}
-
-For achieving the described results, human and rice PPI networks are
-compared using state-of-the-art methods, as well as the proposed ones
-(A2, A3, L3). In the case of the human network, two different versions
-of the human interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
-are used for training and another interactome version is used for
-validation (\emph{HI-III})\cite{Kovacs2019}. For the rice interactome,
-a portion of the known interactions removed and then predicted with
-different techniques based on the known interactions. The main contributions
-of this paper are i) a general framework for link prediction in non-directed
-networks and ii) two applications of this framework to biological
-networks with a structural insight
-
-\begin{comment}
-JF: read throughout the document but edited up here. I prefer to read
-the results before editing the materials and methods section. 
-\end{comment}
-
-The remainder of this paper is organized as follows. Section 2 describes
-the methodological steps and key milestones in the preparation of
-the networks, model parameters and experimental configurations. Section
-3 presents the main results of the models for the human and rice interactomes.
-Section 4 presents the main conclusions of this study. Finally, Appendix
-presents supplementary information and figures which complement the
-results described in this paper. Figures in the Appendix can be identified
-with a prefix (e.g. S\ref{F2}).
-
-\section{Materials and Methods}
-
-\subsection{Data Availability}
-
-The human interactome data was downloaded from the repository of the
-length-3 degree normalized paths methodology \cite{Kovacs2019}: the
-dataset \emph{HI-II-14} and \emph{HI-TESTED} are used for prediction.
-The dataset \emph{HI-III} is available in the same repository and
-is used for validation.
-
-The rice interactome information was downloaded from the STRING database
-\cite{Szklarczyk2019}, corresponding to the subspecies \emph{Oryza
-sativa}. The file contains more than 8 million PPIs from several sources.
-For the purpose of this study, only PPIs with evidence from curated
-databases are used. The resulting network contains $5025$ nodes and
-$164420$ edges.
-
-\subsection{Code Implementation}
-
-The original code implementation in \texttt{C++} from Kovacs et al
-(2019) was adapted to \texttt{Python} (V3.6). For the purpose of algorithmic
-validation, the three methods were implemented from scratch with basic
-functionalities and data structures of the \texttt{Python} language.
-The code is publicly available under the repository github.com/ocinlr/PPIPL3.
-
-\subsection{Data Preprocessing}
-
-Information for the human interactome was used as-is, which corresponds
-to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
-$9433$ and $23322$ interactions, respectively.
-
-For the rice interactome, an additional preprocessing was performed.
-The filtered network for rice consists of $5025$ proteins (nodes)
-and $164420$ interactions (edges) distributed among $178$ connected
-components. The connected component with the greatest number of edges
-was selected. The extracted connected component consists of $n=4390$
-nodes and $m=163319$ edges, which corresponds to $99.33$\% of edges.
-
-\subsection{Edge Prediction}
-
-For predicting interactions for each network, the algorithms described
-below were used. It is important to keep in mind how the PPI network
-$G=(V,E)$ is conceptualized: each node ($v_{i}\in V$) represents
-a protein and each undirected edge ($e_{b}=\{v_{i},v_{j}\},\,e_{b}\in E$)
-represents an interaction among proteins $v_{i}$ and $v_{j}$.
-\begin{description}
-\item [{Common~Neighbors~(A2)}] This method is based on the Triadic Closure
-Principle which states that the more common friends two individuals
-have, the more likely that they know each other. For implementing
-this measure, $A{{}^2}$ matrix is calculated, where $A$ is the adjacency
-matrix of the network.
-\item [{Length-3~Path~Counts~(A3)}] This is a simple implementation
-of the notion of ``if my friends and your friends interact, then
-we might interact too''. For implementing this measure, $A{{}^3}$
-is calculated.
-\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
-overestimate the importance of some edges due to intermediate hubs
-which add many shortcuts in the graph. To address that issue, a degree
-normalization for the path $v_{i}\rightarrow v_{m}\rightarrow v_{n}\rightarrow v_{j}$
-is applied by considering the degree of the intermediate nodes $k_{m}$
-and $k_{n}$, as follows.
-\[
-p_{i,j}=\sum_{m,n\in|V|}\frac{A_{i,m}\cdot A_{m,n}\cdot A_{n,j}}{\sqrt{k_{m}\cdot k_{n}}}
-\]
-\\
-where $A_{i,j}$ represents the value of the adjacency matrix for
-nodes $i$ and $j$.
-\end{description}
-
-\subsection{Sampling Procedure}
-
-For each PPI network, the following procedure was performed 10 times
-to address the stochastic nature of the process:
-\begin{enumerate}
-\item A percentage of interactions is removed at random from the network
-(20\%).
-\item The same amount of removed interactions are then predicted using the
-main methods for prediction mentioned by Kovacs et al (2019): Common
-Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
-and the Length-3 degree-normalized score (\textbf{L3}).
-\item A test dataset is created as follows: all removed edges are included
-(as observed positives for the ML algorithm) and from the predicted
-edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
-the previous classification (observed negatives), a random subset
-is chosen such that the dataset is balanced, that is, the amount of
-observed positive labels is equal to the observed negative labels.
-\item Once the dataset is ready, it is randomly partitioned: 80\% is used
-for \texttt{XGBoost} model training and 20\% is used for validation.
-It is important to have in mind that balanced distribution of the
-positive and negative labels in the datasets was satisfied in training,
-but validation is performed using the remaining unbalanced data.
-\end{enumerate}
-It is worth mentioning that some exploration experiments were carried
-out with interactions removal percentages of 2, 5, 10 and 20\%, as
-well as train-test partitions of 15-85, 80-20 and 75-25, and the explained
-parameter set was selected because it either represented a marginal
-gain (results not shown) or because it was a common parameter selection
-in related literature.
-
-\subsection{Feature Extraction with \texttt{Node2Vec}}
-
-The \texttt{Node2Vec} module was used for extracting features of the
-interactomes. The considerations for the model were:
-\begin{itemize}
-\item All paths in the random walks are equally likely.
-\item Use a modest number of dimensions and threads for calculation.
-\item Since length-3 paths are the defining property in this study, there
-is no necessity for longer walks. However, it is important to try
-out many possible redundant routes and to consider a window of at
-least 4.
-\item Other standard parameters were left with default values.
-\item Edge embeddings were calculated using a geometric ratio of the node
-embeddings.
-\end{itemize}
-
-\subsection{Handcrafted Features}
-
-Due to the poor results of the \emph{raw} Length-3 counting (\textbf{A3}),
-a different approach for this information was carried out: As it still
-gives a lot of information that might be useful for a predictive routine,
-this counting was normalized (dividing by the greatest counting in
-the \textbf{A3} top predictions) and then used as a feature for the
-Machine Learning algorithm. For completeness, \textbf{A2} and \textbf{L3}
-measures were used as a input features. Finally, the case were no
-handcrafted feature was also considered, that is, only the features
-extracted from the structure of the network were considered.
-
-\subsection{Feature to Predict: Existence}
-
-The feature to predict corresponds to the possible existence (\emph{True/False})
-of a link based on the existing information of the network. This is
-evaluated by taking out a fraction of the edges and then trying to
-predict for a given set of possible edges if they have a high probability
-to belong to the original network.
-
-\subsection{Machine Learning Algorithm}
-
-The Extreme Gradient Boosting implementation of gradient boosted trees
-is applied to evaluate the existence of an edge. Gradient boosted
-trees are usually used for supervised learning problems, where the
-training data $X_{i}$ has multiple features and pretends to explain
-(or predict) a target variable $Y_{i}$. The corresponding implementation
-applied for this study is \texttt{XGBoost}.
-
-\subsection{Result Validation}
-
-As mentioned before, 80\% of the final dataset was randomly selected
-and used for training, while the remaining information was used for
-validation. The whole training-validation procedure was applied 10
-times.
-
-The chosen metric for validation was the Area under the Curve (\textbf{AUC})
-of the Receiver Operating Characteristic (\textbf{ROC}). This curve
-corresponds to plot the sensitivity (probability of predicting a real
-positive as positive) against 1-specificity (probability of predicting
-a real negative as positive). AUC values of 1 represent a perfect
-prediction and 0.5 corresponds to a random guess. Normally, values
-over 0.8 of AUC are considered good.
-
-\section{Results and Discussion}
-
-\subsection{Human Interactome}
+\subsection*{Human Interactome}
 
@@
 \caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14} and \emph{HI-TESTED}}
-
-\noindent \centering{}\includegraphics[width=0.48\columnwidth]{hi-ii-14\lyxdot txt}\includegraphics[width=0.48\columnwidth]{hi-tested\lyxdot txt}
+	\includegraphics[width=\textwidth ]{figures/figure2.eps}
+	%\includegraphics[width=0.7\textwidth ]{../hi-tested.txt.}
 \end{figure}
@@
 \caption{\label{T1}Summary statistics for human interactomes}
-\includegraphics[width=1\columnwidth]{T1}
+\includegraphics[width=1\columnwidth]{figures/T1}
 \end{table}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 \noindent \begin{centering}
@@
 \begin{centering}
-\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A2\lyxdot 1}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot A3\lyxdot 1}
+\includegraphics[width=0.48\columnwidth]{figures/Human/Imp\lyxdot A2\lyxdot 1\lyxdot eps}\includegraphics[width=0.48\columnwidth]{figures/Human/Imp\lyxdot A3\lyxdot 1\lyxdot eps}
 \par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{Human/Imp\lyxdot L3\lyxdot 1}
+\centering{}\includegraphics[width=0.48\columnwidth]{figures/Human/Imp\lyxdot L3\lyxdot 1}
 \end{figure}
 
-\subsection{Rice Interactome}
+\subsection*{Rice Interactome}
 
@@
 \caption{\label{T2}Summary statistics for rice interactome}
-\includegraphics[width=1\columnwidth]{T2}
+\includegraphics[width=1\columnwidth]{figures/T2.eps}
 \end{table}
 
-\begin{figure}[H]
+\begin{figure}[h]
 \noindent \begin{centering}
@@
 \par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_ML/ROConlyML_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_ML/CMonlyML_SUMMARY}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Only_ML/ROConlyML_SUMMARY.eps}\includegraphics[width=0.48\columnwidth]{figures/Only_ML/CMonlyML_SUMMARY.eps}
 \end{figure}
@@
 \par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CML3_SUMMARY}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/ROCL3_SUMMARY.eps}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/CML3_SUMMARY.eps}
 \end{figure}
@@
 \par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyL3_SUMMARY}
+\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Only_Metric/ROConlyL3_SUMMARY.eps}\includegraphics[width=0.48\columnwidth]{figures/Only_Metric/CMonlyL3_SUMMARY.eps}
 \end{figure}
@@
 \begin{centering}
-\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
+\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/Imp\lyxdot A2\lyxdot All.eps}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/Imp\lyxdot A3\lyxdot All.eps}
 \par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
+\centering{}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/Imp\lyxdot L3\lyxdot All.eps}
 \end{figure}
-
-\section{Conclusions}
-
-Based on previous work from other authors\cite{Kovacs2019}, the intrinsic
-knowledge of existing protein-protein interaction networks let scientists
-elucidate promising new interactions to explore. In this study, a
-machine learning algorithm was fed with previously elucidated information
-about the network, as well as results form a state-of-the-art edge
-embedding technique. Even if the baseline of the handcrafted features
-(A2, A3, L3) yields good results, boosting that information with the
-vector embedding, which calculates vector representations of the network
-structure improves the overall performance of each method on its own.
-
-On both human interactomes (\emph{HI-II-14} and \emph{HI-TESTED}),
-applying \texttt{XGBoost} with the metrics yield results with high
-AUC for the link prediction problem, and therefore this study extends
-previous work by other authors that only consider the predictive power
-of the metrics alone. In this case, the importance of the handcrafted
-features was very significant and stands out of the Node2Vec features.
-
-On the rice interactome, PPI prediction obtained high AUC values when
-only using the node embedding technique (AUC=0.93) and when only using
-the handcrafted features based on protein neighborhood (AUC \textasciitilde 0.88).
-Even higher AUC results were achieved when the handcrafted features
-and the vector representation information were combined (AUC \textasciitilde 0.98).
-
-Importance values for the A3 method stand out less from the node embedding
-information than in A2 and L3. In this case, the importance gain of
-the handcrafted features in each model was not as high in rice as
-in the human interactome.
-
-With this study, a framework for link prediction on protein-protein
-interactions is presented and validated against the available information
-on rice and human interactomes. However, it might be easily extended
-to other organisms or even to other graphs not necessarily related
-to biology, because the only required information is the network connectivity.
-Satisfactory results might be expected when the framework is applied
-to networks with a dense adjacency matrix, since calculations of L3
-and Node2Vec rely on network exploration for an arbitrary node and
-its degree.
-
-\bibliographystyle{plain}
-\bibliography{refs}
-
-\pagebreak{}
-
-\appendix
-
-\part{Appendix}
-
-\section{Supplementary Information}
-
-\subsection{Datasets}
-
-In the case of the human interactomes, the files were downloaded as-is
-from the repository from Kovacs et al (2019). For the rice interactome,
-the downloaded file was \emph{4530.protein.links.detailed.v11.0.txt.}
-From this file, only interactions which have a curated origin (meaning
-a value different than zero on the column \emph{'databases'}).
-
-\subsection{Node2Vec Parameters}
-
-For all experiments, the following parameters hold:
-\begin{itemize}
-\item For the (\texttt{p=1}, \texttt{q=1})
-\item (\texttt{dimensions=16, workers=4})
-\item (\texttt{walk\_length=5}, \texttt{num\_walks=300}, \texttt{window=5})
-\item (\texttt{min\_count=1}, \texttt{batch\_words=4})
-\item (\texttt{EdgeEmbedding=EdgeHadamardEmbedder})
-\end{itemize}
-
-\subsection{XGBoost parameters}
-\begin{itemize}
-\item \texttt{max\_depth=3}
-\item \texttt{colsample\_bytree=0.6}
-\item \texttt{eval\_metric=''auc''}
-\end{itemize}
-
-\section{Supplementary Figures}
-
-\subsection{Human Interactome}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI1}\emph{HI-II-14} results for \texttt{Node2Vec} with A2
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI2}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI3}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
-feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI1-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-A2 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI2-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-A3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{HI3-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
-L3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3\lyxdot 2}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F8-importance-H2}Importance gain plots for \emph{HI-TESTED}}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A2\lyxdot 2}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A3\lyxdot 2}
-\par\end{centering}
-\centering{}\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot L3\lyxdot 2}
-\end{figure}
-
-\subsection{Rice Interactome}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F2}Results for \texttt{Node2Vec} with A2 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA2_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F3}Results for \texttt{Node2Vec} with A3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA3_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F5}Results for A2 feature only}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA2_SUMMARY}
-\end{figure}
-
-\begin{figure}[H]
-\noindent \begin{centering}
-\caption{\label{F6}Results for A3 feature alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA3_SUMMARY}
-\end{figure}
-
-\end{document}

### 2020-12-28 nicolas/PPIPL3/bmc_template/abstract.tex
Structure change

@@
-\parttitle{Background} %if any
-Abstract A recent study in network-based prediction of protein-protein interactions (PPIs) reveals that two proteins are more likely to interact, the higher the number of paths of length 3 between them (normalized by the geometric average of their interactions).
-This paper extends previous work on mapping binary interactions by taking into account the learning of features (embeddings) of the PPI network.
-In particular, we implement a gradient boosted decision tree model (XGBoost) using handcrafted features (including the normalized measure) and embeddings from an algorithm that generates a low-dimensional representation of nodes (node2vec).
+\parttitle{Background}
+Recent research on network-based prediction of protein-protein
+interactions (PPIs) reveals that two proteins are more likely to interact,
+the higher the number of paths of length 3 between them exist. This paper
+extends previous work on mapping binary interactions between proteins 
+by learning features (i.e. embeddings) of the PPI network. In particular, 
+this paper reports on an application of gradient boosted decision tree
+model (XGBoost) using handcrafted features  such as normalized measures 
+and algorithmic embeddings that generate a low-dimensional representation
+of the proteins and their interactions (node2vec).
 
-\parttitle{Results} %if any
-Our main result shows that while the measure remains an important feature for predicting interactions, better performance is achieved when in addition embedding features are considered.
-The proposed approach is validated for the human and rice interactomes.
-For both cases, the combination of both types of features yield higher AUC values.
+\parttitle{Results}
+The main result shows that while the measure remains an important feature
+for predicting interactions, an improvement in performance is achieved when
+embedding features are considered. The proposed approach is validated for
+the human and rice interactomes. For both cases, the combination of both
+types of features yield higher performance in terms of the Area Under the
+Curve metric (i.e. AUC).
 
-\parttitle{Conclusions} %if any
-As found on this study on both human and rice, when information from handcrafted features based on neighborhood is enhanced with vector representations from random walks, the prediction power of the model improves.
-Besides, a supervised learning model can be trained for predicting unknown interactions based on such information.
-Finally, the developed framework can also be applied to interactomes of other organisms for which PPI networks have recently become available.
+\parttitle{Conclusions}
+As witnessed by both human and rice, when information from 
+handcrafted features based on neighborhood is enhanced with vector
+representations from random walks, the prediction power of the model
+improves. Moreover, a supervised learning model has been trained to predict
+unknown interactions based on such information. Finally, the developed
+framework could also be applied to interactomes of other organisms for which
+PPI networks have recently become available.

### 2020-12-28 nicolas/PPIPL3/bmc_template/case.tex
Structure change

--- 2020-12-11 nicolas/PPIPL3/bmc_template/results.tex
@@
-\subsection*{Human Interactome}
-
-As the first step in the evaluation of A2, A3 and L3 methods for link
-prediction, the full calculation and ranking of all predictions was
-carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
-and \emph{HI-TESTED}) is shown in Figure \ref{fig:HI1}, in which
-the x axis represents a rank position $k$ and the y axis displays
-the precision for the top $k$ predictions of each method, when assuming
-interactome \emph{HI-III} as the validation set.
-
-\begin{figure}[h]
-\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14} and \emph{HI-TESTED}}
-	\includegraphics[width=\textwidth ]{figures/figure2.eps}
-	%\includegraphics[width=0.7\textwidth ]{../hi-tested.txt.}
-\end{figure}
-
-As it can be inferred from the plots, L3-based predictions outperform
-their $A{{}^2}$ counterparts. Results also show that L3-score and
-$A^{3}$predictions follow a very similar trend.
-
-However, it is of our interest to assess if Machine Learning can boost
-the overall performance on the prediction, so the proposed procedure
-was adapted to the human interactome data: Instead of having to remove
-a fraction of edges from the network in order to predict and validate,
-the prediction network and the validation networks are given beforehand.
-The rest of the procedure is carried out as mentioned in Section 2.
-Table \ref{T1} presents the results for the combinations of Node2Vec
-with each feature.
-
-\begin{table}
-\caption{\label{T1}Summary statistics for human interactomes}
-\includegraphics[width=1\columnwidth]{figures/T1}
-\end{table}
-
-In general, performance of all models is above 0.96 in terms of F1-Score.
-Although marginally, the A2 feature-enriched model for HI-TESTED performs
-better than their counterparts based on paths of length 3. The opposite
-is also true for HI-II-14: models enriched with L3 and A3 features
-perform better than their A2 counterpart. Results for the Interactome
-\emph{HI-II-14} with Node2Vec and each metric are presented in Figures
-S\ref{HI1}, S\ref{HI2} and S\ref{HI3}. It can be seen that for
-\emph{HI-II-14}, A2 performs marginally better than L3 when comparing
-precision but worse in terms of recall, although all metrics have
-AUC values of 0.99.
-
-On the other hand, the interactome \emph{HI-TESTED} was assessed with
-the same methodology as \emph{HI-II-14} and results are shown in Figures
-S\ref{HI1-1}, S\ref{HI2-1} and S\ref{HI3-1}. For this interactome,
-A2 performs better in terms of false positives than A3 and L3, but
-in terms of false negatives, A3 actually performs better that A2 and
-L3. In terms of AUC, metrics results are indistinguishable.
-
-An interesting assessment from this human datasets can be observed
-when looking at the importance plots, which show a very biased influence
-of the handcrafted features. Figure \ref{F8-importance-H1} shows
-this behavior for the \emph{HI-II-14} dataset, which is very similar
-to \emph{HI-TESTED} (Figure S\ref{F8-importance-H2}).
-
-\begin{figure}[h]
-\noindent \begin{centering}
-\caption{\label{F8-importance-H1}Importance gain plots for \emph{HI-II-14}}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.48\columnwidth]{figures/Human/Imp\lyxdot A2\lyxdot 1\lyxdot eps}\includegraphics[width=0.48\columnwidth]{figures/Human/Imp\lyxdot A3\lyxdot 1\lyxdot eps}
-\par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{figures/Human/Imp\lyxdot L3\lyxdot 1}
-\end{figure}
-
-\subsection*{Rice Interactome}
+% case.tex
 

### 2020-12-28 nicolas/PPIPL3/bmc_template/conclusions.tex
Structure change

@@
 only using the node embedding technique (AUC=0.93) and when only using
-the handcrafted features based on protein neighborhood (AUC \textasciitilde 0.88).
+the handcrafted features based on protein neighborhood (AUC $\sim$ 0.88).
 Even higher AUC results were achieved when the handcrafted features
-and the vector representation information were combined (AUC \textasciitilde 0.98).
+and the vector representation information were combined (AUC $\sim$ 0.98).
 

### 2020-12-28 nicolas/PPIPL3/bmc_template/framework.tex
Structure change

--- 2020-12-11 nicolas/PPIPL3/bmc_template/results.tex
@@
-\subsection*{Human Interactome}
+% Model framework - similar to materials and methods, but with more paragraph
+% continuity
+
+\subsection*{Framework conception and configuration}
+
+The proposed framework for prediction of protein-protein interactions
+
+\subsection*{Framework evaluation on the human interactome}
 
@@
 \end{figure}
-
-\subsection*{Rice Interactome}
-
-For the rice interactome, different combinations of features were
-trained on the model and validated. The results are shown in Table
-\ref{T2}. First, prediction based on \texttt{Node2Vec} without any
-additional feature included is used as baseline of comparison. Figure
-\ref{F1} shows those results, and one can see that its mean performance
-using the AUC is $0.93$. The results for all 10 repetitions of each
-model are consistent, meaning the variance among each line is not
-significant. The results for each model using only the default features
-are satisfactory. This result can also be assessed when looking at
-the confusion matrix, where a precision of $0.98$ and a recall of
-$0.86$ are obtained.
-
-\begin{table}
-\caption{\label{T2}Summary statistics for rice interactome}
-\includegraphics[width=1\columnwidth]{figures/T2.eps}
-\end{table}
-
-\begin{figure}[h]
-\noindent \begin{centering}
-\caption{\label{F1}Summary ROC curves for \texttt{Node2Vec} model alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Only_ML/ROConlyML_SUMMARY.eps}\includegraphics[width=0.48\columnwidth]{figures/Only_ML/CMonlyML_SUMMARY.eps}
-\end{figure}
-
-The next experiment carried out was to add for each prediction method
-score as a prediction feature, that is, use the score as an additional
-feature for the ML algorithm. Results for the Common Neighbors method
-(\textbf{A2}), which uses paths of length 2, are shown in Figure S\ref{F2}.
-All 10 experiments have small variability among them and perform better
-than the baseline, with AUC of $0.98$. When looking at the confusion
-matrix for this model, most guesses are true positives and true negatives,
-resulting in a precision of $0.99$ and a recall of $0.91$. The same
-analyses are done with the count of paths of length 3 (\textbf{A3})
-and with the degree-normalized length-3 score (\textbf{L3}). These
-results are presented in figures S\ref{F3} and \ref{F4}, respectively.
-The corresponding AUC are $0.97$ and $0.98$.
-
-\begin{figure}[h]
-\noindent \begin{centering}
-\caption{\label{F4}Results for \texttt{Node2Vec} with L3 feature}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/ROCL3_SUMMARY.eps}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/CML3_SUMMARY.eps}
-\end{figure}
-
-The precision when using the count of paths of length 3 (A3) is $0.99$
-while the recall is $0.90$. It is interesting to observe that precision
-decreased $0.16\%$ and recall decreased $0.58\%$. A similar situation
-is obtained for the model with the normalized score (L3), whose precision
-decreased to $0.99$ ($-0.08\%$) and recall decreased to $0.88$
-($-2.38\%$). It can be seen also that the ROC curve achieves high
-values of true positive rate sooner in the A2-featured model than
-on the other models.
-
-Another relevant assessment was carried out to verify the predictive
-power of the metrics. \texttt{XGBoost} models were trained only using
-each metric, for the same 10 experiments. Results are shown in Figures
-S\ref{F5}, S\ref{F6} and \ref{F7}. XGBoost yields lower AUC values
-when only considering A2, A3 and L3. Precision and recall for this
-assessments, although lower than the previous experiments, are still
-considered satisfactory.
-
-\begin{figure}[h]
-\noindent \begin{centering}
-\caption{\label{F7}Results for L3 feature alone}
-\par\end{centering}
-\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{figures/Only_Metric/ROConlyL3_SUMMARY.eps}\includegraphics[width=0.48\columnwidth]{figures/Only_Metric/CMonlyL3_SUMMARY.eps}
-\end{figure}
-
-Finally, an relevant evaluation on the models that include a handcrafted
-feature is necessary: How important is the appended feature for the
-model result? The answer can be observed by the feature importance
-plots for A2, A3 and L3 in Figure \ref{F8-importance}. Each plot
-resembles the importance gain from each feature in the underlying
-decision trees that \texttt{XGBoost} uses. Usually, the more bifurcations,
-the higher gain a feature has, the more relevance a feature has on
-the model prediction itself. All models have the appended feature
-as the most important by a good margin, although margin for A3 is
-not wide as for A2 or L3.
-
-\begin{figure}[h]
-\noindent \begin{centering}
-\caption{\label{F8-importance}Importance gain plots for \texttt{Node2Vec}
-with each feature}
-\par\end{centering}
-\begin{centering}
-\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/Imp\lyxdot A2\lyxdot All.eps}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/Imp\lyxdot A3\lyxdot All.eps}
-\par\end{centering}
-\centering{}\includegraphics[width=0.48\columnwidth]{figures/ML_Metric/Imp\lyxdot L3\lyxdot All.eps}
-\end{figure}

### 2020-12-28 nicolas/PPIPL3/bmc_template/introduction.tex
Structure change

@@
 
-Our work extends previous analyses by using handcrafted features (A2,
+This paper extends previous analyses by using handcrafted features (A2,
 A3, L3) as well as a low-dimensional representation of nodes 
 (\texttt{Node2Vec})\cite{Grover_2016} for PPI networks in the human and rice 
-interactomes. To evaluate the proposed approach, we try to predict interactions
-using a state-of-the-art Machine Learning algorithm called \texttt{XGBoost}\cite{2016ChenXGB},
+interactomes. To evaluate the proposed approach, prediction of interactions is carried on 
+with a state-of-the-art Machine Learning algorithm called \texttt{XGBoost}\cite{2016ChenXGB},
 which uses a gradient boosted decision tree model to weigh the features.
-Our main results show that adding the handcrafted features derived from the
-network connectivity to the Machine Learning model improve the prediction
+The main results show that adding the handcrafted features derived from the
+network connectivity to the Machine Learning model improves the prediction
 power of the models. This property can be assessed when looking at the their
@@
 
-\paragraph*{Document structure:} This paper is organized as follows. \emph{Materials and Methods} describes
-the methodological steps and key milestones in the preparation of
-the networks, model parameters and experimental configurations.
-\emph{Results} presents the main results of the models for the human and rice interactomes.
-\emph{Conclusions} presents the main conclusions of this study. Finally, the \emph{appendix}
+\paragraph*{Document structure:} This paper is organized as follows. \emph{Preliminaries} 
+gives some foundations on protein-protein interactions, graph theory 
+and machine learning. \emph{Prediction Framework} describes
+the methodological steps, key milestones and relevant results in the preparation 
+of the networks, model parameters and experimental configurations for the human
+interactome.
+\emph{Case Study} gives a view at the main aspects when exploring the rice
+interactome with the aforementioned framework, including experimental setup and results.
+\emph{Related Work} compares the framework with some results from the literature, 
+exposing some similarities and differences.
+\emph{Conclusion} presents the main conclusions of this study. Finally, the \emph{Appendix}
 presents supplementary information and figures which complement the

### 2020-12-28 nicolas/PPIPL3/bmc_template/preliminaries.tex
Structure change

% "Preliminaries" refers to fundaments for the new reader on the main topics of
% the paper

Let $G=(V,E), n=|V|, m=|E|$ be an undirected graph describing $n$
proteins and $m$ pairwise interactions between them. $G$ can be represented as a Boolean 
adjacency matrix $A$ of size $n\cdot n$, where the value for a position
$(i,j), 0\leq i,j < n$ is set to $1$ if there is a direct interaction
between proteins $i$ and $j$, or set to $0$ otherwise. As it is an undirected network, $A$ is symmetric.

Given this matrix $A$, to find the number of paths of exactly $k$ steps  in the network, $A$ can be raised to the \textit{k}-th power.

### 2021-01-21 nicolas/PPIPL3/bmc_template/abstract.tex
Recent structure update

@@
 \parttitle{Background}
-Recent research on network-based prediction of protein-protein
-interactions (PPIs) reveals that two proteins are more likely to interact,
-the higher the number of paths of length 3 between them exist. This paper
-extends previous work on mapping binary interactions between proteins 
-by learning features (i.e. embeddings) of the PPI network. In particular, 
-this paper reports on an application of gradient boosted decision tree
-model (XGBoost) using handcrafted features  such as normalized measures 
-and algorithmic embeddings that generate a low-dimensional representation
-of the proteins and their interactions (node2vec).
+Recent research on predicting protein-protein
+interactions (PPIs) indicates that two proteins are more likely to interact if 
+there exists a higher number of paths of length 3 between them. This paper
+extends network-based approaches to map binary interactions between proteins 
+by combining hand-crafted features such as the number of paths between proteins 
+and learned features (i.e. node embeddings) of the PPI network. In particular, 
+the proposed approach introduces a gradient boosted decision tree model 
+(XGBoost) usese node2vec to infer embeddings, that is, a low-dimensional representation
+of the proteins and their interactions.
 
 \parttitle{Results}
-The main result shows that while the measure remains an important feature
-for predicting interactions, an improvement in performance is achieved when
-embedding features are considered. The proposed approach is validated for
-the human and rice interactomes. For both cases, the combination of both
-types of features yield higher performance in terms of the Area Under the
-Curve metric (i.e. AUC).
+The main result shows that while the normalized measure remains an important feature
+for predicting interactions, embeddings may improve performance of predictions by up 
+to 15\% in terms of the Area Under the Curve metric (i.e. AUC). The approach is validated for 
+the human and rice interactomes.
 
 \parttitle{Conclusions}
-As witnessed by both human and rice, when information from 
-handcrafted features based on neighborhood is enhanced with vector
-representations from random walks, the prediction power of the model
-improves. Moreover, a supervised learning model has been trained to predict
-unknown interactions based on such information. Finally, the developed
-framework could also be applied to interactomes of other organisms for which
+As shown by the two case studies in human and rice, combining information from 
+handcrafted features with embeddings enhances the prediction power of the model. 
+Improvements of supervised learning models (such as XGBoost) depend on being able 
+to extract additional information about the local neighborhood of nodes that 
+complement the measure based on the number of paths. Finally, the proposed 
+approach could also be applied to interactomes of a number of other organisms for which
 PPI networks have recently become available.

### 2021-01-21 nicolas/PPIPL3/bmc_template/framework.tex
Recent structure update

@@
 
-\subsection*{Framework conception and configuration}
+\subsection*{Framework Conception and Configuration}
 
-The proposed framework for prediction of protein-protein interactions
+As mentioned before, the functionality of a protein strongly depends on its 3D structure, 
+so that changes in its conformation (mutations) can affect how a protein interacts with its 
+surroundings \cite{Sikosek2014Prot}. In this aspect, $A_G$ gives intrinsic information 
+about complementarity between pairs of proteins and hence paths on the protein network are 
+relevant for the prediction task.
 
-\subsection*{Framework evaluation on the human interactome}
+The proposed framework for prediction of protein-protein interactions consists of the
+combination of the path counting metrics along with the vector representations of the
+edges of the network (node2vec embeddings). Namely, the path counting metrics are either 
+the counting of paths of length 2 (matrix $A_G^2$), the counting of paths of length 3 
+(matrix $A_G^3$) or the degree-normalized counting of paths of length 3 (matrix $L_3$). 
+The information retrieved from these matrices will be addressed from now on as A2, A3 
+and L3, respectively.
 
-As the first step in the evaluation of A2, A3 and L3 methods for link
+Figure \ref{fig:framework} shows the proposed framework for predicting protein-protein interactions. As the 
+framework combines two sources of information, it consists of three processes: (a) Metrics 
+calculation and initial predictions, (b) Node and edge embeddings and (c) Machine Learning 
+prediction.
+
+\begin{figure}[h]
+\caption{\label{fig:framework}Framework scheme}
+	\includegraphics[width=\textwidth ]{figures/framework.eps}
+\end{figure}
+
+The input for the workflow consists of a representation of the connectivity of the network.
+Online databases of protein-protein interactions usually report this information as an 
+adjacency list with some additional attributes about the quality of this interaction. Namely, 
+this data is represented as a list of pairs of identifiers and zero or more attributes for 
+each pair. The framework only needs to know whether an interaction exists and therefore this 
+additional information needs to be discarded in advance. In order to calculate optimally the
+subsequent metrics, the framework creates the adjacency matrix $A_G$ out of the given list. 
+Note that as no attributes from nodes or edges are added, the framework predicts based only 
+on the intrinsic properties of the network related to connectivity.
+
+\subsubsection*{A. Metrics Calculation}
+The goal of this stage is to build matrices $A2$, $A3$ and $L3$ which quantify, 
+the number of paths of length 2, of length 3 and the degree-normalized count of 
+paths of length 3, respectively. For this purpose, two calculations are done: 
+counting the degree of each node ($k_i \in \mathbb{Z}$) and raising $A_G$ to the 
+second to obtain $A2$. Finally, $A3$ and $L3$  are calculated simultaneously. 
+Entries $L3_{i,j}$ of $L3$ can be calculated as
+
+\begin{equation}
+  L3_{i,j}=\sum_{p,q\in V}\frac{{A_G}_{i,p}\cdot {A_G}_{p,q}\cdot {A_G}_{q,j}}{\sqrt{k_{p}\cdot k_{q}}}
+\end{equation}
+
+\subsubsection*{B. node2vec Embeddings}
+The purpose of this step is to retrieve a vector representation of the network 
+structure and thus be able to feed the Machine Learning model and improve its 
+predictions. This is done using a state-of-the-art algorithm called node2vec, 
+which is based on the word2vec technique to learn learn word associations from
+a corpus of text. Instead of sentences, node2vec uses 
+random walks from each node in order to construct a skip-gram model and a neural 
+network to learn features (embeddings) from the nodes\cite{Grover_2016}. These 
+features consist of a $d$-dimensional vector representation of each node.
+Since the framework aims to predict edges, these node embeddings are used as 
+input for the Hadamard operator to calculate edge embeddings. Namely, the edge 
+embedding of edge $c_i,j$ is calculated as the element-wise product of each 
+component of the node embeddings for nodes $i$ and $j$. $d$ is chosen as 16.
+
+\subsubsection*{C. Machine Learning Prediction}
+Finally, this stage aims at gathering information from the previous steps to 
+enrich the prediction of a Machine Learning (ML) model and improve its results.
+
+The selected implementation for this prediction problem is XGBoost, which 
+uses gradient boosted decision trees to learn whether an interaction is feasible. 
+The first step in this process is to combine the two sources of information into 
+Dataframes for each of the metrics: for each edge, the information from one of the 
+metrics (A2, A3 or L3) is appended to the vector representation of $d$ dimensions 
+from node2vec, resulting in 3 different tables. Dataframes including only one 
+source of information (each metric or node2vec data) are also generated and 
+fed to XGBoost for comparison.
+
+For each metric, the initial prediction in step A is used to train the supervised 
+ML model
+
+\subsection*{Framework Evaluation on the Human Interactome}
+
+As the first step in the evaluation of A2, A3 and L3 metrics to assess link
 prediction, the full calculation and ranking of all predictions was
@@
 
-However, it is of our interest to assess if Machine Learning can boost
+However, the interest of this study is to assess if Machine Learning can boost
 the overall performance on the prediction, so the proposed procedure
-was adapted to the human interactome data: Instead of having to remove
+is adapted to the human interactome data: Instead of having to remove
 a fraction of edges from the network in order to predict and validate,
 the prediction network and the validation networks are given beforehand.
-The rest of the procedure is carried out as mentioned in Section 2.
+The rest of the procedure is carried out as mentioned above.
 Table \ref{T1} presents the results for the combinations of Node2Vec

### 2021-01-21 nicolas/PPIPL3/bmc_template/introduction.tex
Recent structure update

@@
 than carrying out tasks as single agents, they are part of dynamic
-networks of protein-protein interactions (PPI) \cite{Lin2017}. Such
+networks of protein-protein interactions (PPIs) \cite{Lin2017}. Such
 networks underlie a variety of interdependent mechanisms, including
-signal transduction, homeostasis control and stress responses. Furthermore,
-PPI networks play an important role in physiological and developmental
+signal transduction, homeostasis control and stress responses, and 
+play an important role in physiological and developmental
 processes such as protein phosphorylation, transcriptional co-factor
@@
 
-A common way to create PPI networks (or validate particular protein-protein
-interactions) is the \emph{Yeast-Two-Hybrid} technique (also known
-as \emph{two-hybrid screening} or \emph{Y2H}). Figure \ref{Y2H}A
-illustrates the biological basis of Y2H: the expression of a specific
-reporter gene is activated by the binding of a DNA-binding Domain
+\emph{Yeast-Two-Hybrid} technique (also known as 
+\emph{two-hybrid screening} or \emph{Y2H}) is a commonly used 
+technique to generate PPI networks. Figure \ref{Y2H}A
+illustrates the biological basis of Y2H. The expression of a specific
+reporter gene is activated by binding a DNA-binding Domain
 (DB) and an Activation Domain (AD) of a Transcription Factor, which
@@
 an interaction between two proteins, the Y2H approach fuses one protein
-to the DB domain (known as \emph{bait}) and the other to the
-AD (known as \emph{prey}). If the proteins interact, the reporter
+(known as \emph{bait}) to the DB domain and the other (known as 
+\emph{prey}) to the AD. If the proteins interact, the reporter
 gene expression is activated by the AD (Fig. \ref{Y2H}B). Otherwise,
@@
 \begin{figure}[h]
-\caption{\label{Y2H}The Yeast-2-Hybrid technique offers an experimental approach
+\caption{\label{Y2H}Yeast-2-Hybrid offers an experimental approach
 for constructing PPI networks.}
@@
 of interactions between proteins have been constructed \cite{Weimann2013Y2H,Rajagopala2015Y2H,Shokri2019TFY2H}.
-However, identifying protein interactions through Y2H experiments is costly in terms of time
-and resources\cite{Laraia2015PPI,Macalino2018PPI}. For an organism with 2000
-unique proteins, for example, a detailed PPI network requires the experimental
-validation of around 2 million interactions. The number of missing
-interactions for pairs of proteins on many organisms require the
-use of computational tools for predicting such interactions.
+However, identifying protein interactions through Y2H experiments is 
+costly both in terms of time
+and resources \cite{Laraia2015PPI,Macalino2018PPI}. For example, 
+constructing a PPI network for an organism with 2000 unique 
+proteins requires the above validation procedure to be repeated 
+around 2 million times. Computational tools for predicting such 
+interactions offer a cpst-effective alternative, especially for 
+organisms that produce a large number of proteins.
 
-Different methods for predicting interactions in PPI networks have
-been proposed \cite{Chang2016PPI,Chen2019PPI,Kotlyar2015PPI}. A 
-common approach for predicting PPIs is based the triadic closure principle
-(TCP). TCP states that the higher the number of common neighboring
+Different methods for predicting interactions based on PPI networks have
+been proposed \cite{Chang2016PPI,Chen2019PPI,Kotlyar2015PPI}. A first 
+approach for predicting PPIs is based the triadic closure principle
+(TCP), which states that the higher the number of common neighboring
 nodes between two nodes, the higher the probability that they interact
-\cite{Goldberg2003SmallWorld}. TCP can be addressed mathematically
-by counting the number of shared neighbors of a pair of nodes, also
-known as the Common Neighbors algorithm. By raising the adjacency
-matrix (A) of the PPI network to the second power (A\texttwosuperior ),
-the provides an insight into which proteins may interact with each other. However, previous studies
-show that this mentioned approach alone usually fails because it does not
-consider the structural and chemical properties of the proteins \cite{Cannistraci2013Networks,Kovacs2019}.
+\cite{Goldberg2003SmallWorld}. TCP is measured by counting the number 
+of shared neighbors of a pair of nodes (denoted as A2), namely by raising the adjacency
+matrix of the PPI network to the second power ($A_G^2$). 
+However, previous studies show that TCP often falls short as it does 
+not consider the structural and chemical properties of the proteins
+\cite{Cannistraci2013Networks}, which are related to 
+complementarity in charges and arrangement of atoms between pairs of proteins.
 
-Kovacs et al (2019) introduce a network-based approach which predicts
+The work in \cite{Kovacs2019} introduces another network-based approach which predicts
 an interaction between two proteins based on the number of paths
-of length 3. The number of paths of length 3 is computed by raising
-the adjacency matrix of the network to the third power (A\textthreesuperior).
-The degree-normalized measure, denoted as L3,  mitigates bias caused by
-intermediate hubs within the paths of length 3. The measure L3 enables
-the approach in Kovacs et al (2019) to outperform previous methods for predicting
+of length 3, in contrast to the number of paths of length 2, denoted as A2. 
+The number of paths of length 3, denoted as A3, is computed by raising
+the adjacency matrix of the network to the third power. Furthermore, the 
+degree-normalized measure of A3, denoted as L3,  mitigates bias caused by
+intermediate hubs within the paths of length 3. Using the measure of L3, 
+the approach in \cite{Kovacs2019} outperforms previous methods for predicting
 binary protein interactions for various organisms, including yeast (\emph{S. cerevisiae}),
 Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
-fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus}) and humans
-\cite{Kovacs2019}.
+fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus}) and humans.
 
 This paper extends previous analyses by using handcrafted features (A2,
-A3, L3) as well as a low-dimensional representation of nodes 
-(\texttt{Node2Vec})\cite{Grover_2016} for PPI networks in the human and rice 
-interactomes. To evaluate the proposed approach, prediction of interactions is carried on 
-with a state-of-the-art Machine Learning algorithm called \texttt{XGBoost}\cite{2016ChenXGB},
-which uses a gradient boosted decision tree model to weigh the features.
-The main results show that adding the handcrafted features derived from the
+A3, L3) as well as learned representations of the proteins and their 
+neighborhoods within the PPI networks in human and rice 
+interactomes (\texttt{node2vec}) \cite{Grover_2016}. To evaluate the 
+proposed approach, predictions of interactions are carried out 
+using a supervised Machine Learning algorithm called \texttt{XGBoost} 
+\cite{2016ChenXGB}. XGBoost is particularly useful since its gradient 
+boosted decision tree model weighs the different features and 
+establishes the importance of each feature. The main results of this paper 
+show that adding the handcrafted features derived from the
 network connectivity to the Machine Learning model improves the prediction
 power of the models. This property can be assessed when looking at the their
-importance in the model.  The main contributions of this paper are i) a
-general framework for link prediction in non-directed networks and ii) two
-applications of this framework to biological networks with a structural
-insight.
+importance in the model. The main contributions of this paper are twofold. 
+First, a general framework for link prediction in non-directed networks is 
+introduced. Second, two applications of this framework to biological networks 
+with a structural insight are presented.
 
-\paragraph*{Document structure:} This paper is organized as follows. \emph{Preliminaries} 
-gives some foundations on protein-protein interactions, graph theory 
-and machine learning. \emph{Prediction Framework} describes
-the methodological steps, key milestones and relevant results in the preparation 
-of the networks, model parameters and experimental configurations for the human
-interactome.
-\emph{Case Study} gives a view at the main aspects when exploring the rice
-interactome with the aforementioned framework, including experimental setup and results.
+\paragraph*{Document structure:} This paper is organized as follows. The 
+section \emph{Preliminaries} gives some foundations on protein-protein 
+interactions, graph theory and machine learning. The section 
+\emph{Prediction Framework} describes the methodological steps, key 
+milestones and relevant results in the preparation of the networks, 
+model parameters and experimental configurations for the human
+interactome. The section \emph{Case Study} gives a view at the main 
+aspects when exploring the rice interactome with the aforementioned 
+framework, including experimental setup and results. The section 
 \emph{Related Work} compares the framework with some results from the literature, 
-exposing some similarities and differences.
-\emph{Conclusion} presents the main conclusions of this study. Finally, the \emph{Appendix}
-presents supplementary information and figures which complement the
-results described in this paper.
+exposing some similarities and differences. The final section presents the 
+main conclusions of this study. Supplementary information and figures which 
+complement the results described in this paper  are presented in the 
+\emph{Appendix}.

### 2021-01-22 nicolas/PPIPL3/bmc_template/framework.tex
updated document structure and script

@@
 
-For each metric, the initial prediction in step A is used to train the supervised 
-ML model
+First, for each metric, the initial prediction in stage A is used to train the 
+supervised ML model. It is important to keep in mind that the training dataset 
+and the validation dataset should be sampled randomly to ensure an even
+distribution of the labeling. However, the intrinsic imbalance in the
+datasets has been addressed with under-sampling, so that the training 
+dataset (80\% of data) has the same amount of interactions labeled true and false. 
+Second, the validation dataset (20\% of data) is introduced in the model and it returns 
+a score in the range $(0,1)$ assessing the probability that an edge is feasible. 
+The considered metric to evaluate the ML model performance is the area under the 
+the curve (AUC) of the Receiver Operating Characteristic (ROC). This curve
+corresponds to plot the sensitivity (probability of predicting a real
+positive as positive) against 1-specificity (probability of predicting
+a real negative as positive). AUC values of 1 represent a perfect
+prediction and 0.5 corresponds to a random guess. Normally, values
+over 0.8 of AUC are considered good.
 
@@
 the overall performance on the prediction, so the proposed procedure 
-is adapted to the human interactome data: Instead of having to remove
-a fraction of edges from the network in order to predict and validate,
-the prediction network and the validation networks are given beforehand.
+is adapted to the human interactome data: the prediction network and 
+the validation networks are given beforehand, so that training is done using 
+the labels from one side and the validation is done with the other side. 
 The rest of the procedure is carried out as mentioned above.

### 2021-01-22 nicolas/PPIPL3/bmc_template/preliminaries.tex
updated document structure and script

@@
 % the paper
+\subsection*{Protein Interactions Networks}
+Let $G=(V,E)$ be an undirected graph describing $n=|V|$ proteins as 
+nodes and $m=|E|$ pairwise interactions between the graph as edges. 
+$G$ can be represented as a Boolean adjacency matrix $A_G$ of size 
+$n^2$, where the value of the entry $(i,j), 0\leq i,j < n$ is set 
+to $1$ if there is a direct interaction between proteins $i$ and $j$ 
+and set to $0$ otherwise. As it is an undirected network, $A_G$ is symmetric.
 
-Let $G=(V,E), n=|V|, m=|E|$ be an undirected graph describing $n$
-proteins and $m$ pairwise interactions between them. $G$ can be represented as a Boolean 
-adjacency matrix $A$ of size $n\cdot n$, where the value for a position
-$(i,j), 0\leq i,j < n$ is set to $1$ if there is a direct interaction
-between proteins $i$ and $j$, or set to $0$ otherwise. As it is an undirected network, $A$ is symmetric.
+Given this matrix $A$, to find the number of paths of exactly $k$ steps 
+in the network, $A$ can be raised to the \textit{k}-th power.
 
-Given this matrix $A$, to find the number of paths of exactly $k$ steps  in the network, $A$ can be raised to the \textit{k}-th power.
+\subsection*{Feature Learning of a Network and node2vec}
+The problem of feature learning in a network can be addressed as a maximum 
+likelihood optimization problem, in which the function (to be learned) 
+$f : V \rightarrow \mathbb{R}^d$ maps from nodes to d-dimensional feature 
+representations. $f$ can also be seen as a matrix of size $|V| \times d$. 
+The optimization problem is complex but tractable when assuming conditional 
+independence and spatial symmetry in feature space. Namely, conditional 
+independence means that the likelihood of observing a neighbor node does 
+not depend on any other neighbor node.
+
+For every node $u \in V$, there exists a network neighborhood, denoted as 
+$N_S (u) \subset V$, generated by a sampling strategy limited to k to be 
+able to compare different strategies fairly.
+When sampling, one possible strategy is to retrieve nodes as close as possible from the source.
+This strategy is known as Breadth-first Sampling (BFS).
+On the other hand, one can move as far as possible from the source, also known as Depth-first Sampling (DFS).
+In this regard, node2vec generates random walks from each node with user-defined weighting for back-edges (return parameter p) and for edges that go away from source (in-out parameter q).
+Based on the information (paths) gathered from these walks, node2vec uses a semi-supervised method to learn rich feature representations for nodes in a network.
+
+Since the purpose of this study is link prediction, another mapping function is defined for the edges, so that its feature representation is a bootstrap over the feature representation of nodes.
+Let $g : V \times V \rightarrow \mathbb{R}^d$, then the function $g$ is defined for any pair of nodes $(u,v)$, even if this edge does not exist in the network.
+For this work, the Hadamard operator is used, which consists of an element-wise product of the node features.
+
+\subsection*{Gradient Boosted Trees and XGBoost}
+(THIS IS PENDING)
